### 2016-04-22 doc/header-emnlp.tex
boilerplate style files and preambles

% preamble from emnlp2016.tex to allow reuse of paper.tex between conferences

\documentclass[11pt,letterpaper]{article}
\usepackage{emnlp2016}
\usepackage{times}
\usepackage{latexsym}

% Uncomment this line for the final submission:
%\emnlpfinalcopy

%  Enter the EMNLP Paper ID here:
\def\emnlppaperid{***}

% To expand the titlebox for more authors, uncomment
% below and set accordingly.
% \addtolength\titlebox{.5in}

\newcommand\BibTeX{B{\sc ib}\TeX}

\title{Instructions for EMNLP 2016 Proceedings\Thanks{This
    document has been adapted from the instructions for earlier ACL
    and NAACL proceedings, including those for
    NAACL-HLT 2016 by Margaret Mitchell and Adam Lopez,
    NAACL HLT15 by Matt Post and Adam Lopez,
    NAACL HLT12 by Nizar Habash and William Schuler,
    NAACL HLT10 by Claudia Leacock and Richard Wicentowski,
    NAACL HLT09 by Joakim Nivre and Noah Smith, 
    for ACL05 by Hwee Tou Ng and Kemal Oflazer,
    for ACL02 by Eugene Charniak and Dekang Lin, and earlier ACL and
    EACL formats.  Those versions were written by several people,
    including John Chen, Henry S. Thompson and Donald Walker.
    Additional elements were taken from the formatting instructions of
    the {\em International Joint Conference on Artificial Intelligence}
    and the {\em Conference on Computer Vision and Pattern Recognition}.}}

% Author information can be set in various styles:
% For several authors from the same institution:
% \author{Author 1 \and ... \and Author n \\
%         Address line \\ ... \\ Address line}
% if the names do not fit well on one line use
%         Author 1 \\ {\bf Author 2} \\ ... \\ {\bf Author n} \\
% For authors from different institutions:
% \author{Author 1 \\ Address line \\  ... \\ Address line
%         \And  ... \And
%         Author n \\ Address line \\ ... \\ Address line}
% To start a seperate ``row'' of authors use \AND, as in
% \author{Author 1 \\ Address line \\  ... \\ Address line
%         \AND
%         Author 2 \\ Address line \\ ... \\ Address line \And
%         Author 3 \\ Address line \\ ... \\ Address line}
% If the title and author information does not fit in the area allocated,
% place \setlength\titlebox{<new height>} right after
% at the top, where <new height> can be something larger than 2.25in
\author{Siddharth Patwardhan \and Daniele Pighin\\
  {\tt publication@emnlp2016.net}}

\date{}

\begin{document}

\maketitle

### 2016-04-22 doc/header-emnlp.tex
adapted preambles to accommodate The One Paper

@@
 
-\title{Instructions for EMNLP 2016 Proceedings\Thanks{This
-    document has been adapted from the instructions for earlier ACL
-    and NAACL proceedings, including those for
-    NAACL-HLT 2016 by Margaret Mitchell and Adam Lopez,
-    NAACL HLT15 by Matt Post and Adam Lopez,
-    NAACL HLT12 by Nizar Habash and William Schuler,
-    NAACL HLT10 by Claudia Leacock and Richard Wicentowski,
-    NAACL HLT09 by Joakim Nivre and Noah Smith, 
-    for ACL05 by Hwee Tou Ng and Kemal Oflazer,
-    for ACL02 by Eugene Charniak and Dekang Lin, and earlier ACL and
-    EACL formats.  Those versions were written by several people,
-    including John Chen, Henry S. Thompson and Donald Walker.
-    Additional elements were taken from the formatting instructions of
-    the {\em International Joint Conference on Artificial Intelligence}
-    and the {\em Conference on Computer Vision and Pattern Recognition}.}}
+\title{Instructions for EMNLP 2016 Proceedings}
 
@@
 
-\begin{document}
+%%%\begin{document}
 
-\maketitle
+%%%\maketitle

### 2016-04-22 doc/paper.tex
adapted preambles to accommodate The One Paper

% Yes, this requires flipping flags to compile for different conferences...
% But you can press Ctrl+T while editing in TeXworks without having to change windows!

% conditional compilation
\def\buildemnlp{0}
\def\buildamta{1}
\def\buildtarget{\buildamta} % required to be a single character for \if

% emnlp2016.sty, included by header-emnlp.tex, does NOT tolerate ANY surrounding \else statements...
% an if/else if/... or ifcase statement would have been preferred...
% but \documentclass is declared in the header file, which results in some hidden error-checking:
    % unsupported \buildtarget will fail to build (``\usepackage before \documentclass'')
    % duplicate target values (e.g., emnlp=amta=0) will fail to build (``two \documentclass commands'')
\if \buildtarget \buildemnlp    
    \input{header-emnlp.tex}
    \newcommand\mycitep[1]{\cite{#1}}     % conference-specific ``glue code''
    \newcommand\mynewcite[1]{\newcite{#1}}
\fi
\if \buildtarget \buildamta    
    \input{header-amta.tex}
    \newcommand\mycitep[1]{\citep{#1}} 
    \newcommand\mynewcite[1]{\cite{#1}}  
\fi

% Add support for snippets of \zh{中文}
\usepackage{CJK}
\usepackage[utf8]{inputenc}
\newcommand\zh[1]{\begin{CJK}{UTF8}{gbsn}#1\end{CJK}}

\begin{document}
\maketitle

\begin{abstract}
Hello, World!

Here is some normal spacing.

\zh{你好，世界！}
\end{abstract}

\section{Test}

The quick brown fox jumps over the lazy dog.

At some point, I will almost definitely be citing Moses \mycitep{moses}

% these are so short that it's probably better to have it in literally instead of using \input
\if \buildtarget \buildamta    
    \small
    \bibliographystyle{apalike}
\else \if \buildtarget \buildemnlp
    %Note that emnlp2016.bst chokes if there are 0 citations in your paper.    
    %\mynewcite{moses}    
    \bibliographystyle{emnlp2016}
\else
    % some extra error checking, which could not be used with emnlp2016.sty
    \GenericError{}{Invalid buildtarget - \buildtarget}
\fi \fi

\bibliography{references}

\end{document}

### 2016-04-25 doc/header-aaai.tex
AAAI style for HCOMP

\def\year{2016}\relax
%File: formatting-instruction.tex
\documentclass[letterpaper]{article}
\usepackage{aaai16}
\usepackage{times}
\usepackage{helvet}
\usepackage{courier}
\frenchspacing
\setlength{\pdfpagewidth}{8.5in}
\setlength{\pdfpageheight}{11in}
\pdfinfo{
/Title (Insert Your Title Here)
/Author (Put All Your Authors Here, Separated by Commas)}
\setcounter{secnumdepth}{0}  
\begin{document}
% The file aaai.sty is the style file for AAAI Press 
% proceedings, working notes, and technical reports.
%
\title{Formatting Instructions \\for Authors Using \LaTeX{}}
\author{AAAI Press\\
Association for the Advancement of Artificial Intelligence\\
2275 East Bayshore Road, Suite 160\\
Palo Alto, California 94303\\
}

### 2016-05-01 data.py
added support for custom templates, and this looks kind of promising

@@ -39,8 +39,8 @@ class NameSetBank(Bank):
             if type(item['tags']) is str:
                 item['tags'] = [item['tags']]
         
-    def _makeN(self, item):
-        raise Exception('Unimplemented in abstract base class' + __class__)
+    #def _makeN(self, item):
+    #    raise Exception('Unimplemented in abstract base class' + __class__)
         
     def all_namesets(self):
         return [NameSet(item['nameset']) for item in self._data()]
@@ -140,6 +140,9 @@ class Template:
         
     def head_symbols(self):
         return [s for s in self.symbols() if 'head' in self.description_for_symbol(s)]
+        
+    def options_for_symbol(self, symbol):
+        return self.__data['symbols'][symbol]['options']
         
     def symbols(self):
         return self.__symbols.keys()
@@ -305,6 +308,7 @@ VERB_FORMS = { lang: VerbFormBank(DATA_DIR + 'verbs_{}.yml'.format(lang)) for la
 
 TEMPLATE_DIR = DATA_DIR + 'templates/'
 CLAUSE_TEMPLATE_BANK = TemplateBank(TEMPLATE_DIR + 'clause_templates.yml')
+CUSTOM_TEMPLATE_BANK = TemplateBank(TEMPLATE_DIR + 'custom_templates.yml')
 NP_TEMPLATE_BANK = TemplateBank(TEMPLATE_DIR + 'np_templates.yml')
 
 

### 2016-05-01 main.py

@@ -23,34 +23,50 @@ assert(__name__ == '__main__') # for now
     # BUT, don't you have to create a public API with which you reach down into the tree?
 # I think that ideally there would be some kind of metadata format that can specify all this 
     
-clause = nodes.node_factory('Clause')
-clause.set_template('transitive')
-clause.set_verb_category('action.possession') 
-clause._create_subnodes()   # TODO: roll this into set_template() - n.b. currently must occur after set_verb_category()?
-                            # - well, the long-term goal is to throw all this ugliness into a metascript...
+def make_transitive_clause():    
+    clause = nodes.node_factory('Clause')
+    clause.set_template('transitive')
+    clause.set_verb_category('action.possession') 
+    clause._create_subnodes()   # TODO: roll this into set_template() - n.b. currently must occur after set_verb_category()?
+                                # - well, the long-term goal is to throw all this ugliness into a metascript...
 
-S, V, O = [clause._get_subnode(sym) for sym in 'SVO']
+    S, V, O = [clause._get_subnode(sym) for sym in 'SVO']
 
-S.set_template('name')
-#O.set_template('name')
-O.set_template('noun'); O.add_options({'tags': ['abstract']})
-for n in [S, O]:
-#    n.set_template('name')
-    n._create_subnodes() 
 
+    S.set_template('name')
+    #O.set_template('name')
+    O.set_template('noun'); O.add_options({'tags': ['object']}); 
+    #O.add_options({'number': ['plural']}) # TODO: how to pluralize more robustly?? set_plural() would not be scalable to other options...
+    for n in [S, O]:
+    #    n.set_template('name')
+        n._create_subnodes() 
+
+        
+    A, B = [np._get_subnode('N') for np in [S, O]] # calling "protected" functions externally is a bad sign/smell]
+    A.add_options({'tags': ['woman']})
+    #S.add_options({'tags': ['woman']}) # gets propagated down now, even if called AFTER A and B have been created
+    #B.add_options({'tags': ['man']}) # hmm... 
+    if B.type() == 'noun':
+        B.set_plural() # this would raise AttributeError on Name anyway
     
-A, B = [np._get_subnode('N') for np in [S, O]]
-#A.add_options({'tags': ['woman']})
-#S.add_options({'tags': ['woman']}) # gets propagated down now, even if called AFTER A and B have been created
[…]

### 2016-05-01 nodes.py

@@ -27,7 +27,7 @@ class Node:
         self.__generated_text = None # { 'en': 'Alice' ... }
         
         # just store them for now and figure out what to do with them later... can always use dict.pop()
-        self.__type = options.pop('type')
+        self.__type = options.pop('type') # n.b. this modifies the original data structure!! 
         self.__options = options # need to instantiate even if empty - that way can query if empty
         
     def add_dependency(self, node, **kwargs):
@@ -38,12 +38,17 @@ class Node:
         assert(type(options) is dict)
         
         # TODO: allow multiple sources for monolingual tags...
-        assert(all(type(v) is list for v in options.values()))
+        #assert(all(type(v) is list for v in options.values()))
         for key, value in options.items():
             old_value = self.__options.get(key) or []
             # TODO: change self.__options to defaultdict(list) and just use +=? meh, doesn't even save any lines
-            self.__options[key] = old_value + value
-      
+            # TODO: handle duplicate options? or does that not matter??
+            if type(value) is list:
+                new_value = value
+            else:
+                new_value = [value]
+            self.__options[key] = old_value + new_value
+            
     def generated_text(self, lang):
         return self.__generated_text.get(lang, '')
     def has_generated_text(self, lang):
@@ -173,9 +178,10 @@ class TemplatedNode(Node):
             self.__subnodes = { s: node_factory(self._type_for_symbol(s))
                 for s in self._symbols() }
                 
+            # language-specific syntax tags 
             for symbol, subnode in self.__subnodes.items():
                 subnode.add_options({'tags': self._tags_for_symbol(symbol)})
-            
+           
             # hook up dependencies between NODES and their data - all nodes must have been instantiated first
             # add_dependency() should allow pronouns to work all the way across the tree...right?
             for s in self._symbols():                
@@ -194,13 +200,16 @@ class TemplatedNode(Node):
     def _tags_for_symbol(self, symbol):
         tags = self._syntax_tags_for_symbol(symbol)
         if tags:
-            return [tags]
+            if type(tags) is list:
+                return tags
+            else:
+                return [tags]
         else:
             return []
             
     ### "protected" - these override the base class ###
     def _generate(self, generators):
-        assert(self._subnodes())
[…]

### 2016-05-01 data.py
pleasingly symmetrical TODO - nouns missing determiners: English singular and Chinese plural

@@ -30,6 +30,7 @@ class NameSetBank(Bank):
         __data = self._data()
         
         # discard dummy entries
+        # TODO: save them elsewhere, if I ever wanted to make them available programmatically - otherwise, they're just glorified comments...
         for i in range(len(__data)-1, -1, -1):
             if all(name == None for name in __data[i]['nameset'].values()):
                 __data.pop(i)            

### 2016-05-01 generator.py

@@ -114,7 +114,7 @@ class Generator:
             template_text = node.get_template_text(lang)
             template = template_text.split()
         
-            # TODO!!! handle modifiers here, right? need to insert them INTO the template...
+            # TODO!!! handle modifiers here, right? need to insert them INTO the template... but in a language-dependent way...
             
             # populate the template - get(key, default value)
             result = [generated_symbols.get(token, token) for token in template ]
@@ -144,7 +144,9 @@ class EnGenerator(Generator):
     def _generate_noun(self, node):
         noun_base = self._get_noun_base(node)
         if node.number() == 'singular':
-            noun = noun_base
+            assert(not node.has_modifiers())
+            assert('object' in node._get_option('tags'))
+            noun = 'the ' + noun_base
         else:
             noun = self.__pluralize_noun(noun_base)
             
@@ -229,7 +231,13 @@ class ZhGenerator(Generator):
     
     def _generate_noun(self, node):
         noun = self._get_noun_base(node)
-        self._generate_node_text(node, noun)
+        if node.number() == 'singular':            
+            self._generate_node_text(node, noun)
+        else:
+            assert('object' in node._get_option('tags')) # for now, assume countable? 一些时间 != times...
+            assert(not node.has_modifiers()) # would need to check modifiers for "pluralizers" like CD
+            self._generate_node_text(node, '一些 ' + noun)
+        
         
     # ah, conjugation-free Chinese...
     def _generate_verb(self, node):

### 2016-05-01 nodes.py

@@ -452,6 +452,9 @@ class LexicalNode(Node):
         
 class GenericNoun(LexicalNode): 
     '''Abstract base class to share code with Name, Noun, Pronoun, ...'''
+    def has_modifiers(self):
+        return bool(self._subnodes())
+    
     def number(self):
         #if 'plural' in self._get_option('tags'): # should this really be lumped with semantic tags? well, it is language-independent...
         number_options = self._get_option('number')

### 2016-05-10 main.py
merged from stash

@@ -13,11 +13,7 @@ assert(__name__ == '__main__') # for now
 
 #seed_rng() # output still not deterministic? loops over dicts are still random
 
-#raise Exception('''
-#NEXT: 
-#loose end - semantic matching for ADJP in general
-#loose end - specifying verb forms literally in templates? (VBG for participle; currently in code)
-#''')
+
 
 ### 1. Specify sentences ###
 
@@ -66,7 +62,11 @@ def make_transitive_clause(number='singular', modifiers=[]):
         modifiers.remove('participle')
         participle = nodes.node_factory('Clause')
         participle.set_template('transitive', readonly=False)
+        
+        # TODO: change to "add_transformation" or "transform"? or should i disallow multiple transformations? at least for my first pass...
         participle.set_transformation('participle')    
+        
+        # TODO: some verbs don't seem to work as participles ("the man having the car") - need to blacklist somehow...
         participle.set_verb_category('emotion.desire')
         PO = participle._get_symbol_subnode('O') # note that current test harness requires all NP's to have specified templates...
         PO.set_template('noun'); PO.add_options({'tags': ['object']})

### 2016-05-10 nodes.py

@@ -9,7 +9,6 @@ import data # read data from databases - giant, read-only globals
 import generator
 import utility
 
-
 UNIMPLEMENTED_EXCEPTION = Exception('Needs to be implemented in derived class')
 
 

### 2016-05-10 doc/paper.tex
clearing stash

@@
 \def\buildamta{1}
-\def\buildtarget{\buildamta} % required to be a single character for \if
+\def\buildhcomp{2}
+\def\buildtarget{\buildemnlp} % required to be a single character for \if
 
@@
     \newcommand\mynewcite[1]{\newcite{#1}}
+    \newcommand\amtaonly[1]{}
+    \newcommand\hcomponly[1]{}
 \fi
@@
     \newcommand\mynewcite[1]{\cite{#1}}  
+    \newcommand\amtaonly[1]{#1}
+    \newcommand\hcomponly[1]{}
+\fi
+\if \buildtarget \buildhcomp    
+    \input{header-aaai.tex}
+    \newcommand\mycitep[1]{\cite{#1}} 
+    \newcommand\mynewcite[1]{\citeauthor{#1}~\shortcite{#1}}  
+    \newcommand\amtaonly[1]{}
+    \newcommand\hcomponly[1]{#1}
 \fi
@@
 \newcommand\zh[1]{\begin{CJK}{UTF8}{gbsn}#1\end{CJK}}
+
+% this would allow clickable urls for my convenience - but many conferences forbid them?
+% \usepackage{hyperref}
 
@@
 \zh{你好，世界！}
+
+Scalable parallel corpus generation
 \end{abstract}
 
-\section{Test}
+Curiosity: how many AMTA pages are the EMNLP instructions? i.e., how long would my 4-page EMNLP paper be?
 
-The quick brown fox jumps over the lazy dog.
+\section{Introduction}
+
+RNN encoder-decoders are not the first corpus-based MT methodology (EBMT then SMT then this), and it is unlikely in this era of Big Data that they will be the last.
+Focusing on corpus generation allows linguists to focus on capturing the most statistically prevalent linguistic phenomena, while insulating them from the strengths and weaknesses of various implementations.
+For instance, EBMT suffers from relatively poor generalization and SMT has problems with nonlocality.
+(RNN built on relatively shaky ground, since word vectors don't QUITE make sense for function words?)
+
+Such synthetic parallel corpora can also be viewed as benchmark tasks (need to cite Facebook bAbI).
+
+One might conjecture that the fundamental problem in natural language processing is one of data representation
+
+\section{Related Work}
+\label{sec:related}
+
+Nice tip: ONE LINE PER SENTENCE.
+
+Unlike more AI-oriented or natural language understanding approaches, we are solely concerned here with the generation of surface forms
 
 At some point, I will almost definitely be citing Moses \mycitep{moses}
+
+the latest version of the Stanford Segmenter \mycitep{segmenter:2005,segmenter:2008}
+
+stanford POS tagger (incomplete bibtex entry?) \mycitep{postagger:2003}
+
+This aspect of our work is perhaps closest in spirit to the equivalence classes used by \mynewcite{brown:99} to reduce the data requirements of an EBMT system \mycitep{ebmtbook}.
+
+In the Moses toolchain, the {\small \tt mkcls} component \mycitep{och:98} performs this process in reverse, binning words into automatically-generated equivalence classes to alleviate data sparsity issues.
+% USUALLY alignment only, CAN be applied to translation... http://www.statmt.org/moses/?n=Advanced.Models#ntoc6
+With some effort, you can use these bins during steps other than alignment, although this is not done by default.
+...
+Moreover, as they are upstream of GIZA++, these bins are constructed at the lexical level, not the phrase level, whereas the templates described below can make substitutions at the level of larger constituents and phrases.
+% empty comment - to give closure to git stash pop
+
+some neural network MT paper?
+
+aha, yoshua bengio's "curse of dimensionality" - becomes a blessing?  
+Unlike many machine learning tasks (image recognition, finance), NLP data are purely artificial/manmade.
+New paradigm?
+Not enough attention has been paid IMO to semi-automated corpus generation.
+Well, typical approaches are paid, like Mechanical Turk...
+but there is a vast untapped source of data in the form of student problem sets, in the spirit of reCAPTCHA (citation!?)
+
+Translation is perhaps the only NLP task for which millions of people already undergo years of training
+
+other approaches to synthetic corpora, which are essentially bootstrapped from other MT systems
+
+Hardly a new idea ... but the emphasis here is on the resulting corpus
+
+maybe the latest Hutchins overview or nutshell?
+
+fire a linguist and accuracy goes up
+
+collobert and weston: learn words from letters. interesting from an AI perspective, seems unlikely to improve performance in the near future
+
+chomsky: transformations, colorless green ideas
+
+size doesn't always matter - those massive OpenSubtitles corpora don't help
+- uh, is the problem quality? should i run some Moses runs with a small portion of MultiUN to see how quality improves?
+
+lots of things are resolved naturally when considering generation instead of discrimination
+- ambiguity (you know what you want to say)
+- segmentation (you know what ``phrases'' you wanted to output)
+
+Can think of this as something intermediate between fully automatic (RBMT, or semi-supervised SMT) and fully hand-crafted (human translation, which is the norm).
+Compositional trees are more automatic-like, but generate more sentence pairs/tuples.
+Complex templates are more hand-crafted, but generate fewer pairs/tuples, and are (presumably) harder to combine compositionally (but are they really!?).
+
+The tree structure we adopt is something intermediate between constituency trees and dependency trees, and it was chosen for ease of implementation.
+
+\section{Model}
+
+\subsection{Simplistic Linguistics}
+We adopt the exceedingly simplistic view of language as nouns, verbs, and descriptions (modifications) thereof.
+
+two sources of complexity
+- lexical (semantic?)
+- syntactic
+
+The fact that verbs can serve as nouns (gerundives) and modifiers (participles and modals) is a significant source of added complexity. (footnote?)
+
+I guess I should use placeholder graphics for now, since deadlines for camera-ready are quite a bit later.
+EMNLP: 6/3 submission, 7/29 notification, 9/23 camera-ready
+AMTA: 6/27 submission, 8/8 notification, 9/12 camera-ready
+
+Testing section reference: \ref{sec:related} - this requires changing secnumdepth=2 in AAAI
+
+\section{Experimental Methods}
+
+citations needed
+IWSLT - point out that this corpus is freely available
+OPUS zh\_zh
+MultiUN
+Moses
+
+parameters used
+
+IWSLT: TED talks are a series of lectures, which makes them something intermediate between the conversational language of movie subtitles and the formal written language of UN resolutions.
+Should I mention that you can kind of discern this from their POS distributions?
+I would probably have to mention something concrete then...
+But yeah, I guess it's worth mentioning if I have space, which I almost certainly do for AMTA
+
+heuristically purged zh\_zh, which is a very noisy corpus
+
+\section{Results}
+
+2016-02-20 learning curves adding only 130,000 in-domain lines improves scores by like 2.0 BLEU points
+
+2016-03-03-225556 1M zh\_zh lines made scores slightly WORSE?
+
+2016-04-29-093202 1M MultiUN lines only improved scores SLIGHTLY (by like 0.4 BLEU points)
+
+Quantity and maybe even quality (you'd expect MultiUN translations to be better?) matters much less than having data in-domain
+
+Of course, I need to redo (ALL?) these jobs with final segmenter choice
+
+Evaluated using the IWSLT 2015 progress data set (tst2014) to allow comparison with both 2015 and 2014 systems.
+
+\section{Future Work}
+
+I.e., shit I REALLY doubt I have time for...
+
+For simplicity, the current implementation is lexically one-to-one; for scalability in the number of languages, you ideally want many-to-many.
+
+For this paper, the lexical database was drawn from the IWSLT training set to maximize bottom-line performance.
+[i.e., i don't have TIME to do a general-purpose database]
+This is indeed one potential use of this system, for training set amplification.
+The envisioned ultimate target domain is (cleaned) everyday speech (maybe movie or TV subtitles?) , for which syntactic and lexical diversity is expected to be relatively low, and unknown words would be particularly suitable for crowdsourcing.
+
+Localization of semantic taxonomy - this would have to be one-to-one, and you'd think it would clutter up the raw text UI...
+
+Instead of the next paragraph, should probably just point out how rudimentary the current ``grammar'' is. 
+Future development may be facilitated by adapting existing work from RBMT or NLG. 
+Okay, I guess just point out this grammar doesn't even handle something as basic as:
+
+Shit tons of linguistic phenomena that aren't even that uncommon, including but not limited to:
+More thorough treatment of articles.
+(More) treatment of prepositional phrases.
+Conjunctions.
+Interrogatives and imperatives.
+Pronouns - implemented as names here...? wait, but what about I/me? ugh
+
+\amtaonly{
+\section{AMTA}
+This text only gets added for AMTA. I think it'd be nice to have this for ``filler'' text
+
+If I really am submitting to this conference too, then I can afford (and probably want to?) draw out the length, 
+} % end \amtaonly
+
+\hcomponly{ % hmm, maybe I should use this stuff to pad AMTA too? well, let's not pad just for padding's sake... spare the reviewers
+\section{HCOMP}
+
+An open challenge to the community - making a crowdsourcing platform
+Should I do a quick back-of-the-envelope calculation on how many child-hours are spent on language-related problem sets?
+
+We believe that for crowdsourcing to be most fruitful, there should a matching incentive to the crowd... 
+At the risk of sounding like a sales pitch, what if the crowd wanted to pay you to give you data, like Tom Sawyer?
+The 2015 paper only tagged like 200 words... well, they were trying something WordNet style... but it just goes to show how slow it is...
+WordNet has shit tons of senses, but only a handful of them are common... we want the most common senses here, not completeness.
+For rare senses, we can make custom templates.
+The semantics we desire are more lightweight (?) - just enough to generate surface forms; don't need to do complex logical reasoning on it.
+
+Rare nouns are the biggest source of lexical diversity.
+VERY expensive to do with a small team.
+Very ripe for crowdsourcing, but not necessarily by MOOCs, as proposed by LearnWork - problem sets seem more appropriate here.
+With a sufficiently compelling problem set app, schools could possibly be willing to pay subscription fees to use this app...
+(hmm, I see a problem - schools don't have that much money...
+well, that's where scale comes into play?) well, at the very least, if you make it free, you still get their data
+The question seems not so much ``should we'' as much as ``why aren't we doing this yet?'' - at least in the context of NLP.
+Unlike some traditional machine learning tasks, which use either sensory or observed data as input, human writers ARE the primary and most natural source of linguistic data.
+You wouldn't want to crowdsource bitmaps of cats.
+And even if you were crowdsourcing photos of cats, or even photo labeling, that's not really something students do beyond preschool.
+But to master one's native tongue takes like 10 years, and there's a BUNCH of unharnessed labor there.
+
+} % end \hcomponly
 
@@
     \bibliographystyle{emnlp2016}
+
+\else \if \buildtarget \buildhcomp    
+    \bibliographystyle{aaai}
 \else
@@
     \GenericError{}{Invalid buildtarget - \buildtarget}
-\fi \fi
+\fi \fi \fi
 

### 2016-05-15 data.py
past tense - implemented as just a tiny stupid change to surface forms for now

@@ -227,6 +227,8 @@ class Template:
     
     def prewords(self, lang):
         return self.__data['langs'][lang].get('prewords', {})
+    def postwords(self, lang):
+        return self.__data['langs'][lang].get('postwords', {})
     
     def symbols(self):
         return self.__symbols.keys()
@@ -456,10 +458,12 @@ class VerbCategory:
         return [VerbSet(item) for item in self.__data['verbsets']]
         
     def tags_for_symbol(self, symbol):
-        return self.__data['tags'].get(symbol)
+        tags = self.__data.get('tags', {})
+        return tags.get(symbol)
         
     def tagged_symbols(self):
-        return self.__data['tags'].keys()
+        tags = self.__data.get('tags', {})
+        return tags.keys()
         
     def template_id(self):
         return self.__data['template']

### 2016-05-15 generator.py

@@ -181,6 +181,13 @@ class Generator:
             if preword:
                 symbol_index = result.index(symbol)
                 result.insert(symbol_index, preword)
+                
+        # TODO: DRY this out with prewords...
+        prewords_per_symbol = node.template_postwords(self.LANG)
+        for symbol, postword in prewords_per_symbol.items():
+            if postword:
+                symbol_index = result.index(symbol)
+                result.insert(symbol_index + 1, postword)
             
         assert(type(result) is list and all(type(item) is str for item in result))
         return result

### 2016-05-15 main.py

@@ -28,15 +28,19 @@ def make_transitive_clause(**kwargs):
     return configure_transitive_clause(clause, **kwargs)
     
 # broken off for reuse in make_meta()
-def configure_transitive_clause(clause, number='singular', modifiers=[], template_readonly=True):
+def configure_transitive_clause(clause, number='singular', modifiers=[], transformations=[], template_readonly=True):
     clause.set_template('transitive', readonly=template_readonly)
     assert(not clause._subnodes())
     
     # must be called after set_template() now, due to current transformation implementation...
     # subnodes are generated after both template and verb category have been set.
     # this gives transformations the chance to modify the template.
-    clause.set_verb_category('action.possession')  #
+    clause.set_verb_category('action')  #
     assert(clause._subnodes())
+    
+    if transformations and not template_readonly:
+        for trans in transformations:
+            clause.add_transformation(trans)
 
     S, V, O = [clause._get_symbol_subnode(sym) for sym in 'SVO']
     S.set_template('noun'); S.add_options({'tags': ['person']})#, 'number': 'plural'})
@@ -164,13 +168,15 @@ def make_meta(**kwargs):
     meta = nodes.node_factory('Clause')
     meta.set_template('meta', readonly=False) # writable to allow verb-bin-specific additions from data
     assert(not meta._subnodes())
-    meta.set_verb_category('emotion.desire.meta') #'cognition.knowledge.infinitive')  
+    meta.set_verb_category('emotion.desire.meta') #'cognition.knowledge')  
     assert(meta._subnodes())
  
     S, C = [meta._get_symbol_subnode(sym) for sym in 'SC']
     #S.set_template('noun');
     S.set_template('name'); #S.add_options({'tags': ['person']})#, 'number': 'plural'})
+    
     configure_transitive_clause(C, template_readonly=False, **kwargs)
+    #C.add_transformation('infinitive') # hey, this works! even though subnodes have already been created
     C.add_transformation('remove punctuation')
     
     #raise Exception('TODO: C-template change for things like "I forced HIM TO VB"')
@@ -181,6 +187,8 @@ def make_meta(**kwargs):
  
 clauses = [ None
     , make_transitive_clause()
+    , make_transitive_clause(template_readonly=False, transformations=['tense.past'])
+    #, make_transitive_clause(template_readonly=False).add_transformation('tense.past') # doesn't work since add_transformation() returns None
     ##, make_transitive_clause(number='plural')
[…]

### 2016-05-15 nodes.py

@@ -239,7 +239,8 @@ class TemplatedNode(Node):
 
     def template_prewords(self, lang):
         return self._template().prewords(lang)
-        
+    def template_postwords(self, lang):
+        return self._template().postwords(lang)        
         
         
     ### "pure virtual" functions - to be implemented in derived classes ###

### 2016-05-15 generator.py
fixed copy/pasted variable name (worked before because both read and write were misnamed)

@@ -183,8 +183,8 @@ class Generator:
                 result.insert(symbol_index, preword)
                 
         # TODO: DRY this out with prewords...
-        prewords_per_symbol = node.template_postwords(self.LANG)
-        for symbol, postword in prewords_per_symbol.items():
+        postwords_per_symbol = node.template_postwords(self.LANG)
+        for symbol, postword in postwords_per_symbol.items():
             if postword:
                 symbol_index = result.index(symbol)
                 result.insert(symbol_index + 1, postword)

### 2016-05-15 data.py
a bit of refactoring - tag verbs as past-tense, and then conditionally generate zh postword based on verb category...still needs a lot of work, but at least this prevents false positives

@@ -193,7 +193,7 @@ class Template:
         
         self.__deps_per_symbol = collections.defaultdict(dict)
         self.__forms_per_symbol = collections.defaultdict(dict) # { V: { en: VBG } }
-        self.__syntax_tags_per_symbol = collections.defaultdict(dict)        
+        self.__syntax_tags_per_symbol = collections.defaultdict(dict)  # { en: { S: subjective } }      
         self.__writable = bool(not readonly)
         
         #self.__symbol_metadata = collections.defaultdict(collections.defaultdict(dict))
@@ -290,6 +290,8 @@ class Template:
         
         # parse tags for each symbol - a bit of non-trivial inversion, so do it here at load time, once-only
         # imperative style is just a bit more DRY and flexible for something like this...
+        
+        # language-specific tags
         for lang, metadata in self.__data['langs'].items():
             if lang in LANGUAGES:                
                 for symbol, tags in metadata.get('tags', {}).items():
@@ -303,6 +305,15 @@ class Template:
                 for symbol, form in metadata.get('forms', {}).items():
                     assert(type(form) is str)
                     self.__forms_per_symbol[symbol][lang] = form
+                    
+        # also append language-independent tags
+        for symbol in self.symbols():
+            tags = self.__data['symbols'][symbol].get('tags')
+            if tags:
+                for lang in LANGUAGES:
+                    old_tags = self.__syntax_tags_per_symbol[symbol].get(lang, [])
+                    self.__syntax_tags_per_symbol[symbol][lang] = old_tags + self.__wrap_as_list(tags)
+                
            
     def __punctuation(self, lang):
         return self.__data['langs'][lang].get('punctuation')
@@ -401,13 +412,26 @@ class Template:
                 
                 
     def __merge_dict(self, destination, input):
+        '''precondition: destination and input are both nested dicts, except for leaf nodes'''
         assert(self._writable())
         assert(type(destination) is dict and type(input) is dict)
         for key, value in input.items():
-            if key in destination.keys() and type(value) is dict:
-                assert(type(destination.get(key)) is dict)
-                self.__merge_dict(destination.get(key), value)                
-            else: # safe to copy in - either there's nothing to clobber, or you're clobbering a scalar
+            # YAML: dicts, lists, and scalars.
+            if key in destination.keys():
+                if type(value) is dict:
+                    assert(type(destination.get(key)) is dict)
+                    self.__merge_dict(destination.get(key), value)   
+                elif type(value) is list:
[…]

### 2016-05-15 generator.py

@@ -176,6 +176,7 @@ class Generator:
             raise Exception('Unimplemented template modification: tried to modify {}'.format(node.type()))
             
         # also handle any prewords and postwords here - since this IS "modify_template()" - although this is not QUITE linguistic modification...
+        # TODO: move this to overridable subroutine, to allow for language-specific semantic checks
         prewords_per_symbol = node.template_prewords(self.LANG)
         for symbol, preword in prewords_per_symbol.items():
             if preword:
@@ -209,28 +210,22 @@ class Generator:
             raise Exception('TODO: unhandled modifiers - {}'.format(modifiers))
             
         return result
+
+    # shared by language-specific subclasses
+    def _modify_clause_with_adverbs(self, node, adverb_nodes, result):
+        assert(type(result) is list and all(type(t) is str for t in result))
+
+        # insert adverb before every head verb in the clause
+        head_symbols = node.head_symbols()
+        assert(all(node._type_for_symbol(head) == 'verb') for head in head_symbols)
         
-    def _modify_clause(self, node):
-        result = self._get_unmodified_template(node)
-        modifiers = list(node.modifiers())
+        adv_strings = [a.generated_text(self.LANG) for a in adverb_nodes]
+        assert(len(adv_strings) is 1) # multiple adverbs doesn't work here for zh...although it might for en? semantic-dep? 
         
-        adverbs = self._pop_modifiers(modifiers, 'adverb')
-        if adverbs:
-            # insert adverb before every head verb in the clause
-            head_symbols = node.head_symbols()
-            assert(all(node._type_for_symbol(head) == 'verb') for head in head_symbols)
+        for head in head_symbols:
+            head_index = result.index(head)
             
-            adv_strings = [a.generated_text(self.LANG) for a in adverbs]
-            assert(len(adv_strings) is 1) # multiple adverbs doesn't work here for zh...although it might for en? semantic-dep? 
-            
-            for head in head_symbols:
-                head_index = result.index(head)
-                
-                result = result[:head_index] + adv_strings + result[head_index:] #self.__conjunction(adv_strings) + result[head_index:]
-            
-        if modifiers:
-            raise Exception('TODO: unhandled modifiers - {}'.format(modifiers))
-        
+            result = result[:head_index] + adv_strings + result[head_index:] #self.__conjunction(adv_strings) + result[head_index:]
         return result
 
     def _modifiers_are_done(self, modifiers):
[…]

### 2016-05-15 nodes.py

@@ -234,6 +234,16 @@ class TemplatedNode(Node):
         if self._can_create_symbol_subnodes():
             self._create_symbol_subnodes()
         
+    def syntax_tags_for_symbol(self, symbol, lang): 
+        # TODO: delegate to Template? kind of assuming knowledge of this data structure here...
+        syntax_tags = [t for t in self._tags_for_symbol(symbol) if type(t) is dict]
+        if syntax_tags:
+            assert(len(syntax_tags) is 1)
+            return syntax_tags[0].get(lang, [])
+        else:
+            return []   
+        
+        
     def template_id(self):     # ugh, exposed for external use
         return self._template_id()
 
@@ -441,8 +451,6 @@ class TransformableNode(ModifierNode):
     #    else:
     #        return TemplatedNode.template_id(self)
     
-    #def transformations(self):
-    #    return self.__transformations    
     def add_transformation(self, transformation_str):        
         assert(not self._template_readonly)
         assert(type(transformation_str) is str)
@@ -570,6 +578,8 @@ class Clause(TransformableNode):
         else:
             raise Exception('incompatible template', id, self._template_id())
             
+    def verb_category_id(self):
+        return self.__verb_category_id
             
     
     # overrides

### 2016-05-15 main.py
small tweak to allow bottom-up tree building, although arguably exposing an overly powerful function on TemplatedNode

@@ -35,7 +35,7 @@ def configure_transitive_clause(clause, number='singular', modifiers=[], transfo
     # must be called after set_template() now, due to current transformation implementation...
     # subnodes are generated after both template and verb category have been set.
     # this gives transformations the chance to modify the template.
-    clause.set_verb_category('action')  #
+    clause.set_verb_category('action')  #'action.possession')#
     assert(clause._subnodes())
     
     if transformations and not template_readonly:
@@ -164,10 +164,14 @@ def make_custom(number='singular', modifiers=[]):
     return custom
  
  
-def make_meta(**kwargs):
+def make_meta(bottom_up=False, **kwargs):
     meta = nodes.node_factory('Clause')
     meta.set_template('meta', readonly=False) # writable to allow verb-bin-specific additions from data
+    
     assert(not meta._subnodes())
+    if bottom_up:
+        C = make_transitive_clause(template_readonly=False, **kwargs)
+        meta.add_symbol_subnode('C', C)    
     meta.set_verb_category('emotion.desire.meta') #'cognition.knowledge')  
     assert(meta._subnodes())
  
@@ -175,7 +179,8 @@ def make_meta(**kwargs):
     #S.set_template('noun');
     S.set_template('name'); #S.add_options({'tags': ['person']})#, 'number': 'plural'})
     
-    configure_transitive_clause(C, template_readonly=False, **kwargs)
+    if not bottom_up:
+        configure_transitive_clause(C, template_readonly=False, **kwargs)
     #C.add_transformation('infinitive') # hey, this works! even though subnodes have already been created
     C.add_transformation('remove punctuation')
     
@@ -200,6 +205,7 @@ clauses = [ None
     , make_transitive_clause(number='plural', modifiers=['adjective', 'determiner', 'participle'])
     , make_transitive_clause(modifiers=['adjective', 'adverb'])
     , make_meta(modifiers=['adjective', 'determiner'])
+    , make_meta(modifiers=['adjective', 'determiner'], bottom_up=True)
     , make_custom()
     #, make_custom(number='plural')
     #, make_custom(modifiers=['determiner'])

### 2016-05-15 nodes.py

@@ -207,6 +207,22 @@ class TemplatedNode(Node):
         for subnode in self._get_headnodes(): #subnode in self._subnodes(): # no, that's WAY too powerful
             subnode.add_options(self._options())
     
+    # still keeping _symbols() "protected" (non-external) as a precaution - should only be called by external tree-builder, which SHOULD know symbols
+    def add_symbol_subnode(self, symbol, subnode):
+        '''Allows trees to be built 'bottom-up' by creating subnodes externally, then tacking them on'''
+        assert(self.has_template())
+        assert(not self.__symbol_subnodes.get(symbol)) # incoming symbol should be new
+        
+        old_symbols = self.__symbol_subnodes.keys()
+        assert(symbol not in old_symbols)
+        
+        # alternative: if this is the final symbol subnode, call the "post-processing" portion of _create_symbol_subnode
+        if len(old_symbols) >= len(self._symbols())-1:
+            raise Exception('Cannot add ALL symbols manually - not yet, anyway')
+            
+        self.__symbol_subnodes[symbol] = subnode
+        
+    
     # used by Generator
     def generated_symbols(self, lang):    
         return { symbol: subnode.generated_text(lang) 
@@ -277,13 +293,17 @@ class TemplatedNode(Node):
     
     def _create_symbol_subnodes(self):
         '''Creates only the subnodes specified by the template - no modifiers'''
-        if not self.__symbol_subnodes and self._can_create_symbol_subnodes():
+        # currently assumes not ALL subnodes are created from bottom up - also asserted in add_subnode()
+        if self._can_create_symbol_subnodes() and not all(s in self.__symbol_subnodes.keys() for s in self._symbols()): #not self.__symbol_subnodes:
             # new implementation ignores multiple calls to this function - to simplify enclosing logic
             #assert(not self.__subnodes) # this function should only be called once, for initialization 
         
             # requires symbols to be unique... but they SHOULD be, since they're in a dict
-            self.__symbol_subnodes = { s: node_factory(self._type_for_symbol(s))
-                for s in self._symbols() }
+            #self.__symbol_subnodes = { s: node_factory(self._type_for_symbol(s))
+            #    for s in self._symbols() }
+            for sym in self._symbols():
+                if sym not in self.__symbol_subnodes.keys():
+                    self.__symbol_subnodes[sym] = node_factory(self._type_for_symbol(sym))
                 
             # the parent-child relationship is not currently expected to change, so it's not part of _bequeath()
             for subnode in self.__symbol_subnodes.values():

### 2016-05-15 doc/paper.tex
still trying to clear stash...this is getting embarrassing
and another empty change to try to appease git stash
minor doc change - just to clear git status

@@
 Moreover, as they are upstream of GIZA++, these bins are constructed at the lexical level, not the phrase level, whereas the templates described below can make substitutions at the level of larger constituents and phrases.
-% empty comment - to give closure to git stash pop
 
@@
 
+WELL... issues of ambiguity are alleviated, but only somewhat (e.g., for a sentence about a bank, the generator should know whether it's a building). 
+BUT ambiguous attachments can still be generated (e.g., the [animal] with the [animal] with the overbite), and it would still take additional logic for the system to recognize this, much less generate all possible translations...
+Maybe just emphasize the fact that lexical ambiguity is virtually eliminated, although structural or semantic ambiguity is still possible in some cases...
+But it still seems like it'll be fewer cases... The [person] saw the dog with(by means of) the [vision implement]
+
 Can think of this as something intermediate between fully automatic (RBMT, or semi-supervised SMT) and fully hand-crafted (human translation, which is the norm).
@@
 The tree structure we adopt is something intermediate between constituency trees and dependency trees, and it was chosen for ease of implementation.
+
+We adopt the pragmatic, ``defeatist'' attitude of modeling only straightforward linguistic phenomena compositionally; if the theory and data structures required to model a phenomenon are sufficiently complicated, everything is probably better off just going directly to a translation via template, instead of trying to concoct a sophisticated theory that has to be consistent with all previously theorized phenomena.
 

### 2016-05-15 main.py
modal surface forms being generated, but still have to perform semantic matching between outer Subject and inner verb

@@ -43,7 +43,7 @@ def configure_transitive_clause(clause, number='singular', modifiers=[], transfo
             clause.add_transformation(trans)
 
     S, V, O = [clause._get_symbol_subnode(sym) for sym in 'SVO']
-    S.set_template('noun'); S.add_options({'tags': ['person']})#, 'number': 'plural'})
+    if S: S.set_template('noun'); S.add_options({'tags': ['person']})#, 'number': 'plural'})
     #S.set_template('name'); #O.add_options({'tags': ['man']})
     #S.set_template('name'); #O.add_options({'tags': ['man']})
     #O.set_template('name')
@@ -75,9 +75,7 @@ def configure_transitive_clause(clause, number='singular', modifiers=[], transfo
         # TODO: some verbs don't seem to work as participles ("the man having the car") - need to blacklist somehow...
         participle.set_verb_category('emotion.desire')
         
-        # must be run before set_verb_category(), which will create subnodes as a side effect
-        # BUT must also be run AFTER set_verb_category(), or else it breaks template matching with verb_category...
-        # maybe i should return to fine-grained control? better: add a user option
+        # currently must be run AFTER set_verb_category(), or else it breaks template matching with verb_category...
         participle.add_transformation('participle') 
         participle.create_symbol_subnodes_manually()
         
@@ -85,7 +83,7 @@ def configure_transitive_clause(clause, number='singular', modifiers=[], transfo
         
         PO = participle._get_symbol_subnode('O') # note that current test harness requires all NP's to have specified templates...
         PO.set_template('noun'); PO.add_options({'tags': ['object']})
-        S.add_modifier(participle)
+        if S: S.add_modifier(participle)
     
     
     # add a determiner. oooooo
@@ -164,7 +162,7 @@ def make_custom(number='singular', modifiers=[]):
     return custom
  
  
-def make_meta(bottom_up=False, **kwargs):
+def make_meta(bottom_up=False, **kwargs): # n.b. these are kwargs for INTERNAL use, not to be passed to node_factory
     meta = nodes.node_factory('Clause')
     meta.set_template('meta', readonly=False) # writable to allow verb-bin-specific additions from data
     
@@ -181,15 +179,55 @@ def make_meta(bottom_up=False, **kwargs):
     
     if not bottom_up:
         configure_transitive_clause(C, template_readonly=False, **kwargs)
[…]

### 2016-05-15 nodes.py

@@ -17,14 +17,15 @@ UNIMPLEMENTED_EXCEPTION = Exception('Needs to be implemented in derived class')
 
 # duh, just have Templated inherit from Node
 class Node: 
-    def __init__(self, **options): # contains data, so probably don't want to use as a mixin (would have to call constructor in derived classes)
+    def __init__(self, **input_options): # contains data, so probably don't want to use as a mixin (would have to call constructor in derived classes)
         self.__dependencies = []
         
         self.__generated_text = None # { 'en': 'Alice' ... }
         self.__is_all_lexicalized = False
         
         # just store them for now and figure out what to do with them later... can always use dict.pop()
-        self.__type = options.pop('type') # n.b. this modifies the input original data structure!! (assumed disposable)
+        options = dict(input_options) # make copy for trashing - would it be clearer to call self.__options.pop()?
+        self.__type = options.pop('type') 
         self.__options = collections.defaultdict(list, options) # need to instantiate even if empty - that way can query if empty
         
     # currently only used for subject/verb agreement? (ADJP uses a bidirectional modifier/target system...)
@@ -534,8 +535,7 @@ class TransformableNode(ModifierNode):
         
     # override as needed - for Clause, this reads from a VerbCategory
     def _transformations_for_subnode(self, symbol):
-        return None
-        
+        pass # returns None
         
     def __apply_single_transformation(self, transformation_str):
         transform = data.TRANSFORMATION_BANK.get_transformation_by_id(transformation_str)
@@ -596,7 +596,7 @@ class Clause(TransformableNode):
                 #self._bequeath_to_subnodes() # now called by _create_symbol_subnodes
 
         else:
-            raise Exception('incompatible template', id, self._template_id())
+            raise Exception('incompatible template', id, self._template_id(), category.template_id())
             
     def verb_category_id(self):
         return self.__verb_category_id

### 2016-05-18 main.py
added subject options finally

@@ -28,7 +28,7 @@ def make_transitive_clause(**kwargs):
     return configure_transitive_clause(clause, **kwargs)
     
 # broken off for reuse in make_meta()
-def configure_transitive_clause(clause, number='singular', modifiers=[], transformations=[], template_readonly=True):
+def configure_transitive_clause(clause, number='singular', subject_type='noun', modifiers=[], transformations=[], template_readonly=True):
     clause.set_template('transitive', readonly=template_readonly)
     assert(not clause._subnodes())
     
@@ -43,7 +43,10 @@ def configure_transitive_clause(clause, number='singular', modifiers=[], transfo
             clause.add_transformation(trans)
 
     S, V, O = [clause._get_symbol_subnode(sym) for sym in 'SVO']
-    if S: S.set_template('noun'); S.add_options({'tags': ['person']})#, 'number': 'plural'})
+    if S: 
+        S.set_template(subject_type); #S.add_options({'tags': ['person']})#, 'number': 'plural'})
+        if subject_type == 'noun':
+            S.add_options({'tags': ['animal']})
     #S.set_template('name'); #O.add_options({'tags': ['man']})
     #S.set_template('name'); #O.add_options({'tags': ['man']})
     #O.set_template('name')
@@ -83,7 +86,10 @@ def configure_transitive_clause(clause, number='singular', modifiers=[], transfo
         
         PO = participle._get_symbol_subnode('O') # note that current test harness requires all NP's to have specified templates...
         PO.set_template('noun'); PO.add_options({'tags': ['object']})
-        if S: S.add_modifier(participle)
+        if S and S.template_id() == 'noun': 
+            S.add_modifier(participle)
+        else:
+            O.add_modifier(participle)
     
     
     # add a determiner. oooooo

### 2016-05-18 doc/paper.tex
singular pronouns without antecedents

@@
 One might conjecture that the fundamental problem in natural language processing is one of data representation
+
+With the ALPAC report's pessimistic outlook on MT arguably responsible for sparking decades of fundamental linguistics research in the tradition of Chomsky, it is only fitting that we now bring linguistics back?
+No, my Return of the Linguist had more to do with how RBMT systems have fallen out of favor in research
 

### 2016-05-18 data.py

@@ -113,6 +113,27 @@ class NounSetBank(WordSetBank):
             
     def _is_dummy(self, datum):
         return all(adj == None for adj in datum['nounset'].values())  
+        
+class PronounSetBank(WordSetBank):
+    # useful when just inserting a random dangling pronoun (sentence level)
+    #def all_nounsets(self):
+    #    result = [NounSet(item['nounset']) for item in self._data()]
+    #    assert(len(result) > 0)
+    #    return result
+    def all_pronsets(self):
+        return [PronounSet(item) for item in self._data()]
+    
+    # TODO: queryable function that just takes metadata and returns the right pronset - for referential pronouns
+        
+    #def find_tagged(self, target_tags):
+    #    '''
+    #    Returns all noun synsets satisfying ALL target tags.
+    #    '''
+    #    return [NounSet(item['nounset']) for item in self._data()
+    #        for tag in item['tags'] if all(TAXONOMY.isa(tag, tt) for tt in target_tags)]
+            
+    def _is_dummy(self, datum):
+        return all(adj == None for adj in datum['pronounset'].values()) 
 
             
 class TemplateBank(Bank):
@@ -146,6 +167,9 @@ class NounFormBank(Bank):
             return NounForms(nf)
         else:
             return None
+            
+class PronounFormBank(Bank):
+    pass
         
 class VerbFormBank(Bank):
     def get(self, word):
@@ -639,6 +663,26 @@ class NounSet(WordSet):
     def noun(self, lang, index):
         return WordSet.word(self, lang, index)
         
+class PronounSet:
+    '''Unlike NounSet, etc., this is the entire list item - duck typing + reach a little deeper'''
+    def __init__(self, data):
+        self.__data = data
+        self.__wordset = WordSet(data['pronounset'])
+        
+    def person(self):
+        return self._data()['person']
+        
+    def _data(self):
+        return self.__data
+    
+    # WordSet facade via duck-typing
+    def num_words(self, lang):
+        return self.__wordset.num_words(lang)
+        
+    def pronoun(self, lang, index): # still delegate to WordSet - to leverage any error-checking there
+        return self.__wordset.word(lang, index)
+        
+        
 class VerbSet(WordSet):
[…]

### 2016-05-18 generator.py

@@ -103,12 +103,14 @@ class Generator:
             name = node.name(self.LANG)
             self._generate_node_text(node, name)   
         elif node_type == 'noun':
-            self._generate_noun(node) # punt to subclass            
+            self._generate_noun(node) # punt to subclass
+        elif node_type == 'pronoun':
+            self._generate_pronoun(node)
         elif node_type == 'verb':
             assert(not node.has_modifiers()) # TODO: insert verb modifiers into Clause template
             self._generate_verb(node)
         else:
-            raise Exception('Unimplemented lexical node type ' + node_type)
+            raise Exception('Unimplemented generation for lexical node type ' + node_type)
             
     def _generate_node_text(self, node, text):
         node.set_generated_text(self.LANG, text)
@@ -277,7 +279,12 @@ class EnGenerator(Generator):
             noun = self.__pluralize_noun(noun_base)
         self._generate_node_text(node, noun)
       
-        
+    def _generate_pronoun(self, node):
+        if node.has_antecedent():
+            raise Exception('TODO: generate pronoun from antecedent')
+        else:
+            pronoun = node.pronoun(self.LANG)
+            node.set_generated_text(self.LANG, pronoun)
         
     def _generate_verb(self, node):
         # should depend on subject
@@ -489,6 +496,15 @@ class ZhGenerator(Generator):
     def _generate_noun(self, node):
         noun = self._get_noun_base(node) # no inflections
         self._generate_node_text(node, noun)
+        
+    def _generate_pronoun(self, node):
+        if node.has_antecedent():
+            raise Exception('TODO: generate pronoun from antecedent')
+        else:
+            pronoun = node.pronoun(self.LANG)
+            if node.number() == 'plural':
+                pronoun += '们'
+            node.set_generated_text(self.LANG, pronoun)
         
     # ah, conjugation-free Chinese...
     def _generate_verb(self, node):

### 2016-05-18 nodes.py

@@ -1064,6 +1064,54 @@ class Noun(GenericNoun):
             candidates = data.NOUNSET_BANK.all_nounsets()
         return candidates
         
+        
+class Pronoun(GenericNoun):
+    def __init__(self, **options):
+        GenericNoun.__init__(self, **options)
+        self.__antecedent = None # Node subclass
+        # TODO: specify person (I, you, he) externally? via options?
+        
+    def has_antecedent(self):
+        has_antecedent = bool(self.__antecedent)
+        # for now, antecedent must mean third-person pronoun... TODO: We (antecedent 
+        assert(not has_antecedent or (has_antecedent and self.person() is 3)) 
+        return has_antecedent
+    def set_antecedent(self, node):
+        assert(issubclass(type(node), NounPhrase) or (issubclass(type(node), TemplatedNode) and node.type() == 'NP'))
+        if node.type() != 'NP':
+            raise Exception('Pronoun must refer to a NounPhrase...right?')
+        self.__antecedent = node 
+
+    def person(self): 
+        if self.has_antecedent():
+            return 3
+        else:
+            assert(self._sample_dataset())
+            pronset = self._sample_dataset()
+            return pronset.person() # TODO: here is where this can finally get set to something other than 3
+
+    # TODO: public option to set number? oh that's via options...
+    
+    def pronoun(self, lang):
+        assert(not self.has_antecedent())
+        pronset = self._sample_dataset()
+
+        assert(pronset.num_words(lang) is 1)
+        return pronset.pronoun(lang, 0)
+        
+    def _get_lexical_candidates(self):
+        assert(not self.has_antecedent()) # should never get here; just read antecedent metadata at generation time
+        return data.PRONSET_BANK.all_pronsets()
+    
+    def _lexicalize(self):
+        # with antecedent, this node is empty and just waits until generation time to read metadata from antecedent
+        if self.has_antecedent(): 
+            raise Exception('Should not lexicalize pronoun with antecedent')            
+        else:
+            GenericNoun._lexicalize(self)
+            
+    
+
 
         
     
@@ -1133,6 +1181,7 @@ FACTORY_MAP = {
     'determiner': Determiner,
     'name': Name, 
     'noun': Noun,
+    'pronoun': Pronoun,
     'verb': Verb
 }
 

### 2016-05-18 data.py
plural pronouns

@@ -143,41 +143,61 @@ class TemplateBank(Bank):
 class TransformationBank(Bank):
     def get_transformation_by_id(self, id):
         return Transformation(self._data()[id])
-        
+
+
+
+class WordFormBank(Bank):
+    def get(self, word):
+        wf = self._data().get(word)
+        if wf:
+            return self.DATA_FACTORY(wf)
+        else:
+            return None
         
 # TODO: need these to implement comparatives/superlatives for en
-class AdjectiveFormBank(Bank):
+class AdjectiveFormBank(WordFormBank):
     pass     
-class AdverbFormBank(Bank):
+class AdverbFormBank(WordFormBank):
     pass
     
+
         
-class DeterminerFormBank(Bank):
-    def get(self, word):
-        df = self._data().get(word)
-        if df:
-            return DeterminerForms(df)
-        else:
-            return None
+class DeterminerFormBank(WordFormBank):
+    def __init__(self, filename):
+        WordFormBank.__init__(self, filename)
+        self.DATA_FACTORY = DeterminerForms
+    #def get(self, word):
+    #    df = self._data().get(word)
+    #    if df:
+    #        return DeterminerForms(df)
+    #    else:
+    #        return None
         
-class NounFormBank(Bank):
-    def get(self, word):
-        nf = self._data().get(word)
-        if nf:
-            return NounForms(nf)
-        else:
-            return None
+class NounFormBank(WordFormBank):
+    def __init__(self, filename):
+        WordFormBank.__init__(self, filename)
+        self.DATA_FACTORY = NounForms
+    #def get(self, word):
+    #    nf = self._data().get(word)
+    #    if nf:
+    #        return NounForms(nf)
+    #    else:
+    #        return None
             
-class PronounFormBank(Bank):
-    pass
-        
-class VerbFormBank(Bank):
+class PronounFormBank(WordFormBank):
+    # need to differentiate between key with empty entry and a missing key
+    # TODO: backport this into WordFormBank? does it work with other subclasses?
     def get(self, word):
-        vf = self._data().get(word)
-        if vf:
-            return VerbForms(vf)
+        if word in self._data():
+            return PronounForms(self._data().get(word))
         else:
             return None
+        
+class VerbFormBank(WordFormBank):
+    def __init__(self, filename):
+        WordFormBank.__init__(self, filename)
+        self.DATA_FACTORY = VerbForms
+    
 
 # not currently inheriting Bank, since this is multi-file...
 class VerbSetBank:
[…]

### 2016-05-18 generator.py

@@ -51,6 +51,7 @@ class Generator:
         
         self._det_form_bank = data.DET_FORMS.get(self.LANG)
         self._noun_form_bank = data.NOUN_FORMS.get(self.LANG)
+        self._pronoun_form_bank = data.PRONOUN_FORMS.get(self.LANG)
         self._verb_form_bank = data.VERB_FORMS.get(self.LANG)
         
         
@@ -283,7 +284,25 @@ class EnGenerator(Generator):
         if node.has_antecedent():
             raise Exception('TODO: generate pronoun from antecedent')
         else:
-            pronoun = node.pronoun(self.LANG)
+            pron_base = node.pronoun(self.LANG)
+            pron_forms = self._pronoun_form_bank.get(pron_base)
+            
+            # so the logic here is language AND data dependent... but I guess I've been doing that the whole time
+            tags = node.tags_for_lang(self.LANG)
+            if 'subjective' in tags:
+                case = 'subjective'
+            elif 'objective' in tags:
+                case = 'objective'
+            else:
+                raise Exception('Unsupported case', tags)
+                
+            if node.number() == 'singular':
+                form = case
+            else:
+                assert(node.number() == 'plural')
+                form = 'PRPS.' + case
+                
+            pronoun = pron_forms.get(form, pron_base)
             node.set_generated_text(self.LANG, pronoun)
         
     def _generate_verb(self, node):
@@ -319,7 +338,8 @@ class EnGenerator(Generator):
             verb_forms = self._verb_form_bank.get(verb_base)
             return verb_forms.get_form(form)
             
-            
+    # TODO: move morphological logic into LexicalNode, which would then have language-dependent code, and grow with # languages?
+        # the alternative, which is currently used, is to punch a new hole in LexicalNode every time you need more metadata...
     def __generate_verb_present(self, verb_node, subject_node):
         # person and number
         # need to choose between the forms of the verb's word
@@ -555,7 +575,7 @@ class ZhGenerator(Generator):
             assert(len(dets) is 1) 
             result.append(dets[0].generated_text(self.LANG))            
         else:
-            if node.number() != 'singular':
+            if node.number() != 'singular' and node.template_id() == 'noun':
                 assert('object' in node._get_option('tags')) # for now, assume countable? 一些时间 != times...
                 assert(not node.has_modifiers()) # would need to check modifiers for "pluralizers" like CD
                 result.append('一些')

### 2016-05-18 main.py

@@ -47,6 +47,9 @@ def configure_transitive_clause(clause, number='singular', subject_type='noun', 
         S.set_template(subject_type); #S.add_options({'tags': ['person']})#, 'number': 'plural'})
         if subject_type == 'noun':
             S.add_options({'tags': ['animal']})
+            
+        if subject_type in ['noun', 'pronoun']:
+            S.add_options({'number': [number]})
     #S.set_template('name'); #O.add_options({'tags': ['man']})
     #S.set_template('name'); #O.add_options({'tags': ['man']})
     #O.set_template('name')
@@ -376,6 +379,8 @@ if __name__ == '__main__':
         , make_meta(modifiers=['adjective', 'determiner'], bottom_up=True)
         , make_modal_topdown()
         , make_modal_bottomup()
+        , make_transitive_clause(subject_type='pronoun')
+        , make_transitive_clause(subject_type='pronoun', number='plural')
         #, make_custom()
         #, make_custom(number='plural')
         #, make_custom(modifiers=['determiner'])

### 2016-05-18 nodes.py

@@ -872,6 +872,18 @@ class LexicalNode(Node):
             raise Exception('out of range: {}/{}'.format(index, len(self.__datasets))) 
             # can't just let the system raise IndexError, because it might not for a WHILE before it actually uses index to access
             
+    def tags_for_lang(self, lang):
+        tags = self._get_option('tags')
+        
+        # start by grabbing any language-independent tags
+        lang_tags = [tag for tag in tags if type(tag) is str]
+        
+        for item in tags:
+            if type(item) is not str:
+                lang_tags += item.get(lang, [])
+        return lang_tags
+        
+            
     def total_num_datasets(self, lang): # TODO: rename this atrocity
         # each dataset object could contain more than one "dataset" (nameset, etc.), e.g., en: [Bob, Robert]
         return sum(ds.num_words(lang) for ds in self.__datasets) 
@@ -991,7 +1003,7 @@ class GenericNoun(LexicalNode):
     def number(self):
         #if 'plural' in self._get_option('tags'): # should this really be lumped with semantic tags? well, it is language-independent...
         number_options = self._get_option('number')        
-        assert(not number_options or not('plural' in number_options and 'singular' in number_options))
+        assert(not number_options or not('plural' in number_options and 'singular' in number_options)) # must be either unspecified or not overspecified
         if number_options and 'plural' in number_options:
             return 'plural'
         else:

### 2016-05-18 main.py
fixed 'regression' (really an unimplemented case that snuck through)

@@ -48,7 +48,7 @@ def configure_transitive_clause(clause, number='singular', subject_type='noun', 
         if subject_type == 'noun':
             S.add_options({'tags': ['animal']})
             
-        if subject_type in ['noun', 'pronoun']:
+        if subject_type in ['pronoun']:
             S.add_options({'number': [number]})
     #S.set_template('name'); #O.add_options({'tags': ['man']})
     #S.set_template('name'); #O.add_options({'tags': ['man']})

### 2016-05-18 data.py
subject-verb semantic matching for pronouns

@@ -120,17 +120,15 @@ class PronounSetBank(WordSetBank):
     #    result = [NounSet(item['nounset']) for item in self._data()]
     #    assert(len(result) > 0)
     #    return result
-    def all_pronsets(self):
-        return [PronounSet(item) for item in self._data()]
+    def all_pronsets(self): # PronounSet wraps the whole list item, not just the 'pronounset' subkey
+        return self.find_tagged([]) #[PronounSet(item) for item in self._data()]
+        # TODO: backport this intraclass DRYing to other WordSetBanks
     
     # TODO: queryable function that just takes metadata and returns the right pronset - for referential pronouns
-        
-    #def find_tagged(self, target_tags):
-    #    '''
-    #    Returns all noun synsets satisfying ALL target tags.
-    #    '''
-    #    return [NounSet(item['nounset']) for item in self._data()
-    #        for tag in item['tags'] if all(TAXONOMY.isa(tag, tt) for tt in target_tags)]
+    
+    def find_tagged(self, target_tags):
+        return [PronounSet(item) for item in self._data() 
+            for tag in item.get('tags', []) if all(TAXONOMY.isa(tag, tt) for tt in target_tags)]
             
     def _is_dummy(self, datum):
         return all(adj == None for adj in datum['pronounset'].values()) 

### 2016-05-18 main.py

@@ -45,7 +45,7 @@ def configure_transitive_clause(clause, number='singular', subject_type='noun', 
     S, V, O = [clause._get_symbol_subnode(sym) for sym in 'SVO']
     if S: 
         S.set_template(subject_type); #S.add_options({'tags': ['person']})#, 'number': 'plural'})
-        if subject_type == 'noun':
+        if subject_type in ['noun', 'pronoun']:
             S.add_options({'tags': ['animal']})
             
         if subject_type in ['pronoun']:
@@ -134,9 +134,9 @@ def configure_transitive_clause(clause, number='singular', subject_type='noun', 
     for d in determiners:
         d.set_num_samples(5)
         
-    #for node in lexical_nodes:
-    #    if node.type() == 'name':
-    #        node.set_num_samples(5)
+    for node in lexical_nodes:
+        if node.type() == 'pronoun':
+            node.set_num_samples(10)
     
     
 

### 2016-05-18 nodes.py

@@ -1067,7 +1067,7 @@ class Noun(GenericNoun):
         self.add_options({'number': ['plural']})
         
     def _get_lexical_candidates(self):
-        semantic_tags = [tag for tag in self._get_option('tags') if type(tag) is str] 
+        semantic_tags = self.semantic_tags() #[tag for tag in self._get_option('tags') if type(tag) is str]  #
 
         #assert(len(semantic_tags) <= 1)
         if semantic_tags:
@@ -1107,18 +1107,29 @@ class Pronoun(GenericNoun):
     def pronoun(self, lang):
         assert(not self.has_antecedent())
         pronset = self._sample_dataset()
-
+        
         assert(pronset.num_words(lang) is 1)
         return pronset.pronoun(lang, 0)
         
     def _get_lexical_candidates(self):
-        assert(not self.has_antecedent()) # should never get here; just read antecedent metadata at generation time
-        return data.PRONSET_BANK.all_pronsets()
+        if self.has_antecedent():
+            # currently should never get here; just read antecedent metadata at generation time
+            raise Exception('currently unused')
+            # OR maybe this should get called after all, but at generation time?
+                # call _get_lexical_candidates() from pronoun()?
+            
+            # TODO: subject-verb semantic matching with antecedents...
+                # to do this properly, should really pull tags from antecedent, then feed to VERB at lexicalization time...
+                # for now, just rely on the tree writer...
+        else:
+            return data.PRONSET_BANK.find_tagged(self.semantic_tags() or [])
+
     
     def _lexicalize(self):
         # with antecedent, this node is empty and just waits until generation time to read metadata from antecedent
         if self.has_antecedent(): 
             raise Exception('Should not lexicalize pronoun with antecedent')            
+            # wait, doesn't this function get called, but should just do nothing?
         else:
             GenericNoun._lexicalize(self)
             

### 2016-05-18 nodes.py
d'oh, infinite recursion

@@ -1085,8 +1085,7 @@ class Pronoun(GenericNoun):
         
     def has_antecedent(self):
         has_antecedent = bool(self.__antecedent)
-        # for now, antecedent must mean third-person pronoun... TODO: We (antecedent 
-        assert(not has_antecedent or (has_antecedent and self.person() is 3)) 
+        #assert(not has or (has and self.person() is 3)) # duh, infinite recursion
         return has_antecedent
     def set_antecedent(self, node):
         assert(issubclass(type(node), NounPhrase) or (issubclass(type(node), TemplatedNode) and node.type() == 'NP'))
@@ -1096,6 +1095,7 @@ class Pronoun(GenericNoun):
 
     def person(self): 
         if self.has_antecedent():
+            # TODO: first or second person antecedent (Alice and I, we ... )
             return 3
         else:
             assert(self._sample_dataset())

### 2016-05-19 nodes.py
added comment (giving up on pronouns for now)

@@ -282,7 +282,7 @@ class TemplatedNode(Node):
                 # language-specific syntax tags 
                 subnode.add_options({'tags': self._tags_for_symbol(symbol)})
                    
-                # literal lexical forms hard-coded into template - propagate to lexical node
+                # literal lexical forms (VBG, VBP, etc.) hard-coded into template - propagate to lexical node
                 # rebuild data structure using data API, instead of assuming particular form of raw data. probably no slower than deepcopy...?
                 literal_forms = { lang: self._template().literal_form_for_symbol(symbol, lang)
                     for lang in utility.LANGUAGES
@@ -1089,7 +1089,7 @@ class Pronoun(GenericNoun):
     def __init__(self, **options):
         GenericNoun.__init__(self, **options)
         self.__antecedent = None # Node subclass
-        # TODO: specify person (I, you, he) externally? via options?
+        # TODO: specify person (I, you, he) externally using option pronoun.person (can't be a tag though - would clash with semantic check)
         
     # TODO: merge this with the "dependencies" system (is that relegated to template-specified dependencies at this point?)
     def has_antecedent(self):

### 2016-05-20 data.py
topicalization, with pronoun antecedent specified in template

@@ -240,11 +240,20 @@ class Template:
         # "declarations"
         self.__data = {}
         self.__symbols = {}
-        
-        self.__deps_per_symbol = collections.defaultdict(dict)
-        self.__forms_per_symbol = collections.defaultdict(dict) # { V: { en: VBG } }
-        self.__syntax_tags_per_symbol = collections.defaultdict(dict)  # { en: { S: subjective } }      
         self.__writable = bool(not readonly)
+        
+        self.__forms_per_symbol = collections.defaultdict(dict) # { V: { en: VBG } }
+        self.__lang_indep_deps_per_symbol = collections.defaultdict(list)
+        self.__per_lang_deps_per_symbol = collections.defaultdict(dict)
+        self.__syntax_tags_per_symbol = collections.defaultdict(dict)  # { en: { S: subjective } }      
+                
+        # comprehensive list of data structures to be reparsed on template change
+        if self._writable():
+            self.__parsed = [self.__forms_per_symbol, 
+                             self.__lang_indep_deps_per_symbol,
+                             self.__per_lang_deps_per_symbol, 
+                             self.__syntax_tags_per_symbol]
+
         
         #self.__symbol_metadata = collections.defaultdict(collections.defaultdict(dict))
         # e.g., symbol_metadata[symbol]['tags']
@@ -261,9 +270,6 @@ class Template:
         
     def description_for_symbol(self, symbol):
         return self.__data['symbols'][symbol].get('description') or []
-        
-    def deps_for_symbol(self, symbol):
-        return self.__deps_per_symbol[symbol]
         
     def ghosts(self):    
         result = {}
@@ -282,13 +288,19 @@ class Template:
         
     def head_symbols(self):
         return [s for s in self.symbols() if 'head' in self.description_for_symbol(s)]
-        
+      
+    def lang_indep_deps_for_symbol(self, symbol):
+        return self.__lang_indep_deps_per_symbol[symbol]
+      
     def literal_form_for_symbol(self, symbol, lang):
         forms_per_lang = self.__forms_per_symbol.get(symbol, {}) 
         return forms_per_lang.get(lang) 
         
     def options_for_symbol(self, symbol):
         return self.__data['symbols'][symbol]['options']
+    
+    def per_lang_deps_for_symbol(self, symbol):
+        return self.__per_lang_deps_per_symbol[symbol].items()
     
     def prewords(self, lang):
         return self.__data['langs'][lang].get('prewords', {})
[…]

### 2016-05-20 main.py

@@ -33,15 +33,16 @@ def configure_transitive_clause(clause, number='singular', subject_type='noun', 
     clause.set_template('transitive', readonly=template_readonly)
     assert(not clause._subnodes())
     
+    # oops, this should be called before subnodes are created
+    if transformations and not template_readonly:
+        for trans in transformations:
+            clause.add_transformation(trans)
+    
     # must be called after set_template() now, due to current transformation implementation...
     # subnodes are generated after both template and verb category have been set.
     # this gives transformations the chance to modify the template.
     clause.set_verb_category('action')  #'action.possession')#
     assert(clause._subnodes())
-    
-    if transformations and not template_readonly:
-        for trans in transformations:
-            clause.add_transformation(trans)
 
     S, V, O = [clause._get_symbol_subnode(sym) for sym in 'SVO']
     if S: 
@@ -390,11 +391,13 @@ if __name__ == '__main__':
         , make_transitive_clause(subject_type='pronoun', number='plural')
         , make_transitive_clause(object_type='pronoun', modifiers=['participle'])
         , make_transitive_clause(object_type='pronoun', modifiers=['participle'], participle_object_type='name')
+        , make_transitive_clause(transformations=['topicalization'], template_readonly=False)
         #, make_custom()
         #, make_custom(number='plural')
         #, make_custom(modifiers=['determiner'])
         #, make_custom(number='plural', modifiers=['determiner'])    
         ]
+
     
     outputs = { lang: open('output_{}.txt'.format(lang), 'w', encoding='utf8') for lang in LANGUAGES }
     for c in clauses:

### 2016-05-20 nodes.py

@@ -314,14 +314,20 @@ class TemplatedNode(Node):
             # hook up dependencies between NODES and their data - all nodes must have been instantiated first
             # add_dependency() should allow pronouns to work all the way across the tree...right?
             # TODO: move this to _bequeath()? should deps be allowed to change from transformations?
-            for s in self._symbols():                
-                for lang, deps in self._deps_for_symbol(s).items():
-                    for d in deps:
-                        self.__symbol_subnodes[s].add_dependency(self.__symbol_subnodes[d], lang=lang)
+            for sym in self._symbols():
+                for lang, dependents in self._per_lang_deps_for_symbol(sym):
+                    for dep in dependents:
+                        assert(self._get_symbol_subnode(dep) and self._get_symbol_subnode(sym))
+                        self._get_symbol_subnode(sym).add_dependency(self._get_symbol_subnode(dep), lang=lang)
             
-            self.__headnodes = [self._get_symbol_subnode(s) for s in self.__template.head_symbols()]
+            self.__headnodes = [self._get_symbol_subnode(sym) for sym in self.__template.head_symbols()]
             assert(len(self.__headnodes) <= 1)            
             # TODO: alter head upon transformation
+            
+            # after a bit of ugliness in Template (and calls to it), all-lang deps can be specified in data now
+            for sym in self._symbols():
+                for dep in self._lang_indep_deps_for_symbol(sym):
+                    self._get_symbol_subnode(sym).add_dependency(self._get_symbol_subnode(dep))
 
             self._bequeath_to_subnodes()
 
@@ -353,19 +359,24 @@ class TemplatedNode(Node):
 
 
     ### "protected" functions - available to derived classes ###
-
-    
-    def _deps_for_symbol(self, symbol):
-        return self.__template.deps_for_symbol(symbol)
+    def _get_headnodes(self):
+        return self.__headnodes
         
     def _get_symbol_subnode(self, symbol):
         return self.__symbol_subnodes.get(symbol)
         
+    def _lang_indep_deps_for_symbol(self, symbol):
+        return self.__template.lang_indep_deps_for_symbol(symbol)
+
+    def _per_lang_deps_for_symbol(self, symbol):
+        result = self.__template.per_lang_deps_for_symbol(symbol) #.items()        
+        assert(all(type(item) is tuple for item in result))
+        return result
+        
     def _pop_symbol_subnode(self, symbol):
         return self.__symbol_subnodes.pop(symbol)
[…]

### 2016-05-20 main.py
updated TODO comments

@@ -407,16 +407,7 @@ if __name__ == '__main__':
         o.close()
 
 
-#with open('output.txt', 'w', encoding='utf8') as output:
-#    for lang in LANGUAGES:
-#        output.write(clause.generated_text(lang) + '\n')
-
-
-    
-# TODO (ugh basically EVERYTHING is left to do...)
-    # transformations (tense, participle, etc.)
-    # modifiers
-    # additional modifications or constraints like plural or tag
+# TODO: external mapping of noun classes (animal, person, object) to eligible adjectives (color, etc.) - internal would take longer
 
 #
 ## TODO: have the Clause grab its own data from a TemplateBank

### 2016-05-23 nouns.py
quickie script to get summary statistics about nounsets - to check progress while scaling up in data

@@ -1,0 +1,23 @@
+'''
+Inspect nounsets.yml
+'''
+
+from data import NOUNSET_BANK
+import collections
+import utility
+
+
+if __name__ == '__main__':
+    nounset__data = NOUNSET_BANK._data()
+    #import yaml; print(yaml.dump(nounset__data))
+    
+    tag_counts = collections.Counter()
+    
+    for nounset in nounset__data:
+        for tag in nounset['tags']:
+            tag_counts[tag] += 1
+            
+    for tag, count in utility.nested_sort(tag_counts.items()):
+        print('{:>9} {}'.format(count, tag))
+        
+    # TODO: syntax checking

### 2016-05-23 utility.py

@@ -7,6 +7,20 @@ DATA_DIR = 'datasets/'
 LANGUAGES = ['en', 'zh']
 USE_RANDOM = True
 
+import operator # itemgetter
+
+def nested_sort(items):
+    '''
+    Sort (string, number) tuples deterministically
+    - descending in number
+    - within each number, ascending in string
+    '''
+    return (
+        sorted(
+            sorted(items, key=operator.itemgetter(0)),      # inner ascending sort by secondary key (string)
+            reverse=True, 
+            key=operator.itemgetter(1)))                    # outer descending sort by primary key (number)
+
 # TODO: call this once at startup
 # BUT there might still be nondeterminism in terms of dict key ordering...
 def seed_rng():

### 2016-05-23 nouns.py
write missing noun forms to file - this should streamline workflow

@@ -1,23 +1,39 @@
 '''
-Inspect nounsets.yml
+Inspect nounsets.yml - intended for development use
 '''
 
-from data import NOUNSET_BANK
+from data import NOUNSET_BANK, NOUN_FORMS
 import collections
 import utility
 
 
 if __name__ == '__main__':
     nounset__data = NOUNSET_BANK._data()
-    #import yaml; print(yaml.dump(nounset__data))
     
-    tag_counts = collections.Counter()
+    finished_word_forms = { lang: NOUN_FORMS[lang]._data().keys() for lang in utility.LANGUAGES }
+    missing_word_forms = collections.defaultdict(list)
+    tag_counts = collections.Counter()    
+    
+    
     
     for nounset in nounset__data:
         for tag in nounset['tags']:
             tag_counts[tag] += 1
             
+        for lang in utility.LANGUAGES:
+            word = nounset['nounset'][lang]
+            if word not in finished_word_forms[lang]: #.get(word):
+                missing_word_forms[lang].append(word)
+            
+    # record missing monolingual morphological forms 
+    with open('missing.txt', 'w', encoding='utf8') as output:
+         for lang, missing_list in missing_word_forms.items():
+             output.write('\n\nWords missing from nouns_{}.yml\n'.format(lang))
+             output.write('\n'.join(sorted(missing_list)))
+            
     for tag, count in utility.nested_sort(tag_counts.items()):
         print('{:>9} {}'.format(count, tag))
         
-    # TODO: syntax checking
+    
+        
+    # TODO: syntax checking of individual yaml files

### 2016-05-23 doc/paper.tex
topicalization, with pronoun antecedent specified in template
noun data entry - with a productivity-increasing script

@@
 Such synthetic parallel corpora can also be viewed as benchmark tasks (need to cite Facebook bAbI).
+This is actually probably more important than you'd think, because looking at IWSLT15, some of the human translations are VERY loose.
+So there's another element of noise in the data, and it's one that no amount of downstream machine learning can get rid of.
+(Although people have tried corpus evaluation...)
 
@@
 No, my Return of the Linguist had more to do with how RBMT systems have fallen out of favor in research
+
+My main point seems to be that nobody makes high quality and quantity parallel corpora JUST for machine translation, and it seems like that's sorely needed for both training and benchmarking.
+Even the fanciest new machine learning model is dead in the water if it doesn't have enough data.
 
@@
 We adopt the pragmatic, ``defeatist'' attitude of modeling only straightforward linguistic phenomena compositionally; if the theory and data structures required to model a phenomenon are sufficiently complicated, everything is probably better off just going directly to a translation via template, instead of trying to concoct a sophisticated theory that has to be consistent with all previously theorized phenomena.
+
+This is (probably? ref??) unlike other NLG work in that our emphasis is on surface forms, not the underlying data representation.
 

### 2016-05-23 data.py

@@ -635,14 +635,19 @@ class WordForms:
         
     def _get(self, key, default=None):
         return self.__data.get(key, default)
+
         
 class DeterminerForms(WordForms):
     def get(self, key, default=None):
         return self._get(key, default)
         
 class NounForms(WordForms):
-    def get(self, key, default=None):
-        return self._get(key, default) # hmm, all this does is expose the base class's raw accessor...
+    def get(self, key, default=None):        
+        try:
+            result = self._get(key, default) # hmm, all this does is expose the base class's raw accessor...
+        except AttributeError:
+            result = default # allow keys with scalars, like 人: ren2
+        return result            
         
 class PronounForms(WordForms):
     def get(self, key, default=None):

### 2016-05-23 nouns.py

@@ -34,6 +34,9 @@ if __name__ == '__main__':
     for tag, count in utility.nested_sort(tag_counts.items()):
         print('{:>9} {}'.format(count, tag))
         
+    print('Total number of nounsets:', len(nounset__data))
+
+        
     
         
     # TODO: syntax checking of individual yaml files

### 2016-05-28 main.py
fixed harness regression for custom templates

@@ -1,7 +1,8 @@
 import os
 
-import data # may take a while
+import data # may take a while, depending on data set size
 
+import datetime
 import generator
 import nodes
 import random
@@ -340,6 +341,8 @@ def randomly_configure_node(node, **kwargs):
         randomly_configure_clause(node, **kwargs)
     elif node.type() == 'NP':
         randomly_configure_np(node, **kwargs)
+    #elif node.type() == 'ADJP':
+    #    randomly_configure_adjp(node, **kwargs)
     else:
         raise Exception('Unhandled node type, presumably non-lexical', type(node))
         
@@ -434,17 +437,19 @@ def randomly_configure_np(np, **kwargs):
     
     
 def make_random_adjp(target, **kwargs):
+    adjp = nodes.node_factory('ADJP')
+    randomly_configure_adjp(adjp, target=target, **kwargs)
+    return adjp
 
+def randomly_configure_adjp(adjp, target=None, **kwargs):
     # first determine the type you want - determiners done separately
     if utility.rand() <= 0.9:
         template_id = 'adjective'
     else:
-        template_id = 'noun'        
-    
-    adjp = nodes.node_factory('ADJP')
+        template_id = 'noun' 
+        
     adjp.set_template(template_id)
     # TODO: semantic adjective tags - probably just a very loose mapping here for now, like object: [color, size, ...]
-    return adjp
 
 def make_random_determiner(**kwargs):
     det = nodes.node_factory('ADJP')
@@ -489,7 +494,8 @@ def randomly_configure_custom(custom):
     
     templated_subnodes = [sn for sn in custom._subnodes() if issubclass(type(sn), nodes.TemplatedNode)]
     for subnode in templated_subnodes:
-        randomly_configure_node(subnode, stack_depth=2)
+        if not subnode.has_template(): # might have specified template manually in data
+            randomly_configure_node(subnode, stack_depth=2)
     
 
 
@@ -577,24 +583,18 @@ def generate_all(clause, outputs=None, blow_it_up=False):
             
         # TODO: do this somewhere else instead of tacking it on at the end??
         # ugh, sentence casing is important for multi-sentence lines... 
-        print(sentence_case(clause.generated_text('en')))
-        
+        if True: #not utility.PRODUCTION:
+            print(sentence_case(clause.generated_text('en')))
+
         if outputs:
             for lang in LANGUAGES:
[…]

### 2016-05-28 utility.py

@@ -5,10 +5,14 @@ if sys.version_info < (3, 0):
 # "global" read-only variables
 DATA_DIR = 'datasets/'
 LANGUAGES = ['en', 'zh']
+PRODUCTION = False
 USE_RANDOM = True
 
 import operator # itemgetter
 import random
+
+if PRODUCTION:
+    USE_RANDOM = True
 
 def nested_sort(items):
     '''

### 2016-05-28 main.py
removed testing code

@@ -583,7 +583,7 @@ def generate_all(clause, outputs=None, blow_it_up=False):
             
         # TODO: do this somewhere else instead of tacking it on at the end??
         # ugh, sentence casing is important for multi-sentence lines... 
-        if True: #not utility.PRODUCTION:
+        if not utility.PRODUCTION:
             print(sentence_case(clause.generated_text('en')))
 
         if outputs:

### 2016-05-28 main.py
cleaning up to allow checkout for production

@@ -615,18 +615,17 @@ def make_test_clauses():
         , make_meta(modifiers=['adjective', 'determiner'], bottom_up=True)
         , make_modal_topdown()
         , make_modal_bottomup()
+        , make_transitive_clause(number='plural', modifiers=['adjective', 'determiner'])
         #, make_transitive_clause(subject_type='pronoun')
         #, make_transitive_clause(subject_type='pronoun', number='plural')
         #, make_transitive_clause(object_type='pronoun', modifiers=['participle'])
         #, make_transitive_clause(object_type='pronoun', modifiers=['participle'], participle_object_type='name')
-        #, make_transitive_clause(transformations=['topicalization'], template_readonly=False)
-        , make_custom()
-        , make_custom(number='plural')
-        , make_custom(modifiers=['determiner'])
-        , make_custom(number='plural', modifiers=['determiner']) 
+        , make_transitive_clause(transformations=['topicalization'], template_readonly=False)
+        #, make_custom()
+        #, make_custom(number='plural')
+        #, make_custom(modifiers=['determiner'])
+        #, make_custom(number='plural', modifiers=['determiner']) 
         ]
-
-    #test_clauses = []
         
     return test_clauses
         

### 2016-05-28 main.py
streamlined for testing new data

@@ -10,6 +10,16 @@ import utility
 from utility import LANGUAGES, seed_rng
 
 
+# collected flags for quick toggling during testing
+if utility.PRODUCTION:
+    FIXED_ROLL = FIXED_TEMPLATE = MUTE_OLD_TEST = None
+else:
+    FIXED_ROLL = FIXED_TEMPLATE = MUTE_OLD_TEST = None
+    FIXED_ROLL = 0.75 # 0-0.8 for Clause, 0.8-0.9 for Custom, 0.9-1.0 for C1 C2 - see make_random_sentence()
+    FIXED_TEMPLATE = '把'
+    MUTE_OLD_TEST = True
+
+    
 
 #assert(__name__ == '__main__') # for now # disabled for capitalization hack
 
@@ -310,7 +320,9 @@ def modifiable_template_ids():
 def make_random_sentence():
     '''Make a random clause or custom, and modify it in random places too.'''
     roll = utility.rand()
-    #roll = 0.85#utility.rand()
+    if FIXED_ROLL:
+        roll = FIXED_ROLL
+        
     if roll <= 0.8:
         clause = nodes.node_factory('Clause')    
         randomly_configure_clause(clause)
@@ -365,7 +377,8 @@ def randomly_configure_clause(clause, stack_depth=1, **kwargs):
         candidates += ['把']
         
     template_id = utility.pick_random(candidates)
-    #template_id = '把' 
+    if FIXED_TEMPLATE:
+        template_id = FIXED_TEMPLATE
     
     # randomly add a transformation, like topicalization or past tense
     # in current implementation, topicalization breaks if this is done after verb_category is set 
@@ -488,8 +501,10 @@ def make_random_participle(target, max_runs=10, stack_depth=1, **kwargs):
 def randomly_configure_custom(custom):
     candidates = [id for id in data.CUSTOM_TEMPLATE_BANK.all_template_ids() 
                     if id not in { 'multiple' }] # blacklist some that are not intended for random generation here...
+                    
     template_id = random.choice(candidates)
-    #template_id = 'test'
+    if FIXED_TEMPLATE:
+        template_id = FIXED_TEMPLATE
     custom.set_template(template_id)
     
     templated_subnodes = [sn for sn in custom._subnodes() if issubclass(type(sn), nodes.TemplatedNode)]
@@ -632,7 +647,8 @@ def make_test_clauses():
         
 def run_test():
     test_clauses = make_test_clauses()
[…]

### 2016-05-28 generator.py
ugly hack to prevent double-noun modifiers, like 'car car' (vs. 'clown car')

@@ -73,6 +73,12 @@ class Generator:
         self.__counter = 0 # for multiple passes through the tree (generating dependencies)
         
 
+    # en abhors "car car", and "车车" sounds like baby talk
+    # TODO: this doesn't fix the tree... it just prevents it from generating strings like that...
+    def _can_modify_with_noun(self, target_node, noun_modifier_node):
+        target_nouns = [child.generated_text(self.LANG) for child in target_node._get_headnodes()]
+        return noun_modifier_node.generated_text(self.LANG) not in target_nouns
+            
         
         
     def _generate_lexical(self, node):
@@ -415,7 +421,10 @@ class EnGenerator(Generator):
             
         # nouns (clown car)
         nouns = self._pop_modifiers(modifiers, 'noun')
-        result += [n.generated_text(self.LANG) for n in nouns]            
+        #result += [n.generated_text(self.LANG) for n in nouns]   
+        for n in nouns:
+            if self._can_modify_with_noun(node, n):
+                result.append(n.generated_text(self.LANG))
                     
         result += template
 
@@ -615,7 +624,10 @@ class ZhGenerator(Generator):
                     
         # nouns (clown car)
         nouns = self._pop_modifiers(modifiers, 'noun')
-        result += [n.generated_text(self.LANG) for n in nouns]                
+        #result += [n.generated_text(self.LANG) for n in nouns]                
+        for n in nouns:
+            if self._can_modify_with_noun(node, n):
+                result.append(n.generated_text(self.LANG))
 
         assert(self._modifiers_are_done(modifiers))
         

### 2016-05-28 main.py

@@ -677,7 +677,7 @@ def run_production():
     
     
 if __name__ == '__main__':
-    seed_rng() # for reproducibility
+    #seed_rng() # for reproducibility? meh, doesn't work since I loop over dict keys, which is not deterministic
 
     # hmm, should I really be using singletons for this?
     analyzer = generator.analyzer

### 2016-05-28 generator.py
also suppress duplicate adjectives

@@ -416,7 +416,8 @@ class EnGenerator(Generator):
         adjs = self._pop_modifiers(modifiers, 'adjective')
         if adjs:
             # TODO: adjective ordering
-            adj_strings = [a.generated_text(self.LANG) for a in adjs]
+            # remove duplicate adjectives. TODO: fix the tree as well, instead of just suppressing output
+            adj_strings = list(set(a.generated_text(self.LANG) for a in adjs)) # back-converted to list for use by conjunction
             result += self.__conjunction(adj_strings) 
             
         # nouns (clown car)
@@ -610,8 +611,12 @@ class ZhGenerator(Generator):
             
             # single-character adj's go next to the target, and if multiple, with 和 intervening...
             adj_strings = [a.generated_text(self.LANG) for a in adjs]
-            single_char_adjs = [s for s in adj_strings if len(s) is 1]
-            multi_char_adjs = [s for s in adj_strings if s not in single_char_adjs]
+            
+            # TODO: handle duplicated adjectives correctly - would require changing tree
+            # prevent redundant adjectives for now, which read differently in zh
+            # note that this doesn't fix the tree - it just suppresses the output
+            single_char_adjs = list(set(s for s in adj_strings if len(s) is 1))
+            multi_char_adjs = list(set(s for s in adj_strings if s not in single_char_adjs))
             
             for adj_str in multi_char_adjs:
                 result += [adj_str, '的']

### 2016-05-28 data.py
groundwork

@@ -40,13 +40,27 @@ class WordSetBank(Bank):
 
 
 class AdjectiveSetBank(WordSetBank):
+    def __init__(self, filename):
+        WordSetBank.__init__(self, filename)
+        self.__all_tags = set.union(*[set(item.tags()) for item in self.all_adjsets() if item.tags()])
+
     def all_adjsets(self):
         result = [AdjectiveSet(item) for item in self._data()]
         assert(len(result) > 0)
         return result
         
-    def find_tagged(self, target_tag):
-        raise Exception('TODO: unimplemented stub') # n.b. the "tags" key is currently optional in adjsets.yml
+    def all_tags(self):
+        return self.__all_tags
+        
+        
+    def find_tagged_with_any(self, target_tags):
+        '''
+        Unlike NounSetBank and VerbSetBank, this returns the "union" of all tags instead of the intersection.
+        This is because adjectives are generally specified w.r.t. an existing target, which can take many possible modifier types (color, size, etc.)
+        '''
+        #raise Exception('TODO: unimplemented stub') # n.b. the "tags" key is currently optional in adjsets.yml
+        return [adjset for adjset in self.all_adjsets() #AdjectiveSet(item) for item in self._data() 
+                if any(tt in adjset.tags() for tt in target_tags) or not adjset.tags()]        
         
     def _is_dummy(self, datum):
         return all(adj == None for adj in datum['adjset'].values())
@@ -721,6 +735,12 @@ class WordSet:
         #raise UNIMPLEMENTED_EXCEPTION
         return self._data()[self._wordset_key()][lang] or []
         
+    def _wrap_as_list(self, datum):
+        if type(datum) is list:
+            return datum
+        else:
+            return [datum]
+        
     # ugh, expose to derived classes...
     def _data(self):
         return self.__data
@@ -728,6 +748,13 @@ class WordSet:
 class AdjectiveSet(WordSet):
     def adjective(self, lang, index):
         return WordSet.word(self, lang, index)
+        
+    def tags(self):
+        tags = self._data().get('tags')
+        if tags:
+            return self._wrap_as_list(tags)
+        else:
+            return [] # if you try to return wrap(data.get('tags')), you can wind up with [None]
         
     def _wordset_key(self):
         return 'adjset'

### 2016-05-28 nodes.py

@@ -1061,9 +1061,10 @@ class Adjective(LexicalNode):
         tags = [t for t in (self._get_option('tags') or []) if type(t) is str]
         assert(len(tags) <= 1)
         if tags:            
-            candidates = data.ADJSET_BANK.find_tagged(tags[0])
+            candidates = data.ADJSET_BANK.find_tagged_with_any(tags)
         else:
             candidates = data.ADJSET_BANK.all_adjsets()
+            
         return candidates
         
 class Determiner(LexicalNode):

### 2016-05-28 check_adjs.py
more sophisticated adj-data checking

@@ -4,6 +4,13 @@ import utility
 
 UNINFLECTED_LANGUAGES = ['zh']
 LANGS = [lang for lang in utility.LANGUAGES if lang not in UNINFLECTED_LANGUAGES]
+NONE_STR = '<None>'
+
+def wrap_as_list(datum):
+    if type(datum) is list:
+        return datum
+    else:
+        return [datum]
 
 
 if __name__ == '__main__':
@@ -12,12 +19,20 @@ if __name__ == '__main__':
     
     finished_word_forms = { lang: adjform__data[lang].keys() for lang in LANGS  }
     missing_word_forms = collections.defaultdict(set)
-    #tag_counts = collections.Counter()    
+    tag_counts = collections.Counter()    
     
     adjsets = []
     for adjset in adjset__data:
-        #for tag in nounset['tags']:
-        #    tag_counts[tag] += 1
+        for tag in wrap_as_list(adjset.get('tags')):
+            if tag: 
+                tag_counts[tag] += 1
+            elif tag == None:
+                tag_counts[NONE_STR] += 1
+            else:
+                raise Exception('malformed tag?', tag)
+            
+        if not adjset.get('tags'):
+            tag_counts[NONE_STR] += 1
             
         for lang in LANGS:
             word = adjset['adjset'][lang]
@@ -29,3 +44,8 @@ if __name__ == '__main__':
              missing_str = '\n\nWords missing from adjs_{}.yml\n'.format(lang) + '\n'.join(sorted(missing_list))
              output.write(missing_str)
              print(missing_str)
+             
+        for tag, count in utility.nested_sort(tag_counts.items()):
+             print('{:>9} {}'.format(count, tag))
+
+

### 2016-05-28 data.py
DIRTY CODE - not sure where I am anymore

@@ -36,7 +36,8 @@ class WordSetBank(Bank):
         # TODO: save them elsewhere, if I ever wanted to make them available programmatically - otherwise, they're just glorified comments...
         for i in range(len(__data)-1, -1, -1):
             if self._is_dummy(__data[i]):
-                __data.pop(i) 
+                __data.pop(i)
+
 
 
 class AdjectiveSetBank(WordSetBank):
@@ -52,15 +53,21 @@ class AdjectiveSetBank(WordSetBank):
     def all_tags(self):
         return self.__all_tags
         
-        
-    def find_tagged_with_any(self, target_tags):
-        '''
-        Unlike NounSetBank and VerbSetBank, this returns the "union" of all tags instead of the intersection.
-        This is because adjectives are generally specified w.r.t. an existing target, which can take many possible modifier types (color, size, etc.)
-        '''
+    def find_tagged(self, target_tags):
         #raise Exception('TODO: unimplemented stub') # n.b. the "tags" key is currently optional in adjsets.yml
         return [adjset for adjset in self.all_adjsets() #AdjectiveSet(item) for item in self._data() 
-                if any(tt in adjset.tags() for tt in target_tags) or not adjset.tags()]        
+                if all(tt in adjset.tags() for tt in target_tags) or not adjset.tags()] 
+        
+      
+    # I don't think I should use this - it's way too confusing
+    #def find_tagged_with_any(self, target_tags):
+    #    '''
+    #    Unlike NounSetBank and VerbSetBank, this returns the "union" of all tags instead of the intersection.
+    #    This is because adjectives are generally specified w.r.t. an existing target, which can take many possible modifier types (color, size, etc.)
+    #    '''
+    #    #raise Exception('TODO: unimplemented stub') # n.b. the "tags" key is currently optional in adjsets.yml
+    #    return [adjset for adjset in self.all_adjsets() #AdjectiveSet(item) for item in self._data() 
+    #            if any(tt in adjset.tags() for tt in target_tags) or not adjset.tags()]        
         
     def _is_dummy(self, datum):
         return all(adj == None for adj in datum['adjset'].values())
@@ -123,7 +130,8 @@ class NounSetBank(WordSetBank):
         Returns all noun synsets satisfying ALL target tags.
         '''
         return [NounSet(item) for item in self._data()
-            for tag in item['tags'] if all(TAXONOMY.isa(tag, tt) for tt in target_tags)]
[…]

### 2016-05-28 nodes.py

@@ -988,6 +988,7 @@ class LexicalNode(Node):
     # protected - used by derived classes
     def _pick_samples(self, candidates):
         num_candidates = len(candidates)
+        if not num_candidates: print('lexicalnode._pick_samples', candidates, self.type(), self._options())
         assert(num_candidates > 0)
         if self.num_samples() > num_candidates:
             self.set_num_samples(num_candidates)
@@ -1059,9 +1060,10 @@ class Adjective(LexicalNode):
         
     def _get_lexical_candidates(self):
         tags = [t for t in (self._get_option('tags') or []) if type(t) is str]
-        assert(len(tags) <= 1)
+        #assert(len(tags) <= 1)
         if tags:            
-            candidates = data.ADJSET_BANK.find_tagged_with_any(tags)
+            candidates = data.ADJSET_BANK.find_tagged(tags)
+            #print('adj._get_lex', candidates)
         else:
             candidates = data.ADJSET_BANK.all_adjsets()
             
@@ -1160,6 +1162,7 @@ class Noun(GenericNoun):
             candidates = data.NOUNSET_BANK.find_tagged(semantic_tags)   
         else:
             candidates = data.NOUNSET_BANK.all_nounsets()
+
         return candidates
         
 class Pronoun(GenericNoun):

### 2016-05-28 doc/paper.tex
added a few adjectives:

@@
 (RNN built on relatively shaky ground, since word vectors don't QUITE make sense for function words?)
+Seems somewhat foolhardy to marginalize the domain experts
+Significantly more engineering (man-hours?) seems to go into the machine-learning component.
+(Well, granted, there are human translators who are spending many man-hours generating sentence pairs.
+But they're not intended for MT.)
+Traditionally, corpus construction has been considered prohibitively expensive, ``scavenged'' from other sources like parliamentary proceedings or lecture subtitles.
+But must this be so? Vast untapped resource of students or whatever.
 
@@
 Even the fanciest new machine learning model is dead in the water if it doesn't have enough data.
+
+Return of the linguist: it is relatively easy to come up with examples that industrial-grade systems fail. (hmm,)
+he was shown the car
+他 给 展示 了 车?
+
+\section{Single-Line Scoring}
+
+It is a testament to how indisidious these errors can be that they persisted this long (tst2010 was distributed as recently as IWSLT2015
+This underscores the need for human post-editing, at the very least, for test sets.
+In a shared evaluation task with many language pairs, the most practical way for this is probably crowdsourcing.
+
+It also demonstrates the tremendous imbalance between linguistic effort and machine learning effort/investment
 

### 2016-05-28 data.py
hacked together adj-noun matching on the tree-generator side, just pumping specific tags onto the noun randomly

@@ -43,10 +43,15 @@ class WordSetBank(Bank):
 class AdjectiveSetBank(WordSetBank):
     def __init__(self, filename):
         WordSetBank.__init__(self, filename)
-        self.__all_tags = set.union(*[set(item.tags()) for item in self.all_adjsets() if item.tags()])
+        self.__all_tags = set.union(*[set(item.tags()) for item in self.__all_adjsets() if item.tags()])
 
-    def all_adjsets(self):
-        result = [AdjectiveSet(item) for item in self._data()]
+    # adjsets effectively have two types of tags
+    # the standard tag types, which you are eliminated via find_tagged (all_tags() would return "big", but find_tagged('color') would not)
+    # adjective sets with tags like "target.inanimate" can ONLY be found through find_tagged('target.inanimate')
+        # in other words, there are certain adjectives that you can only opt INTO, not just opt out of
+    def all_unrestricted_adjsets(self):
+        #result = [AdjectiveSet(item) for item in self._data()]
+        result = [adjset for adjset in self.__all_adjsets() if not any(t.startswith('target.') for t in adjset.tags())]
         assert(len(result) > 0)
         return result
         
@@ -55,9 +60,11 @@ class AdjectiveSetBank(WordSetBank):
         
     def find_tagged(self, target_tags):
         #raise Exception('TODO: unimplemented stub') # n.b. the "tags" key is currently optional in adjsets.yml
-        return [adjset for adjset in self.all_adjsets() #AdjectiveSet(item) for item in self._data() 
-                if all(tt in adjset.tags() for tt in target_tags) or not adjset.tags()] 
-        
+        return [adjset for adjset in self.__all_adjsets() #
+                if all(tt in adjset.tags() for tt in target_tags)]# or not adjset.tags()] 
+       
+    def __all_adjsets(self):
+        return [AdjectiveSet(item) for item in self._data()]
       
     # I don't think I should use this - it's way too confusing
     #def find_tagged_with_any(self, target_tags):
@@ -120,18 +127,28 @@ class NameSetBank(WordSetBank):
         return all(adj == None for adj in datum['nameset'].values()) 
             
 class NounSetBank(WordSetBank):
+    def __init__(self, filename):
+        WordSetBank.__init__(self, filename)
+        self.__all_tags = set.union(*[set(item.tags()) for item in self.all_nounsets() if item.tags()])
+        
+    def all_tags(self):
+        return self.__all_tags
+
[…]

### 2016-05-28 main.py

@@ -403,6 +403,20 @@ def randomly_configure_clause(clause, stack_depth=1, **kwargs):
     
     
     
+def add_random_tag_to_np(np):
+    '''
+    Do np's existing tags preclude adding a (presumably more specific) tag?
+    I.e., is "specific_tag" a subset of the existing tags?
+    '''    
+    specific_tag = utility.pick_random(list(data.NOUNSET_BANK.all_tags()))
+    
+    # this is thinking too hard, right? a tag is addable as long as it doesn't result in zero candidates
+    #if all(data.TAXONOMY.isa(specific_tag, t) for t in np.semantic_tags()):
+    #    np.add_options({'tags': [specific_tag]})
+    if data.NOUNSET_BANK.find_tagged(np.semantic_tags() + [specific_tag]):
+        np.add_options({'tags': [specific_tag]})
+    
+    
 def randomly_configure_np(np, **kwargs):
     # pure rand() doesn't give fine-grained control over the distribution
     #template_id = utility.pick_random(data.NP_TEMPLATE_BANK.all_template_ids())
@@ -427,6 +441,12 @@ def randomly_configure_np(np, **kwargs):
     # generator only supports objects right now (otherwise determiners get tricky... like "water" or "cloth")
     if template_id == 'noun':
         np.add_options({'tags': ['object']})
+        
+        # pick random, specific tags from the nounbank and add them to np, if possible.
+        # we'll make up for lexical variety by making lots and lots of trees, instead of lexically varying one tree many times
+        for i in range(3):
+            if utility.rand() <= 0.6:
+                add_random_tag_to_np(np)
         
         # at most one determiner (syntactic constraint)
         if utility.rand() <= 0.5:
@@ -462,9 +482,27 @@ def randomly_configure_adjp(adjp, target=None, **kwargs):
         template_id = 'adjective'
     else:
         template_id = 'noun' 
-        
     adjp.set_template(template_id)
-    # TODO: semantic adjective tags - probably just a very loose mapping here for now, like object: [color, size, ...]
+
+    if target and template_id == 'adjective':
+        
+        # first blacklist any adjectives with required tags (target.inanimate, etc.) if the required tag is not in the target
+        modifier_tags = data.ADJSET_BANK.all_tags()
+        target_tags = target.semantic_tags()
+        
+        for mt in [mt for mt in modifier_tags if mt.startswith('target.')]:
+            required_target_type = mt.split('.', maxsplit=1)[1]
+            
[…]

### 2016-05-28 nodes.py

@@ -988,7 +988,6 @@ class LexicalNode(Node):
     # protected - used by derived classes
     def _pick_samples(self, candidates):
         num_candidates = len(candidates)
-        if not num_candidates: print('lexicalnode._pick_samples', candidates, self.type(), self._options())
         assert(num_candidates > 0)
         if self.num_samples() > num_candidates:
             self.set_num_samples(num_candidates)
@@ -1063,9 +1062,8 @@ class Adjective(LexicalNode):
         #assert(len(tags) <= 1)
         if tags:            
             candidates = data.ADJSET_BANK.find_tagged(tags)
-            #print('adj._get_lex', candidates)
         else:
-            candidates = data.ADJSET_BANK.all_adjsets()
+            candidates = data.ADJSET_BANK.all_unrestricted_adjsets() # adjs are hacked slightly differently - see data.py
             
         return candidates
         

### 2016-05-28 main.py
syntax error:

@@ -501,7 +501,7 @@ def randomly_configure_adjp(adjp, target=None, **kwargs):
         # TODO: target-specific blacklists, like removing the "color" tag for any "abstract" targets (optional! colorless green ideas)        
         
         # limit adjective choices at lexicalization time
-        adjp.add_options([utility.pick_random(list(modifier_tags))]})
+        adjp.add_options([utility.pick_random(list(modifier_tags))])
 
 
 def make_random_determiner(**kwargs):

### 2016-05-30 check_adjs.py
finished adj forms from last batch

@@ -57,6 +57,10 @@ if __name__ == '__main__':
     for tag, count in utility.nested_sort(tag_counts.items()):
         print('{:>9} {}'.format(count, tag))
 
+    print('\n"Boolean" adjs in en that have explicitly forbidden JJR and JJS')
+    for en_word, word_data in sorted(adjform__data['en'].items()):
+        if word_data and any(form in word_data and word_data[form] is None  for form in ['JJR', 'JJS']):
+            print(en_word)
     
     if untagged_word_forms['en']:
         print('\nUntagged words (en) - consult WordNet?')
@@ -65,9 +69,10 @@ if __name__ == '__main__':
                     
     with open('missing_adjs.txt', 'w', encoding='utf8') as output:
         for lang, missing_list in missing_word_forms.items():
-             missing_str = '\n\nWords missing from adjs_{}.yml\n'.format(lang) + '\n'.join(sorted(missing_list))
-             output.write(missing_str)
-             print(missing_str)
+            missing_str = '\n\nWords missing from adjs_{}.yml\n'.format(lang) + '\n'.join(sorted(missing_list))
+            output.write(missing_str)
+            print(missing_str)
+
 
 
             

### 2016-05-30 check_nouns.py
updated noun checker to handle multiple-word synsets

@@ -6,6 +6,11 @@ from data import NOUNSET_BANK, NOUN_FORMS
 import collections
 import utility
 
+def wrap_as_list(datum):
+    if type(datum) is list:
+        return datum
+    else:
+        return [datum]
 
 if __name__ == '__main__':
     nounset__data = NOUNSET_BANK._data()
@@ -21,9 +26,15 @@ if __name__ == '__main__':
             tag_counts[tag] += 1
             
         for lang in utility.LANGUAGES:
-            word = nounset['nounset'][lang]
-            if word not in finished_word_forms[lang]: #.get(word):
-                missing_word_forms[lang].append(word)
+            try:
+                word_data = nounset['nounset'][lang]
+            except KeyError:
+                print(nounset)
+                raise
+                
+            for word in wrap_as_list(word_data):
+                if word not in finished_word_forms[lang]: #.get(word):
+                    missing_word_forms[lang].append(word)
             
     # record missing monolingual morphological forms 
     with open('missing_nouns.txt', 'w', encoding='utf8') as output:

### 2016-05-30 check_nouns.py
updated noun checker to handle multiple taglists too - should really route this through WordSet...

@@ -12,6 +12,7 @@ def wrap_as_list(datum):
     else:
         return [datum]
 
+# TODO: have these checkers use WordSet instead
 if __name__ == '__main__':
     nounset__data = NOUNSET_BANK._data()
     
@@ -22,8 +23,22 @@ if __name__ == '__main__':
     
     
     for nounset in nounset__data:
-        for tag in nounset['tags']:
-            tag_counts[tag] += 1
+        #for tag in nounset['tags']:
+        #    tag_counts[tag] += 1
+        for tag in wrap_as_list(nounset.get('tags')):
+            if tag: 
+                if type(tag) is str:
+                    tag_counts[tag] += 1
+                elif type(tag) is list:
+                    for subtag in tag:
+                        assert(type(subtag) is str)
+                        tag_counts[subtag] += 1
+                else:
+                    raise Exception('malformed tag? expected str or list', tag)
+            elif tag is None:
+                tag_counts[NONE_STR] += 1
+            else:
+                raise Exception('malformed tag? "False" but not None', tag)
             
         for lang in utility.LANGUAGES:
             try:

### 2016-05-30 generator.py
consolidated duplicate-adjective code into base class Generator - now THIS is scalable with languages

@@ -272,11 +272,29 @@ class Generator:
         '''Removes modifiers with matching template ID (changes list in place) and returns them'''
         assert(type(modifiers) is list)
         assert(all(issubclass(type(m), nodes.ModifierNode) for m in modifiers))
-        popped = [m for m in modifiers if m.template_id() == id_to_pop]
-        for p in popped:
+        
+        pop_list = [m for m in modifiers if m.template_id() == id_to_pop]
+        
+        # prevent redundant adjectives for now, which read differently in zh. note that this doesn't fix the tree - it just suppresses the output
+        # note that duplicated nouns are suppressed by _can_modify_with_noun above.
+        # TODO: handle duplicated adjectives at tree-time somehow - when adjsets are chosen, you'd have to inspect them nonlocally.
+            # zh duplication should be handled with single-node + a flag or something
+        if id_to_pop == 'adjective':
+        
+            # pop_list = nodes with first occurrence of each adj string
+            old_pop_list = pop_list
+            words = [node.generated_text(self.LANG) for node in pop_list]
+            pop_list = [old_pop_list[i] for i in [words.index(w) for w in set(words)]]
+            
+            # discard all occurrences after the first - do not return them to caller for generation
+            for p in old_pop_list:
+                if p not in pop_list:
+                    modifiers.remove(p)
+        
+        for p in pop_list:
             modifiers.remove(p)
             assert(modifiers.count(p) is 0)
-        return popped
+        return pop_list
         
         
     
@@ -430,8 +448,7 @@ class EnGenerator(Generator):
         adjs = self._pop_modifiers(modifiers, 'adjective')
         if adjs:
             # TODO: adjective ordering
-            # remove duplicate adjectives. TODO: fix the tree as well, instead of just suppressing output
-            adj_strings = list(set(a.generated_text(self.LANG) for a in adjs)) # back-converted to list for use by conjunction
+            adj_strings = [a.generated_text(self.LANG) for a in adjs]
             result += self.__conjunction(adj_strings) 
             
         # nouns (clown car)
@@ -644,11 +661,8 @@ class ZhGenerator(Generator):
             # single-character adj's go next to the target, and if multiple, with 和 intervening...
             adj_strings = [a.generated_text(self.LANG) for a in adjs]
             
-            # TODO: handle duplicated adjectives correctly - would require changing tree
-            # prevent redundant adjectives for now, which read differently in zh
[…]

### 2016-05-30 main.py

@@ -504,7 +504,7 @@ def randomly_configure_np(np, **kwargs):
             np.add_modifier(determiner)
                 
         
-        # TODO: disallow multiple identical adjectives (the big and big person)
+        # TODO: disallow multiple identical adjectives (the big and big person) in the TREE - currently done at the generator level...
         adjp_count = 0
         for i in range(5): # TODO: zh gets awkward with more than 2 adjectives, esp. single-char...
             if utility.rand() < 0.2: 

### 2016-05-30 check_nouns.py
remove duplicate nouns

@@ -17,7 +17,7 @@ if __name__ == '__main__':
     nounset__data = NOUNSET_BANK._data()
     
     finished_word_forms = { lang: NOUN_FORMS[lang]._data().keys() for lang in utility.LANGUAGES }
-    missing_word_forms = collections.defaultdict(list)
+    missing_word_forms = collections.defaultdict(set)
     tag_counts = collections.Counter()    
     
     
@@ -49,7 +49,7 @@ if __name__ == '__main__':
                 
             for word in wrap_as_list(word_data):
                 if word not in finished_word_forms[lang]: #.get(word):
-                    missing_word_forms[lang].append(word)
+                    missing_word_forms[lang].add(word)
             
     # record missing monolingual morphological forms 
     with open('missing_nouns.txt', 'w', encoding='utf8') as output:

### 2016-05-30 data.py
automatically pluralize en nouns ending in y/ies

@@ -155,7 +155,7 @@ class NounSetBank(WordSetBank):
         '''
         return [NounSet(item) for item in self._data()
             #for tag in item['tags'] if all(TAXONOMY.isa(tag, tt) for tt in target_tags)] # listcomps are really prone to subtle logic errors
-            if all( any(TAXONOMY.isa(tag, tt) for tag in item.get('tags', [])) for tt in target_tags ) or not item.get('tags')]
+            if all( any(TAXONOMY.isa(tag, tt) for tag in (item.get('tags') or [])) for tt in target_tags ) or not item.get('tags')]
             
     # overrides (concrete implementations)
     def _all_datasets(self):

### 2016-05-30 generator.py

@@ -499,9 +499,14 @@ class EnGenerator(Generator):
     def __pluralize_noun(self, noun_base):
         noun_forms = self._noun_form_bank.get(noun_base)
         if noun_forms and noun_forms.get('NNS'):
-            return noun_forms.get('NNS')
+            result = noun_forms.get('NNS')
         else:
-            return noun_base + 's'
+            if noun_base.endswith('y'):
+                result = noun_base[:-1] + 'ies'
+            else:
+                result = noun_base + 's'
+            
+        return result
             
             
     # name modification is actually kind of annoying

### 2016-05-30 generator.py
and nouns that end in s

@@ -503,6 +503,8 @@ class EnGenerator(Generator):
         else:
             if noun_base.endswith('y'):
                 result = noun_base[:-1] + 'ies'
+            elif noun_base.endswith('s'):
+                result = noun_base + 'es'
             else:
                 result = noun_base + 's'
             

### 2016-05-30 data.py
one last batch of nouns before moving on to rare nouns

@@ -44,11 +44,25 @@ class WordSetBank(Bank):
             if self._is_dummy(__data[i]):
                 __data.pop(i)
 
-        self.__all_tags = None # set
+        self.__all_tags = set() # set
                 
     def all_tags(self):
-        if self.__all_tags is None: # lazy initialization to avoid choking on unimplemented _all_datasets()
-            self.__all_tags = set.union(*[set(item.tags()) for item in self._all_datasets() if item.tags()])
+        if not self.__all_tags: # lazy initialization to avoid choking on unimplemented _all_datasets()
+            for item in self._all_datasets():
+                tag_data = item.tags()
+                if tag_data:
+                    assert(type(tag_data) is list)
+                    if all(type(t) is str for t in tag_data):
+                        tag_lists = [tag_data]
+                    else:
+                        tag_lists = tag_data
+                        
+                    assert(all(type(tag_list) is list for tag_list in tag_lists))
+                    assert(all(type(tag) is str for tag_list in tag_lists for tag in tag_list))
+                    
+                    for tag_list in tag_lists:
+                        self.__all_tags.update(set(tag_list))
+
         return self.__all_tags.copy() # gets trashed all OVER the place downstream. this was an insidious bug.
                 
     #def _all_datasets(self): # duck typing will raise a more informative exception (tells you WHICH subclass needs implementing)
@@ -145,7 +159,7 @@ class NameSetBank(WordSetBank):
             
 class NounSetBank(WordSetBank):
     def all_nounsets(self):
-        result = self.find_tagged([]) #[NounSet(item) for item in self._data()]
+        result = [NounSet(item) for item in self._data()] #self.find_tagged([]) #
         assert(len(result) > 0)
         return result
         
@@ -153,10 +167,11 @@ class NounSetBank(WordSetBank):
         '''
         Returns all noun synsets satisfying ALL target tags.
         '''
-        return [NounSet(item) for item in self._data()
-            #for tag in item['tags'] if all(TAXONOMY.isa(tag, tt) for tt in target_tags)] # listcomps are really prone to subtle logic errors
-            if all( any(TAXONOMY.isa(tag, tt) for tag in (item.get('tags') or [])) for tt in target_tags ) or not item.get('tags')]
-            
+        #return [NounSet(item) for item in self._data()
+        #    #for tag in item['tags'] if all(TAXONOMY.isa(tag, tt) for tt in target_tags)] # listcomps are really prone to subtle logic errors
[…]

### 2016-05-30 generator.py

@@ -567,8 +567,8 @@ class ZhGenerator(Generator):
                     assert(all(type(item) is str for item in measure_words_from_file))
                     candidates = measure_words_from_file
                 else:
-                    assert(type(measure_words_from_file) is dict) # I suppose it could be a number or a bool...
-                    raise Exception('M: expected str or list (YAML)')
+                    assert(type(measure_words_from_file) in [dict, type(None)]) # I suppose it could be a number or a bool...
+                    raise Exception('M: expected str or list (YAML)', noun_form.get('pinyin'))
                     
                 if utility.rand() <= 0.9:
                     measure_word = utility.pick_random(measure_words_from_file) # allows multiple M's per word

### 2016-05-30 taxonomy.py

@@ -14,8 +14,12 @@ class Taxonomy:
     '''
     def __init__(self, filename):
         self.__data = collections.defaultdict(set)
-        tree = yaml_reader.read_file(filename)
-        self.__index_tree(tree, set())
+        #tree = yaml_reader.read_file(filename)
+        #self.__index_tree(tree, set())
+        
+        tree_list = yaml_reader.read_file(filename)
+        for tree in tree_list:
+            self.__index_tree(tree, set())
         
     # not thread safe - writes to self.__data - but you wouldn't want to multithread this...
     def __index_tree(self, tree, parents):        
@@ -84,6 +88,9 @@ def test():
             self.assertTrue(t.isa('man', 'man'))
             self.assertFalse(t.isa('man', 'woman'))
             self.assertTrue(t.isa('man', 'animate'))
+            self.assertTrue(t.isa('man', 'physical.object'))
+            self.assertTrue(t.isa('physical.object', 'object'))
+            self.assertTrue(t.isa('man', 'object'))
             
         def test__common_ancestors(self):
             self.assertTrue('physical' in self.__t._common_ancestors('organ', 'animal'))
@@ -100,6 +107,7 @@ def test():
             self.assertTrue(self.__t.canbe('object', 'organ'))
             self.assertFalse(self.__t.canbe('event', 'physical'))
             self.assertFalse(self.__t.canbe('organ', 'abstract'))
+            
             
       
     suite = unittest.TestLoader().loadTestsFromTestCase(TestCase)

### 2016-05-30 doc/paper.tex
random ideas

@@
 
+AHA, what you've really done is traded in-house linguists who coded rules for outsourced linguists who churn out data.
+Is that really optimal?
+Better off bringing corpus writers back in-house, where possible.
+
 Such synthetic parallel corpora can also be viewed as benchmark tasks (need to cite Facebook bAbI).
@@
 Even the fanciest new machine learning model is dead in the water if it doesn't have enough data.
+
+disproportionate amount of attention being paid to the machine learning
+not enough attention being paid to the corpora themselves
 
@@
 Moreover, as they are upstream of GIZA++, these bins are constructed at the lexical level, not the phrase level, whereas the templates described below can make substitutions at the level of larger constituents and phrases.
+Furthermore, it is run at training time and cannot help with unknown words at test time (is this true??)
+The only real cure for unknown words is moar data
 
@@
 but there is a vast untapped source of data in the form of student problem sets, in the spirit of reCAPTCHA (citation!?)
+NLP is unique among machine learning tasks in that humans are not only consumers, but also producers of data - the alpha and the omega, the beginning and the end.
+MT is further unique among NLP in that both input and output representations are widely accepted (modulo dialectical variations) - the source and target languages themselves.
+People on the street would agree to the representations.
 

### 2016-05-30 main.py
crude wrapper script for large-scale, multiprocess generation

@@ -1,11 +1,11 @@
+import argparse
+import datetime
 import os
+import random
 
 import data # may take a while, depending on data set size
-
-import datetime
 import generator
 import nodes
-import random
 import utility
 from utility import LANGUAGES, seed_rng
 
@@ -21,7 +21,7 @@ else:
     #FIXED_ROLL = 0.75 # 0-0.8 for Clause, 0.8-0.9 for Custom, 0.9-1.0 for C1 C2 - see make_random_sentence()
     #FIXED_TEMPLATE = '把'
     SKIP_OLD_TEST = True # whether or run old test clauses; irrelevant for production
-    NUM_RANDOM_TEST = 50
+    NUM_RANDOM_TEST = 500
 
     
 
@@ -451,7 +451,7 @@ def add_random_tag_to_np(np):
     specific_tag = utility.pick_random(list(data.NOUNSET_BANK.all_tags()))
     if FIXED_NP_TAG:
         specific_tag = FIXED_NP_TAG
-    
+
     if (    all(data.TAXONOMY.canbe(specific_tag, np_tag) for np_tag in np.semantic_tags()) and # new tag not forbidden by tree
             data.NOUNSET_BANK.find_tagged(np.semantic_tags() + [specific_tag])):  # adding new tag doesn't kill all candidates
         np.add_options({'tags': [specific_tag]})
@@ -770,8 +770,9 @@ def run_test():
         o.close()
         
         
-def run_production():
-    output_prefix = datetime.datetime.isoformat(datetime.datetime.now()).replace('T', '-').replace(':', '')[:-7]
+def run_production(output_prefix):
+    if not output_prefix:
+        output_prefix = datetime.datetime.isoformat(datetime.datetime.now()).replace('T', '-').replace(':', '')[:-7]
     outputs = { lang: open('{}-generated.{}'.format(output_prefix, lang), 'w', encoding='utf8') for lang in LANGUAGES }
     for i in range(0, NUM_SENTENCES): # for large corpus, you want to "stream" the trees instead of storing them all
         # TODO: pick, say, just one random lexical node (make sure it's Noun/Verb/Adj) and blow it up - guarantee some "parallel" sentences
@@ -784,6 +785,10 @@ def run_production():
     
     
 if __name__ == '__main__':
+    parser = argparse.ArgumentParser()
+    parser.add_argument('--output') # can only specify default value 
+    args = parser.parse_args()
+
[…]

### 2016-05-30 run_production.sh

@@ -1,0 +1,16 @@
+#!/bin/bash
+
+# a multiprocess wrapper script for generating a LOT of examples
+
+NUM_CORES=$(nproc --all)
+num_threads=$((NUM_CORES * 1))
+
+jobtime=$(date +%F-%H%M%S)
+for thread in $(seq 1 $num_threads); do
+    python3 main.py --output /tmp/$jobtime.part$thread &
+done
+wait
+
+for lang in en zh; do
+    cat /tmp/$jobtime.part*.$lang > $jobtime-generated.$lang
+done

### 2016-05-30 data.py
forgot to add multiple-tag technology to NameSetBank...

@@ -151,8 +151,10 @@ class NameSetBank(WordSetBank):
     def find_tagged(self, target_tags):
         # TODO: handle single-tag case where it's not even a list?
         # TODO: share code with NounSetBank? is it worth sacrificing clarity just to DRY out 2 lines?
-        return [NameSet(item) for item in self._data() 
-            for tag in item['tags'] if all(TAXONOMY.isa(tag, tt) for tt in target_tags)]
+        #return [NameSet(item) for item in self._data() 
+        #    for tag in item['tags'] if all(TAXONOMY.isa(tag, tt) for tt in target_tags)]
+        return [item for item in self.all_namesets() 
+            for tag in item.tags() if all(TAXONOMY.isa(tag, tt) for tt in target_tags)]
             
     def _is_dummy(self, datum):
         return all(adj == None for adj in datum['nameset'].values()) 

### 2016-05-30 data.py
skip en plural and zh measure word data entry for now

@@ -232,6 +232,9 @@ class WordFormBank(Bank):
             return self.DATA_FACTORY(wf)
         else:
             return None
+            
+    def has_word(self, word):
+        return word in self._data()
         
 # TODO: need these to implement comparatives/superlatives for en
 class AdjectiveFormBank(WordFormBank):

### 2016-05-30 generator.py

@@ -501,6 +501,9 @@ class EnGenerator(Generator):
         if noun_forms and noun_forms.get('NNS'):
             result = noun_forms.get('NNS')
         else:
+            if utility.CHECK_DATABASE:
+                assert(self._noun_form_bank.has_word(noun_base)) # original concept was blank entry to verify regularity
+            
             if noun_base.endswith('y'):
                 result = noun_base[:-1] + 'ies'
             elif noun_base.endswith('s'):
@@ -553,9 +556,19 @@ class ZhGenerator(Generator):
         if target.number() == 'singular':
             noun = self._get_noun_base(target)
             noun_form = self._noun_form_bank.get(noun)
-            
-            # TODO: allow measure word omission (e.g. 这 世界 - only allowed for some words?)
-            measure_words_from_file = noun_form.get('M', '个')
+
+            # workaround to allow skipping some less important data entry for now
+            if utility.CHECK_DATABASE:
+                assert(noun_form) # would call None.get() if noun is missing from nouns_zh.yml
+                measure_words_from_file = noun_form.get('M', '个') 
+            else:
+                if noun_form: 
+                    measure_words_from_file = noun_form.get('M', '个')
+                else:
+                    assert(noun_form is None)
+                    measure_words_from_file = '个'
+
+            # TODO: allow measure word omission (e.g. 这 世界 - only allowed for some words?)            
             if measure_words_from_file == '个':
                 measure_word = '个'
             else:

### 2016-05-30 main.py

@@ -21,7 +21,7 @@ else:
     #FIXED_ROLL = 0.75 # 0-0.8 for Clause, 0.8-0.9 for Custom, 0.9-1.0 for C1 C2 - see make_random_sentence()
     #FIXED_TEMPLATE = '把'
     SKIP_OLD_TEST = True # whether or run old test clauses; irrelevant for production
-    NUM_RANDOM_TEST = 500
+    NUM_RANDOM_TEST = 50
 
     
 

### 2016-05-30 utility.py

@@ -3,6 +3,7 @@ if sys.version_info < (3, 0):
     raise Exception('This project requires Python 3 - mainly for better Unicode support')
     
 # "global" read-only variables
+CHECK_DATABASE = False # skip some data entry for now
 DATA_DIR = 'datasets/'
 LANGUAGES = ['en', 'zh']
 PRODUCTION = True

### 2016-05-30 main.py
make noun-noun modification more common, to allow non-countable nouns to sneak into the corpus

@@ -533,7 +533,7 @@ def make_random_adjp(target, **kwargs):
 
 def randomly_configure_adjp(adjp, target=None, **kwargs):
     # first determine the type you want - determiners done separately
-    if utility.rand() <= 0.95: # TODO: semantic matching to make N-N modification less awkward and confusing
+    if utility.rand() <= 0.8: # TODO: semantic matching to make N-N modification less awkward and confusing
         template_id = 'adjective'
     else:
         template_id = 'noun' 

### 2016-05-31 doc/paper.tex
okay, NOW the common-noun batch is done
hammering out bits of the first section of the introduction

@@
 \begin{abstract}
-Hello, World!
+An increased focus on
 
-Here is some normal spacing.
+is proposed.
 
-\zh{你好，世界！}
+Preliminary results do not show much improvement.
+This is because we need to scale up?
+Future work is proposed.
 
-Scalable parallel corpus generation
 \end{abstract}
 
-Curiosity: how many AMTA pages are the EMNLP instructions? i.e., how long would my 4-page EMNLP paper be?
+\section{Motivation} % Previously called ``Introduction''
 
-\section{Introduction}
+As early as Jelinek's (in)famous 1980's quote, ``{\em Every time I fire a linguist, my performance goes up},'' both linguists and linguistics have been increasingly marginalized in natural language processing.
+More recently, for example, \mynewcite{collobert 2011 footnote 8} mention the possibility of ``{\em [ideally]...to learn from letter sequences rather than words},'' illustrating an AI-motivated desire to move away from even fundamental linguistic concepts.
+A specific case of this trend is the gradual shift away from classical linguistic rule-based machine translation (MT) \mycitep{hutchins} to  contemporary machine learning approaches, such as staistical phrase-based systems like Moses \mycitep{moses} and more recent deep learning approaches \mycitep{sutskever, bengio}.
 
-RNN encoder-decoders are not the first corpus-based MT methodology (EBMT then SMT then this), and it is unlikely in this era of Big Data that they will be the last.
-Focusing on corpus generation allows linguists to focus on capturing the most statistically prevalent linguistic phenomena, while insulating them from the strengths and weaknesses of various implementations.
-For instance, EBMT suffers from relatively poor generalization and SMT has problems with nonlocality.
-(RNN built on relatively shaky ground, since word vectors don't QUITE make sense for function words?)
-Seems somewhat foolhardy to marginalize the domain experts
-Significantly more engineering (man-hours?) seems to go into the machine-learning component.
-(Well, granted, there are human translators who are spending many man-hours generating sentence pairs.
-But they're not intended for MT.)
-Traditionally, corpus construction has been considered prohibitively expensive, ``scavenged'' from other sources like parliamentary proceedings or lecture subtitles.
-But must this be so? Vast untapped resource of students or whatever.
+In the end, though, machine learning is learning from data, and even the most bleeding-edge MT system relies crucially on multilingual parallel corpora.
+In this paper, we propose recalling linguists from their increasing exile from MT to handle what they know best: language.
+Specifically, we believe that machine translation could benefit from high-quality, ``designer'' parallel corpora that have been hand-crafted by linguists.
+This could take the form of amplified training sets in the short run, and perhaps a ``universal translation memory'' in the long run.
 
-AHA, what you've really done is traded in-house linguists who coded rules for outsourced linguists who churn out data.
-Is that really optimal?
-Better off bringing corpus writers back in-house, where possible.
+This division of labor would let linguists be linguists, with MT system development proceeding in parallel, consuming the linguists' work downstream.
 
-Such synthetic parallel corpora can also be viewed as benchmark tasks (need to cite Facebook bAbI).
-This is actually probably more important than you'd think, because looking at IWSLT15, some of the human translations are VERY loose.
-So there's another element of noise in the data, and it's one that no amount of downstream machine learning can get rid of.
-(Although people have tried corpus evaluation...)
+UGH I have nothing really concrete to show here... this is like an opinion paper at this point...
+Although I guess I do have this stupid system I worked on, and I'll push it to the web, and they can laugh at how pathetic it is.
 
-One might conjecture that the fundamental problem in natural language processing is one of data representation
+Significantly more developer-months go into the MT system.\footnote{Of course, hundreds of translator-hours }
 
-With the ALPAC report's pessimistic outlook on MT arguably responsible for sparking decades of fundamental linguistics research in the tradition of Chomsky, it is only fitting that we now bring linguistics back?
-No, my Return of the Linguist had more to do with how RBMT systems have fallen out of favor in research
+niceties like sentence alignment  
+(Well, granted, there are human translators who are spending many man-hours generating sentence pairs. But they're not intended for MT.)
+There have also been attempts to
 
@@
 
-Return of the linguist: it is relatively easy to come up with examples that industrial-grade systems fail. (hmm,)
-he was shown the car
-他 给 展示 了 车?
+I think this first part of the introduction should be short, and then some subsections below?
+
+Traditionally, corpus construction has been considered prohibitively expensive, ``scavenged'' from other sources like parliamentary proceedings or lecture subtitles.
+But must this be so? Vast untapped resource of students or whatever.
+
+No, my Return of the Linguist had more to do with how RBMT systems have fallen out of favor in research
+banishment has been premature? weren't using the linguist in the right way?
 
 \section{Single-Line Scoring}
+
+The machine translation task we consider in this paper is the IWSLT shared task
+
+A brief analysis of corpus quality
+
+\mycitep{multeval}
+
+Do I want to display an example bad-sentence pair? 
+Some of the lines are duplicated. Examining 
+Just mention for now that there is an entire lecture that is bad
+Also hand-purged several other sentences
+
+TODO: table of scores after purging
+
+This results in a significant TER score change that is almost as large as the score difference between the highest and lowest ranking system from the evaluation campaign.  
+Interestingly, BLEU scores stayed relatively robust.
 
@@
 It also demonstrates the tremendous imbalance between linguistic effort and machine learning effort/investment
+
+\subsection{Upsides, Potential Applications}
+
+Such synthetic parallel corpora can also be viewed as benchmark tasks (need to cite Facebook bAbI).
+This is actually probably more important than you'd think, because looking at IWSLT15, some of the human translations are VERY loose.
+So there's another element of noise in the data, and it's one that no amount of downstream machine learning can get rid of.
+(Although people have tried corpus evaluation...)
+
+% interesting, but probably confusing to readers?
+%In the language of agile software development, test-driven MT system development, in which linguists pose sentence pairs, and MT developers work to translate them
+
+Focusing on corpus generation would allow linguists to focus their attention on significant linguistic phenomena
+
+methods every few years; linguistic theories, representations every decade or so; parallel corpora live forever
+
+lexical ambiguity is naturally alleviated
+
+RBMT - countless developer-years of expertise
+
+unlike all the different competing methods, it seems that a ``universal translation memory'' would be inherently collaborative (methods are competitive by nature, whereas data can be collaborative)
+
+%, and with the two meeting at the golden railroad spike of the parallel corpus. Apparently I've liked this stupid thing for 10+ years
+% on capturing the most statistically prevalent linguistic phenomena, while insulating them from the strengths and weaknesses of various MT implementations.
+(wait, is that a good thing?
+don't you need some feedback as to what KIND of corpus the MT system needs?)
+%With the ALPAC report's pessimistic outlook on MT arguably responsible for sparking decades of fundamental linguistics research in the tradition of Chomsky, it is only fitting that we now bring linguistics back?
+%RNN encoder-decoders are not the first corpus-based MT methodology (EBMT then SMT then this), and it is unlikely in this era of Big Data that they will be the last.
+% For instance, EBMT suffers from relatively poor generalization and SMT has problems with nonlocality.
+% (RNN built on relatively shaky ground, since word vectors don't QUITE make sense for function words?)
+% Seems somewhat foolhardy to marginalize the domain experts (no, lots of deep learning advances have been made by IGNORING expert-created representations)
+% One might conjecture that the fundamental problem in natural language processing is one of data representation
+%Return of the linguist: it is relatively easy to come up with examples that industrial-grade systems fail. (hmm,)
+%he was shown the car \zh{他 给 展示 了 车}?
 
@@
 
-stanford POS tagger (incomplete bibtex entry?) \mycitep{postagger:2003}
-
 This aspect of our work is perhaps closest in spirit to the equivalence classes used by \mynewcite{brown:99} to reduce the data requirements of an EBMT system \mycitep{ebmtbook}.
@@
 % USUALLY alignment only, CAN be applied to translation... http://www.statmt.org/moses/?n=Advanced.Models#ntoc6
+% uh, so do people already do this??
 With some effort, you can use these bins during steps other than alignment, although this is not done by default.
@@
 maybe the latest Hutchins overview or nutshell?
-
-fire a linguist and accuracy goes up
-
-collobert and weston: learn words from letters. interesting from an AI perspective, seems unlikely to improve performance in the near future
 
@@
 
+used for corpus analysis stanford POS tagger (incomplete bibtex entry?) \mycitep{postagger:2003}
+
 \subsection{Simplistic Linguistics}
@@
 
+Out of domain corpus
+
 citations needed
@@
 
+The corpus, which is freely available, consists of multilingual subtitles of TED talks, which are prepared lectures on a wide variety of subjects.
+
 IWSLT: TED talks are a series of lectures, which makes them something intermediate between the conversational language of movie subtitles and the formal written language of UN resolutions.
-Should I mention that you can kind of discern this from their POS distributions?
-I would probably have to mention something concrete then...
-But yeah, I guess it's worth mentioning if I have space, which I almost certainly do for AMTA
 
@@
 
-\section{Future Work}
+\section{Proposed Future Work}
 
 I.e., shit I REALLY doubt I have time for...
-
-For simplicity, the current implementation is lexically one-to-one; for scalability in the number of languages, you ideally want many-to-many.
 
@@
 
-Localization of semantic taxonomy - this would have to be one-to-one, and you'd think it would clutter up the raw text UI...
+crowdsourcing for ``amateur linguists''
+
+AHA, what you've really done is traded in-house linguists who coded rules for outsourced linguists who churn out data.
+Is that really optimal?
+Better off bringing corpus writers back in-house, where possible.
+
+estimate about 1 noun per minute (my pathetic work speed)
 
@@
 
+In the short term
+Increasing lexical variation by expanding the lexicon, especially for uncommon words in the training corpus, to alleviate data sparsity.
+Increasing syntactic variation in a high-quality way by harvesting more templates, in the old EBMT tradition
+Work along these lines is currently underway.
+
 \amtaonly{
@@
 If I really am submitting to this conference too, then I can afford (and probably want to?) draw out the length,
+
+Should I mention that you can kind of discern differences between IWSLT, movie subtitles, and MultiUN from their POS distributions? I would probably have to mention something concrete then...? 
 } % end \amtaonly
@@
 \section{HCOMP}
+
+wait, the HCOMP ``short paper'' needs to be 2 pages at MOST, including references
 

### 2016-05-31 doc/paper.tex
section 1.0: introduction/motivation

@@
 
+A particular system for doing this is presented.
+It allows syntactic complexity via rules.
+It also allows high-quality examples via templates, with minor syntactic modifications via rules.
+
 Preliminary results do not show much improvement.
 This is because we need to scale up?
+
 Future work is proposed.
+In particular, we believe this is amenable to crowdsourcing by ``amateur linguists''.
 
@@
 
-\section{Motivation} % Previously called ``Introduction''
+\section{Motivation}
 
 As early as Jelinek's (in)famous 1980's quote, ``{\em Every time I fire a linguist, my performance goes up},'' both linguists and linguistics have been increasingly marginalized in natural language processing.
-More recently, for example, \mynewcite{collobert 2011 footnote 8} mention the possibility of ``{\em [ideally]...to learn from letter sequences rather than words},'' illustrating an AI-motivated desire to move away from even fundamental linguistic concepts.
-A specific case of this trend is the gradual shift away from classical linguistic rule-based machine translation (MT) \mycitep{hutchins} to  contemporary machine learning approaches, such as staistical phrase-based systems like Moses \mycitep{moses} and more recent deep learning approaches \mycitep{sutskever, bengio}.
+More recently, for example, \mynewcite{collobert 2011 footnote 8} mention the possibility of ``{\em [ideally]...to learn from letter sequences rather than words},'' illustrating an AI-motivated desire to move away from even the most fundamental of linguistic concepts.
+A specific instance of this overall trend is the gradual shift of machine translation (MT) away from classical linguistic rule-based approaches \mycitep{hutchins} to contemporary machine learning approaches, which underlie statistical phrase-based systems like Moses \mycitep{moses} and more recent neural network systems \mycitep{sutskever, bengio}.
+But in the end, even the most cutting-edge machine learning MT system relies crucially on multilingual parallel corpora, typically taken from external sources like multilingual parliamentary proceedings.
+This results in what we perceive to be a disproportionate amount of effort being dedicated to the machine learning component of machine translation, with the ever-important input corpora being treated almost as an afterthought.\footnote{
+    Of course, these parallel corpora represent hundreds of translator-hours of work. 
+    However, these translations are generally optimized for human consumption, not machine learning, with no regard for such niceties as sentence alignment.
+    This is illustrated briefly in Section \ref{case_study}.
+    }
 
-In the end, though, machine learning is learning from data, and even the most bleeding-edge MT system relies crucially on multilingual parallel corpora.
-In this paper, we propose recalling linguists from their increasing exile from MT to handle what they know best: language.
-Specifically, we believe that machine translation could benefit from high-quality, ``designer'' parallel corpora that have been hand-crafted by linguists.
-This could take the form of amplified training sets in the short run, and perhaps a ``universal translation memory'' in the long run.
+In this paper, we propose bringing linguists back from a premature exile to craft high-quality, ``designer'' parallel corpora for machine translation.
+Such a division of labor would let linguists be linguists, allowing them to focus on corpus quality, size, and coverage, with MT system development proceeding in parallel.  
+In a sense, the use of external corpora has ``outsourced'' the rule-based MT linguist's job to third-party translators, who implicitly encode linguistic information in parallel corpora.  
+What we propose is to bring these jobs back ``in-house'', in the updated context of corpus-based MT.
 
-This division of labor would let linguists be linguists, with MT system development proceeding in parallel, consuming the linguists' work downstream.
+% TODO: i haven't talked yet about 
+% oh is that the ``upsides'' section?
 
-UGH I have nothing really concrete to show here... this is like an opinion paper at this point...
-Although I guess I do have this stupid system I worked on, and I'll push it to the web, and they can laugh at how pathetic it is.
+\subsection{Case Study}
+\label{case_study}
 
-Significantly more developer-months go into the MT system.\footnote{Of course, hundreds of translator-hours }
-
-niceties like sentence alignment  
-(Well, granted, there are human translators who are spending many man-hours generating sentence pairs. But they're not intended for MT.)
-There have also been attempts to
-
-My main point seems to be that nobody makes high quality and quantity parallel corpora JUST for machine translation, and it seems like that's sorely needed for both training and benchmarking.
-Even the fanciest new machine learning model is dead in the water if it doesn't have enough data.
-
-disproportionate amount of attention being paid to the machine learning
-not enough attention being paid to the corpora themselves
-
-I think this first part of the introduction should be short, and then some subsections below?
-
-Traditionally, corpus construction has been considered prohibitively expensive, ``scavenged'' from other sources like parliamentary proceedings or lecture subtitles.
-But must this be so? Vast untapped resource of students or whatever.
-
-No, my Return of the Linguist had more to do with how RBMT systems have fallen out of favor in research
-banishment has been premature? weren't using the linguist in the right way?
-
-\section{Single-Line Scoring}
-
-The machine translation task we consider in this paper is the IWSLT shared task
+As a concrete example, we briefly consider the official baseline results from the Chinese to English IWSLT shared task studied in Section \ref{experiments}.
 
@@
 
-TODO: table of scores after purging
+TODO: table of scores after purging. This is pretty crucial
 
@@
 
+More distressing is the fact that this data set has been used as the official development/tuning set for the past several IWSLT evaluation campaigns.
+
+It is unclear what effect such noisy corpora would have on other systems, 
+especially when used for tuning
+
 \subsection{Upsides, Potential Applications}
+
+My main point seems to be that nobody makes high quality and quantity parallel corpora JUST for machine translation, and it seems like that's sorely needed for both training and benchmarking.
+
+Traditionally, corpus construction has been considered prohibitively expensive, ``scavenged'' from other sources like parliamentary proceedings or lecture subtitles.
+But we argue here that it need not be so expensive.
 
@@
 
-Focusing on corpus generation would allow linguists to focus their attention on significant linguistic phenomena
+This could take the form of amplified training sets in the short run, and perhaps a ``universal translation memory'' in the long run.
 
@@
 
-\subsection{Simplistic Linguistics}
+%\subsection{Simplistic Linguistics}
 We adopt the exceedingly simplistic view of language as nouns, verbs, and descriptions (modifications) thereof.
@@
 
-I guess I should use placeholder graphics for now, since deadlines for camera-ready are quite a bit later.
-EMNLP: 6/3 submission, 7/29 notification, 9/23 camera-ready
-AMTA: 6/27 submission, 8/8 notification, 9/12 camera-ready
-
 Testing section reference: \ref{sec:related} - this requires changing secnumdepth=2 in AAAI
 
-\section{Experimental Methods}
+%\subsection{Implementation Details}
+Python
+YAML for data: very well suited. full Unicode support, and you don't even need to type quotation marks to enclose strings
+Crude though it may be, we plan on eventually open-sourcing both the generator and its accompanying data
+
+\section{Preliminary Experiments}
+\label{experiments}
+
+\subsection{Methods}
+As mentioned in 
+The IWSLT 2015 version of the WIT$^3$ corpus, which consists of multilingual subtitles of TED talks, covering a wide variety of subjects.
+
+As prepared lectures, they fall somewhere between the conversational language of movie subtitles and the formal written language of UN resolutions.
+The latter are freely available online from the OPUS project \mycitep{opus}
+
+We consider the Chinese to English machine translation task, using the
 
@@
 citations needed
-IWSLT - point out that this corpus is freely available
-OPUS zh\_zh
-MultiUN
-Moses
+IWSLT \mycitep{wit3} - point out that this corpus is freely available
+OPUS zh\_zh \mycitep{opensubtitles} MultiUN\mycitep{multiun} 
+Moses \mycitep{moses}
+KenLM \mycitep{kenlm}, which we found to be faster and stabler at decoding time.
 
@@
 
-The corpus, which is freely available, consists of multilingual subtitles of TED talks, which are prepared lectures on a wide variety of subjects.
+The corpus, which is freely available, consists
 
-IWSLT: TED talks are a series of lectures, which makes them something intermediate between the conversational language of movie subtitles and the formal written language of UN resolutions.
+IWSLT: TED talks are a series of lectures, which makes them
 
@@
 
-\section{Results}
+\subsection{Results}
 
@@
 
+perhaps using example-based MT techniques \mycitep{ebmtbook} and crowdsourcing to generate them at scale.
+
 For this paper, the lexical database was drawn from the IWSLT training set to maximize bottom-line performance.
@@
 
-crowdsourcing for ``amateur linguists''
+expert-sourcing by linguists, using the computer to lever up - computer assisted translation, in a sense; maybe computer-amplified
+crowdsourcing by ``amateur linguists''
 
-AHA, what you've really done is traded in-house linguists who coded rules for outsourced linguists who churn out data.
-Is that really optimal?
-Better off bringing corpus writers back in-house, where possible.
+But must this be so? Vast untapped resource of students or whatever.
 
 estimate about 1 noun per minute (my pathetic work speed)
+
+locally weighted linear regression? online learning?
+but corpus generation at test time is probably just as noisy as any discriminative method
 
@@
 Work along these lines is currently underway.
+
+% TODO: Wit^3 is not fully capitalized in stupid bibliography
+% TODO: wtf, ``chinese'' is not capitalized in stupid bibliography
 

### 2016-05-31 doc/paper.tex
halfway through tst2010 case study

@@
 
-In this paper, we propose bringing linguists back from a premature exile to craft high-quality, ``designer'' parallel corpora for machine translation.
-Such a division of labor would let linguists be linguists, allowing them to focus on corpus quality, size, and coverage, with MT system development proceeding in parallel.  
+In this paper, we propose bringing linguists back from a premature exile to craft high-quality, ``designer'' synthetic parallel corpora for machine translation.
+Such a division of labor would let linguists be linguists, with MT system development proceeding in parallel.  
 In a sense, the use of external corpora has ``outsourced'' the rule-based MT linguist's job to third-party translators, who implicitly encode linguistic information in parallel corpora.  
@@
 
-% TODO: i haven't talked yet about 
+% TODO: i haven't talked yet about the unreasonable effectiveness of in-domain data...
 % oh is that the ``upsides'' section?
 
+% this is some stupid little side note I just noticed a couple weeks ago, but it is expanding to a huge size, because it's like the only concrete ``positive'' result I have...
 \subsection{Case Study}
@@
 
-As a concrete example, we briefly consider the official baseline results from the Chinese to English IWSLT shared task studied in Section \ref{experiments}.
+As a concrete example to illustrate the need for higher-quality parallel corpora, we briefly consider the official Chinese to English baseline translation of the {\small \tt tst2010} data set\footnote{https://wit3.fbk.eu/score.php?release=2014-01} from the IWSLT shared task studied in Section \ref{experiments}.
+As a very crude way of assessing translation quality, we used the official evaluation software, MultEval \mycitep{multeval} to compute scores for each translated line individually.
+Upon manual examination, it was found that the lines with the worst individual TER scores \mycitep{ter}, some as high as 300\%, almost all corresponded to significant translation errors in the Chinese side of the gold corpus.
 
-A brief analysis of corpus quality
+Not only misaligned sentences, but also apparent data entry errors,
 
-\mycitep{multeval}
+Unreasonable for a human to duplicate, much less MT systems.
 
-Do I want to display an example bad-sentence pair? 
-Some of the lines are duplicated. Examining 
+One would expect such outliers to be relatively well-tolerated in training sets
+But the distressing part is that the {\small \tt tst2010} data set has been used as the official development/tuning set for the past several IWSLT evaluation campaigns.
+It is unclear what effect such noisy corpora would have on other systems, 
+especially when used for tuning
+This results in a significant TER score change that is almost as large as the score difference between the highest and lowest ranking system from the evaluation campaign.  
+BLEU scores, on the other hand, did not change much.
+
+Removing TED talk number 767 (over 200 lines), as well as 15 hand-picked lines with bad TER scores,
+
+% TODO: table and reference
+
+% TODO: display an example bad-sentence pair?
+
+Examining
+
 Just mention for now that there is an entire lecture that is bad
@@
 
-This results in a significant TER score change that is almost as large as the score difference between the highest and lowest ranking system from the evaluation campaign.  
-Interestingly, BLEU scores stayed relatively robust.
+It is a testament to how indisidious these errors can be that they persisted this long (tst2010 was distributed as recently as IWSLT2015
 
-It is a testament to how indisidious these errors can be that they persisted this long (tst2010 was distributed as recently as IWSLT2015
-This underscores the need for human post-editing, at the very least, for test sets.
-In a shared evaluation task with many language pairs, the most practical way for this is probably crowdsourcing.
+Not even an international conference is immune to such issues
 
@@
 
-More distressing is the fact that this data set has been used as the official development/tuning set for the past several IWSLT evaluation campaigns.
-
-It is unclear what effect such noisy corpora would have on other systems, 
-especially when used for tuning
+Higher corpus quality is desirable.
+Larger in-domain corpus quantity is desirable too - see learning curve controls below?
 
@@
 
-%, and with the two meeting at the golden railroad spike of the parallel corpus. Apparently I've liked this stupid thing for 10+ years
-% on capturing the most statistically prevalent linguistic phenomena, while insulating them from the strengths and weaknesses of various MT implementations.
-(wait, is that a good thing?
-don't you need some feedback as to what KIND of corpus the MT system needs?)
-%With the ALPAC report's pessimistic outlook on MT arguably responsible for sparking decades of fundamental linguistics research in the tradition of Chomsky, it is only fitting that we now bring linguistics back?
-%RNN encoder-decoders are not the first corpus-based MT methodology (EBMT then SMT then this), and it is unlikely in this era of Big Data that they will be the last.
-% For instance, EBMT suffers from relatively poor generalization and SMT has problems with nonlocality.
-% (RNN built on relatively shaky ground, since word vectors don't QUITE make sense for function words?)
-% Seems somewhat foolhardy to marginalize the domain experts (no, lots of deep learning advances have been made by IGNORING expert-created representations)
-% One might conjecture that the fundamental problem in natural language processing is one of data representation
-%Return of the linguist: it is relatively easy to come up with examples that industrial-grade systems fail. (hmm,)
-%he was shown the car \zh{他 给 展示 了 车}?
+% moreover, you're at the mercy of your third-party translators - if you want more data, you gotta wait.
 
@@
 
-Nice tip: ONE LINE PER SENTENCE.
+The basic
+
+The idea underlying generation system was essentially proposed 
+This aspect of our work is perhaps closest in spirit to the equivalence classes used by \mynewcite{brown:99} to reduce the data requirements of an EBMT system \mycitep{ebmtbook}.
+
+system described below is to have linguists write a template and then levering up with the computer to add lexical and syntactic variation.  
+As a simple example, 
+{\small \tt The NP jumped over the moon}
 
 Unlike more AI-oriented or natural language understanding approaches, we are solely concerned here with the generation of surface forms
-
-At some point, I will almost definitely be citing Moses \mycitep{moses}
-
-the latest version of the Stanford Segmenter \mycitep{segmenter:2005,segmenter:2008}
-
-This aspect of our work is perhaps closest in spirit to the equivalence classes used by \mynewcite{brown:99} to reduce the data requirements of an EBMT system \mycitep{ebmtbook}.
 
@@
 
+the latest version of the Stanford Segmenter \mycitep{segmenter:2005,segmenter:2008}
+
 citations needed
@@
 
+Crowdsourcing also has its place in post-editing, at the very least, for test sets.
+In a shared evaluation task with many language pairs, the most practical way for this is probably crowdsourcing.
+
 But must this be so? Vast untapped resource of students or whatever.
@@
 Work along these lines is currently underway.
+
+%RNN encoder-decoders are not the first corpus-based MT methodology (EBMT then SMT then this), and it is unlikely in this era of Big Data that they will be the last.
 

### 2016-05-31 doc/paper.tex
sloppy looking table seems acceptable in 2015 proceedings

@@
 
-As early as Jelinek's (in)famous 1980's quote, ``{\em Every time I fire a linguist, my performance goes up},'' both linguists and linguistics have been increasingly marginalized in natural language processing.
+As early as Jelinek's (in)famous quote from the 1980's that ``{\em Every time I fire a linguist, my performance goes up},'' both linguists and linguistics have been increasingly marginalized in natural language processing.
 More recently, for example, \mynewcite{collobert 2011 footnote 8} mention the possibility of ``{\em [ideally]...to learn from letter sequences rather than words},'' illustrating an AI-motivated desire to move away from even the most fundamental of linguistic concepts.
@@
 
+% TODO: don't need to go into this much detail, do you? This would be the first section to trim, if it runs OVER length
+% To find an example of suboptimally translated parallel corpora in the literature, one need not look far.
 As a concrete example to illustrate the need for higher-quality parallel corpora, we briefly consider the official Chinese to English baseline translation of the {\small \tt tst2010} data set\footnote{https://wit3.fbk.eu/score.php?release=2014-01} from the IWSLT shared task studied in Section \ref{experiments}.
@@
 
-One would expect such outliers to be relatively well-tolerated in training sets
-But the distressing part is that the {\small \tt tst2010} data set has been used as the official development/tuning set for the past several IWSLT evaluation campaigns.
+% TODO: table with change; also list best and worst 
+% caption should mention that the difference between 
+    % ALTHOUGH you would expect that it would improve all models, it's not clear whether 
+Removing TED talk number 767 (over 200 lines), as well as 15 hand-picked lines with bad TER scores,
+
+First, note that the scores improve, which is consistent with the observation that
+
+good machine translations of bad Chinese sentences.
+
+This results in a significant TER score change that is almost as large as the score difference between the highest and lowest ranking system from the evaluation campaign.  
+BLEU scores \mycitep{bleu}, on the other hand, did not change much.
+
+One would expect such errors to be relatively well-tolerated in large training corpora
+But one bad 200-line talk constitutes a significant fraction of {\small \tt tst2010}, which is only 1500 lines total.
+More distressing is the fact is that the {\small \tt tst2010} data set has been used as the official development/tuning set for the past several IWSLT evaluation campaigns.
 It is unclear what effect such noisy corpora would have on other systems, 
 especially when used for tuning
-This results in a significant TER score change that is almost as large as the score difference between the highest and lowest ranking system from the evaluation campaign.  
-BLEU scores, on the other hand, did not change much.
-
-Removing TED talk number 767 (over 200 lines), as well as 15 hand-picked lines with bad TER scores,
 
 % TODO: table and reference
+\begin{table}[]
+\centering
+\label{table:tst2010}
+\begin{tabular}{|l|l|l|}
+\hline
+ & BLEU & TER  \\ \hline
+Full {\small \tt tst2010} & 11.2 & 77.0  \\ \hline
+Purged {\small \tt tst2010} & 11.8 & 71.3  \\ \hline
+%\hline
+%{\small \tt tst2014} &  &   \\ \hline
+% &  &   \\ \hline
+\end{tabular}
+\caption{By comparison, the difference between the baseline TER score and (not shown) } % TODO: caption
+\end{table}
 
@@
 
+Translation is perhaps the only NLP task for which millions of people already undergo years of training
+
+Unlike many machine learning tasks (image recognition, finance), NLP data are purely artificial/manmade.
+New paradigm?
+Not enough attention has been paid IMO to semi-automated corpus generation.
+Well, typical approaches are paid, like Mechanical Turk...
+but there is a vast untapped source of data in the form of student problem sets, in the spirit of reCAPTCHA (citation!?)
+NLP is unique among machine learning tasks in that humans are not only consumers, but also producers of data - the alpha and the omega, the beginning and the end.
+MT is further unique among NLP in that both input and output representations are widely accepted (modulo dialectical variations) - the source and target languages themselves.
+People on the street would agree to the representations.
+
 \section{Related Work}
@@
 The only real cure for unknown words is moar data
-
-some neural network MT paper?
-
-aha, yoshua bengio's "curse of dimensionality" - becomes a blessing?  
-Unlike many machine learning tasks (image recognition, finance), NLP data are purely artificial/manmade.
-New paradigm?
-Not enough attention has been paid IMO to semi-automated corpus generation.
-Well, typical approaches are paid, like Mechanical Turk...
-but there is a vast untapped source of data in the form of student problem sets, in the spirit of reCAPTCHA (citation!?)
-NLP is unique among machine learning tasks in that humans are not only consumers, but also producers of data - the alpha and the omega, the beginning and the end.
-MT is further unique among NLP in that both input and output representations are widely accepted (modulo dialectical variations) - the source and target languages themselves.
-People on the street would agree to the representations.
-
-Translation is perhaps the only NLP task for which millions of people already undergo years of training
 
@@
 Evaluated using the IWSLT 2015 progress data set (tst2014) to allow comparison with both 2015 and 2014 systems.
+
+training curves (just to prove my controls aren't wrong)
+my synthetic corpora don't help. 
+but in-domain corpus helps significantly more than out-of-domain. (this is a commonly reported result)
+
+Bottom line
+My corpora don't really seem to help, but neither do out-of-domain corpora.
+
+this will easily take up a whole column, because it includes 2 tables...
 

### 2016-06-01 doc/paper.tex
added some references

@@
 As early as Jelinek's (in)famous quote from the 1980's that ``{\em Every time I fire a linguist, my performance goes up},'' both linguists and linguistics have been increasingly marginalized in natural language processing.
-More recently, for example, \mynewcite{collobert 2011 footnote 8} mention the possibility of ``{\em [ideally]...to learn from letter sequences rather than words},'' illustrating an AI-motivated desire to move away from even the most fundamental of linguistic concepts.
-A specific instance of this overall trend is the gradual shift of machine translation (MT) away from classical linguistic rule-based approaches \mycitep{hutchins} to contemporary machine learning approaches, which underlie statistical phrase-based systems like Moses \mycitep{moses} and more recent neural network systems \mycitep{sutskever, bengio}.
+More recently, for example, \mynewcite{collobert:2011} mention the possibility of ``{\em [ideally]...to learn from letter sequences rather than words},'' illustrating an AI-motivated desire to move away from even the most fundamental of linguistic concepts.
+A specific instance of this overall trend is the gradual shift of machine translation (MT) away from classical linguistic rule-based approaches \mycitep{hutchins:2005} to contemporary machine learning approaches, which underlie statistical phrase-based systems like Moses \mycitep{moses} and more recent neural network systems \mycitep{sutskever:2014,bengio:2014}.
 But in the end, even the most cutting-edge machine learning MT system relies crucially on multilingual parallel corpora, typically taken from external sources like multilingual parliamentary proceedings.
@@
     % ALTHOUGH you would expect that it would improve all models, it's not clear whether 
-Removing TED talk number 767 (over 200 lines), as well as 15 hand-picked lines with bad TER scores,
+%Removing TED talk number 767 (over 200 lines), as well as 15 hand-picked lines with bad TER scores,
 
-First, note that the scores improve, which is consistent with the observation that
+%First, note that the scores improve, which is consistent with the observation that 
+% something to keep in mind in case reviewers ask
 
 good machine translations of bad Chinese sentences.
-
-This results in a significant TER score change that is almost as large as the score difference between the highest and lowest ranking system from the evaluation campaign.  
-BLEU scores \mycitep{bleu}, on the other hand, did not change much.
 
@@
 
+%This results in a significant TER score change that is almost as large as the score difference between the highest and lowest ranking system from the evaluation campaign.  
+%BLEU scores \mycitep{bleu}, on the other hand, did not change much.
+
+% no, omit this, because DUH, if you change the test set, the scores will change. (it's like changing your ruler.)
 % TODO: table and reference
-\begin{table}[]
-\centering
-\label{table:tst2010}
-\begin{tabular}{|l|l|l|}
-\hline
- & BLEU & TER  \\ \hline
-Full {\small \tt tst2010} & 11.2 & 77.0  \\ \hline
-Purged {\small \tt tst2010} & 11.8 & 71.3  \\ \hline
+%\begin{table}[]
+%\centering
+%\label{table:tst2010}
+%\begin{tabular}{|l|l|l|}
 %\hline
-%{\small \tt tst2014} &  &   \\ \hline
-% &  &   \\ \hline
-\end{tabular}
-\caption{By comparison, the difference between the baseline TER score and (not shown) } % TODO: caption
-\end{table}
+% & BLEU & TER  \\ \hline
+%Full {\small \tt tst2010} & 11.2 & 77.0  \\ \hline
+%Purged {\small \tt tst2010} & 11.8 & 71.3  \\ \hline
+%%\hline
+%%{\small \tt tst2014} &  &   \\ \hline
+%% &  &   \\ \hline
+%\end{tabular}
+%\caption{By comparison, the difference between the baseline TER score and (not shown) } % TODO: caption
+%\end{table}
 

### 2016-06-01 doc/paper.tex
section 1.1: case study (just stupid tst2010 - at least SOMETHING concrete)

@@
 \begin{abstract}
-An increased focus on
+An increased focus on a
 
@@
 
-As early as Jelinek's (in)famous quote from the 1980's that ``{\em Every time I fire a linguist, my performance goes up},'' both linguists and linguistics have been increasingly marginalized in natural language processing.
+As early as Jelinek's (in)famous quote during the 1980's that ``{\em Every time I fire a linguist, my performance goes up},'' both linguists and linguistics have been increasingly marginalized in natural language processing.
 More recently, for example, \mynewcite{collobert:2011} mention the possibility of ``{\em [ideally]...to learn from letter sequences rather than words},'' illustrating an AI-motivated desire to move away from even the most fundamental of linguistic concepts.
@@
 But in the end, even the most cutting-edge machine learning MT system relies crucially on multilingual parallel corpora, typically taken from external sources like multilingual parliamentary proceedings.
-This results in what we perceive to be a disproportionate amount of effort being dedicated to the machine learning component of machine translation, with the ever-important input corpora being treated almost as an afterthought.\footnote{
+This results in what we perceive to be a disproportionate amount of effort being dedicated to the machine learning component of machine translation, with input corpora being treated almost as an afterthought.\footnote{
     Of course, these parallel corpora represent hundreds of translator-hours of work. 
@@
 % To find an example of suboptimally translated parallel corpora in the literature, one need not look far.
-As a concrete example to illustrate the need for higher-quality parallel corpora, we briefly consider the official Chinese to English baseline translation of the {\small \tt tst2010} data set\footnote{https://wit3.fbk.eu/score.php?release=2014-01} from the IWSLT shared task studied in Section \ref{experiments}.
-As a very crude way of assessing translation quality, we used the official evaluation software, MultEval \mycitep{multeval} to compute scores for each translated line individually.
-Upon manual examination, it was found that the lines with the worst individual TER scores \mycitep{ter}, some as high as 300\%, almost all corresponded to significant translation errors in the Chinese side of the gold corpus.
 
-Not only misaligned sentences, but also apparent data entry errors,
+As an example of suboptimal parallel corpora being used for machine translation, consider the official Chinese to English baseline translation of the {\small \tt tst2010} data set\footnote{https://wit3.fbk.eu/score.php?release=2014-01} from the TED talks corpus, described in greater detail in Section \ref{experiments}.
+As a very crude way of assessing translation quality, we used the official evaluation software MultEval \mycitep{multeval} to compute scores for each translated line individually.
+Upon manual inspection, the worst scoring individual lines (with TER scores over 150\%) were almost all found to correspond to poor human translations on the Chinese side of the gold corpus.\footnote{
+    High TER scores indicated misaligned sentences as well as clerical errors.
+In particular, the lecture with {\small \tt talkid} 767 had dozens of erroneous lines, in some of the worst cases having the same Chinese sentence copy/pasted three times on the same line.
+    The baseline Moses system tries valiantly, but it ultimately scores poorly for trying to produce a good machine translation of a poor human translation.
+    Alternatively, one can interpret these errors as an additional source of noise in the human encoding of English into Chinese.
+}
+Such translations might suffice for viewing translated videos monolingually, but they are far from ideal for training a machine translation system.
 
-Unreasonable for a human to duplicate, much less MT systems.
+One might expect such errors to be relatively well-tolerated in large training corpora. 
+Nevertheless, this is particularly unsettling because the {\small \tt tst2010} data set has been used as the official development/tuning set for the past several IWSLT evaluation campaigns. 
+It is also unclear what effect such noisy test corpora have on MT system evaluation and comparison.
+
+In any case, the fact that these errors have persisted this long ({\small \tt tst2010} has been distributed with only minor changes for every IWSLT Chinese-English evaluation campaign since 2012) is a testament to how insidious these errors can be, even for a leading international conference.
+At the very least, this suggests a need for more careful post-editing of bitexts, particularly for development and test corpora.
+More ideal, though, would be the creation of parallel corpora for the express purpose of training, tuning, and evaluating data-driven MT systems.
+
+% TODO: display an example bad-sentence pair? If I don't, then what was the point of getting zh fonts working?? sigh
+
+% I guess don't be concrete - don't say it... because then, you have to be more concrete about it...
+	% I can elaborate if reviewers ask... but for now, let's just be concise...
 
@@
 % caption should mention that the difference between 
-    % ALTHOUGH you would expect that it would improve all models, it's not clear whether 
+    % ALTHOUGH you would expect that it would improve all models, it's not clear whether...err? hard to be precise here.
 %Removing TED talk number 767 (over 200 lines), as well as 15 hand-picked lines with bad TER scores, 
-
 %First, note that the scores improve, which is consistent with the observation that 
 % something to keep in mind in case reviewers ask
-
-good machine translations of bad Chinese sentences.
-
-One would expect such errors to be relatively well-tolerated in large training corpora
-But one bad 200-line talk constitutes a significant fraction of {\small \tt tst2010}, which is only 1500 lines total.
-More distressing is the fact is that the {\small \tt tst2010} data set has been used as the official development/tuning set for the past several IWSLT evaluation campaigns.
-It is unclear what effect such noisy corpora would have on other systems, 
-especially when used for tuning
-
 %This results in a significant TER score change that is almost as large as the score difference between the highest and lowest ranking system from the evaluation campaign.  
 %BLEU scores \mycitep{bleu}, on the other hand, did not change much.
-
 % no, omit this, because DUH, if you change the test set, the scores will change. (it's like changing your ruler.)
@@
 %\end{table}
+% TODO: table of scores after purging. This is pretty crucial? why?
+	% i think i THOUGHT this was crucial because i had to back up my claims concretely...
+	% but the claim from the data is tenuous at best - of course the scores change if the test set changes! so WHAT?
+	% BESIDES, I tried training and tuning with purged corpora, and the difference was minimal. it's quite possible that people just don't CARE
 
-% TODO: display an example bad-sentence pair?
+\subsection{Upsides, Potential Applications - RETITLE}
 
-Examining
-
-Just mention for now that there is an entire lecture that is bad
-Also hand-purged several other sentences
-
-TODO: table of scores after purging. This is pretty crucial
-
-It is a testament to how indisidious these errors can be that they persisted this long (tst2010 was distributed as recently as IWSLT2015
-
-Not even an international conference is immune to such issues
-
-It also demonstrates the tremendous imbalance between linguistic effort and machine learning effort/investment
-
-Higher corpus quality is desirable.
+Okay, so Higher corpus quality is desirable.
 Larger in-domain corpus quantity is desirable too - see learning curve controls below?
-
-\subsection{Upsides, Potential Applications}
 
@@
 
-Such synthetic parallel corpora can also be viewed as benchmark tasks (need to cite Facebook bAbI).
+Such synthetic parallel corpora can also be viewed as benchmark tasks (need to cite \mycitep{Facebook bAbI}).
 This is actually probably more important than you'd think, because looking at IWSLT15, some of the human translations are VERY loose.
@@
 
-maybe the latest Hutchins overview or nutshell?
-
-chomsky: transformations, colorless green ideas
-
 size doesn't always matter - those massive OpenSubtitles corpora don't help
@@
 
-\section{Model}
+\section{Our Primitive Multilingual Generator}
 
@@
 We adopt the exceedingly simplistic view of language as nouns, verbs, and descriptions (modifications) thereof.
+
+We focus on declarative indicative present active sentences in the hopes that they can be transformed into other forms in the long run.
+Similarly, although we enforce some semantic constraints, our system generally produces nonsensical sentences in the spirit of ``Colorless green ideas sleep furiously''.
+chomsky: transformations, colorless green ideas?
 
@@
 The IWSLT 2015 version of the WIT$^3$ corpus, which consists of multilingual subtitles of TED talks, covering a wide variety of subjects. 
+considered briefly in Section \ref{case_study}
 
 As prepared lectures, they fall somewhere between the conversational language of movie subtitles and the formal written language of UN resolutions.
-The latter are freely available online from the OPUS project \mycitep{opus}
+The latter are freely available online from the OPUS project \mycitep{opus:2012} as the OpenSubtitles\footnote{http://www.opensubtitles.org/} 2016 \mycitep{opensubtitles:2016} and the MultiUN \mycitep{multiun} corpora, respectively.
 
 We consider the Chinese to English machine translation task, using the
+
+first
 
@@
 IWSLT \mycitep{wit3} - point out that this corpus is freely available
-OPUS zh\_zh \mycitep{opensubtitles} MultiUN\mycitep{multiun} 
+OPUS 
+zh\_zh corpus from the OpenSubtitles 2016 corpus , with lines cleaned using some ad hoc heuristics
+MultiUN 
 Moses \mycitep{moses}
-KenLM \mycitep{kenlm}, which we found to be faster and stabler at decoding time.
 
-parameters used
+We configured Moses as described in \mycitep{wit3}, with the exception of changing the target monolingual language model toolkit to KenLM \mycitep{kenlm}, which we found to be faster and stabler than IRSTLM \mycitep{irstlm} at decoding time.
+For the remaining unspecified model settings, we used the {\small \tt grow-diag} alignment heuristic recommended by \mynewcite{segmenter:2008} and the {\small \tt mslr-fe} configuration for lexicalized reordering.
 
@@
 Interrogatives and imperatives.
-Pronouns - implemented as names here...? wait, but what about I/me? ugh
 

### 2016-06-01 doc/paper.tex
added some references and started Potential Benefits

@@
     However, these translations are generally optimized for human consumption, not machine learning, with no regard for such niceties as sentence alignment.
-    This is illustrated briefly in Section \ref{case_study}.
+    This is illustrated briefly in Section \ref{subsec:case_study}.
     }
@@
 
+% although it is ``concrete'', this section will probably be the first to get cut if I run OVER length (which had previously been unimaginable, but is starting to look more and more likely)
 % this is some stupid little side note I just noticed a couple weeks ago, but it is expanding to a huge size, because it's like the only concrete ``positive'' result I have...
 \subsection{Case Study}
-\label{case_study}
+\label{subsec:case_study}
 
@@
 
-As an example of suboptimal parallel corpora being used for machine translation, consider the official Chinese to English baseline translation of the {\small \tt tst2010} data set\footnote{https://wit3.fbk.eu/score.php?release=2014-01} from the TED talks corpus, described in greater detail in Section \ref{experiments}.
+As an example of suboptimal parallel corpora being used for machine translation, consider the official Chinese to English baseline translation of the {\small \tt tst2010} data set\footnote{https://wit3.fbk.eu/score.php?release=2014-01} from the TED talks corpus, described in greater detail in Section \ref{sec:experiments}.
 As a very crude way of assessing translation quality, we used the official evaluation software MultEval \mycitep{multeval} to compute scores for each translated line individually.
@@
 
-\subsection{Upsides, Potential Applications - RETITLE}
+\subsection{Potential Benefits}
 
-Okay, so Higher corpus quality is desirable.
+Large, high quality synthetic parallel corpora could be useful as:
+
+\begin{itemize}
+\item Training sets for corpus-based MT systems, particularly for low-resource language pairs
+\item Benchmark tasks for MT evaluation, in the spirit of the Facebook bAbI tasks in question answering \mycitep{bAbI}
+\item Translation memories for computer-aided translation
+\item ``Language archives'': New NLP methods are developed every few years; linguistic theories change every decade or two; parallel corpora live forever (e.g., the Rosetta Stone).
+\end{itemize}
+% sigh, this set of bullets seems much clearer than all those stupid sentences that came before
+
 Larger in-domain corpus quantity is desirable too - see learning curve controls below?
-
-My main point seems to be that nobody makes high quality and quantity parallel corpora JUST for machine translation, and it seems like that's sorely needed for both training and benchmarking.
-
-Traditionally, corpus construction has been considered prohibitively expensive, ``scavenged'' from other sources like parliamentary proceedings or lecture subtitles.
-But we argue here that it need not be so expensive.
-
-Such synthetic parallel corpora can also be viewed as benchmark tasks (need to cite \mycitep{Facebook bAbI}).
-This is actually probably more important than you'd think, because looking at IWSLT15, some of the human translations are VERY loose.
-So there's another element of noise in the data, and it's one that no amount of downstream machine learning can get rid of.
-(Although people have tried corpus evaluation...)
-
-% interesting, but probably confusing to readers?
-%In the language of agile software development, test-driven MT system development, in which linguists pose sentence pairs, and MT developers work to translate them
-
-This could take the form of amplified training sets in the short run, and perhaps a ``universal translation memory'' in the long run.
-
-methods every few years; linguistic theories, representations every decade or so; parallel corpora live forever
 
@@
 
-unlike all the different competing methods, it seems that a ``universal translation memory'' would be inherently collaborative (methods are competitive by nature, whereas data can be collaborative)
+in the spirit of example-based MT system \mycitep{ebmtbook}
 
@@
 
+% should this go in the ``conclusion''?
+Okay, so our ultimate goal is translation of everyday conversational sentences
+
+This could take the form of amplified training sets in the short run, and perhaps a ``universal translation memory'' in the long run.
+unlike all the different competing methods, it seems that a ``universal translation memory'' would be inherently collaborative (methods are competitive by nature, can't combine naturally, outside of ensemble methods, whereas data provides a natural avenue for large-scale collaboration)
+
+WELL... issues of ambiguity are alleviated, but only somewhat (e.g., for a sentence about a bank, the generator should know whether it's a building). 
+BUT ambiguous attachments can still be generated (e.g., the [animal] with the [animal] with the overbite), and it would still take additional logic for the system to recognize this, much less generate all possible translations...
+Maybe just emphasize the fact that lexical ambiguity is virtually eliminated, although structural or semantic ambiguity is still possible in some cases...
+But it still seems like it'll be fewer cases... The [person] saw the dog with(by means of) the [vision implement]
+
+Can think of this as something intermediate between fully automatic (RBMT, or semi-supervised SMT) and fully hand-crafted (human translation, which is the norm).
+Compositional trees are more automatic-like, but generate more sentence pairs/tuples.
+Complex templates are more hand-crafted, but generate fewer pairs/tuples, and are (presumably) harder to combine compositionally (but are they really!?).
+
+Section \ref{sec:related} discusses previous related work,
+Section \ref{sec:generator} describes our first attempt at implementing a parallel corpus generator,
+Section \ref{sec:experiments} gives preliminary results, 
+and Section \ref{sec:future} proposes future work.
+
 \section{Related Work}
@@
 The idea underlying generation system was essentially proposed
-This aspect of our work is perhaps closest in spirit to the equivalence classes used by \mynewcite{brown:99} to reduce the data requirements of an EBMT system \mycitep{ebmtbook}.
 
@@
 
-other approaches to synthetic corpora, which are essentially bootstrapped from other MT systems
+Previous attempts at creating synthetic parallel corpora have used rule-based MT (RBMT) systems \mycitep{hu:2007,dugast:2008,murakami:2009,rubino:2014}
+
+but the input to the RBMT systems was again dictated by external corpora -- domain overlap was prioritized over linguistic consideration of what sentences the RBMT system could translate most accurately.
+
+Our approach,
+
+not designed from the ground up. We consider a more {\em ab initio} approach
+
+This aspect of our work is perhaps closest in spirit to the equivalence classes used by \mynewcite{brown:99} to reduce the data requirements of an EBMT system \mycitep{ebmtbook}.
+
+But we propose allowing more rule-based modifications of these templates
+Also, our goal is to generate a large, wide-coverage, high-quality corpus, not jumping directly to translation from examples as in EBMT systems
 
@@
 
-WELL... issues of ambiguity are alleviated, but only somewhat (e.g., for a sentence about a bank, the generator should know whether it's a building). 
-BUT ambiguous attachments can still be generated (e.g., the [animal] with the [animal] with the overbite), and it would still take additional logic for the system to recognize this, much less generate all possible translations...
-Maybe just emphasize the fact that lexical ambiguity is virtually eliminated, although structural or semantic ambiguity is still possible in some cases...
-But it still seems like it'll be fewer cases... The [person] saw the dog with(by means of) the [vision implement]
-
-Can think of this as something intermediate between fully automatic (RBMT, or semi-supervised SMT) and fully hand-crafted (human translation, which is the norm).
-Compositional trees are more automatic-like, but generate more sentence pairs/tuples.
-Complex templates are more hand-crafted, but generate fewer pairs/tuples, and are (presumably) harder to combine compositionally (but are they really!?).
-
 The tree structure we adopt is something intermediate between constituency trees and dependency trees, and it was chosen for ease of implementation.
@@
 
-\section{Our Primitive Multilingual Generator}
+\section{A Primitive Multilingual Generator}
+\label{generator}
 
@@
 Python
-YAML for data: very well suited. full Unicode support, and you don't even need to type quotation marks to enclose strings
+YAML for data: very well suited. full Unicode support, and you don't even need to type quotation marks to enclose strings.
+with an eye toward scalability
 Crude though it may be, we plan on eventually open-sourcing both the generator and its accompanying data
@@
 \section{Preliminary Experiments}
-\label{experiments}
+\label{sec:experiments}
 
@@
 The IWSLT 2015 version of the WIT$^3$ corpus, which consists of multilingual subtitles of TED talks, covering a wide variety of subjects. 
-considered briefly in Section \ref{case_study}
+considered briefly in Section \ref{subsec:case_study}
 
@@
 \section{Proposed Future Work}
+\label{sec:future}
 
@@
 
+parallel corpus construction has been considered prohibitively expensive, hence the use of pre-translated sources like parliamentary proceedings
+we believe it is worth an initial investment to create a scalable system
+
 expert-sourcing by linguists, using the computer to lever up - computer assisted translation, in a sense; maybe computer-amplified
@@
 But must this be so? Vast untapped resource of students or whatever.
+an open challenge to the NLP community at large
+with a sufficiently compelling and thus widely used educational app 
+a potentially game-changing source of supervised data for NLP in general
 

### 2016-06-01 doc/paper.tex
halfway through Related Work - moving along at a nice clip now

@@
 
-As early as Jelinek's (in)famous quote during the 1980's that ``{\em Every time I fire a linguist, my performance goes up},'' both linguists and linguistics have been increasingly marginalized in natural language processing.
+As early as Jelinek's (in)famous quote during the 1980's that ``{\em Every time I fire a linguist, my performance goes up},'' both linguists and linguistics have been increasingly marginalized in natural language processing (NLP).
 More recently, for example, \mynewcite{collobert:2011} mention the possibility of ``{\em [ideally]...to learn from letter sequences rather than words},'' illustrating an AI-motivated desire to move away from even the most fundamental of linguistic concepts.
@@
 \subsection{Potential Benefits}
+\label{subsec:benefits}
 
@@
 
-Larger in-domain corpus quantity is desirable too - see learning curve controls below?
+We believe that natural language processing in general and machine translation in particular are particularly amenable to artifical data generation. 
+Unlike many machine learning applications (image recognition, finance), NLP data sets are manmade to begin with; humans themselves {\em are} the primary natural producers and consumers of raw written language data.
+Translation is further privileged as perhaps the only NLP task for which millions of people (foreign-language learners) already undergo years of formal training, which has implications for scalability via crowdsourcing (Section \ref{sec:future}).
 
-lexical ambiguity is naturally alleviated
+Other practical considerations shine a favorable light on parallel corpus generation.
+As the inverse problem of translation, multilingual generation is also an easier task in some senses. 
+For example, lexical ambiguity can be largely alleviated,\footnote{
+    Ideally the generator would have {\em a priori} knowledge about whether it is writing a sentence pair about, say,  a (financial) bank or a (river) bank.
+} and segmentation of languages like Chinese \mycitep{segmenter:2005,segmenter:2008} becomes a non-issue.
+Parallel corpora also provide a natural avenue through which to interface contemporary MT systems with the countless linguist-years of expertise that have been invested into rule-based MT systems, as discussed in Section \ref{sec:related}.
+And unlike the various competing MT methods, which are by nature competitive, corpora are inherently collaborative.
 
-RBMT - countless developer-years of expertise
-
-in the spirit of example-based MT system \mycitep{ebmtbook}
-
-% moreover, you're at the mercy of your third-party translators - if you want more data, you gotta wait.
-
-Translation is perhaps the only NLP task for which millions of people already undergo years of training
-
-Unlike many machine learning tasks (image recognition, finance), NLP data are purely artificial/manmade.
-New paradigm?
-Not enough attention has been paid IMO to semi-automated corpus generation.
-Well, typical approaches are paid, like Mechanical Turk...
-but there is a vast untapped source of data in the form of student problem sets, in the spirit of reCAPTCHA (citation!?)
-NLP is unique among machine learning tasks in that humans are not only consumers, but also producers of data - the alpha and the omega, the beginning and the end.
-MT is further unique among NLP in that both input and output representations are widely accepted (modulo dialectical variations) - the source and target languages themselves.
-People on the street would agree to the representations.
-
-% should this go in the ``conclusion''?
-Okay, so our ultimate goal is translation of everyday conversational sentences
-
-This could take the form of amplified training sets in the short run, and perhaps a ``universal translation memory'' in the long run.
-unlike all the different competing methods, it seems that a ``universal translation memory'' would be inherently collaborative (methods are competitive by nature, can't combine naturally, outside of ensemble methods, whereas data provides a natural avenue for large-scale collaboration)
-
-WELL... issues of ambiguity are alleviated, but only somewhat (e.g., for a sentence about a bank, the generator should know whether it's a building). 
-BUT ambiguous attachments can still be generated (e.g., the [animal] with the [animal] with the overbite), and it would still take additional logic for the system to recognize this, much less generate all possible translations...
-Maybe just emphasize the fact that lexical ambiguity is virtually eliminated, although structural or semantic ambiguity is still possible in some cases...
-But it still seems like it'll be fewer cases... The [person] saw the dog with(by means of) the [vision implement]
-
-Can think of this as something intermediate between fully automatic (RBMT, or semi-supervised SMT) and fully hand-crafted (human translation, which is the norm).
-Compositional trees are more automatic-like, but generate more sentence pairs/tuples.
-Complex templates are more hand-crafted, but generate fewer pairs/tuples, and are (presumably) harder to combine compositionally (but are they really!?).
-
+This paper is organized as follows.
 Section \ref{sec:related} discusses previous related work,
@@
 
-The basic
+The idea of using synthetic parallel corpora in machine translation is hardly new.
+Previous attempts at creating synthetic parallel corpora have used rule-based MT (RBMT) systems \mycitep{hu:2007,dugast:2008,murakami:2009,rubino:2014} but the input to the RBMT systems was again dictated by external corpora in the desire to produce immediate results.
+A longer-term approach that may pay dividends (Section \ref{subsec:benefits}) would be to run RBMT systems on specially chosen sentences for which they are known to produce high-quality translations, or perhaps eventually directly encoding their rules into carefully crafted parallel corpora.
 
-The idea underlying generation system was essentially proposed
+Our parallel corpus generation system (Section \ref{sec:generator}) is very similar to the recursive equivalence class approach to example-based MT \mycitep{ebmtbook} used by \mynewcite{brown:99}. 
+Our generator also uses a somewhat more varied grammar that also admits the use of participial noun modifiers.
+But more importantly, we believe this line of work was ahead of its time and consider our main contribution to be the reevaluation of its value to the contemporary machine learning approach to MT.
 
-system described below is to have linguists write a template and then levering up with the computer to add lexical and syntactic variation.  
-As a simple example, 
-{\small \tt The NP jumped over the moon}
+Equivalence classing is itself an integral part of the default pipeline of Moses \mycitep{moses}, the open-source statistical MT system, with the {\small \tt mkcls} component \mycitep{och:98} binning words into automatically-generated equivalence classes to alleviate data sparsity issues.
 
-Unlike more AI-oriented or natural language understanding approaches, we are solely concerned here with the generation of surface forms
+Moreover, as they are upstream of GIZA++, these bins are constructed at the lexical level, not the phrase level, whereas the templates described below can make substitutions at the level of larger constituents and phrases.
 
-In the Moses toolchain, the {\small \tt mkcls} component \mycitep{och:98} performs this process in reverse, binning words into automatically-generated equivalence classes to alleviate data sparsity issues.
+Furthermore, it is run at training time and cannot help with unknown words at test time (is this true??)
+The only real cure for unknown words is moar data
+
+no straightforward way to apply linguistic modifiers to the classes, relying instead on the syntactic variety of the data corpus to capture linguistic effects.
+
 % USUALLY alignment only, CAN be applied to translation... http://www.statmt.org/moses/?n=Advanced.Models#ntoc6
@@
 ...
-Moreover, as they are upstream of GIZA++, these bins are constructed at the lexical level, not the phrase level, whereas the templates described below can make substitutions at the level of larger constituents and phrases.
-Furthermore, it is run at training time and cannot help with unknown words at test time (is this true??)
-The only real cure for unknown words is moar data
 
-Previous attempts at creating synthetic parallel corpora have used rule-based MT (RBMT) systems \mycitep{hu:2007,dugast:2008,murakami:2009,rubino:2014}
+\section{A Primitive Multilingual Generator}
+\label{sec:generator}
 
-but the input to the RBMT systems was again dictated by external corpora -- domain overlap was prioritized over linguistic consideration of what sentences the RBMT system could translate most accurately.
+As a simple example, 
+{\small \tt The NP jumped over the moon}
 
-Our approach,
+system described below is to have linguists write a template and then levering up with the computer to add lexical and syntactic variation.
 
-not designed from the ground up. We consider a more {\em ab initio} approach
+The rudimentary tree structure we adopt is something intermediate between constituency trees and dependency trees, and it was chosen for ease of implementation. 
+Modifiers are constructed as constituent-like, but also maintain dependency links to their target words.
+[ugh should i bother with this?]
+especially if we are just concerned with producing high-quality surface forms, instead of performing complex logical reasoning on the underlying data representation
 
-This aspect of our work is perhaps closest in spirit to the equivalence classes used by \mynewcite{brown:99} to reduce the data requirements of an EBMT system \mycitep{ebmtbook}.
+used for corpus analysis stanford POS tagger (incomplete bibtex entry?) \mycitep{postagger:2003}
 
-But we propose allowing more rule-based modifications of these templates
-Also, our goal is to generate a large, wide-coverage, high-quality corpus, not jumping directly to translation from examples as in EBMT systems
+Unlike AI-oriented or natural language understanding approaches, our primary goal is to generate parallel corpora of surface forms, offering us a significant amount of flexibility on the method of generation.
+adopt elements of both rule-based and example-based MT systems
 
-Hardly a new idea ... but the emphasis here is on the resulting corpus
+Our first attempt, described in Section \ref{sec:generator},
 
-size doesn't always matter - those massive OpenSubtitles corpora don't help
-- uh, is the problem quality? should i run some Moses runs with a small portion of MultiUN to see how quality improves?
+rules: syntactic variety
 
-lots of things are resolved naturally when considering generation instead of discrimination
-- ambiguity (you know what you want to say)
-- segmentation (you know what ``phrases'' you wanted to output)
-
-The tree structure we adopt is something intermediate between constituency trees and dependency trees, and it was chosen for ease of implementation.
+in the spirit of example-based MT system \mycitep{ebmtbook}
 
 We adopt the pragmatic, ``defeatist'' attitude of modeling only straightforward linguistic phenomena compositionally; if the theory and data structures required to model a phenomenon are sufficiently complicated, everything is probably better off just going directly to a translation via template, instead of trying to concoct a sophisticated theory that has to be consistent with all previously theorized phenomena.
-
-This is (probably? ref??) unlike other NLG work in that our emphasis is on surface forms, not the underlying data representation.
-
-\section{A Primitive Multilingual Generator}
-\label{generator}
-
-used for corpus analysis stanford POS tagger (incomplete bibtex entry?) \mycitep{postagger:2003}
+The tradeoff is that 
+Compositional trees are more automatic-like, but generate more sentence pairs/tuples.
+Complex templates are more hand-crafted, but generate fewer pairs/tuples, and are (presumably) harder to combine compositionally (but are they really!?).
 
@@
 Similarly, although we enforce some semantic constraints, our system generally produces nonsensical sentences in the spirit of ``Colorless green ideas sleep furiously''.
-chomsky: transformations, colorless green ideas?
+based very loosely on the early work of chomsky: transformations, colorless green ideas?
 
@@
 Moses \mycitep{moses}
+
+Since we are interested in the relative effect an additional synthetic parallel corpus
+we choose a simple baseline system and fix other variables
 
@@
 
+moreover, you're at the mercy of your third-party translators - if you want more data, you gotta wait.
+But with a sufficiently compelling educational app/platform, you wouldn't have to wait OR pay for more data
+
 Crowdsourcing also has its place in post-editing, at the very least, for test sets.
@@
 a potentially game-changing source of supervised data for NLP in general
+
+Well, typical approaches are paid, like Mechanical Turk... but there is a vast untapped source of data in the form of student problem sets
+as with the success story of reCAPTCHA \mycitep{recaptcha}, we believe this will scale best when there is a natural harmony between data supply and demand
 

### 2016-06-02 doc/paper.tex
finished Related Work. I guess.

@@
 What we propose is to bring these jobs back ``in-house'', in the updated context of corpus-based MT.
-
-% TODO: i haven't talked yet about the unreasonable effectiveness of in-domain data...
-% oh is that the ``upsides'' section?
 
@@
 
+% hmm, this would up being disturbingly short...
 \section{Related Work}
@@
 Our generator also uses a somewhat more varied grammar that also admits the use of participial noun modifiers.
-But more importantly, we believe this line of work was ahead of its time and consider our main contribution to be the reevaluation of its value to the contemporary machine learning approach to MT.
+But more importantly, we believe this line of work was ahead of its time and consider our main contribution to be bringing attention to its potential value to the contemporary machine learning approach to MT.
 
-Equivalence classing is itself an integral part of the default pipeline of Moses \mycitep{moses}, the open-source statistical MT system, with the {\small \tt mkcls} component \mycitep{och:98} binning words into automatically-generated equivalence classes to alleviate data sparsity issues.
+Equivalence classing is in fact an integral part of the standard pipeline of Moses \mycitep{moses}, the open-source statistical MT system. 
+To add ``an `example-based' touch to the statistical approach,'' the {\small \tt mkcls} component \mycitep{och:98} clusters words into automatically-generated equivalence classes to ameliorate data sparsity issues.
+However, these clusters are trained only at the word level and not for phrases, and the clusters themselves are typically only used during alignment.
+Furthermore, since they are learned in an unsupervised manner from the parallel corpus, they can still benefit from the addition of more high-quality training examples, synthetic or otherwise.
 
-Moreover, as they are upstream of GIZA++, these bins are constructed at the lexical level, not the phrase level, whereas the templates described below can make substitutions at the level of larger constituents and phrases.
+% POTENTIAL REVIEWER QUESTION (assuming they even understand wtf I was talking about here): well, but you can add word cluster ids as factors/features...
+(http://www.statmt.org/moses/?n=Advanced.Models#ntoc6 - simply annotate words )
+% my weak responses
+    % supervised data almost always trumps unsupervised data, especially in NLP. there is no real substitute for REAL, in-domain data
+    % also, you would still have to require the models to to learn from these noisy clusterings
+    % actually, word+mkcls is still a LEXICALIZED translation factor; that's probably why it doesn't really help during the translation phase
+        % class-LM (target sequence model over cluster-ids) seems like it might be more useful....BUT, this is for MONOLINGUAL target-side data, 
+            % which generally doesn't need the help
+            % which is a strange thing to do to begin with, taking clusterings that were trained on expensive bilingual corpora and using them on plentiful monolingual corpora
+% a possibly stronger response: automatic clusters are also of lower quality than hand-crafted template variations, and there will likely be clustering errors
+% side note: if mkcls clusters were so awesome, they'd be used as standard factors... so you'd have to think that they're not
 
-Furthermore, it is run at training time and cannot help with unknown words at test time (is this true??)
-The only real cure for unknown words is moar data
+%%%%% WAIT a second, doesn't this in principle just leave it up to Moses to pick up ``class phrases''?
 
-no straightforward way to apply linguistic modifiers to the classes, relying instead on the syntactic variety of the data corpus to capture linguistic effects.
-
-% USUALLY alignment only, CAN be applied to translation... http://www.statmt.org/moses/?n=Advanced.Models#ntoc6
-% uh, so do people already do this??
-With some effort, you can use these bins during steps other than alignment, although this is not done by default.
-...
+%% uh...
 
@@
 \label{sec:generator}
+
+% let's figure out how to do a ``figure''
 

### 2016-06-02 doc/paper.tex
fixed non-ascii in references (grr, only incompatible once \zh is used) and revised introduction somewhat

@@
 An increased focus on a
-
+S \zh{我把} O V
 is proposed.
@@
 
-In this paper, we propose bringing linguists back from a premature exile to craft high-quality, ``designer'' synthetic parallel corpora for machine translation.
+One of the major advantages of data-driven MT is that it removes the need for linguists to develop a new set of rules for an MT system to support a new language.
+In a sense, though, the rule-based MT linguist's job has really been  ``outsourced'' to third-party translators, who implicitly encode linguistic information in parallel corpora.
+We believe there is still a significant ``in-house'' role for linguists to play in corpus-based machine translation, and that is to direct the crafting of high-quality, ``designer'' synthetic parallel corpora, leveraging technologies like crowdsourcing to maximize throughput.
 Such a division of labor would let linguists be linguists, with MT system development proceeding in parallel.
-In a sense, the use of external corpora has ``outsourced'' the rule-based MT linguist's job to third-party translators, who implicitly encode linguistic information in parallel corpora.  
-What we propose is to bring these jobs back ``in-house'', in the updated context of corpus-based MT.
 
@@
 
-% hmm, this would up being disturbingly short...
+% hmm, this wound up being disturbingly short...
 \section{Related Work}
@@
 Our parallel corpus generation system (Section \ref{sec:generator}) is very similar to the recursive equivalence class approach to example-based MT \mycitep{ebmtbook} used by \mynewcite{brown:99}. 
-Our generator also uses a somewhat more varied grammar that also admits the use of participial noun modifiers.
+Our generator has more of a rule-based flavor, using a somewhat more varied grammar that also admits the use of participial noun modifiers.
 But more importantly, we believe this line of work was ahead of its time and consider our main contribution to be bringing attention to its potential value to the contemporary machine learning approach to MT.
@@
 
-%%%%% WAIT a second, doesn't this in principle just leave it up to Moses to pick up ``class phrases''?
-
-%% uh...
-
 \section{A Primitive Multilingual Generator}
@@
 
-% let's figure out how to do a ``figure''
+Unlike AI-oriented or natural language understanding approaches, our primary goal is to generate parallel corpora of surface forms, offering us a significant amount of flexibility as to the method of generation.
 
-As a simple example, 
-{\small \tt The NP jumped over the moon}
+\small 
+\begin{center}
+{\tt S V O .}
+
+%S \zh{把} O V
+
+{\tt S \zh{我把} O V}
+
+%\zh{把} {\tt O V} \zh{}
+%\\
+%S 
+\end{center}
+\normalsize
+
+We adopt a pragmatic attitude of only trying to model straightforward linguistic phenomena compositionally; if the theory and data structures required to model a linguistic phenomenon are sufficiently complicated,
+
+everything is probably better off just going directly to a translation via template, instead of trying to concoct a sophisticated theory that has to be consistent with all previously theorized phenomena.
+
+At the time of this writing, our system can 
+Literal 
+sentence templates are all
+
+Our parallel corpus generator starts with
+
+native speakers, linguists or otherwise
+
+S and O can be expanded
+
+We can then allow the generator to vary both the syntactic and lexical content of
 
@@
 
-Unlike AI-oriented or natural language understanding approaches, our primary goal is to generate parallel corpora of surface forms, offering us a significant amount of flexibility on the method of generation.
 adopt elements of both rule-based and example-based MT systems
@@
 
-We adopt the pragmatic, ``defeatist'' attitude of modeling only straightforward linguistic phenomena compositionally; if the theory and data structures required to model a phenomenon are sufficiently complicated, everything is probably better off just going directly to a translation via template, instead of trying to concoct a sophisticated theory that has to be consistent with all previously theorized phenomena.
 The tradeoff is that 
@@
 Out of domain corpus
+
+we of 
+the {\small \tt tmcombine} script \mycitep{tmcombine} included with Moses
 
@@
 
+Sufficient quantities of supervised data seem to underlie high or state of the art performance in NLP
+
 estimate about 1 noun per minute (my pathetic work speed)
@@
 
-%RNN encoder-decoders are not the first corpus-based MT methodology (EBMT then SMT then this), and it is unlikely in this era of Big Data that they will be the last.
+RNN encoder-decoders are not the first corpus-based MT methodology (EBMT then SMT then this), and it is unlikely in this era of Big Data that they will be the last.
 
@@
     %Note that emnlp2016.bst chokes if there are 0 citations in your paper.    
-    %\mynewcite{moses}    
+    \mynewcite{moses}    
     \bibliographystyle{emnlp2016}

### 2016-06-02 doc/paper.tex
finished a sloppy pass at Generator section, and at this point it's pretty clear I'm going to run OVER length... but yeah, this last part is much faster, although the Future Work section is going to be slow again...

@@
 As early as Jelinek's (in)famous quote during the 1980's that ``{\em Every time I fire a linguist, my performance goes up},'' both linguists and linguistics have been increasingly marginalized in natural language processing (NLP).
-More recently, for example, \mynewcite{collobert:2011} mention the possibility of ``{\em [ideally]...to learn from letter sequences rather than words},'' illustrating an AI-motivated desire to move away from even the most fundamental of linguistic concepts.
+More recently, for example, \mynewcite{collobert:2011} mention the possibility ``{\em [ideally]...to learn from letter sequences rather than words},'' illustrating an AI-motivated desire to move away from even the most fundamental of linguistic concepts.
 A specific instance of this overall trend is the gradual shift of machine translation (MT) away from classical linguistic rule-based approaches \mycitep{hutchins:2005} to contemporary machine learning approaches, which underlie statistical phrase-based systems like Moses \mycitep{moses} and more recent neural network systems \mycitep{sutskever:2014,bengio:2014}.
@@
 
-% hmm, this wound up being disturbingly short...
 \section{Related Work}
@@
 Equivalence classing is in fact an integral part of the standard pipeline of Moses \mycitep{moses}, the open-source statistical MT system. 
-To add ``an `example-based' touch to the statistical approach,'' the {\small \tt mkcls} component \mycitep{och:98} clusters words into automatically-generated equivalence classes to ameliorate data sparsity issues.
+During training, the {\small \tt mkcls} component adds ``{\em an `example-based' touch to the statistical approach}'' by clustering words into automatically-generated equivalence classes to ameliorate data sparsity issues \mycitep{och:98}.
 However, these clusters are trained only at the word level and not for phrases, and the clusters themselves are typically only used during alignment.
@@
 
-\section{A Primitive Multilingual Generator}
+% these links are too long for EMNLP. TODO: add them as footnotes for AMTA?
+% http://www.fb10.uni-bremen.de/anglistik/langpro/kpml/genbank/generation-bank.html
+% http://www.fb10.uni-bremen.de/anglistik/langpro/kpml/ genbank/R3b12-English/Docu/ENGLISH-nigel-exerciseset-mismatches-19981209/EX-SET-1.html
+In terms of freely available multilingual natural language generators, the most well-developed one we could find was KPML \mycitep{kpml}.
+But sentence writing is fairly involved because the focus seems to be on having a well-developed linguistic theory to explain the surface forms.
+In contrast, our primary goal is to generate parallel corpora of surface forms, affording us a significant amount of flexibility as to the method of generation.
+
+In the next section, we describe our own primitive attempt at a multilingual generator. 
+While we try to model straightforward linguistic phenomena compositionally, in the tradition of example-based MT \mycitep{ebmtbook}, we adopt the pragmatic attitude that sometimes the most effective way to represent a translation phenomenon is simply to write it down.
+Thus, with the long-term goal of scalability, we attempt to construct a system sufficiently flexible to write sentence pairs in large quantities through syntactic variations and also with ``hand-crafted'' quality through example-based templates.
+
+\section{A Nascent Multilingual Generator}
 \label{sec:generator}
 
-Unlike AI-oriented or natural language understanding approaches, our primary goal is to generate parallel corpora of surface forms, offering us a significant amount of flexibility as to the method of generation.
+As an example of how our synthetic parallel corpus generation system works, consider the following English-Chinese template pair:
 
-\small 
-\begin{center}
-{\tt S V O .}
+%\small
+\begin{center} \begin{small} 
+{\tt S V O .} \\
+{\tt S \zh{把} O V \zh{。}}
+\end{small} \end{center} 
+%\normalsize
 
-%S \zh{把} O V
+\noindent Our system then expands the subject {\small \tt S} and object {\small \tt O} as noun phrases, adding adjectival and participial modifiers as desired. 
+Semantic constraints can be imposed from the verb {\small \tt V} as well as externally.
+The generator can then vary the template lexically\footnote{Rewrite rule probabilities were set by hand; a more principled approach would, say, use rewrite probabilities from a generative PCFG parser.} and syntactically, allowing a large number of sentence pairs to be generated from a single template.
+We can also trade quantity for quality by adding more literals to the template, in the spirit of example-based MT.
 
-{\tt S \zh{我把} O V}
-
-%\zh{把} {\tt O V} \zh{}
-%\\
-%S 
-\end{center}
-\normalsize
-
-We adopt a pragmatic attitude of only trying to model straightforward linguistic phenomena compositionally; if the theory and data structures required to model a linguistic phenomenon are sufficiently complicated,
-
-everything is probably better off just going directly to a translation via template, instead of trying to concoct a sophisticated theory that has to be consistent with all previously theorized phenomena.
-
-At the time of this writing, our system can 
-Literal 
-sentence templates are all
-
-Our parallel corpus generator starts with
-
-native speakers, linguists or otherwise
-
-S and O can be expanded
-
-We can then allow the generator to vary both the syntactic and lexical content of
-
-system described below is to have linguists write a template and then levering up with the computer to add lexical and syntactic variation.
-
-The rudimentary tree structure we adopt is something intermediate between constituency trees and dependency trees, and it was chosen for ease of implementation. 
-Modifiers are constructed as constituent-like, but also maintain dependency links to their target words.
-[ugh should i bother with this?]
-especially if we are just concerned with producing high-quality surface forms, instead of performing complex logical reasoning on the underlying data representation
-
-used for corpus analysis stanford POS tagger (incomplete bibtex entry?) \mycitep{postagger:2003}
-
-adopt elements of both rule-based and example-based MT systems
-
-Our first attempt, described in Section \ref{sec:generator},
-
-rules: syntactic variety
-
-in the spirit of example-based MT system \mycitep{ebmtbook}
-
-The tradeoff is that 
-Compositional trees are more automatic-like, but generate more sentence pairs/tuples.
-Complex templates are more hand-crafted, but generate fewer pairs/tuples, and are (presumably) harder to combine compositionally (but are they really!?).
-
+% TODO: looks like this is GOING to run over length. save some of this for AMTA, which is going to run way UNDER length
 %\subsection{Simplistic Linguistics}
-We adopt the exceedingly simplistic view of language as nouns, verbs, and descriptions (modifications) thereof.
-
-We focus on declarative indicative present active sentences in the hopes that they can be transformed into other forms in the long run.
-Similarly, although we enforce some semantic constraints, our system generally produces nonsensical sentences in the spirit of ``Colorless green ideas sleep furiously''.
-based very loosely on the early work of chomsky: transformations, colorless green ideas?
-
-two sources of complexity
-- lexical (semantic?)
-- syntactic
-
-The fact that verbs can serve as nouns (gerundives) and modifiers (participles and modals) is a significant source of added complexity. (footnote?)
-
-Testing section reference: \ref{sec:related} - this requires changing secnumdepth=2 in AAAI
+We adopt an exceedingly simplistic view of language as nouns, verbs, and their modifications and elaborations.
+Inspired by early work on transformations by \mynewcite{chomsky}, we focus on declarative indicative present active sentences in the hopes that they can be transformed into other forms in the long run.
+Similarly, although we enforce some semantic constraints, our system generally produces nonsensical sentences like ``{\em Colorless green ideas sleep furiously.}''
+The fact that verb phrases can also serve as nouns (gerundives) and modifiers (participles) appears to be a common source of complexity, and we made it a priority to implement participial modifiers.
+To capture recursive structures like participles, we adopt a rudimentary tree representation that is something intermediate between constituency trees and dependency trees, with constituent-like modifiers that also maintain dependency-like links to their targets.
 
 %\subsection{Implementation Details}
-Python
-YAML for data: very well suited. full Unicode support, and you don't even need to type quotation marks to enclose strings.
-with an eye toward scalability
-Crude though it may be, we plan on eventually open-sourcing both the generator and its accompanying data
+% TODO: mention Python?
+One implementation detail worth mentioning is the choice of YAML for writing sentence templates and lexical data. 
+In particular, YAML has native Unicode support and does not even require quotation marks to enclose strings.
+
+% TODO: should I bother with this? 
+Crude though it may be, we plan on eventually open-sourcing both our generator and its accompanying data
+
+% oh this part really is just flying now, because there's no uncertainty as to what I DID
 
@@
 \subsection{Methods}
+
+% TODO: filler for AMTA? used for corpus analysis stanford POS tagger (incomplete bibtex entry?) \mycitep{postagger:2003}
+
 As mentioned in 

### 2016-06-02 doc/paper.tex
ALMOST finished with Results

@@
 \begin{abstract}
-An increased focus on a
-S \zh{我把} O V
-is proposed.
+The large-scale generation of high-quality synthetic parallel corpora for data-driven machine translation is proposed.
 
-A particular system for doing this is presented.
-It allows syntactic complexity via rules.
-It also allows high-quality examples via templates, with minor syntactic modifications via rules.
+A preliminary system for doing this is presented,
 
-Preliminary results do not show much improvement.
+syntactic complexity in the spirit of rule-based machine translation,
+
+as well as high-quality sentence pairs via templates, in the spirit of example-based machine translation.
+
+Early results do not show much improvement.
 This is because we need to scale up?
 
-Future work is proposed.
-In particular, we believe this is amenable to crowdsourcing by ``amateur linguists''.
+Future work is proposed, in particular, 
+In particular, we believe this is amenable to crowdsourcing by ``amateur linguists.''
 
@@
 
-Large, high quality synthetic parallel corpora could be useful as:
+Large, high-quality synthetic parallel corpora could be useful as:
 
@@
 Our generator has more of a rule-based flavor, using a somewhat more varied grammar that also admits the use of participial noun modifiers.
-But more importantly, we believe this line of work was ahead of its time and consider our main contribution to be bringing attention to its potential value to the contemporary machine learning approach to MT.
+But more importantly, we believe this line of work was ahead of its time and consider our main contribution to be bringing attention to its potential value to contemporary MT.
 
@@
 
-% TODO: looks like this is GOING to run over length. save some of this for AMTA, which is going to run way UNDER length
+% TODO: looks like this is GOING to run over length. save some if not all of this for AMTA, which is going to run way UNDER length
 %\subsection{Simplistic Linguistics}
@@
 % TODO: should I bother with this? 
-Crude though it may be, we plan on eventually open-sourcing both our generator and its accompanying data
+Crude though it may be, we plan on eventually open-sourcing both our generator and its accompanying data.
 
 % oh this part really is just flying now, because there's no uncertainty as to what I DID
+
+% running over length, but let's just write it all down and then see what to cut (hint: that stupid tst2010 stuff. maybe just summarize it in a footnote)
 
@@
 
-\subsection{Methods}
+%\subsection{Methods} % no space? just do one paragraph each?
+We consider training set amplification as a first application.
+Since we are interested in the relative effect of our synthetic parallel corpus, we chose a simple baseline system and fix all other variables.
+Namely, we used the baseline configuration of Moses \mycitep{moses} as described in \mycitep{iwslt2015}, with the exception of changing the target language model to KenLM \mycitep{kenlm}, which we found to be faster and stabler than IRSTLM \mycitep{irstlm} at decoding time.
+For the remaining unspecified model settings, we used the {\small \tt grow-diag} alignment heuristic recommended by \mynewcite{segmenter:2008} and the {\small \tt mslr-fe} configuration for lexicalized reordering.
+Chinese text was preprocessed using the latest version of the Stanford Segmenter \mycitep{segmenter:2005,segmenter:2008}.
+Phrase and lexicalized reordering tables were interpolated using the {\small \tt tmcombine} script included with Moses \mycitep{tmcombine}, with the interpolation tuned on the {\small \tt dev2010} data set.
+BLEU scores were computed as described by \mynewcite{iwslt2015} using the IWSLT 2015 progress data set ({\small \tt tst2014}).\footnote{We chose the {\small \tt tst2014} test set to facilitate comparison with systems from both IWSLT 2014 and 2015 in future work.}
+
+%\subsection{Task}
 
 % TODO: filler for AMTA? used for corpus analysis stanford POS tagger (incomplete bibtex entry?) \mycitep{postagger:2003}
+We evaluate the contribution of our synthetic parallel corpora using the Chinese to English task of IWSLT 2015 \mycitep{iwslt2015}, mentioned briefly in Section \ref{subsec:case_study}.
+This task uses the WIT$^3$ corpus \mycitep{wit3}, which consists of multilingual subtitles of TED talks covering a wide variety of subjects. 
+As a control, we also add the freely available MultiUN corpus \mycitep{multiun} from the OPUS project \mycitep{opus:2012}.
 
-As mentioned in 
-The IWSLT 2015 version of the WIT$^3$ corpus, which consists of multilingual subtitles of TED talks, covering a wide variety of subjects. 
-considered briefly in Section \ref{subsec:case_study}
+% TODO: add OpenSubtitles 2016 results? they don't really add much to the discussion - I can just mention them briefly, maybe even in a footnote
+% oh okay, this is nice - if I don't include OpenSubtitles, then I don't have to talk about the stupid purging either
+%heuristically purged zh\_zh, which is a very noisy corpus
+%zh\_zh corpus from the OpenSubtitles 2016 corpus , with lines cleaned using some ad hoc heuristics
 
-As prepared lectures, they fall somewhere between the conversational language of movie subtitles and the formal written language of UN resolutions.
-The latter are freely available online from the OPUS project \mycitep{opus:2012} as the OpenSubtitles\footnote{http://www.opensubtitles.org/} 2016 \mycitep{opensubtitles:2016} and the MultiUN \mycitep{multiun} corpora, respectively.
+%\subsection{Results}
 
-We consider the Chinese to English machine translation task, using the
+% yeah sure, let's add Bing/Google results just for kicks, then take them out later
 
-first
+% definitely want 2 tables?
+    % one for learning curves, to show that iwslt > zh_zh > multiun - domain overlap is greater
 
-Out of domain corpus
+    % one for 100%, to show that the out of domain corpora aren't helping that much. need moar in-domain data (probably)
+    % well, maybe i'll do away with this one
 
-we of 
-the {\small \tt tmcombine} script \mycitep{tmcombine} included with Moses
+    % I guess don't bother showing single-corpus results for zh_zh and multiun, because then I'd have to show MY single-corpus results, which are even more embarrassing than the horribleness I've written up here
 
-the latest version of the Stanford Segmenter \mycitep{segmenter:2005,segmenter:2008}
+% when i actually go and READ the table, it's clear that I don't even want to LOOK at more than one metric. it's information overload.
+%\begin{table}
+%\centering
+%\label{table:tst2010}
+%\begin{tabular}{|l|r|r|r|}
+%\hline
+%                    & BLEU  & NIST & TER    \\ \hline
+%IWSLT15 official    & 11.43 & 4.67 & 72.65  \\ \hline
+%Our baseline        & 11.54 & 4.43 & 72.43  \\ \hline
+%  + OpenSubtitles   & 11.54 & 4.43 & 72.43  \\ \hline
+%  + MultiUN         & 11.67 & 4.48 & 72.60  \\ \hline
+%  + generated       & 11.46 & 4.39 & 72.51  \\ \hline
+%% &  &   \\ \hline
+%\end{tabular}
+%\caption{}
+% 
+%\end{table}
 
-citations needed
-IWSLT \mycitep{wit3} - point out that this corpus is freely available
-OPUS 
-zh\_zh corpus from the OpenSubtitles 2016 corpus , with lines cleaned using some ad hoc heuristics
-MultiUN 
-Moses \mycitep{moses}
+% side-by-side tables as in emnlp2016.tex
+\begin{table}
+\label{tab:results}
+\centering
+\small
+\begin{tabular}{cc}
 
-Since we are interested in the relative effect an additional synthetic parallel corpus
-we choose a simple baseline system and fix other variables
+\begin{tabular}{|l|r|}
+\hline
+                    & BLEU   \\ \hline
+25\% corpus         &  9.95  \\ \hline
+  + 75\%            & \textbf{11.47}  \\ \hline
+%  + Subtitles       & 10.57  \\ \hline
+  + MultiUN         & 10.76  \\ \hline
+  + Generated       &  9.91  \\ \hline
+\end{tabular}
 
-We configured Moses as described in \mycitep{wit3}, with the exception of changing the target monolingual language model toolkit to KenLM \mycitep{kenlm}, which we found to be faster and stabler than IRSTLM \mycitep{irstlm} at decoding time.
-For the remaining unspecified model settings, we used the {\small \tt grow-diag} alignment heuristic recommended by \mynewcite{segmenter:2008} and the {\small \tt mslr-fe} configuration for lexicalized reordering.
+\begin{tabular}{|l|r|}
+\hline                    & BLEU   \\ \hline
+Official            & 11.43  \\ \hline
+Our baseline        & 11.54  \\ \hline
+%  + Subtitles       & 11.54  \\ \hline
+  + MultiUN         & \textbf{11.67}  \\ \hline
+  + Generated       & 11.46  \\ \hline
+\end{tabular} & % oh this is how you get 2 tables side by side
 
-The corpus, which is freely available, consists
+\end{tabular} % cc (i.e., the big super-table)
 
-IWSLT: TED talks are a series of lectures, which makes them
+\caption{
+    (Left) Chinese-to-English BLEU scores for our baseline Moses configuration with phrase and reordering tables trained on a random 25\% subset of the IWSLT15 training corpus and interpolated (+) with tables for the remaining 75\% (\textasciitilde 150K) of the in-domain corpus, the MultiUN corpus (a \textasciitilde 1M line subset), and a synthetic corpus from our generator (\textasciitilde 5M lines).
+    (Right) Same as on the left, but starting with tables for the entire IWSLT15 corpus.
+    Official published scores of the IWSLT15 baseline system are also given for comparison.    
+    Qualitatively similar trends for NIST and TER scores were also observed (omitted for clarity).
+%    The ``baseline'' row gives 
+%    MultiUN is actually more data because it has more words per line.
+% automatically cleaned version of the en-zh_zh language pair of OpenSubtitles 2016, roughly 1M lines.
+% and a similar number of lines from MultiUN, which is actually more words. but it doesn't do as well.
+} % caption
 
-heuristically purged zh\_zh, which is a very noisy corpus
+\end{table}
 
-\subsection{Results}
+% preemptively answering reviewer question: yeah yeah, the trained tmcombine weights are like 0.001 for my generated corpora
+The results of these experiments are shown in Table \ref{tab:results}.
+Our generated corpora do not make a significant impact on the final BLEU score,\footnote{
+    Results for the most recently generated synthetic corpus are shown; scores slightly higher than our baseline have been obtained on older generated corpora.
+    We believe both to be within the error bars of the MERT tuning procedure.
+}
+indicating that our generator has a long way to go.
 
-2016-02-20 learning curves adding only 130,000 in-domain lines improves scores by like 2.0 BLEU points
+%%%%%%%%%%%%%%%%%%%%%
+TODOOOOOOOOOOOO
+% TODO: HOW SO?
+wanted universal translation memory
+much less ambitious goal of training set amplification
+- out of domain data only boosts performance slightly
+phrase table interpolation using the tmcombine script included with Moses
+The results show that our generator has a long way to go.
+Quantity and maybe even quality (you'd expect MultiUN translations to be better?) seems to matter much less than whether the additional data is in-domain
+So lots of different basic syntactic variation didn't help
+vocabulary is much too small?
+(only about 100 nouns currently.
+we used Stanford POS tagger \mycitep{postagger:2003} to find most common nouns, since different corpora tend to translate words differently, and the mini-goal here was training set amplification)
+more templates
+work on that is currently in progress. % I have never liked the sound of ``currently underway''
+%%%%%%%%%%%%%%%%%%%%%%%%
 
-2016-03-03-225556 1M zh\_zh lines made scores slightly WORSE?
+For the table on the left, note that adding in-domain data improves performance significantly more than a much larger amount of out-of-domain data (MultiUN), a commonly reported phenomenon \mycitep{tmcombine}.
+As prepared lectures, TED talks fall somewhere between the conversational language of movie subtitles and the formal written language of UN resolutions; data from the OpenSubtitles 2016 corpus \mycitep{opensubtitles:2016}\footnote{http://www.opensubtitles.org/} from the OPUS project are similarly ineffective at improving performance (scores not shown).
+For the table on the right, note also that the same out-of-domain corpus now gives a much more modest performance increase.
+These observations suggest that amplification of in-domain corpora should still be beneficial, particularly for low-resource language pairs or domains.
 
-2016-04-29-093202 1M MultiUN lines only improved scores SLIGHTLY (by like 0.4 BLEU points)
+% no, I can't mention this, because it's just saying, ``duh, hope it's in-domain!''
+%For a scaled-up corpus generator, one anticipates that the main concern would be to avoid skewing the phrase table with the ``unnatural'' synthetic corpus.
 
-Quantity and maybe even quality (you'd expect MultiUN translations to be better?) matters much less than having data in-domain
-
-Of course, I need to redo (ALL?) these jobs with final segmenter choice
-
-Evaluated using the IWSLT 2015 progress data set (tst2014) to allow comparison with both 2015 and 2014 systems.
-
-training curves (just to prove my controls aren't wrong)
-my synthetic corpora don't help. 
-but in-domain corpus helps significantly more than out-of-domain. (this is a commonly reported result)
-
-Bottom line
-My corpora don't really seem to help, but neither do out-of-domain corpora.
-
-this will easily take up a whole column, because it includes 2 tables...
+% TODO: maybe add Bing and Google results, since nobody seems to report these? save this for AMTA? (haha not gonna have room, sucka)
 

### 2016-06-03 doc/paper.tex
quick and sloppy finish to Results

@@
     \newcommand\amtaonly[1]{}
+    \newcommand\emnlponly[1]{#1}
     \newcommand\hcomponly[1]{}
@@
     \newcommand\amtaonly[1]{#1}
+    \newcommand\emnlponly[1]{}
     \newcommand\hcomponly[1]{}
@@
     \newcommand\amtaonly[1]{}
+    \newcommand\emnlponly[1]{}
     \newcommand\hcomponly[1]{#1}
@@
 % this is some stupid little side note I just noticed a couple weeks ago, but it is expanding to a huge size, because it's like the only concrete ``positive'' result I have...
+%\amtaonly{
 \subsection{Case Study}
@@
 	% BESIDES, I tried training and tuning with purged corpora, and the difference was minimal. it's quite possible that people just don't CARE
+%} % \amtaonly
 
@@
 %\subsection{Simplistic Linguistics}
+%\amtaonly{
 We adopt an exceedingly simplistic view of language as nouns, verbs, and their modifications and elaborations.
@@
 To capture recursive structures like participles, we adopt a rudimentary tree representation that is something intermediate between constituency trees and dependency trees, with constituent-like modifiers that also maintain dependency-like links to their targets.
+%}
 
@@
 The results of these experiments are shown in Table \ref{tab:results}.
-Our generated corpora do not make a significant impact on the final BLEU score,\footnote{
-    Results for the most recently generated synthetic corpus are shown; scores slightly higher than our baseline have been obtained on older generated corpora.
-    We believe both to be within the error bars of the MERT tuning procedure.
-}
-indicating that our generator has a long way to go.
-
-%%%%%%%%%%%%%%%%%%%%%
-TODOOOOOOOOOOOO
-% TODO: HOW SO?
-wanted universal translation memory
-much less ambitious goal of training set amplification
-- out of domain data only boosts performance slightly
-phrase table interpolation using the tmcombine script included with Moses
-The results show that our generator has a long way to go.
-Quantity and maybe even quality (you'd expect MultiUN translations to be better?) seems to matter much less than whether the additional data is in-domain
-So lots of different basic syntactic variation didn't help
-vocabulary is much too small?
-(only about 100 nouns currently.
-we used Stanford POS tagger \mycitep{postagger:2003} to find most common nouns, since different corpora tend to translate words differently, and the mini-goal here was training set amplification)
-more templates
-work on that is currently in progress. % I have never liked the sound of ``currently underway''
-%%%%%%%%%%%%%%%%%%%%%%%%
-
 For the table on the left, note that adding in-domain data improves performance significantly more than a much larger amount of out-of-domain data (MultiUN), a commonly reported phenomenon \mycitep{tmcombine}.
@@
 
-% no, I can't mention this, because it's just saying, ``duh, hope it's in-domain!''
-%For a scaled-up corpus generator, one anticipates that the main concern would be to avoid skewing the phrase table with the ``unnatural'' synthetic corpus.
+Our generated corpora do not make a significant impact on the final BLEU score,\footnote{
+    Results for the most recently generated synthetic corpus are shown; scores slightly higher than our baseline have been obtained on older generated corpora.
+    Both of these outcomes seem to be within the error bars of the MERT tuning procedure.
+}
+indicating that our parallel corpus generator has a long way to go.
+Development on our generator has so far focused on ``rule-based'' syntactic diversity and growing a lexical database tailored to the training set, 
+Given these results, a change of direction seems in order; work on harvesting more templates from the IWSLT15 training corpus is currently in progress.
 

### 2016-06-03 doc/paper.tex
first pass at Future Work; abstract under construction

@@
     \newcommand\mynewcite[1]{\newcite{#1}}
-    \newcommand\amtaonly[1]{}
+    \newcommand\amtaonly[1]{#1} % currently being used to trim down to 4 pages
     \newcommand\emnlponly[1]{#1}
@@
 \begin{abstract}
+\amtaonly{UNDER CONSTRUCTION}
+
 The large-scale generation of high-quality synthetic parallel corpora for data-driven machine translation is proposed.
 
-A preliminary system for doing this is presented,
+A preliminary system for doing this is presented, striving toward syntactic complexity, in the spirit of rule-based machine translation, and high-quality sentence pairs via templates, in the spirit of example-based machine translation.
 
-syntactic complexity in the spirit of rule-based machine translation,
-
-as well as high-quality sentence pairs via templates, in the spirit of example-based machine translation.
-
-Early results do not show much improvement.
-This is because we need to scale up?
-
-Future work is proposed, in particular, 
-In particular, we believe this is amenable to crowdsourcing by ``amateur linguists.''
+Early results have yet to show improvement. [This is because we need to scale up?]
+Future work is proposed. 
+We believe this is particularly amenable to crowdsourcing by ``amateur linguists.''
 
@@
     However, these translations are generally optimized for human consumption, not machine learning, with no regard for such niceties as sentence alignment.
-    This is illustrated briefly in Section \ref{subsec:case_study}.
+    \amtaonly{This is illustrated briefly in Section \ref{subsec:case_study}.}
     }
+
+% it's TEMPTING to say this, but the same is essentially true of all supervised data sets - they're pretty much all human-labeled...
+%It is surprising to say the least to contrast 
+%technological sophistication
+%
+%with the direct human translation 
+%whereas parallel corpus acquisition is still more or less entirely dependent on .
 
@@
 % this is some stupid little side note I just noticed a couple weeks ago, but it is expanding to a huge size, because it's like the only concrete ``positive'' result I have...
-%\amtaonly{
+\amtaonly{
 \subsection{Case Study}
@@
 	% BESIDES, I tried training and tuning with purged corpora, and the difference was minimal. it's quite possible that people just don't CARE
-%} % \amtaonly
+} % \amtaonly
 
@@
 %\subsection{Simplistic Linguistics}
-%\amtaonly{
+\amtaonly{
 We adopt an exceedingly simplistic view of language as nouns, verbs, and their modifications and elaborations.
@@
 To capture recursive structures like participles, we adopt a rudimentary tree representation that is something intermediate between constituency trees and dependency trees, with constituent-like modifiers that also maintain dependency-like links to their targets.
-%}
+}
 
@@
 %\subsection{Methods} % no space? just do one paragraph each?
-We consider training set amplification as a first application.
+We now consider training set amplification as a first application.
 Since we are interested in the relative effect of our synthetic parallel corpus, we chose a simple baseline system and fix all other variables.
@@
 % TODO: filler for AMTA? used for corpus analysis stanford POS tagger (incomplete bibtex entry?) \mycitep{postagger:2003}
-We evaluate the contribution of our synthetic parallel corpora using the Chinese to English task of IWSLT 2015 \mycitep{iwslt2015}, mentioned briefly in Section \ref{subsec:case_study}.
+The contribution of our synthetic parallel corpora was evaluated using the Chinese to English task of IWSLT 2015 \mycitep{iwslt2015}\emnlponly{.}\amtaonly{, mentioned briefly in Section \ref{subsec:case_study}.}
 This task uses the WIT$^3$ corpus \mycitep{wit3}, which consists of multilingual subtitles of TED talks covering a wide variety of subjects. 
@@
 
-I.e., shit I REALLY doubt I have time for...
+\amtaonly{UNDER CONSTRUCTION (basically threw everything into loosely strung sentences - will revise tomorrow}
 
-perhaps using example-based MT techniques \mycitep{ebmtbook} and crowdsourcing to generate them at scale.
+\amtaonly{misc. points that don't have full paragraphs...}
 
-For this paper, the lexical database was drawn from the IWSLT training set to maximize bottom-line performance.
-[i.e., i don't have TIME to do a general-purpose database]
-This is indeed one potential use of this system, for training set amplification.
-The envisioned ultimate target domain is (cleaned) everyday speech (maybe movie or TV subtitles?) , for which syntactic and lexical diversity is expected to be relatively low, and unknown words would be particularly suitable for crowdsourcing.
+We have not yet implemented the vast majority linguistic constructions in our rudimentary/primitive generator (for example, past tense is only partially supported).
+Future development may be expedited by adapting work from existing rule-based MT or natural language generation systems. 
+Dynamic corpus generation may also have a role to play in online or curriculum learning settings.
 
-parallel corpus construction has been considered prohibitively expensive, hence the use of pre-translated sources like parliamentary proceedings
-we believe it is worth an initial investment to create a scalable system
+Deep learning methods like RNN encoder-decoders \mycitep{sutskever:2014,bengio:2014} are just the latest in a succession of corpus-based approaches, following in the footsteps of example-based and statistical machine translation, and it is unlikely in this era of Big Data that they will be the last.
+But they all share a reliance on parallel corpora, and we believe the task of parallel corpus generation will remain relevant for the foreseeable future.
 
-expert-sourcing by linguists, using the computer to lever up - computer assisted translation, in a sense; maybe computer-amplified
-crowdsourcing by ``amateur linguists''
-
-moreover, you're at the mercy of your third-party translators - if you want more data, you gotta wait.
-But with a sufficiently compelling educational app/platform, you wouldn't have to wait OR pay for more data
-
-Crowdsourcing also has its place in post-editing, at the very least, for test sets.
+We have so far been largely concerned with using the computer to increase human translation throughput.
+But as alluded to in Section \ref{subsec:benefits}, parallel corpora have the innate potential for open collaboration; not only can they be  expert-sourced to linguists, but they can also crowdsourced to ``amateur linguists.''
+One particularly promising application domain is conversational language, for which syntactic and lexical diversity is expected to be relatively low compared with more formal domains. 
+Multilingual lexical databases, in particular, would be particularly suitable for crowdsourcing, e.g., for slang terms.
+At the very least, crowdsourcing should be strongly considered for post-editing small corpora used as potentially noise-sensitive components of shared tasks, like development or test sets.
 In a shared evaluation task with many language pairs, the most practical way for this is probably crowdsourcing.
 
-But must this be so? Vast untapped resource of students or whatever.
-an open challenge to the NLP community at large
-with a sufficiently compelling and thus widely used educational app 
-a potentially game-changing source of supervised data for NLP in general
+We believe the most promising path to large-scale parallel corpus creation is through educational crowdsourcing.
+As with the success story of reCAPTCHA \mycitep{recaptcha}, we believe this will scale best when there is a natural harmony between data supply and demand.
+Unlike other staple NLP tasks like parsing or part-of-speech tagging, MT is particularly well-suited to this, with English being taught as a foreign language in classrooms around the world.
+The data representation (English itself) is widely agreed upon and there is an existing market, a vast untapped resource of student problem sets that is yet to be harnessed by machine learning.
+\amtaonly{As a back of the envelope estimate, at a translation rate of about 1 noun per minute (based on our own data input experience), even for a modest user base of 10,000 users, a ten-minute problem set could in principle yield 100,000 data samples.
+It seems like the bottleneck would be judicious problem set design.}
 
-Well, typical approaches are paid, like Mechanical Turk... but there is a vast untapped source of data in the form of student problem sets
-as with the success story of reCAPTCHA \mycitep{recaptcha}, we believe this will scale best when there is a natural harmony between data supply and demand
+Large quantities of supervised data underlie state of the art performance in many NLP tasks.
+While current paid approaches like Amazon's Mechanical Turk are already a viable way of obtaining crowdsourced data, we believe it would be well worth the initial investment of time and resources to build a sufficiently compelling and widely used educational app capable of harvesting supervised natural language data from its natural source.
+The cost per data point could be significantly reduced or perhaps even become negative (a profitable app), although the harvested data itself would clearly have value and would quite possibly be worth subsidizing.
+We consider the construction of this potentially game-changing source of supervised NLP data to be an open challenge to the NLP community in general and large industrial practitioners in particular.
 
-Sufficient quantities of supervised data seem to underlie high or state of the art performance in NLP
-
-estimate about 1 noun per minute (my pathetic work speed)
-
-locally weighted linear regression? online learning?
-but corpus generation at test time is probably just as noisy as any discriminative method
-
-Instead of the next paragraph, should probably just point out how rudimentary the current ``grammar'' is. 
-Future development may be facilitated by adapting existing work from RBMT or NLG. 
-Okay, I guess just point out this grammar doesn't even handle something as basic as:
-
-Shit tons of linguistic phenomena that aren't even that uncommon, including but not limited to:
-More thorough treatment of articles.
-(More) treatment of prepositional phrases.
-Conjunctions.
-Interrogatives and imperatives.
-
-In the short term
-Increasing lexical variation by expanding the lexicon, especially for uncommon words in the training corpus, to alleviate data sparsity.
-Increasing syntactic variation in a high-quality way by harvesting more templates, in the old EBMT tradition
-Work along these lines is currently underway.
-
-RNN encoder-decoders are not the first corpus-based MT methodology (EBMT then SMT then this), and it is unlikely in this era of Big Data that they will be the last.
+% TODO: should I note somewhere how much mature the ML component is,
 
@@
 
-\amtaonly{
-\section{AMTA}
-This text only gets added for AMTA. I think it'd be nice to have this for ``filler'' text
-
-If I really am submitting to this conference too, then I can afford (and probably want to?) draw out the length,
-
-Should I mention that you can kind of discern differences between IWSLT, movie subtitles, and MultiUN from their POS distributions? I would probably have to mention something concrete then...? 
-} % end \amtaonly
+%\amtaonly{
+%\section{AMTA}
+%This text only gets added for AMTA. I think it'd be nice to have this for ``filler'' text
+%
+%If I really am submitting to this conference too, then I can afford (and probably want to?) draw out the length, 
+%
+%Should I mention that you can kind of discern differences between IWSLT, movie subtitles, and MultiUN from their POS distributions? I would probably have to mention something concrete then...? 
+%} % end \amtaonly
 
@@
     %Note that emnlp2016.bst chokes if there are 0 citations in your paper.    
-    \mynewcite{moses}    
+    %\mynewcite{moses}    
     \bibliographystyle{emnlp2016}

### 2016-06-03 doc/paper.tex
a halfway presentable abstract

@@
     \newcommand\mynewcite[1]{\newcite{#1}}
-    \newcommand\amtaonly[1]{#1} % currently being used to trim down to 4 pages
+    \newcommand\amtaonly[1]{} % currently being used to trim down to 4 pages
     \newcommand\emnlponly[1]{#1}
@@
 
+% TODO: follow the 300-word website limit as a guideline to abstract length? check pdf guide
 \begin{abstract}
-\amtaonly{UNDER CONSTRUCTION}
-
-The large-scale generation of high-quality synthetic parallel corpora for data-driven machine translation is proposed.
-
-A preliminary system for doing this is presented, striving toward syntactic complexity, in the spirit of rule-based machine translation, and high-quality sentence pairs via templates, in the spirit of example-based machine translation.
-
-Early results have yet to show improvement. [This is because we need to scale up?]
-Future work is proposed. 
-We believe this is particularly amenable to crowdsourcing by ``amateur linguists.''
-
+The large-scale generation of synthetic parallel corpora for data-driven machine translation is proposed.
+A rudimentary system for doing this is presented, with mechanisms for syntactic complexity in the spirit of classical rule-based machine translation and high-quality sentence pairs in the spirit of example-based machine translation.
+Preliminary experiments integrating these synthetic corpora into a shared task are reported.
+Promising future directions are discussed, in particular, large-scale contributions from ``amateur linguists'' via crowdsourcing.
 \end{abstract}
@@
 
+% TODO: mention that official results are on a slightly different Moses configuration? mehhhhh
 \caption{

### 2016-06-03 doc/paper.tex
kind of revised Future Work. meh, seems good enough...

@@
 Inspired by early work on transformations by \mynewcite{chomsky}, we focus on declarative indicative present active sentences in the hopes that they can be transformed into other forms in the long run.
-Similarly, although we enforce some semantic constraints, our system generally produces nonsensical sentences like ``{\em Colorless green ideas sleep furiously.}''
+Similarly, although we enforce some semantic constraints, our system generally produces nonsensical sentences like Chomsky's celebrated ``{\em Colorless green ideas sleep furiously.}''
 The fact that verb phrases can also serve as nouns (gerundives) and modifiers (participles) appears to be a common source of complexity, and we made it a priority to implement participial modifiers.
@@
 % TODO: mention Python?
+We plan on eventually open-sourcing both our generator and its accompanying data.
 One implementation detail worth mentioning is the choice of YAML for writing sentence templates and lexical data. 
@@
 % TODO: should I bother with this?
-Crude though it may be, we plan on eventually open-sourcing both our generator and its accompanying data.
 
@@
 %\subsection{Methods} % no space? just do one paragraph each?
-We now consider training set amplification as a first application.
-Since we are interested in the relative effect of our synthetic parallel corpus, we chose a simple baseline system and fix all other variables.
-Namely, we used the baseline configuration of Moses \mycitep{moses} as described in \mycitep{iwslt2015}, with the exception of changing the target language model to KenLM \mycitep{kenlm}, which we found to be faster and stabler than IRSTLM \mycitep{irstlm} at decoding time.
+We now consider training set amplification as a first application, using the Chinese to English MT task of IWSLT 2015 \mycitep{iwslt2015}\emnlponly{.}\amtaonly{, mentioned briefly in Section \ref{subsec:case_study}.}
+This task uses the WIT$^3$ corpus \mycitep{wit3}, which consists of multilingual subtitles of TED talks covering a wide variety of subjects. 
+The freely available MultiUN corpus \mycitep{multiun} from the OPUS project \mycitep{opus:2012} was also used for comparison.
+
+To investigate the relative effect of our synthetic parallel corpora, we chose a simple baseline system and fixed all other variables.
+Namely, we used the baseline configuration of Moses \mycitep{moses} as described in \mycitep{iwslt2015}, with the exception of changing the target language model to KenLM \mycitep{kenlm}, which we found to be faster and stabler at decoding time than IRSTLM \mycitep{irstlm}.
 For the remaining unspecified model settings, we used the {\small \tt grow-diag} alignment heuristic recommended by \mynewcite{segmenter:2008} and the {\small \tt mslr-fe} configuration for lexicalized reordering.
@@
 % TODO: filler for AMTA? used for corpus analysis stanford POS tagger (incomplete bibtex entry?) \mycitep{postagger:2003}
-The contribution of our synthetic parallel corpora was evaluated using the Chinese to English task of IWSLT 2015 \mycitep{iwslt2015}\emnlponly{.}\amtaonly{, mentioned briefly in Section \ref{subsec:case_study}.}
-This task uses the WIT$^3$ corpus \mycitep{wit3}, which consists of multilingual subtitles of TED talks covering a wide variety of subjects. 
-As a control, we also add the freely available MultiUN corpus \mycitep{multiun} from the OPUS project \mycitep{opus:2012}.
 
@@
 \begin{table}
-\label{tab:results}
 \centering
@@
 % TODO: mention that official results are on a slightly different Moses configuration? mehhhhh
-\caption{
+\caption{ \label{tab:results} % oh, for reference numbering, label goes INSIDE caption
     (Left) Chinese-to-English BLEU scores for our baseline Moses configuration with phrase and reordering tables trained on a random 25\% subset of the IWSLT15 training corpus and interpolated (+) with tables for the remaining 75\% (\textasciitilde 150K) of the in-domain corpus, the MultiUN corpus (a \textasciitilde 1M line subset), and a synthetic corpus from our generator (\textasciitilde 5M lines).
-    (Right) Same as on the left, but starting with tables for the entire IWSLT15 corpus.
+    (Right) Same as the left, but starting with tables trained on the entire IWSLT15 corpus.
     Official published scores of the IWSLT15 baseline system are also given for comparison.    
@@
 % preemptively answering reviewer question: yeah yeah, the trained tmcombine weights are like 0.001 for my generated corpora
-The results of these experiments are shown in Table \ref{tab:results}.
+The results of these experiments are shown in Table~\ref{tab:results}.
 For the table on the left, note that adding in-domain data improves performance significantly more than a much larger amount of out-of-domain data (MultiUN), a commonly reported phenomenon \mycitep{tmcombine}.
-As prepared lectures, TED talks fall somewhere between the conversational language of movie subtitles and the formal written language of UN resolutions; data from the OpenSubtitles 2016 corpus \mycitep{opensubtitles:2016}\footnote{http://www.opensubtitles.org/} from the OPUS project are similarly ineffective at improving performance (scores not shown).
-For the table on the right, note also that the same out-of-domain corpus now gives a much more modest performance increase.
-These observations suggest that amplification of in-domain corpora should still be beneficial, particularly for low-resource language pairs or domains.
+As prepared lectures, TED talks fall somewhere between the conversational language of movie subtitles and the formal written language of UN resolutions; data from the OpenSubtitles\footnote{http://www.opensubtitles.org/} 2016 corpus \mycitep{opensubtitles:2016} from the OPUS project were similarly less effective at improving performance than a much smaller amount of in-domain data (scores were similar to MultiUN and are not shown).
+For the table on the right, note the diminishing returns of out-of-domain data as the amount of in-domain data increases.
+These observations underscore the importance of domain overlap and hint at the  potential value to be found in the amplification of in-domain corpora.
 
-Our generated corpora do not make a significant impact on the final BLEU score,\footnote{
-    Results for the most recently generated synthetic corpus are shown; scores slightly higher than our baseline have been obtained on older generated corpora.
-    Both of these outcomes seem to be within the error bars of the MERT tuning procedure.
+Our generated corpora, however, do not make a significant impact on the final BLEU score,\footnote{
+    The sub-baseline scores shown are for our most recently generated synthetic corpus; scores slightly higher than our baseline have also been obtained on older generated corpora.
+    Both of these outcomes seem to be within experimental error bars.
 }
 indicating that our parallel corpus generator has a long way to go.
-Development on our generator has so far focused on ``rule-based'' syntactic diversity and growing a lexical database tailored to the training set, 
-Given these results, a change of direction seems in order; work on harvesting more templates from the IWSLT15 training corpus is currently in progress.
+Development on our generator has so far focused on ``rule-based'' syntactic diversity and growing a lexical database tailored to the training set. 
+Given the resulting scores, a change of direction seems in order; work on harvesting more templates from the IWSLT15 training corpus is currently in progress.
 
@@
 
-Deep learning methods like RNN encoder-decoders \mycitep{sutskever:2014,bengio:2014} are just the latest in a succession of corpus-based approaches, following in the footsteps of example-based and statistical machine translation, and it is unlikely in this era of Big Data that they will be the last.
-But they all share a reliance on parallel corpora, and we believe the task of parallel corpus generation will remain relevant for the foreseeable future.
+Deep learning methods like RNN encoder-decoders \mycitep{sutskever:2014,bengio:2014} are but the latest in a succession of corpus-based approaches to MT, following in the footsteps of example-based and statistical machine translation; in this era of Big Data it is unlikely that they will be the last.
+But all such methods share a reliance on parallel corpora, and we believe parallel corpus generation will remain relevant pursuit as even newer methods emerge.
 
-We have so far been largely concerned with using the computer to increase human translation throughput.
-But as alluded to in Section \ref{subsec:benefits}, parallel corpora have the innate potential for open collaboration; not only can they be  expert-sourced to linguists, but they can also crowdsourced to ``amateur linguists.''
+As alluded to in Section \ref{subsec:benefits}, parallel corpora have the innate potential for open collaboration; not only can they be expert-sourced to linguists, but they can also crowdsourced to ``amateur linguists.''
 One particularly promising application domain is conversational language, for which syntactic and lexical diversity is expected to be relatively low compared with more formal domains. 
-Multilingual lexical databases, in particular, would be particularly suitable for crowdsourcing, e.g., for slang terms.
+Multilingual lexical databases for slang terms, in particular, would be particularly suitable for crowdsourcing.
 At the very least, crowdsourcing should be strongly considered for post-editing small corpora used as potentially noise-sensitive components of shared tasks, like development or test sets.
-In a shared evaluation task with many language pairs, the most practical way for this is probably crowdsourcing.
+In a shared evaluation task with many language pairs, the most practical way for this is likely to be crowdsourcing.
 
@@
 Unlike other staple NLP tasks like parsing or part-of-speech tagging, MT is particularly well-suited to this, with English being taught as a foreign language in classrooms around the world.
-The data representation (English itself) is widely agreed upon and there is an existing market, a vast untapped resource of student problem sets that is yet to be harnessed by machine learning.
+The data format (English itself) is widely agreed upon, and there is even an existing market, a vast untapped resource of student problem sets that is yet to be harnessed by machine learning.
 \amtaonly{As a back of the envelope estimate, at a translation rate of about 1 noun per minute (based on our own data input experience), even for a modest user base of 10,000 users, a ten-minute problem set could in principle yield 100,000 data samples.
@@
 
-Large quantities of supervised data underlie state of the art performance in many NLP tasks.
+Large quantities of supervised data underlie state-of-the-art performance in many NLP tasks.
 While current paid approaches like Amazon's Mechanical Turk are already a viable way of obtaining crowdsourced data, we believe it would be well worth the initial investment of time and resources to build a sufficiently compelling and widely used educational app capable of harvesting supervised natural language data from its natural source.
-The cost per data point could be significantly reduced or perhaps even become negative (a profitable app), although the harvested data itself would clearly have value and would quite possibly be worth subsidizing.
-We consider the construction of this potentially game-changing source of supervised NLP data to be an open challenge to the NLP community in general and large industrial practitioners in particular.
+The cost per data point could be significantly reduced or perhaps even become negative (a profitable app), although the value of the harvested data itself might be worth subsidizing.
+We consider the construction of such a potentially game-changing source of supervised NLP data to be an open challenge to the NLP community in general.
 

### 2016-06-03 doc/paper.tex
scanned through my notes for anything else I might want to add

@@
 
-% TODO: follow the 300-word website limit as a guideline to abstract length? check pdf guide
+% website limits abstract to 300-word, but that's WAY too long. Mine is about 70 words, and it's short but not too short.
 \begin{abstract}
@@
     }
+% TODO: modernization of corpus acquisition / bring machine learning to bear here too
 
@@
 
+% TODO (AMTA padding?): it's rather like having an athlete with an Olympic-level modern training regimen, when it comes to diet/nutrition, they'll just eat whatever they can find on the streets.
+% Long and rich tradition of decades, if not centuries of linguists, "chefs" - want to exploit this somehow
+
 One of the major advantages of data-driven MT is that it removes the need for linguists to develop a new set of rules for an MT system to support a new language.
@@
 Such a division of labor would let linguists be linguists, with MT system development proceeding in parallel.  
+% TODO: let machines be machines, do what machines do best
 
@@
 
-% TODO: don't need to go into this much detail, do you? This would be the first section to trim, if it runs OVER length
 % To find an example of suboptimally translated parallel corpora in the literature, one need not look far.
-
+% TODO: summarize this in a footnote for emnlp? is there room?
 As an example of suboptimal parallel corpora being used for machine translation, consider the official Chinese to English baseline translation of the {\small \tt tst2010} data set\footnote{https://wit3.fbk.eu/score.php?release=2014-01} from the TED talks corpus, described in greater detail in Section \ref{sec:experiments}.
@@
 
-% TODO: table with change; also list best and worst 
 % caption should mention that the difference between 
@@
 % no, omit this, because DUH, if you change the test set, the scores will change. (it's like changing your ruler.)
-% TODO: table and reference
+% table and reference
 %\begin{table}[]
@@
 %\end{tabular}
-%\caption{By comparison, the difference between the baseline TER score and (not shown) } % TODO: caption
+%\caption{By comparison, the difference between the baseline TER score and (not shown) } 
 %\end{table}
@@
 
+% TODO: a reference to ``blessing of dimensionality'' for the inverse problem? how it's relatively easy to generate an exponentially large number of sentences?
 Other practical considerations shine a favorable light on parallel corpus generation.
@@
     Ideally the generator would have {\em a priori} knowledge about whether it is writing a sentence pair about, say,  a (financial) bank or a (river) bank.
-} and segmentation of languages like Chinese \mycitep{segmenter:2005,segmenter:2008} becomes a non-issue.
+}, segmentation of languages like Chinese \mycitep{segmenter:2005,segmenter:2008} becomes a non-issue, and there is the potential for generating annotated corpora with zero annotation error.
 Parallel corpora also provide a natural avenue through which to interface contemporary MT systems with the countless linguist-years of expertise that have been invested into rule-based MT systems, as discussed in Section \ref{sec:related}.
@@
 In terms of freely available multilingual natural language generators, the most well-developed one we could find was KPML \mycitep{kpml}.
-But sentence writing is fairly involved because the focus seems to be on having a well-developed linguistic theory to explain the surface forms.
-In contrast, our primary goal is to generate parallel corpora of surface forms, affording us a significant amount of flexibility as to the method of generation.
+But making sentences using KPML is a lengthy process, involving a well-developed linguistic theory to explain the surface forms.
+In contrast, our primary goal is to generate large parallel corpora of surface forms quickly. 
+This also affords us a significant amount of flexibility as to the method of generation.
+% TODO: for our purposes, the linguistic correctness of a generator (hokey hidden structures) is less important than whether it generates well-translated sentences.
 
@@
 \label{sec:generator}
+
+% TODO: didn't mention how you could easily create infinitely many sentences like ``the number of cats is N'', but they'll be soo correlated that they have zero translation value
 
@@
 
-% TODO: looks like this is GOING to run over length. save some if not all of this for AMTA, which is going to run way UNDER length
+% looks like this is GOING to run over length. save some if not all of this for AMTA, which is going to run way UNDER length
 %\subsection{Simplistic Linguistics}
@@
 
-% TODO: should I bother with this?
-
 % oh this part really is just flying now, because there's no uncertainty as to what I DID
-
-% running over length, but let's just write it all down and then see what to cut (hint: that stupid tst2010 stuff. maybe just summarize it in a footnote)
 

### 2016-06-04 doc/paper.tex
epicure's suggestions, and also my read through

@@
 More recently, for example, \mynewcite{collobert:2011} mention the possibility ``{\em [ideally]...to learn from letter sequences rather than words},'' illustrating an AI-motivated desire to move away from even the most fundamental of linguistic concepts.
-A specific instance of this overall trend is the gradual shift of machine translation (MT) away from classical linguistic rule-based approaches \mycitep{hutchins:2005} to contemporary machine learning approaches, which underlie statistical phrase-based systems like Moses \mycitep{moses} and more recent neural network systems \mycitep{sutskever:2014,bengio:2014}.
+A specific instance of this overall trend is the gradual shift of machine translation (MT) away from classical linguistic rule-based approaches to contemporary machine learning approaches \mycitep{hutchins:2005}, which underlie statistical phrase-based systems like Moses \mycitep{moses} and more recent neural network systems \mycitep{sutskever:2014,bengio:2014}.
 But in the end, even the most cutting-edge machine learning MT system relies crucially on multilingual parallel corpora, typically taken from external sources like multilingual parliamentary proceedings.
@@
     Of course, these parallel corpora represent hundreds of translator-hours of work. 
-    However, these translations are generally optimized for human consumption, not machine learning, with no regard for such niceties as sentence alignment.
+    However, such translations are optimized for human consumption, not machine learning, with no regard for such niceties as sentence alignment.
     \amtaonly{This is illustrated briefly in Section \ref{subsec:case_study}.}
@@
 
-\subsection{Potential Benefits}
+\amtaonly{\subsection{Potential Benefits}}
 \label{subsec:benefits}
@@
 \item Training sets for corpus-based MT systems, particularly for low-resource language pairs
-\item Benchmark tasks for MT evaluation, in the spirit of the Facebook bAbI tasks in question answering \mycitep{bAbI}
-\item Translation memories for computer-aided translation
-\item ``Language archives'': New NLP methods are developed every few years; linguistic theories change every decade or two; parallel corpora live forever (e.g., the Rosetta Stone).
+\item Benchmark tasks for MT evaluation, in the spirit of Facebook's bAbI project \mycitep{bAbI}
+\item Translation memories for computer-assisted translation
+\item ``Language archives'': New NLP methods are developed every few years; linguistic theories change every decade or so; parallel corpora live forever (like the Rosetta Stone).
 \end{itemize}
@@
 Other practical considerations shine a favorable light on parallel corpus generation.
-As the inverse problem of translation, multilingual generation is also an easier task in some senses. 
+As the ``inverse'' of translation, multilingual generation is also an easier task in some senses.
 For example, lexical ambiguity can be largely alleviated,\footnote{
-    Ideally the generator would have {\em a priori} knowledge about whether it is writing a sentence pair about, say,  a (financial) bank or a (river) bank.
-}, segmentation of languages like Chinese \mycitep{segmenter:2005,segmenter:2008} becomes a non-issue, and there is the potential for generating annotated corpora with zero annotation error.
-Parallel corpora also provide a natural avenue through which to interface contemporary MT systems with the countless linguist-years of expertise that have been invested into rule-based MT systems, as discussed in Section \ref{sec:related}.
-And unlike the various competing MT methods, which are by nature competitive, corpora are inherently collaborative.
+    Ideally the generator would ``know'' {\em a priori} whether it is writing a sentence about, say, (financial) banks or (river) banks.
+} segmentation of languages like Chinese \mycitep{segmenter:2005,segmenter:2008} becomes a non-issue, and there is the potential for generating annotated corpora with zero annotation error.
+Parallel corpora also provide a natural avenue to interface contemporary MT systems with the countless linguist-years of expertise that have been invested into rule-based MT systems.
+%, as discussed in Section \ref{sec:related}.
+% that's not really discussed there anymore, is it?
+And unlike the various different MT methods, which are by nature competitive, corpora are inherently collaborative.
 
@@
 Section \ref{sec:related} discusses previous related work,
-Section \ref{sec:generator} describes our first attempt at implementing a parallel corpus generator,
+Section \ref{sec:generator} describes our rudimentary parallel corpus generator,
 Section \ref{sec:experiments} gives preliminary results, 
@@
 The idea of using synthetic parallel corpora in machine translation is hardly new.
-Previous attempts at creating synthetic parallel corpora have used rule-based MT (RBMT) systems \mycitep{hu:2007,dugast:2008,murakami:2009,rubino:2014} but the input to the RBMT systems was again dictated by external corpora in the desire to produce immediate results.
+Previous attempts at creating synthetic parallel corpora have used rule-based MT (RBMT) systems \mycitep{hu:2007,dugast:2008,murakami:2009,rubino:2014}, but the input to the RBMT systems was again dictated by external corpora in the desire to produce immediate results.
 A longer-term approach that may pay dividends (Section \ref{subsec:benefits}) would be to run RBMT systems on specially chosen sentences for which they are known to produce high-quality translations, or perhaps eventually directly encoding their rules into carefully crafted parallel corpora.
@@
 % http://www.fb10.uni-bremen.de/anglistik/langpro/kpml/ genbank/R3b12-English/Docu/ENGLISH-nigel-exerciseset-mismatches-19981209/EX-SET-1.html
-In terms of freely available multilingual natural language generators, the most well-developed one we could find was KPML \mycitep{kpml}.
-But making sentences using KPML is a lengthy process, involving a well-developed linguistic theory to explain the surface forms.
+In terms of freely available multilingual natural language generators, the most well-developed one we found was KPML \mycitep{kpml}.
+But crafting sentences using KPML is a lengthy process, involving a well-developed linguistic theory to explain the surface forms.
 In contrast, our primary goal is to generate large parallel corpora of surface forms quickly. 
-This also affords us a significant amount of flexibility as to the method of generation.
+% This affords us an amount of flexibility as to the method of generation. % redundant and doesn't really flow
 % TODO: for our purposes, the linguistic correctness of a generator (hokey hidden structures) is less important than whether it generates well-translated sentences.
 
-In the next section, we describe our own primitive attempt at a multilingual generator. 
-While we try to model straightforward linguistic phenomena compositionally, in the tradition of example-based MT \mycitep{ebmtbook}, we adopt the pragmatic attitude that sometimes the most effective way to represent a translation phenomenon is simply to write it down.
-Thus, with the long-term goal of scalability, we attempt to construct a system sufficiently flexible to write sentence pairs in large quantities through syntactic variations and also with ``hand-crafted'' quality through example-based templates.
+In the next section, we describe our own primitive attempt at building a multilingual generator from scratch. 
+While we try to model straightforward linguistic phenomena compositionally, we also adopt the pragmatic attitude, in the tradition of example-based MT \mycitep{ebmtbook}, that sometimes the most effective way to represent a translation phenomenon is simply to write it down.
+Thus, with the long-term goal of scalability, we attempt to construct a system sufficiently flexible to write some sentence pairs in large quantities through syntactic variations and others with ``hand-crafted'' quality through example-based templates.
 
@@
 Semantic constraints can be imposed from the verb {\small \tt V} as well as externally.
-The generator can then vary the template lexically\footnote{Rewrite rule probabilities were set by hand; a more principled approach would, say, use rewrite probabilities from a generative PCFG parser.} and syntactically, allowing a large number of sentence pairs to be generated from a single template.
+The generator can then vary the template lexically\footnote{Rewrite probabilities were set by hand; a more principled approach might use, say, a generative PCFG parser.} and syntactically, allowing a large number of sentence pairs to be generated from a single template.
 We can also trade quantity for quality by adding more literals to the template, in the spirit of example-based MT.
@@
 %\subsection{Simplistic Linguistics}
+We adopt an extremely simplistic view of language as nouns, verbs, and their modifications and elaborations.
 \amtaonly{
-We adopt an exceedingly simplistic view of language as nouns, verbs, and their modifications and elaborations.
 Inspired by early work on transformations by \mynewcite{chomsky}, we focus on declarative indicative present active sentences in the hopes that they can be transformed into other forms in the long run.
 Similarly, although we enforce some semantic constraints, our system generally produces nonsensical sentences like Chomsky's celebrated ``{\em Colorless green ideas sleep furiously.}''
-The fact that verb phrases can also serve as nouns (gerundives) and modifiers (participles) appears to be a common source of complexity, and we made it a priority to implement participial modifiers.
-To capture recursive structures like participles, we adopt a rudimentary tree representation that is something intermediate between constituency trees and dependency trees, with constituent-like modifiers that also maintain dependency-like links to their targets.
-}
+}The fact that verb phrases can also serve as nouns (gerundives) and modifiers (participles) appears to be a major source of syntactic complexity, and we made it a priority to implement participial modifiers.
+To capture recursive structures like participles, we adopt a rudimentary tree representation that is something intermediate between constituency and dependency trees, with constituent-like modifiers that also maintain dependency-like links to their targets.
+We plan on eventually open-sourcing both our generator code and its accompanying data.
 
 %\subsection{Implementation Details}
+\amtaonly{
 % TODO: mention Python?
-We plan on eventually open-sourcing both our generator and its accompanying data.
 One implementation detail worth mentioning is the choice of YAML for writing sentence templates and lexical data. 
 In particular, YAML has native Unicode support and does not even require quotation marks to enclose strings.
+}
 
@@
 Phrase and lexicalized reordering tables were interpolated using the {\small \tt tmcombine} script included with Moses \mycitep{tmcombine}, with the interpolation tuned on the {\small \tt dev2010} data set.
-BLEU scores were computed as described by \mynewcite{iwslt2015} using the IWSLT 2015 progress data set ({\small \tt tst2014}).\footnote{We chose the {\small \tt tst2014} test set to facilitate comparison with systems from both IWSLT 2014 and 2015 in future work.}
+BLEU scores were computed as described by \mynewcite{iwslt2015} using the IWSLT 2015 progress data set ({\small \tt tst2014}).\footnote{We chose the {\small \tt tst2014} test set to facilitate comparison with published results from both IWSLT 2014 and 2015 in future work.}
 
@@
 % side-by-side tables as in emnlp2016.tex
-\begin{table}
+\begin{table}[ht] % h for ``here'' - float the table right here in the text, instead of at the top of page
 \centering
@@
 For the table on the left, note that adding in-domain data improves performance significantly more than a much larger amount of out-of-domain data (MultiUN), a commonly reported phenomenon \mycitep{tmcombine}.
-As prepared lectures, TED talks fall somewhere between the conversational language of movie subtitles and the formal written language of UN resolutions; data from the OpenSubtitles\footnote{http://www.opensubtitles.org/} 2016 corpus \mycitep{opensubtitles:2016} from the OPUS project were similarly less effective at improving performance than a much smaller amount of in-domain data (scores were similar to MultiUN and are not shown).
+TED talks, which are prepared lectures, fall somewhere between the conversational language of movie subtitles and the formal written language of UN resolutions.
+As such, the OpenSubtitles\footnote{http://www.opensubtitles.org/} 2016 corpus \mycitep{opensubtitles:2016} from the OPUS project was also out-of-domain; results for this corpus were similar to MultiUN and are not shown.
 For the table on the right, note the diminishing returns of out-of-domain data as the amount of in-domain data increases.
@@
 
-Our generated corpora, however, do not make a significant impact on the final BLEU score,\footnote{
+Our generated corpora do not yet make a significant impact on the final BLEU score,\footnote{
     The sub-baseline scores shown are for our most recently generated synthetic corpus; scores slightly higher than our baseline have also been obtained on older generated corpora.
@@
 Development on our generator has so far focused on ``rule-based'' syntactic diversity and growing a lexical database tailored to the training set. 
-Given the resulting scores, a change of direction seems in order; work on harvesting more templates from the IWSLT15 training corpus is currently in progress.
+Given the resulting scores, a change of direction seems in order; work on harvesting more ``example-based'' templates from the IWSLT15 training corpus is currently in progress.
 
@@
 
-\amtaonly{UNDER CONSTRUCTION (basically threw everything into loosely strung sentences - will revise tomorrow}
-
-\amtaonly{misc. points that don't have full paragraphs...}
-
-We have not yet implemented the vast majority linguistic constructions in our rudimentary/primitive generator (for example, past tense is only partially supported).
+Many linguistic constructions have yet to be implemented in our rudimentary/primitive generator (for example, past tense is only partially supported).
 Future development may be expedited by adapting work from existing rule-based MT or natural language generation systems. 
-Dynamic corpus generation may also have a role to play in online or curriculum learning settings.
+Dynamic, on-the-fly corpus generation may also have a role to play in online or curriculum learning settings.
 
 Deep learning methods like RNN encoder-decoders \mycitep{sutskever:2014,bengio:2014} are but the latest in a succession of corpus-based approaches to MT, following in the footsteps of example-based and statistical machine translation; in this era of Big Data it is unlikely that they will be the last.
-But all such methods share a reliance on parallel corpora, and we believe parallel corpus generation will remain relevant pursuit as even newer methods emerge.
+But all such methods share a reliance on parallel corpora, and we believe parallel corpus generation will remain a relevant pursuit as even newer methods emerge.
 
@@
 At the very least, crowdsourcing should be strongly considered for post-editing small corpora used as potentially noise-sensitive components of shared tasks, like development or test sets.
-In a shared evaluation task with many language pairs, the most practical way for this is likely to be crowdsourcing.
+For shared evaluation tasks with many language pairs, crowdsourcing may also be the only practical way to do this.
 
@@
 Large quantities of supervised data underlie state-of-the-art performance in many NLP tasks.
-While current paid approaches like Amazon's Mechanical Turk are already a viable way of obtaining crowdsourced data, we believe it would be well worth the initial investment of time and resources to build a sufficiently compelling and widely used educational app capable of harvesting supervised natural language data from its natural source.
-The cost per data point could be significantly reduced or perhaps even become negative (a profitable app), although the value of the harvested data itself might be worth subsidizing.
+While current paid approaches like Amazon's Mechanical Turk are already a viable way of obtaining crowdsourced data, we believe it would be well worth the initial investment of time and resources to build a educational app compelling enough to become widely adopted.
+This could then serve as a platform from which to harvest supervised natural language data from its natural source. 
+Compared with current paid approaches, the cost per datum could be significantly reduced or perhaps even become negative (a profitable app), although the value of the harvested data itself might be worth subsidizing.
 We consider the construction of such a potentially game-changing source of supervised NLP data to be an open challenge to the NLP community in general.

### 2016-06-06 main.py
fixed modal ghost-subject for random generation. tweaked nounsets accordingly to avoid zero-candidates

@@ -428,6 +428,15 @@ def randomly_configure_clause(clause, stack_depth=1, **kwargs):
     
     # TODO: nested modals with same verb read like duplicated verbs in zh... should be alleviated by scaling # verbs up...    
     clause.set_template(template_id, readonly=readonly)
+    
+    # modal ghost subject part 2 of 2 (in execution order)
+    # workaround to set ghost node from any enclosing modal here - can only be called after template has been set
+    # tag propagation should occur fine from this verb to ghost node, even if not fully instantiated...
+        # MAY result in null nodes?
+    subject_from_enclosing_modal = kwargs.pop('subject_from_enclosing_modal', None)
+    if subject_from_enclosing_modal:
+        clause.add_ghostnode('S', subject_from_enclosing_modal, kind='linked')
+    
     clause.set_random_verb_category()  
     
     # randomly add a transformation - these REQUIRE verb_category to have been set...
@@ -438,6 +447,24 @@ def randomly_configure_clause(clause, stack_depth=1, **kwargs):
     
     # finally, recurse to children
     templated_subnodes = [sn for sn in clause._subnodes() if issubclass(type(sn), nodes.TemplatedNode)]
+    
+    # modal ghost subject part 1 of 2 (in execution order)
+    # SPECIAL CASE: complement of modal verb (S V VP[complement]) needs to set up ghost subject node from enclosing nodal
+    # - ghost node can only be set AFTER C's template has been set - currently passing it in externally via kwargs
+    if template_id == 'modal':
+        S, C = [clause._get_symbol_subnode(sym) for sym in 'SC']
+        if S:
+            assert(S in templated_subnodes)
+        else:
+            assert(stack_depth > 1) # S is a ghost node from an outer modal that needs to be passed further down
+            S = clause._get_symbol_ghostnode('S')
+        assert(S)
+
+        templated_subnodes.remove(C)
+        assert(C and C not in templated_subnodes)        
+        
+        randomly_configure_clause(C, stack_depth=stack_depth+1, subject_from_enclosing_modal=S, **kwargs)
+    
     for subnode in templated_subnodes:
         randomly_configure_node(subnode, stack_depth=stack_depth+1, **kwargs)
     
@@ -466,13 +493,14 @@ def randomly_configure_np(np, **kwargs):
         forbidden_templates = np._get_option('forbidden templates')
     except KeyError:
[…]

### 2016-06-06 data.py
refactored VerbCategory and VerbSetBank to share code with PP

@@ -286,9 +286,15 @@ class VerbFormBank(WordFormBank):
         WordFormBank.__init__(self, filename)
         self.DATA_FACTORY = VerbForms
     
-
-# not currently inheriting Bank, since this is (potentially) multi-file...
-class VerbSetBank:
+    
+# not currently inheriting Bank, since this is (potentially) multi-file...    
+class CategoryBank:
+    '''
+    base class for verbsets and prepsets, both of which have
+    - semantic data, with most syntactic data coming from a separate template file (clause_templates.yml and pp_templates.yml)
+        - might want to separate into per-category files, eventually
+    - lexical data
+    '''
     def __init__(self, path):
         self.__data = {}        
         if os.path.isdir(path):  
@@ -301,20 +307,20 @@ class VerbSetBank:
             raise Exception('A path that is neither dir nor file - how zen...', path)
             
         self.__all_templates = { datum['template'] for datum in self.__data.values() }
-
+        
     def all_templates(self):
         '''Returns all available templates - mainly for testing purposes (tiny verbset list)'''
         return self.__all_templates.copy()
-            
+        
     def categories(self):
         return list(self.__data.keys())
-       
+        
     def get_category(self, category):
-        return VerbCategory(self.__data[category])
+        return self.DATA_FACTORY(self.__data[category])
         
     def get_categories_by_template(self, template_id):
         return sorted([cat for cat in self.categories() if self.__data[cat]['template'] == template_id])
-    
+        
     def __add_file(self, filename):
         new_data = read_file(filename)
         
@@ -322,6 +328,27 @@ class VerbSetBank:
         assert(type(new_data) is dict)
         assert(not any(key in self.__data.keys() for key in new_data.keys()))        
         self.__data.update(new_data)
+
+
+class PrepSetBank(CategoryBank):
+    def __init__(self, path):
+        CategoryBank.__init__(self, path)
+        self.DATA_FACTORY = PrepCategory    
+        
+class VerbSetBank(CategoryBank):
+    def __init__(self, path):
+        CategoryBank.__init__(self, path)
+        self.DATA_FACTORY = VerbCategory
+
+
+            
+
+       
+
[…]

### 2016-06-07 data.py
allow empty strings in wordsets

@@ -951,7 +951,7 @@ class PronounSet(WordSet):
 
 class PrepositionSet(WordSet):
     def _get_word_data(self, lang):
-        return self._data()[lang] or []
+        return self._data()[lang] #or []
         
 class VerbSet(WordSet):
     # verb categories' structure is different 

### 2016-06-07 nodes.py

@@ -53,7 +53,7 @@ class Node:
     def generated_text(self, lang):
         return self.__generated_text[lang] # throw an exception to reveal premature accesses #.get(lang, '')
     def has_generated_text(self, lang):
-        return bool(self.__generated_text.get(lang))
+        return self.__generated_text.get(lang) is not None
     def set_generated_text(self, lang, text):
         self.__generated_text[lang] = text
     def reset_generated_text(self):

### 2016-06-07 generator.py
verb-modifying prepositions, conditioning on verb category. surprisingly fast, after the rest of the PP infrastructure had been laid

@@ -419,6 +419,19 @@ class EnGenerator(Generator):
         adverbs = self._pop_modifiers(modifiers, 'adverb')
         if adverbs:
             result = self._modify_clause_with_adverbs(node, adverbs, result) # shared from base class
+
+        prepositions = self._pop_modifiers(modifiers, 'pp.advp.targeting.clause')
+        if prepositions:
+            for pp in prepositions:
+                pp_text = pp.generated_text(self.LANG)
+            
+                # unlike NP, the unmodified contents of a clause template are generally nontrivial, hence these acrobatics...
+                # TODO: hmm, this "reverses" the order of the PP's if there is more than one...
+                if 'O' in result:
+                    result.insert(result.index('O')+1, pp_text)                    
+                else:
+                    for head in node.head_symbols():
+                        result.insert(result.index(head)+1, pp_text)
             
         if modifiers:
             raise Exception('TODO: unhandled modifiers - {}'.format(modifiers))
@@ -427,7 +440,7 @@ class EnGenerator(Generator):
 
             
     def _modify_np(self, node):
-        template = self._get_unmodified_template(node)
+        template = self._get_unmodified_template(node) # TODO: handle custom NP templates...
         
         result = [] # most of this logic recycles nicely from the version where modifiers were owned by the LexicalNode itself
         modifiers = list(node.modifiers()) # make a copy (trashed immediately)
@@ -652,6 +665,12 @@ class ZhGenerator(Generator):
                 result = ['当时'] + result
             else:
                 raise Exception('TODO: check adverbs for time phrases')
+          
+        prepositions = self._pop_modifiers(modifiers, 'pp.advp.targeting.clause')
+        if prepositions:            
+            # TODO: query PP metadata for non-default generation orders. I suppose I could tag P
+            # hijack old adverb code (looks like 地 gets added in ADVP for adverbs). this WORKS!?    
+            result = self._modify_clause_with_adverbs(node, prepositions, result)
           
         if modifiers:
             raise Exception('TODO: unhandled modifiers - {}'.format(modifiers))

### 2016-06-07 nodes.py

@@ -837,13 +837,23 @@ class PrepositionalPhrase(TransformableNode):
         if not TransformableNode.can_modify(self, lexical_target):
             return False
         
-        #assert(issubtype(type(target_node), GenericNoun)) # for verbs, you have to look at verb category...
-        assert(lexical_target.type() == 'noun')
-        current_tags = lexical_target.semantic_tags()
-        requested_tags = sum([gt.semantic_tags() for gt in self._ghost_targets()], [])
-        
-        # don't muck with the taxonomy; just directly check whether any nouns exist fulfilling these tags (based on random np technology)        
-        return bool(data.NOUNSET_BANK.find_tagged(current_tags + requested_tags))
+        if lexical_target.type() == 'noun':
+            assert(self.template_id() == 'pp.adjp')
+            current_tags = lexical_target.semantic_tags()
+            requested_tags = sum([gt.semantic_tags() for gt in self._ghost_targets()], [])
+            
+            # don't muck with the taxonomy; just directly check whether any nouns exist fulfilling these tags (based on random np technology)        
+            return bool(data.NOUNSET_BANK.find_tagged(current_tags + requested_tags))
+            
+        elif lexical_target.type() == 'verb':
+            assert(self.template_id() == 'pp.advp.targeting.clause')
+            
+            # should be okay as long as it's a supported verb category, right? that's it?
+            # back-hacks to clause parent...
+            return lexical_target.parent().verb_category_id() in self._template().categories()
+            
+        else:
+            raise Exception('unsupported lexical target type', lexical_target.type())
         
     def set_template(self, id): #, readonly=True):
         TransformableNode.set_template(self, id, readonly=False) # always allow additions to PP from category

### 2016-06-08 main.py
adverbial preposition infrastructure

@@ -148,6 +148,19 @@ def configure_transitive_clause(clause, number='singular', subject_type='noun', 
         PO.set_template('noun')
         PO.add_options({'tags': ['object']})
     
+        # prepositions can modify verbs too
+        pp2 = nodes.node_factory('PP')
+        pp2.set_template('pp.advp.targeting.clause')
+        pp2.set_prep_category('with.using.advp')
+        
+        # testing: won't be able to modify unless random (VERB) category is compatible        
+        if category in pp2._template().categories(): 
+            clause.add_modifier(pp2)
+            PO = pp2._get_symbol_subnode('O')
+            PO.set_template('noun')
+            PO.add_options({'tags': ['object']})
+            
+        
     
     # add a determiner. oooooo
     # TODO: does this logic really belong here?
@@ -456,9 +469,19 @@ def randomly_configure_clause(clause, stack_depth=1, **kwargs):
         assert(clause.verb_category_id())
         if clause.verb_category_id() in generator.PAST_TENSE_WHITELIST and utility.rand() <= 0.75:
             clause.add_transformation('tense.past')
+            
+    # adverbial preposition
+    # TODO: whitelisted transformations for prepositions (e.g., past tense)
+    if not clause.transformation_list() and stack_depth < 3:# and utility.rand() <= 0.25:
+        pp = make_random_pp_advp(clause, stack_depth=stack_depth, **kwargs)
+        if pp:
+            clause.add_modifier(pp)
     
     # finally, recurse to children
-    templated_subnodes = [sn for sn in clause._subnodes() if issubclass(type(sn), nodes.TemplatedNode)]
+    # skips pre-configured modifier nodes, unlike _subnodes()
+    #templated_subnodes = [sn for sn in clause._subnodes() if issubclass(type(sn), nodes.TemplatedNode)]
+    symbol_subnodes = [clause._get_symbol_subnode(sym) for sym in clause._symbols()] 
+    templated_subnodes = [sn for sn in symbol_subnodes if issubclass(type(sn), nodes.TemplatedNode)]
     
     # modal ghost subject part 1 of 2 (in execution order)
     # SPECIAL CASE: complement of modal verb (S V VP[complement]) needs to set up ghost subject node from enclosing nodal
@@ -480,6 +503,24 @@ def randomly_configure_clause(clause, stack_depth=1, **kwargs):
     # TODO: configure head first? last? or is the current random order perfectly fine?
     for subnode in templated_subnodes:
[…]

### 2016-06-08 data.py
disabled testing code

@@ -976,7 +976,7 @@ DET_FORMS = { lang: DeterminerFormBank(DATA_DIR + 'dets_{}.yml'.format(lang)) fo
 NAME_BANK = NameSetBank(DATA_DIR + 'namesets.yml')
 # TODO: load language-specific name files, for any languages that might have noun declensions
 
-NOUNSET_BANK = NounSetBank(DATA_DIR + 'nounsets-test.yml')
+NOUNSET_BANK = NounSetBank(DATA_DIR + 'nounsets.yml')
 NOUN_FORMS = { lang: NounFormBank(DATA_DIR + 'nouns_{}.yml'.format(lang)) for lang in LANGUAGES }
 
 PREPSET_BANK = PrepositionSetBank(DATA_DIR + 'prepsets.yml')

### 2016-06-09 data.py
moved target category into prepset body - it doesn't need to hook into any existing Template infrastructure

@@ -816,6 +816,15 @@ class PrepCategory(Category):
     def all_prepsets(self):
         return [PrepositionSet(item) for item in self._data()['prepsets']]
         
+    def target_categories(self):
+        assert(self.template_id() == 'pp.advp.targeting.clause')
+        categories = self._data().get('target categories')
+        if categories:
+            return wrap_as_list(categories)
+        else:
+            return []
+
+
 class WordForms:
     '''Language-specific morphological forms (e.g., nouns_en.yml)'''
     def __init__(self, data):

### 2016-06-09 nodes.py

@@ -850,7 +850,8 @@ class PrepositionalPhrase(TransformableNode):
             
             # should be okay as long as it's a supported verb category, right? that's it?
             # back-hacks to clause parent...                                                     # allow all-category PP's, like "in NP(location)"
-            return (lexical_target.parent().verb_category_id() in self._template().categories()) #or self._template().categories() == ['*']
+            #return (lexical_target.parent().verb_category_id() in self._template().categories()) #or self._template().categories() == ['*']
+            return lexical_target.parent().verb_category_id() in self.__prep_category.target_categories()
             
         else:
             raise Exception('unsupported lexical target type', lexical_target.type())

### 2016-06-09 data.py
just like that, indirect objects seem to work

@@ -813,8 +813,20 @@ class VerbCategory(Category):
         return result #transforms.get(symbol)
 
 class PrepCategory(Category):
+    def __init__(self, data):
+        Category.__init__(self, data)
+        self.__ppform = {}
+
     def all_prepsets(self):
         return [PrepositionSet(item) for item in self._data()['prepsets']]
+        
+    def ppform(self, lang):
+        if lang not in self.__ppform:
+            forms_for_all_langs = self._data().get('ppforms', {})
+            forms = wrap_as_list(forms_for_all_langs.get(lang, ['standard']))
+            self.__ppform[lang] = pick_random(forms)            
+        return self.__ppform[lang]
+        
         
     def target_categories(self):
         assert(self.template_id() == 'pp.advp.targeting.clause')

### 2016-06-09 generator.py

@@ -162,6 +162,9 @@ class Generator:
             
             # populate the template - get(key, default value)
             result = [generated_symbols.get(token, token) for token in template ]
+
+            # remove any empty strings (e.g., from zh indirect object)
+            result = [item for item in result if item]
             
             if node._get_option('capitalization hack'): # currently from custom_templates.yml/multiple
                 result[0] = main.sentence_case(result[0])
@@ -423,15 +426,21 @@ class EnGenerator(Generator):
         prepositions = self._pop_modifiers(modifiers, 'pp.advp.targeting.clause')
         if prepositions:
             for pp in prepositions:
-                pp_text = pp.generated_text(self.LANG)
-            
-                # unlike NP, the unmodified contents of a clause template are generally nontrivial, hence these acrobatics...
-                # TODO: hmm, this "reverses" the order of the PP's if there is more than one...
-                if 'O' in result:
-                    result.insert(result.index('O')+1, pp_text)                    
-                else:
+                if pp.ppform(self.LANG) == 'indirect object':
+                    pp_obj = pp.generated_text(self.LANG).split(' ', maxsplit=1)[1]
                     for head in node.head_symbols():
-                        result.insert(result.index(head)+1, pp_text)
+                        result.insert(result.index(head)+1, pp_obj)                
+                else: 
+                    assert(pp.ppform(self.LANG) == 'standard')
+                    pp_text = pp.generated_text(self.LANG)
+
+                    # unlike NP, the unmodified contents of a clause template are generally nontrivial, hence these acrobatics...
+                    # TODO: hmm, this "reverses" the order of the PP's if there is more than one...
+                    if 'O' in result:
+                        result.insert(result.index('O')+1, pp_text)                    
+                    else:
+                        for head in node.head_symbols():
+                            result.insert(result.index(head)+1, pp_text)
             
         if modifiers:
             raise Exception('TODO: unhandled modifiers - {}'.format(modifiers))
@@ -668,9 +677,17 @@ class ZhGenerator(Generator):
           
         prepositions = self._pop_modifiers(modifiers, 'pp.advp.targeting.clause')
         if prepositions:            
+            if len(prepositions) is 1 and prepositions[0].ppform(self.LANG) == 'indirect object':
+                pp_text = prepositions[0].generated_text(self.LANG)                
+                result.insert(result.index('O'), pp_text)
+                
[…]

### 2016-06-09 main.py

@@ -11,7 +11,7 @@ from utility import LANGUAGES, seed_rng
 
 
 # collected flags for quick toggling during testing
-FIXED_ADJ_TAG = FIXED_NP_TAG = FIXED_ROLL = FIXED_TEMPLATE = None
+FIXED_ADJ_TAG = FIXED_NP_TAG = FIXED_ROLL = FIXED_TEMPLATE = FIXED_VERB_CATEGORY = None
 SKIP_OLD_TEST = True
 if utility.PRODUCTION:
     NUM_SENTENCES = 1000000
@@ -437,7 +437,7 @@ def randomly_configure_clause(clause, stack_depth=1, **kwargs):
             candidates += [key] * (value-1)
         
     template_id = utility.pick_random(candidates)
-    if FIXED_TEMPLATE:
+    if FIXED_TEMPLATE and not utility.PRODUCTION:
         template_id = FIXED_TEMPLATE
     
     # randomly add a transformation, like topicalization or past tense
@@ -462,7 +462,11 @@ def randomly_configure_clause(clause, stack_depth=1, **kwargs):
     if subject_from_enclosing_modal:
         clause.add_ghostnode('S', subject_from_enclosing_modal, kind='linked')
     
-    clause.set_random_verb_category()  
+    if FIXED_VERB_CATEGORY and not utility.PRODUCTION:
+        clause.set_verb_category(FIXED_VERB_CATEGORY)
+    else:
+        clause.set_random_verb_category()  
+    
     
     # randomly add a transformation - these REQUIRE verb_category to have been set...
     if do_transform and not clause.transformation_list(): # TODO: multiple transformations (might also have been applied by enclosing meta, etc.)
@@ -548,15 +552,11 @@ def randomly_configure_np(np, stack_depth=1, **kwargs):
     except KeyError:
         forbidden_templates = []
     
-    roll = utility.rand()
-    if roll < 0.05 and 'pronoun' not in forbidden_templates and data.PRONSET_BANK.find_tagged(np.semantic_tags()):
-        template_id = 'pronoun'
-    elif 0.05 <= roll < 0.2 and 'name' not in forbidden_templates and data.NAME_BANK.find_tagged(np.semantic_tags()):
-        template_id = 'name' # shouldn't make it THAT common        
-    else: 
-        assert(data.NOUNSET_BANK.find_tagged(np.semantic_tags()))
-        template_id = 'noun'
-            
[…]

### 2016-06-09 nodes.py

@@ -816,6 +816,9 @@ class PrepositionalPhrase(TransformableNode):
         assert(category_candidates)
         category = utility.pick_random(category_candidates)
         self.set_prep_category(category)
+        
+    def ppform(self, lang):
+        return self.__prep_category.ppform(lang)
     
         
     # overrides

### 2016-06-09 main.py
add indirect object MOST of the time, at randomizer level

@@ -469,14 +469,18 @@ def randomly_configure_clause(clause, stack_depth=1, **kwargs):
     
     
     # randomly add a transformation - these REQUIRE verb_category to have been set...
-    if do_transform and not clause.transformation_list(): # TODO: multiple transformations (might also have been applied by enclosing meta, etc.)
+    if False: #do_transform and not clause.transformation_list(): # TODO: multiple transformations (might also have been applied by enclosing meta, etc.)
         assert(clause.verb_category_id())
         if clause.verb_category_id() in generator.PAST_TENSE_WHITELIST and utility.rand() <= 0.75:
             clause.add_transformation('tense.past')
             
     # adverbial preposition
     # TODO: whitelisted transformations for prepositions (e.g., past tense)
-    if not clause.transformation_list() and stack_depth < 3:# and utility.rand() <= 0.25:
+    if clause.verb_category_id().endswith('indirect.object') and utility.rand() <= 0.9: # SOMETIMES you just "give", without an IO...
+        pp = make_random_pp_advp(clause, stack_depth=stack_depth, category='to.recipient.advp.indirect.object', **kwargs)
+        assert(pp)
+        clause.add_modifier(pp)
+    elif not clause.transformation_list() and stack_depth < 3 and utility.rand() <= 0.25:
         pp = make_random_pp_advp(clause, stack_depth=stack_depth, **kwargs)
         if pp:
             clause.add_modifier(pp)
@@ -511,14 +515,17 @@ def randomly_configure_clause(clause, stack_depth=1, **kwargs):
     
     
 # based on make_random_participle() and configure_random_clause() technology
-def make_random_pp_advp(target, max_runs=10, stack_depth=1, **kwargs):
+def make_random_pp_advp(target, max_runs=10, stack_depth=1, category=None, **kwargs):
     assert(target.type() == 'Clause')
         
     target_heads = target._get_headnodes()
     for i in range(max_runs):
         preposition = nodes.node_factory('PP')
         preposition.set_template('pp.advp.targeting.clause')
-        preposition.set_random_category()
+        if category:
+            preposition.set_prep_category(category)
+        else:
+            preposition.set_random_category()
         
         if all(preposition.can_modify(head) for head in target_heads):
             randomly_configure_node(preposition._get_symbol_subnode('O'), stack_depth=stack_depth+1, **kwargs)

### 2016-06-10 data.py
a couple more adverbial prepositions, and laid groundwork for verbs that must be ditransitive (like told)

@@ -311,6 +311,9 @@ class CategoryBank:
     def get_categories_by_template(self, template_id):
         return sorted([cat for cat in self.categories() if self.__data[cat]['template'] == template_id])
         
+    def _data(self):
+        return self.__data
+        
     def __add_file(self, filename):
         new_data = read_file(filename)
         
@@ -325,6 +328,20 @@ class PrepositionSetBank(CategoryBank):
         CategoryBank.__init__(self, path)
         self.DATA_FACTORY = PrepCategory    
         
+        # try paying this cost up front instead of fumbling randomly in main/randomizer
+        self.__categories_per_target_category = collections.defaultdict(set)
+        for category in self.categories():
+            category_data = PrepCategory(self._data()[category])
+            if category_data.template_id() == 'pp.advp.targeting.clause':
+                for target_category in category_data.target_categories():
+                    self.__categories_per_target_category[target_category].add(category)
+        
+    def get_categories_by_template_and_target_category(self, template_id, target_category):
+        return [cat for cat in self.get_categories_by_template(template_id) 
+                    if cat in self.__categories_per_target_category[target_category]]
+
+                    
+    
 class VerbSetBank(CategoryBank):
     def __init__(self, path):
         CategoryBank.__init__(self, path)

### 2016-06-10 main.py

@@ -419,8 +419,9 @@ def randomly_configure_clause(clause, stack_depth=1, **kwargs):
         candidates = modifiable_template_ids()
         
     # don't go too deep down the rabbit hole (human speakers and writers typically don't)
+    if stack_depth > 1:
+        candidates.remove('meta')    
     if stack_depth > 2:
-        candidates.remove('meta')
         candidates.remove('modal')
         
     # add back in some custom templates as desired... not very scalable...
@@ -437,7 +438,7 @@ def randomly_configure_clause(clause, stack_depth=1, **kwargs):
             candidates += [key] * (value-1)
         
     template_id = utility.pick_random(candidates)
-    if FIXED_TEMPLATE and not utility.PRODUCTION:
+    if FIXED_TEMPLATE and FIXED_TEMPLATE in candidates and not utility.PRODUCTION:
         template_id = FIXED_TEMPLATE
     
     # randomly add a transformation, like topicalization or past tense
@@ -476,7 +477,8 @@ def randomly_configure_clause(clause, stack_depth=1, **kwargs):
             
     # adverbial preposition
     # TODO: whitelisted transformations for prepositions (e.g., past tense)
-    if clause.verb_category_id().endswith('indirect.object') and utility.rand() <= 0.9: # SOMETIMES you just "give", without an IO...
+    # TODO: move these verb-specific PP's to a subroutine
+    if clause.verb_category_id().endswith('indirect.object') and utility.rand() <= 0.9: # SOMETIMES you just "give", without an IO... OR should this just always be on? (template the other case?)
         pp = make_random_pp_advp(clause, stack_depth=stack_depth, category='to.recipient.advp.indirect.object', **kwargs)
         assert(pp)
         clause.add_modifier(pp)
@@ -515,7 +517,8 @@ def randomly_configure_clause(clause, stack_depth=1, **kwargs):
     
     
 # based on make_random_participle() and configure_random_clause() technology
-def make_random_pp_advp(target, max_runs=10, stack_depth=1, category=None, **kwargs):
+# max_runs reduced from 10 - do the category search in PrepositionalPhrase node instead of blindly here, but still worry about other checks
+def make_random_pp_advp(target, max_runs=5, stack_depth=1, category=None, **kwargs):
     assert(target.type() == 'Clause')
         
     target_heads = target._get_headnodes()
@@ -525,12 +528,12 @@ def make_random_pp_advp(target, max_runs=10, stack_depth=1, category=None, **kwa
         if category:
             preposition.set_prep_category(category)
         else:
[…]

### 2016-06-10 nodes.py

@@ -811,11 +811,22 @@ class PrepositionalPhrase(TransformableNode):
             
             
     # TODO: DRY this out with Clause.set_random_verb_category()
-    def set_random_category(self):
-        category_candidates = data.PREPSET_BANK.get_categories_by_template(self._template_id())
-        assert(category_candidates)
-        category = utility.pick_random(category_candidates)
-        self.set_prep_category(category)
+    def set_random_category(self, target_verbs = []):
+        if target_verbs:
+            assert(self.type() == 'ADVP') # categories are also used for ADJP! (NP-PP semantics)
+            #category_candidates = [cand for cand in category_candidates  
+            # if all(verb.parent().verb_category_id() in data.PREPSET_BANK.get_category(cand).target_categories() for verb in target_verbs)]
+            assert(len(target_verbs) is 1) # well, really assuming verb category is the same for all verbs
+            verb = target_verbs[0]
+            category_candidates = data.PREPSET_BANK.get_categories_by_template_and_target_category(
+                                                self._template_id(), verb.parent().verb_category_id())
+        else:
+            category_candidates = data.PREPSET_BANK.get_categories_by_template(self._template_id())
+            assert(category_candidates)        
+
+        if category_candidates:
+            category = utility.pick_random(category_candidates)
+            self.set_prep_category(category)
         
     def ppform(self, lang):
         return self.__prep_category.ppform(lang)
@@ -850,7 +861,8 @@ class PrepositionalPhrase(TransformableNode):
             
         elif lexical_target.type() == 'verb':
             assert(self.template_id() == 'pp.advp.targeting.clause')
-            
+            assert(self.has_prep_category())
+
             # should be okay as long as it's a supported verb category, right? that's it?
             # back-hacks to clause parent...                                                     # allow all-category PP's, like "in NP(location)"
             #return (lexical_target.parent().verb_category_id() in self._template().categories()) #or self._template().categories() == ['*']
@@ -984,8 +996,9 @@ class NounPhrase(TemplatedNode):
         # NR
         # VP-NP - who knows!
         
-        
-        
+
+
+# TODO: disallow adjectives that are "included" in the noun? like "? ??"        
 class AdjectivePhrase(ModifierNode):
     def __init__(self, **kwargs): 
         ModifierNode.__init__(self, data.ADJP_TEMPLATE_BANK, **kwargs)

### 2016-06-25 doc/paper.tex
just one more hack
various new doc ideas for amta
adverbial preposition infrastructure
add indirect object MOST of the time, at randomizer level
various comments

@@
 \def\buildhcomp{2}
-\def\buildtarget{\buildemnlp} % required to be a single character for \if
+\def\buildtarget{\buildamta} % required to be a single character for \if
+
+\title{Return of the Linguist:\\Toward scalable parallel corpus generation for machine translation}
 
@@
     \newcommand\hcomponly[1]{}
+
+    % needed by Multeval tables
+    \usepackage{amstext}
+    \usepackage{multirow}
 \fi
@@
 % Long and rich tradition of decades, if not centuries of linguists, "chefs" - want to exploit this somehow
+% fueled by data
 
@@
 
+\amtaonly{
+
+\begin{table}[htb]
+\begin{center}
+%\begin{footnotesize}
+\begin{tabular}{|l|l|l|l|l|l|}
+\hline
+\bf Metric & \bf System & \bf Avg & \bf $\overline{s}_{\text{sel}}$ & \bf $s_{\text{Test}}$ & \bf $p$-value \\
+\hline
+\multirow{2}{*}{BLEU $\uparrow$}
+& baseline & 11.5 & 0.3 & - & - \\
+& system 1 & 10.9 & 0.3 & - & 0.00 \\
+\hline
+\multirow{2}{*}{TER $\downarrow$}
+& baseline & 76.0 & 0.8 & - & - \\
+& system 1 & 82.8 & 1.0 & - & 0.00 \\
+\hline
+\end{tabular}
+%\end{footnotesize}
+\end{center}
+\caption{\label{tab:scores} Metric scores for all systems generated using MultEval 0.5.1 \mycitep{multeval}: jBLEU V0.1.1 (an exact reimplementation of NIST's mteval-v13.pl without tokenization); Translation Error Rate (TER) V0.8.0.
+p-values are relative to baseline and indicate whether a difference of this magnitude (between the baseline and the system on that line) is likely to be generated again by some random process (a randomized optimizer).
+Metric scores are averages over multiple runs.
+$s_{sel}$ indicates the variance due to test set selection and has nothing to do with optimizer instability.}
+\end{table}
+
+} % amtaonly
+
 \amtaonly{\subsection{Potential Benefits}}
@@
 
+% TODO: corpus generation is (almost like?) a sub-problem
 % TODO: a reference to ``blessing of dimensionality'' for the inverse problem? how it's relatively easy to generate an exponentially large number of sentences?
+% TODO: tree-based methods become a lot easier too? I even construct a random stupid tree myself...
 Other practical considerations shine a favorable light on parallel corpus generation.
@@
 The idea of using synthetic parallel corpora in machine translation is hardly new.
+% TODO: it's not domain limitation that matters - it's the fact that they're not amplifying, just mapping sentences 1-1
 Previous attempts at creating synthetic parallel corpora have used rule-based MT (RBMT) systems \mycitep{hu:2007,dugast:2008,murakami:2009,rubino:2014}, but the input to the RBMT systems was again dictated by external corpora in the desire to produce immediate results.
@@
 % side note: if mkcls clusters were so awesome, they'd be used as standard factors... so you'd have to think that they're not
+% TODO: propose ``next-generation mkcls''? meh that's so ridiculously vague...
 
@@
 % looks like this is GOING to run over length. save some if not all of this for AMTA, which is going to run way UNDER length
+% TODO: maybe these nonsense sentences shed some light as to what is NOT a sensible, ``natural'' sentence...
 %\subsection{Simplistic Linguistics}
@@
 
+% TODO: mention that LM was held constant - since monolingual (even in-domain?) corpora are typically available at large scale
 %\subsection{Task}
@@
 % TODO: filler for AMTA? used for corpus analysis stanford POS tagger (incomplete bibtex entry?) \mycitep{postagger:2003}
+
+% TODO: propose using sub-sentence timing information?
+you already HAVE relatively fine-grained alignment information...
+but the corpus crawler just throws that away.
+Then again, you might get stuff like cross-fragment alignments...
 
@@
     (Left) Chinese-to-English BLEU scores for our baseline Moses configuration with phrase and reordering tables trained on a random 25\% subset of the IWSLT15 training corpus and interpolated (+) with tables for the remaining 75\% (\textasciitilde 150K) of the in-domain corpus, the MultiUN corpus (a \textasciitilde 1M line subset), and a synthetic corpus from our generator (\textasciitilde 5M lines).
-    (Right) Same as the left, but starting with tables trained on the entire IWSLT15 corpus.
+    (Right) Same as the left, but starting from tables trained on the entire IWSLT15 corpus.
     Official published scores of the IWSLT15 baseline system are also given for comparison.    
@@
 % TODO: maybe add Bing and Google results, since nobody seems to report these? save this for AMTA? (haha not gonna have room, sucka)
+
+% TODO: current SMT is still really expert-sourced to professional translators. crowdsourcing translations takes things to their logical extreme.
 
@@
 We consider the construction of such a potentially game-changing source of supervised NLP data to be an open challenge to the NLP community in general.
+
+% TODO: I suspect MT requirements are very similar to that of, say, AP Latin - where the best translations are the most LITERAL, not necessarily the most fluent. 
+% That also means that you might have to create TWO parallel corpora per language pair? To translate idioms in the source as compositional in the target...
+% this would be closer to training data, since human translators are lazy too.
+
+% TODO: Tarzan English (see OneNote)
 

### 2016-06-27 doc/paper.tex
seriously? they EXTENDED the deadline by 2 weeks...

@@
     However, such translations are optimized for human consumption, not machine learning, with no regard for such niceties as sentence alignment.
-    \amtaonly{This is illustrated briefly in Section \ref{subsec:case_study}.}
+    \amtaonly{
+        This is not unlike an athlete having an Olympic-level training regimen, fueled by food scrounged off the streets.
+
+        An example of suboptimal parallel corpora being used for machine translation is given in Section \ref{subsec:case_study}.
+        At the very least, it suggests a need for more careful post-editing of bitexts, particularly for development and test corpora.
+        More ideal, though, would be the creation of parallel corpora for the express purpose of training, tuning, and evaluating data-driven MT systems.
     }    
+} % footnote
 % TODO: modernization of corpus acquisition / bring machine learning to bear here too
@@
 
-% TODO (AMTA padding?): it's rather like having an athlete with an Olympic-level modern training regimen, when it comes to diet/nutrition, they'll just eat whatever they can find on the streets.
-% Long and rich tradition of decades, if not centuries of linguists, "chefs" - want to exploit this somehow
-% fueled by data
-
 One of the major advantages of data-driven MT is that it removes the need for linguists to develop a new set of rules for an MT system to support a new language.
@@
 
-% although it is ``concrete'', this section will probably be the first to get cut if I run OVER length (which had previously been unimaginable, but is starting to look more and more likely)
-% this is some stupid little side note I just noticed a couple weeks ago, but it is expanding to a huge size, because it's like the only concrete ``positive'' result I have...
-\amtaonly{
-\subsection{Case Study}
-\label{subsec:case_study}
-
-% To find an example of suboptimally translated parallel corpora in the literature, one need not look far.
-% TODO: summarize this in a footnote for emnlp? is there room?
-As an example of suboptimal parallel corpora being used for machine translation, consider the official Chinese to English baseline translation of the {\small \tt tst2010} data set\footnote{https://wit3.fbk.eu/score.php?release=2014-01} from the TED talks corpus, described in greater detail in Section \ref{sec:experiments}.
-As a very crude way of assessing translation quality, we used the official evaluation software MultEval \mycitep{multeval} to compute scores for each translated line individually.
-Upon manual inspection, the worst scoring individual lines (with TER scores over 150\%) were almost all found to correspond to poor human translations on the Chinese side of the gold corpus.\footnote{
-    High TER scores indicated misaligned sentences as well as clerical errors.
-In particular, the lecture with {\small \tt talkid} 767 had dozens of erroneous lines, in some of the worst cases having the same Chinese sentence copy/pasted three times on the same line.
-    The baseline Moses system tries valiantly, but it ultimately scores poorly for trying to produce a good machine translation of a poor human translation.
-    Alternatively, one can interpret these errors as an additional source of noise in the human encoding of English into Chinese.
-}
-Such translations might suffice for viewing translated videos monolingually, but they are far from ideal for training a machine translation system.
-
-One might expect such errors to be relatively well-tolerated in large training corpora. 
-Nevertheless, this is particularly unsettling because the {\small \tt tst2010} data set has been used as the official development/tuning set for the past several IWSLT evaluation campaigns. 
-It is also unclear what effect such noisy test corpora have on MT system evaluation and comparison.
-
-In any case, the fact that these errors have persisted this long ({\small \tt tst2010} has been distributed with only minor changes for every IWSLT Chinese-English evaluation campaign since 2012) is a testament to how insidious these errors can be, even for a leading international conference.
-At the very least, this suggests a need for more careful post-editing of bitexts, particularly for development and test corpora.
-More ideal, though, would be the creation of parallel corpora for the express purpose of training, tuning, and evaluating data-driven MT systems.
-
-% TODO: display an example bad-sentence pair? If I don't, then what was the point of getting zh fonts working?? sigh
-
-% I guess don't be concrete - don't say it... because then, you have to be more concrete about it...
-	% I can elaborate if reviewers ask... but for now, let's just be concise...
-
-% caption should mention that the difference between 
-    % ALTHOUGH you would expect that it would improve all models, it's not clear whether...err? hard to be precise here.
-%Removing TED talk number 767 (over 200 lines), as well as 15 hand-picked lines with bad TER scores, 
-%First, note that the scores improve, which is consistent with the observation that 
-% something to keep in mind in case reviewers ask
-%This results in a significant TER score change that is almost as large as the score difference between the highest and lowest ranking system from the evaluation campaign.  
-%BLEU scores \mycitep{bleu}, on the other hand, did not change much.
-% no, omit this, because DUH, if you change the test set, the scores will change. (it's like changing your ruler.)
-% table and reference
-%\begin{table}[]
-%\centering
-%\label{table:tst2010}
-%\begin{tabular}{|l|l|l|}
-%\hline
-% & BLEU & TER  \\ \hline
-%Full {\small \tt tst2010} & 11.2 & 77.0  \\ \hline
-%Purged {\small \tt tst2010} & 11.8 & 71.3  \\ \hline
-%%\hline
-%%{\small \tt tst2014} &  &   \\ \hline
-%% &  &   \\ \hline
-%\end{tabular}
-%\caption{By comparison, the difference between the baseline TER score and (not shown) } 
-%\end{table}
-% TODO: table of scores after purging. This is pretty crucial? why?
-	% i think i THOUGHT this was crucial because i had to back up my claims concretely...
-	% but the claim from the data is tenuous at best - of course the scores change if the test set changes! so WHAT?
-	% BESIDES, I tried training and tuning with purged corpora, and the difference was minimal. it's quite possible that people just don't CARE
-} % \amtaonly
-
-\amtaonly{
-
-\begin{table}[htb]
-\begin{center}
-%\begin{footnotesize}
-\begin{tabular}{|l|l|l|l|l|l|}
-\hline
-\bf Metric & \bf System & \bf Avg & \bf $\overline{s}_{\text{sel}}$ & \bf $s_{\text{Test}}$ & \bf $p$-value \\
-\hline
-\multirow{2}{*}{BLEU $\uparrow$}
-& baseline & 11.5 & 0.3 & - & - \\
-& system 1 & 10.9 & 0.3 & - & 0.00 \\
-\hline
-\multirow{2}{*}{TER $\downarrow$}
-& baseline & 76.0 & 0.8 & - & - \\
-& system 1 & 82.8 & 1.0 & - & 0.00 \\
-\hline
-\end{tabular}
-%\end{footnotesize}
-\end{center}
-\caption{\label{tab:scores} Metric scores for all systems generated using MultEval 0.5.1 \mycitep{multeval}: jBLEU V0.1.1 (an exact reimplementation of NIST's mteval-v13.pl without tokenization); Translation Error Rate (TER) V0.8.0.
-p-values are relative to baseline and indicate whether a difference of this magnitude (between the baseline and the system on that line) is likely to be generated again by some random process (a randomized optimizer).
-Metric scores are averages over multiple runs.
-$s_{sel}$ indicates the variance due to test set selection and has nothing to do with optimizer instability.}
-\end{table}
-
-} % amtaonly
-
-\amtaonly{\subsection{Potential Benefits}}
-\label{subsec:benefits}
+%\amtaonly{\subsection{Potential Benefits}}
+\label{subsec:benefits} % ha, this works even if it's not a subsection - LaTeX just figures out what sec/subsec it's in
 
@@
 Section \ref{sec:generator} describes our rudimentary parallel corpus generator,
-Section \ref{sec:experiments} gives preliminary results, 
+Section \ref{sec:experiments} gives experimental results, 
 and Section \ref{sec:future} proposes future work.
@@
 % these links are too long for EMNLP. TODO: add them as footnotes for AMTA?
-% http://www.fb10.uni-bremen.de/anglistik/langpro/kpml/genbank/generation-bank.html
-% http://www.fb10.uni-bremen.de/anglistik/langpro/kpml/ genbank/R3b12-English/Docu/ENGLISH-nigel-exerciseset-mismatches-19981209/EX-SET-1.html
+%
+
 In terms of freely available multilingual natural language generators, the most well-developed one we found was KPML \mycitep{kpml}.
-But crafting sentences using KPML is a lengthy process, involving a well-developed linguistic theory to explain the surface forms.
+But crafting sentences using KPML is a lengthy process,\amtaonly{\footnote{
+    For examples, see http://www.fb10.uni-bremen.de/anglistik/langpro/kpml/genbank/generation-bank.html}} involving a well-developed linguistic theory to explain the surface forms.
 In contrast, our primary goal is to generate large parallel corpora of surface forms quickly. 
@@
 Semantic constraints can be imposed from the verb {\small \tt V} as well as externally.
-The generator can then vary the template lexically\footnote{Rewrite probabilities were set by hand; a more principled approach might use, say, a generative PCFG parser.} and syntactically, allowing a large number of sentence pairs to be generated from a single template.
+The generator can then vary the template lexically\footnote{
+    Rewrite probabilities were set by hand; a more principled approach might use, say, a generative PCFG parser.} and syntactically, allowing a large number of sentence pairs to be generated from a single template.
 We can also trade quantity for quality by adding more literals to the template, in the spirit of example-based MT.
 
-% looks like this is GOING to run over length. save some if not all of this for AMTA, which is going to run way UNDER length
 % TODO: maybe these nonsense sentences shed some light as to what is NOT a sensible, ``natural'' sentence...
@@
 
-\section{Preliminary Experiments}
-\label{sec:experiments}
+\section{Experimental Results}
 
 %\subsection{Methods} % no space? just do one paragraph each?
+\amtaonly{\subsection{Training Set Amplification}}
+\label{sec:experiments}
 We now consider training set amplification as a first application, using the Chinese to English MT task of IWSLT 2015 \mycitep{iwslt2015}\emnlponly{.}\amtaonly{, mentioned briefly in Section \ref{subsec:case_study}.}
@@
 Namely, we used the baseline configuration of Moses \mycitep{moses} as described in \mycitep{iwslt2015}, with the exception of changing the target language model to KenLM \mycitep{kenlm}, which we found to be faster and stabler at decoding time than IRSTLM \mycitep{irstlm}.
+Since monolingual corpora are typically available at scale, target language models were obtained using the same in-domain monolingual corpus for all runs.
 For the remaining unspecified model settings, we used the {\small \tt grow-diag} alignment heuristic recommended by \mynewcite{segmenter:2008} and the {\small \tt mslr-fe} configuration for lexicalized reordering.
 Chinese text was preprocessed using the latest version of the Stanford Segmenter \mycitep{segmenter:2005,segmenter:2008}.
-Phrase and lexicalized reordering tables were interpolated using the {\small \tt tmcombine} script included with Moses \mycitep{tmcombine}, with the interpolation tuned on the {\small \tt dev2010} data set.
-BLEU scores were computed as described by \mynewcite{iwslt2015} using the IWSLT 2015 progress data set ({\small \tt tst2014}).\footnote{We chose the {\small \tt tst2014} test set to facilitate comparison with published results from both IWSLT 2014 and 2015 in future work.}
-
-% TODO: mention that LM was held constant - since monolingual (even in-domain?) corpora are typically available at large scale
-%\subsection{Task}
-
-% TODO: filler for AMTA? used for corpus analysis stanford POS tagger (incomplete bibtex entry?) \mycitep{postagger:2003}
+Phrase and lexicalized reordering tables were interpolated using the {\small \tt tmcombine} script included with Moses \mycitep{tmcombine}, with the interpolation optimized on the {\small \tt dev2010} data set.
+Following \mynewcite{iwslt2015}, the model was tuned on the {\small \tt tst2010} data set using the MERT procedure included in Moses.
+For Table~\ref{tab:results}, BLEU scores were computed as described by \mynewcite{iwslt2015} using the IWSLT 2015 progress data set ({\small \tt tst2014}).\footnote{
+    We chose the {\small \tt tst2014} test set to facilitate comparison with published results from both IWSLT 2014 and 2015 in future work.
+} % might be nice to throw this sentence into the Table caption, but LATeX chokes on the \footnote{}
 
@@
 %zh\_zh corpus from the OpenSubtitles 2016 corpus , with lines cleaned using some ad hoc heuristics
-
-%\subsection{Results}
-
-% yeah sure, let's add Bing/Google results just for kicks, then take them out later
-
-% definitely want 2 tables?
-    % one for learning curves, to show that iwslt > zh_zh > multiun - domain overlap is greater
-
-    % one for 100%, to show that the out of domain corpora aren't helping that much. need moar in-domain data (probably)
-    % well, maybe i'll do away with this one
-
-    % I guess don't bother showing single-corpus results for zh_zh and multiun, because then I'd have to show MY single-corpus results, which are even more embarrassing than the horribleness I've written up here
-
-% when i actually go and READ the table, it's clear that I don't even want to LOOK at more than one metric. it's information overload.
-%\begin{table}
-%\centering
-%\label{table:tst2010}
-%\begin{tabular}{|l|r|r|r|}
-%\hline
-%                    & BLEU  & NIST & TER    \\ \hline
-%IWSLT15 official    & 11.43 & 4.67 & 72.65  \\ \hline
-%Our baseline        & 11.54 & 4.43 & 72.43  \\ \hline
-%  + OpenSubtitles   & 11.54 & 4.43 & 72.43  \\ \hline
-%  + MultiUN         & 11.67 & 4.48 & 72.60  \\ \hline
-%  + generated       & 11.46 & 4.39 & 72.51  \\ \hline
-%% &  &   \\ \hline
-%\end{tabular}
-%\caption{}
-% 
-%\end{table}
 
@@
   + 75\%            & \textbf{11.47}  \\ \hline
-%  + Subtitles       & 10.57  \\ \hline
   + MultiUN         & 10.76  \\ \hline
@@
 Our baseline        & 11.54  \\ \hline
-%  + Subtitles       & 11.54  \\ \hline
   + MultiUN         & \textbf{11.67}  \\ \hline
@@
     Qualitatively similar trends for NIST and TER scores were also observed (omitted for clarity).
-%    The ``baseline'' row gives 
-%    MultiUN is actually more data because it has more words per line.
-% automatically cleaned version of the en-zh_zh language pair of OpenSubtitles 2016, roughly 1M lines.
-% and a similar number of lines from MultiUN, which is actually more words. but it doesn't do as well.
 } % caption
@@
 Given the resulting scores, a change of direction seems in order; work on harvesting more ``example-based'' templates from the IWSLT15 training corpus is currently in progress.
+\amtaonly{As a first step in this direction, we examine the gold corpora qualitatively.}
 
 % TODO: maybe add Bing and Google results, since nobody seems to report these? save this for AMTA? (haha not gonna have room, sucka)
+% but should also add IWSLT-best results if you're doing that - and I don't see the point, unless my results were anywhere NEAR them
+% nah, if at all, just mention in a table caption briefly somewhere
+
+% new results for AMTA, and adding the old ``case study''
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\amtaonly{
+\subsection{Qualitative Analysis of Corpora}
+\label{subsec:case_study}
+
+Consider the official Chinese to English baseline translation of the {\small \tt tst2010} data set\footnote{
+    \label{foot:tst2010}https://wit3.fbk.eu/score.php?release=2014-01
+}.
+As a crude way of assessing translation quality, we used the official evaluation software MultEval \mycitep{multeval} to compute scores for each translated line individually.
+Upon manual inspection, the worst scoring individual lines (especially above 150\%) were almost all found to correspond to poor human translations on the Chinese side of the gold corpus.
+This is particularly unsettling because the {\small \tt tst2010} data set has been used as the official development/tuning set for the past several IWSLT evaluation campaigns. 
+The fact that these errors have persisted this long ({\small \tt tst2010} has been distributed with only minor changes for every IWSLT Chinese-English evaluation campaign since 2012) is a testament to how insidious these errors can be, even for a prominent international workshop.
+
+% TODO: display an example bad-sentence pair? If I don't, then what was the point of getting zh fonts working?? sigh
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{|l|l|}
+\hline
+Gold zh & \zh{安德森：是六万美元。} \\ \hline
+Gold en & 60,000. \\ \hline
+Moses output & CA : That is 60,000 dollars .  \\ \hline
+\end{tabular}
+\caption{ \label{tab:bad_pair}
+   A ``poorly translated'' sentence pair from {\small \tt tst2010}, with a TER score of 250\%.    
+}
+\end{table}
+
+An example is given in Table~\ref{tab:bad_pair}.
+The baseline Moses system tries valiantly, but it ultimately scores poorly for producing a good machine translation of a liberal human translation.
+Alternatively, one can interpret these errors as an additional source of noise in the human encoding of English into Chinese.
+It is not entirely clear how the MT process is affected by this kind of noise.
+
+The most effective automated ways we found to identify lines in dire need of post-editing were:
+\begin{itemize}
+\item 
+Duplicated Chinese segments on the same line. 
+This was presumably caused by having the same sentence displayed at multiple time points in the video subtitles. 
+Often, would also indicate passages that were poorly aligned at the sentence level.
+Such translations might suffice for viewing translated videos monolingually, but they are far from ideal for training a machine translation system.
+
+\item 
+Missing segments (usually English) when compared with the original lecture transcripts online.\footnote{http://www.ted.com/} 
+This issue seems to have been introduced in the IWSLT15 Chinese-English corpus. 
+When applied to the IWSLT14 corpus, most of these would be ``false positives'' indicating post-editing of musical symbols and other artifacts.
+True positives were often indicative of misaligned sentences.
+\end{itemize}
+
+\subsection{Experiments on Post-Edited Corpora}
+\label{subsec:post_edited}
+
+% SLIGHTLY inconsistent formatting, since now I have discussion ABOVE the table instead of below...
+% but this creates a natural break for the discussion 
+Table~\ref{tab:tst2010} shows the results of ``lightly'' post-editing {\small \tt tst2010}. 
+On the Chinese side, duplicated segments, applause lines, and translator notes were removed. 
+Speaker annotations were normalized on both sides, as well as sentence alignments wherever possible.
+Small but noticeable differences in scores are observed. 
+(By comparison, the difference between the best and worst scoring Chinese-to-English systems in IWSLT14 \mycitep{iwslt2014} was 6 BLEU points and 7 TER points.)
+In this case, the post-edited tuning set seems to guide the MERT optimization procedure of Moses to a slightly different local optimum.
+
+% my baseline - 2016-05-15-213619
+% tst2010 fixed lite - 2016-06-17-131927
+\begin{table}[ht]
+\centering
+\begin{tabular}{|l|r|r|r|}
+\hline
+                    & BLEU  & NIST & TER    \\ \hline
+(a) Official IWSLT15    & 11.43 & \textbf{4.67} & 72.65  \\ \hline
+(b) Our baseline        & 11.54 & 4.43 & \textbf{72.43} \\ \hline
+(c) Post-edited {\small \tt tst2010} & \textbf{11.91} & 4.33 & 75.34 \\ \hline
+% &  &   \\ \hline
+\end{tabular}
+    % TODO: fairly labor intensive, not THAT much payoff? but there is definitely a payoff, and if you could crowd-lever it...
+
+\caption{ \label{tab:tst2010}
+    Chinese-to-English translation scores evaluated on the {\small \tt tst2014} data set, with experimental parameters as described in Section~\ref{sec:experiments}. 
+    (a) Official published IWSLT15 baseline scores \mycitep{iwslt2015}, for comparison.
+    (b) Our baseline Moses configuration, which uses KenLM instead of IRSTLM, as described previously.
+    (c) Same as (b), but with a post-edited tuning set {\small \tt tst2010}. 
+} % caption
+\end{table}
+
+For the rest of this section, we show the results of applying the same post-editing procedure to the test and training sets as well, with the additional step of restoring missing segments for both languages.
+To preserve the integrity of {\small \tt tst2014}, evaluation is now performed using MultEval \mycitep{multeval} on the {\small \tt tst2010} data set itself, with the MERT procedure tuned on the {\small \tt dev2010} data set instead.
+These are the data sets and evaluation software used by the official benchmark cited in footnote \ref{foot:tst2010}.
+MultEval has the added benefit of providing bootstrapped error bar estimates.
+
+% TODO TODO TODO: rerun these jobs with GIZA++ instead of mgiza! TINY difference in error bars, but should try to be as careful as possible
+    % I suppose I MIGHT consider this if the paper is ACCEPTED, haha
+\begin{table}[htb]
+\begin{center}
+%\begin{footnotesize}
+\begin{tabular}{|l|l|l|l|l|l|}
+\hline
+\bf Metric & \bf System & \bf Avg & \bf $\overline{s}_{\text{sel}}$ & \bf $s_{\text{Test}}$ & \bf $p$-value \\
+\hline
+\multirow{2}{*}{BLEU $\uparrow$}
+& baseline & 11.5 & 0.3 & - & - \\
+& system 1 & 10.9 & 0.3 & - & 0.00 \\
+\hline
+\multirow{2}{*}{TER $\downarrow$}
+& baseline & 76.0 & 0.8 & - & - \\
+& system 1 & 82.8 & 1.0 & - & 0.00 \\
+\hline
+\end{tabular}
+%\end{footnotesize}
+\end{center}
+\caption{\label{tab:scores} Metric scores for all systems generated using MultEval 0.5.1 \mycitep{multeval}: jBLEU V0.1.1 (an exact reimplementation of NIST's mteval-v13.pl without tokenization); Translation Error Rate (TER) V0.8.0.
+p-values are relative to baseline and indicate whether a difference of this magnitude (between the baseline and the system on that line) is likely to be generated again by some random process (a randomized optimizer).
+Metric scores are averages over multiple runs.
+$s_{sel}$ indicates the variance due to test set selection and has nothing to do with optimizer instability.}
+\end{table}
+
+Input-domain noise can play a role similar to regularization in machine learning - get a reference from denoising autoencoders?
+We do indeed find that making the tuning set translations more literal also tends to decrease performance, but this could be because of premature stopping.
+Also, the amount of noise added is generally some small amount of Gaussian noise, and it is unclear
+
+For tst2014, revising the DEV set tst2010 instead leads to different local minima instead. 
+A better explanation is that the effect of this amount and kind of noise is ill-defined.
+
+% Data Snooping - dedicate this to its own section, to make clear that these are not ``REAL'' results
+% we perform snooping on a dedicated snooping set (as if that makes it right...)
+
+} % amtaonly
 
@@
 % TODO: wtf, ``chinese'' is not capitalized in stupid bibliography
+
+% TODO: SMT as dynamic-length or variable-length EBMT. just a qualitative observation of mine...but I suppoose, thinking about SMT mechanistically, that's what it is, really
 

### 2016-06-28 doc/paper.tex
added MultEval tables

@@
 To preserve the integrity of {\small \tt tst2014}, evaluation is now performed using MultEval \mycitep{multeval} on the {\small \tt tst2010} data set itself, with the MERT procedure tuned on the {\small \tt dev2010} data set instead.
-These are the data sets and evaluation software used by the official benchmark cited in footnote \ref{foot:tst2010}.
+Our baseline model is also trained on the IWSLT corpus Chinese-English training corpus from 2014 instead of 2015.
+These data sets and this evaluation software were used by the official benchmark cited in footnote \ref{foot:tst2010}.
 MultEval has the added benefit of providing bootstrapped error bar estimates.
 
-% TODO TODO TODO: rerun these jobs with GIZA++ instead of mgiza! TINY difference in error bars, but should try to be as careful as possible
-    % I suppose I MIGHT consider this if the paper is ACCEPTED, haha
+% TODO: I am here. the rest of the stuff from here to ``Proposed Future Work'' is not finished yet. Just draft entries...
+
+% In Table~\ref{tab:multeval_tst2010} 
+In the tables below with MultEval results, metric scores for all systems were generated using MultEval 0.5.1 \mycitep{multeval}: jBLEU V0.1.1 (an exact reimplementation of NIST's mteval-v13.pl without tokenization); Translation Error Rate (TER) V0.8.0.
+The $p$-values are relative to the baseline and indicate whether a difference of this magnitude (between the baseline and the system on that line) is likely to be generated again by some random process (a randomized optimizer).
+Metric scores are averages over three runs per system. 
+$s_{sel}$ indicates the variance due to test set selection and has nothing to do with optimizer instability.\footnote{
+    This paragraph and the tables with MultEval results are adopted from \LaTeX~code automatically generated by MultEval.
+}
+
+% \pagebreak % soooooo ugly
+
+% well, looks like I'm stuck doing discussion THEN table... (tab:multeval_tst2010 won't fit on the same page if before)
+Let's discuss Table~\ref{tab:multeval_tst2010}.
+Train2015 adds more data, but it's noisy and does not produce a statistically significant change in scores.
+On the other hand, post-edited {\small \tt train2015} produces a significant improvement in scores.
+(Quality matters - not just quantity, which is something we've been saying. Post-editing makes a significant difference.)
+% TODO: post-editing DOES provide performance boost. But it's labor intensive, and a SMALL boost...
+Unlike Table~\ref{tab:tst2010}, post-editing the MERT tuning set {\small \tt dev2010} actually makes things uniformly worse.
+The effect of this kind of noise in tuning corpora appears to be ill-defined.
+Input-domain noise can play a role similar to regularization in machine learning - get a reference from denoising autoencoders?
+We do indeed find that making the tuning set translations more literal also tends to decrease performance, but this could be because of premature stopping.
+Also, the amount of noise added is generally some small amount of Gaussian noise, and it is unclear
+
 \begin{table}[htb]
@@
 \hline
-\multirow{2}{*}{BLEU $\uparrow$}
-& baseline & 11.5 & 0.3 & - & - \\
-& system 1 & 10.9 & 0.3 & - & 0.00 \\
+\multirow{5}{*}{BLEU $\uparrow$}
+& Official IWSLT14 baseline  & 11.2 & -   & -   & - \\
+& Our baseline & 11.4 & 0.3 & 0.2 & - \\
+& Trained on 2015 corpus ({\small \tt train2015}) & 11.5 & 0.3 & 0.3 & 0.30 \\
+& Post-edited: {\small \tt train2015} & \textbf{11.7} & 0.3 & 0.1 & 0.01 \\
+& Post-edited: {\small \tt dev2010} & 10.8 & 0.3 & 0.3 & 0.00 \\
+%& Post-edited: {\small \tt train2015} and {\small \tt dev2010} & 11.0 & 0.3 & 0.2 & 0.00 \\
 \hline
-\multirow{2}{*}{TER $\downarrow$}
-& baseline & 76.0 & 0.8 & - & - \\
-& system 1 & 82.8 & 1.0 & - & 0.00 \\
+\multirow{5}{*}{TER $\downarrow$}
+& Official IWSLT14 baseline & 77.0 & -   & -   & - \\
+& Our baseline & 76.2 & 0.8 & 0.3 & - \\
+& Trained on {\small \tt train2015} & 76.4 & 0.8 & 0.5 & 0.26 \\
+& Post-edited: {\small \tt train2015} & \textbf{75.6} & 0.8 & 0.0 & 0.00 \\
+& Post-edited: {\small \tt dev2010} & 83.4 & 1.0 & 0.5 & 0.00 \\
+%& system 4 & 82.7 & 1.0 & 0.8 & 0.00 \\
 \hline
@@
 \end{center}
-\caption{\label{tab:scores} Metric scores for all systems generated using MultEval 0.5.1 \mycitep{multeval}: jBLEU V0.1.1 (an exact reimplementation of NIST's mteval-v13.pl without tokenization); Translation Error Rate (TER) V0.8.0.
-p-values are relative to baseline and indicate whether a difference of this magnitude (between the baseline and the system on that line) is likely to be generated again by some random process (a randomized optimizer).
-Metric scores are averages over multiple runs.
-$s_{sel}$ indicates the variance due to test set selection and has nothing to do with optimizer instability.}
+
+\caption{\label{tab:multeval_tst2010} % \caption[LoF entry] would allow multi-paragraph caption... \hspace{0.6 cm} to indent
+%\-\caption[LoF entry]{ \label{tab:multeval_tst2010}
+MultEval results for systems evaluated on the unmodified {\small \tt tst2010} data set.
+Official baseline results (with IRSTLM instead of KenLM as discussed previously) are also given for comparison.
+%UHHHH do I have anything else to say here in the immediate caption? 
+%I'll interpret results in the text body.
+%But this caption needs to be kind of short, because otherwise it winds up on a different page than its discussion... although at this rate, it doesn't really matter - that'll happen anyway
+} % caption
 \end{table}
 
-Input-domain noise can play a role similar to regularization in machine learning - get a reference from denoising autoencoders?
-We do indeed find that making the tuning set translations more literal also tends to decrease performance, but this could be because of premature stopping.
-Also, the amount of noise added is generally some small amount of Gaussian noise, and it is unclear
+% TODO: Adding train2015 + 
+% this is pointless to say, unless I also have unmodified train2015 + dev2010?
 
-For tst2014, revising the DEV set tst2010 instead leads to different local minima instead. 
-A better explanation is that the effect of this amount and kind of noise is ill-defined.
+\pagebreak
+Now we discuss Table~\ref{tab:multeval_tst2010-fixed}.
+This time there is a statistically significant improvement in TER for train2015, but not for BLEU.
+Post-editing train2015 again leads to further improvement.
+More surprisingly, post-edited dev2010 gets slightly better BLEU at the expense of much worse TER; this is more like Table~\ref{tab:tst2010} than Table~\ref{tab:multeval_tst2010}.
+This is counter-intuitive, since you would think that removing sentences with duplicated segments from tst2010 would improve TER scores, not worsen them.
+This emphasizes the fact that tuning set noise can affect bottom-line performance in relatively unpredictable ways.
 
+\begin{table}[htb]
+\begin{center}
+%\begin{footnotesize}
+\begin{tabular}{|l|l|l|l|l|l|}
+\hline
+\bf Metric & \bf System (all with post-edited {\small \tt tst2010}) & \bf Avg & \bf $\overline{s}_{\text{sel}}$ & \bf $s_{\text{Test}}$ & \bf $p$-value \\
+\hline
+\multirow{5}{*}{BLEU $\uparrow$}
+& Our baseline & 11.4 & 0.3 & 0.2 & - \\
+& Trained on {\small \tt train2015} & 11.4 & 0.3 & 0.2 & 0.44 \\
+& Post-edited: {\small \tt train2015} & 11.6 & 0.3 & 0.2 & 0.01 \\
+& Post-edited: {\small \tt dev2010} & \textbf{11.9} & 0.3 & 0.1 & 0.00 \\
+%& Post-edited: {\small \tt train2015} and {\small \tt dev2010} & \textbf{12.2} & 0.3 & 0.1 & 0.00 \\
+\hline
+\multirow{5}{*}{TER $\downarrow$}
+& Our baseline & 70.6 & 0.4 & 0.5 & - \\
+& Trained on {\small \tt train2015} & 70.1 & 0.4 & 0.3 & 0.00 \\
+& Post-edited: {\small \tt train2015} & \textbf{69.7} & 0.4 & 0.2 & 0.00 \\
+& Post-edited: {\small \tt dev2010} & 75.1 & 0.5 & 0.3 & 0.00 \\
+%& Post-edited: {\small \tt train2015} and {\small \tt dev2010} & 74.7 & 0.5 & 0.2 & 0.00 \\
+\hline
+\end{tabular}
+%\end{footnotesize}
+\end{center}
+
+\caption{\label{tab:multeval_tst2010-fixed} % \caption[LoF entry] would allow multi-paragraph caption... \hspace{0.6 cm} to indent
+%\-\caption[LoF entry]{ \label{tab:multeval_tst2010}
+MultEval results for systems evaluated on the post-edited {\small \tt tst2010} data set.
+} % caption
+\end{table}
+
+\subsection{Data Snooping}
 % Data Snooping - dedicate this to its own section, to make clear that these are not ``REAL'' results
-% we perform snooping on a dedicated snooping set (as if that makes it right...)
+we perform snooping on a dedicated snooping set (as if that makes it right...)
+
+uhhh there's not enough room for a table and discussion, is there? i guess i'd have to pad with some philosophical discussion about snooping
+
+\begin{table}[htb]
+\begin{center}
+%\begin{footnotesize}
+\begin{tabular}{|l|l|l|l|l|l|}
+\hline
+\bf Metric & \bf System & \bf Avg & \bf $\overline{s}_{\text{sel}}$ & \bf $s_{\text{Test}}$ & \bf $p$-value \\
+\hline
+\multirow{3}{*}{BLEU $\uparrow$}
+& baseline & 11.4 & 0.3 & 0.2 & - \\
+& system 1 & 36.8 & 0.6 & 0.0 & 0.00 \\
+& system 2 & 11.2 & 0.3 & 0.1 & 0.03 \\
+\hline
+\multirow{3}{*}{TER $\downarrow$}
+& baseline & 76.2 & 0.8 & 0.3 & - \\
+& system 1 & 48.3 & 0.7 & 0.0 & 0.00 \\
+& system 2 & 76.3 & 0.7 & 0.1 & 0.46 \\
+\hline
+\end{tabular}
+%\end{footnotesize}
+\end{center}
+
+\caption{\label{tab:multeval_tst2010-fixed} % \caption[LoF entry] would allow multi-paragraph caption... \hspace{0.6 cm} to indent
+%\-\caption[LoF entry]{ \label{tab:multeval_tst2010}
+MultEval results for systems evaluated on the post-edited {\small \tt tst2010} data set.
+System 1: trained on the concatenation of train14 and tst2010. SNOOPING!
+System 2: snooped out a bunch of mistranslated or OOV words from baseline translation and generated sentences with those words using the generator (remember that thing?)
+} % caption
+\end{table}
+
+The rationale for not bothering with tmcombine was that the additional data was very small in comparison.
+Shouldn't skew the tables that much or whatever.
+Yet there is still the ``unreasonable effectiveness of data'' or whatever - but really it's just overfitting...
+
+Snooping data set establishes an ``upper bound'' of sorts of how much performance boost you could get by simply adding more data.
+It's a measure of the inherent limitation (bias?) of the model being used.
+Generated data doesn't make a dent, and in fact makes things slightly worse (although TER is within error bars).
+So nonsense sentences of OOV words does NOT help - or even if it does, the nonsense-noise drowns it out
 

### 2016-07-03 doc/paper.tex
small changes/comments

@@
 
+% TODO: adding (fairly arbitrary) out-of-domain corpus is really more like using a different model (or maybe bagging?)
+
 % TODO: mention that official results are on a slightly different Moses configuration? mehhhhh
@@
 Now we discuss Table~\ref{tab:multeval_tst2010-fixed}.
+Baseline's TER score changes by about 10\% relative to TAble~\ref{tab:multeval_tst2010}, but BLEU doesn't change as much.
+Apparently, BLEU is more robust to relatively minor changes?
 This time there is a statistically significant improvement in TER for train2015, but not for BLEU.
@@
 
+% NOTE
+% +0.3 BLEU -0.6 TER from post-editing the entire training corpus, esp the new 2015 additions (~25000 lines)
+% +0.1 BLEU -0.2 TER from ``snoop-editing'' Moses output lexical items only. cheating takes you FAR...
 \begin{table}[htb]
@@
 \hline
-\multirow{5}{*}{BLEU $\uparrow$}
+\multirow{4}{*}{BLEU $\uparrow$}
 & Our baseline & 11.4 & 0.3 & 0.2 & - \\
@@
 \hline
-\multirow{5}{*}{TER $\downarrow$}
+\multirow{4}{*}{TER $\downarrow$}
 & Our baseline & 70.6 & 0.4 & 0.5 & - \\
@@
 
+% TODO: I think another thing that SHOULD be done, but isn't commonly done, is that for a given DOMAIN, characterize inter-annotator differences in terms of , what kind of BLEU/TER scores do typical human translations get?
+
 %\amtaonly{

### 2016-07-11 doc/paper.tex
finished draft sentences for snooping section

--- 2016-06-27 doc/paper.tex
@@
     \usepackage{multirow}
+
 \fi
@@
 In particular, YAML has native Unicode support and does not even require quotation marks to enclose strings.
-}
+
+% TODO: YAML code examples. couldn't get verbatim + zh to work...
+%To maximize human translator productivity, 
+%meta-templates, which can then be easily parsed into templates to be fed into the generator.
+%An example is shown in Figure~\ref{fig:yaml_template}.
+} % amtaonly
 
@@
 Given the resulting scores, a change of direction seems in order; work on harvesting more ``example-based'' templates from the IWSLT15 training corpus is currently in progress.
-\amtaonly{As a first step in this direction, we examine the gold corpora qualitatively.}
+\amtaonly{
+As a first step in this direction, we examine the gold corpora qualitatively.}
 
@@
 This was presumably caused by having the same sentence displayed at multiple time points in the video subtitles. 
-Often, would also indicate passages that were poorly aligned at the sentence level.
+Often, this would also indicate passages that were poorly aligned at the sentence level.
 Such translations might suffice for viewing translated videos monolingually, but they are far from ideal for training a machine translation system.
@@
 Missing segments (usually English) when compared with the original lecture transcripts online.\footnote{http://www.ted.com/} 
-This issue seems to have been introduced in the IWSLT15 Chinese-English corpus. 
-When applied to the IWSLT14 corpus, most of these would be ``false positives'' indicating post-editing of musical symbols and other artifacts.
-True positives were often indicative of misaligned sentences.
+This issue seems to have been introduced in the IWSLT15 Chinese-English corpus, and was also indicative of misaligned sentences.
+%When applied to the IWSLT14 corpus, most of these would be ``false positives'' indicating post-editing of musical symbols and other artifacts.
+%True positives were often indicative of misaligned sentences.
 \end{itemize}
@@
 % but this creates a natural break for the discussion 
-Table~\ref{tab:tst2010} shows the results of ``lightly'' post-editing {\small \tt tst2010}. 
+Table~\ref{tab:tst2010} shows the scores obtained by ``lightly'' post-editing the MERT tuning set {\small \tt tst2010} as follows. 
 On the Chinese side, duplicated segments, applause lines, and translator notes were removed. 
 Speaker annotations were normalized on both sides, as well as sentence alignments wherever possible.
-Small but noticeable differences in scores are observed. 
-(By comparison, the difference between the best and worst scoring Chinese-to-English systems in IWSLT14 \mycitep{iwslt2014} was 6 BLEU points and 7 TER points.)
+As a result, small but noticeable differences in scores are observed. 
+(For comparison, the difference between the best and worst scoring Chinese-to-English systems in IWSLT14 \mycitep{iwslt2014} was 6 BLEU points and 7 TER points.)
 In this case, the post-edited tuning set seems to guide the MERT optimization procedure of Moses to a slightly different local optimum.
@@
 To preserve the integrity of {\small \tt tst2014}, evaluation is now performed using MultEval \mycitep{multeval} on the {\small \tt tst2010} data set itself, with the MERT procedure tuned on the {\small \tt dev2010} data set instead.
-These are the data sets and evaluation software used by the official benchmark cited in footnote \ref{foot:tst2010}.
+Our baseline model is also trained on the IWSLT corpus Chinese-English training corpus from 2014 instead of 2015.
+These data sets and this evaluation software were used by the official benchmark cited in footnote \ref{foot:tst2010}.
 MultEval has the added benefit of providing bootstrapped error bar estimates.
 
-% TODO TODO TODO: rerun these jobs with GIZA++ instead of mgiza! TINY difference in error bars, but should try to be as careful as possible
-    % I suppose I MIGHT consider this if the paper is ACCEPTED, haha
+In tables with MultEval scores below, metric scores for all systems were generated using MultEval 0.5.1 \mycitep{multeval}: jBLEU V0.1.1 (an exact reimplementation of NIST's mteval-v13.pl without tokenization); Translation Error Rate (TER) V0.8.0.
+The $p$-values are relative to the baseline and indicate whether a difference of this magnitude (between the baseline and the system on that line) is likely to be generated again by some random process (a randomized optimizer).
+Metric scores are averages over three runs per system. 
+$s_{sel}$ indicates the variance due to test set selection and has nothing to do with optimizer instability.\footnote{
+    This paragraph and the tables with MultEval results are adopted from \LaTeX~code automatically generated by MultEval.
+}
+
+% \pagebreak % soooooo ugly
+
 \begin{table}[htb]
@@
 \hline
-\multirow{2}{*}{BLEU $\uparrow$}
-& baseline & 11.5 & 0.3 & - & - \\
-& system 1 & 10.9 & 0.3 & - & 0.00 \\
+\multirow{5}{*}{BLEU $\uparrow$}
+& Official IWSLT14 baseline  & 11.2 & -   & -   & - \\
+& Our baseline & 11.4 & 0.3 & 0.2 & - \\
+& Tuning: {\small \tt dev2010} (post-edited) & 10.8 & 0.3 & 0.3 & 0.00 \\
+& Training: 2015 corpus & 11.5 & 0.3 & 0.3 & 0.30 \\
+& Training: 2015 corpus (post-edited) & \textbf{11.7} & 0.3 & 0.1 & 0.01 \\
+%& Post-edited: {\small \tt train2015} and {\small \tt dev2010} & 11.0 & 0.3 & 0.2 & 0.00 \\
 \hline
-\multirow{2}{*}{TER $\downarrow$}
-& baseline & 76.0 & 0.8 & - & - \\
-& system 1 & 82.8 & 1.0 & - & 0.00 \\
+\multirow{5}{*}{TER $\downarrow$}
+& Official IWSLT14 baseline & 77.0 & -   & -   & - \\
+& Our baseline & 76.2 & 0.8 & 0.3 & - \\
+& Tuning: {\small \tt dev2010} (post-edited) & 83.4 & 1.0 & 0.5 & 0.00 \\
+& Training: 2015 corpus & 76.4 & 0.8 & 0.5 & 0.26 \\
+& Training: 2015 corpus (post-edited) & \textbf{75.6} & 0.8 & 0.0 & 0.00 \\
+%& system 4 & 82.7 & 1.0 & 0.8 & 0.00 \\
 \hline
@@
 \end{center}
-\caption{\label{tab:scores} Metric scores for all systems generated using MultEval 0.5.1 \mycitep{multeval}: jBLEU V0.1.1 (an exact reimplementation of NIST's mteval-v13.pl without tokenization); Translation Error Rate (TER) V0.8.0.
-p-values are relative to baseline and indicate whether a difference of this magnitude (between the baseline and the system on that line) is likely to be generated again by some random process (a randomized optimizer).
-Metric scores are averages over multiple runs.
-$s_{sel}$ indicates the variance due to test set selection and has nothing to do with optimizer instability.}
+
+\caption{\label{tab:multeval_tst2010} % \caption[LoF entry] would allow multi-paragraph caption... \hspace{0.6 cm} to indent
+%\-\caption[LoF entry]{ \label{tab:multeval_tst2010}
+MultEval results for systems evaluated on the unmodified {\small \tt tst2010} data set, trained on the 2014 training corpus, and tuned on {\small \tt dev2010} unless indicated otherwise.
+Official baseline results (with IRSTLM instead of KenLM as discussed previously) are also given for comparison.
+%UHHHH do I have anything else to say here in the immediate caption? 
+%I'll interpret results in the text body.
+%But this caption needs to be kind of short, because otherwise it winds up on a different page than its discussion... although at this rate, it doesn't really matter - that'll happen anyway
+} % caption
 \end{table}
 
-Input-domain noise can play a role similar to regularization in machine learning - get a reference from denoising autoencoders?
-We do indeed find that making the tuning set translations more literal also tends to decrease performance, but this could be because of premature stopping.
-Also, the amount of noise added is generally some small amount of Gaussian noise, and it is unclear
+Table~\ref{tab:multeval_tst2010} gives results evaluated on the original {\small \tt tst2010} data set.
+The IWSLT 2015 training corpus {\small \tt train2015} did not produce a significant improvement over the IWSLT 2014 baseline.
+Although {\small \tt train2015} includes an additional \textasciitilde 30,000 lines of in-domain data, increasing the corpus size by \textasciitilde 17\%, it is also plagued by the new problem of missing fragments and resulting misaligned sentences mentioned in Section~\ref{subsec:post_edited}.
+Accordingly, post-editing \textasciitilde 7500 lines (3.6\%) of {\small \tt train2015} produced a small but significant improvement in scores,\footnote{
+This might look like a proportional improvement in BLEU, but gains were found to come in discrete steps during the editing process.
+This is thought to be due to the non-parametric nature of SMT, with changes in scores only occurring when there are appreciable changes to the parts of the probability tables that are relevant to the test set.
+} demonstrating the importance of corpus quality, not just quantity.
+Finally, unlike Table~\ref{tab:tst2010}, post-editing the MERT tuning development set {\small \tt dev2010} surprisingly makes things strictly worse.
+The effect of corrective post-editing tuning corpora appears to be relatively unpredictable, guiding the MERT optimizer to different local optima with varying out-of-sample performance.
+Whereas input-domain noise can play a role similar to regularization in methods such as denoising autoencoders \mycitep{bengio:2008}, the effect of ``large-amplitude'' noise from mistranslated or misaligned parallel corpora is unclear, and it may warrant further investigation in other models.
 
-For tst2014, revising the DEV set tst2010 instead leads to different local minima instead. 
-A better explanation is that the effect of this amount and kind of noise is ill-defined.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\noindent\makebox[\linewidth]{\rule{\paperwidth}{0.4pt}}
+TODO: I am here. the rest of the stuff from here to ``Proposed Future Work'' is not finished yet. Just draft entries...
 
+% TODO: Adding train2015 + 
+% this is pointless to say, unless I also have unmodified train2015 + dev2010?
+
+% NOTE
+% +0.3 BLEU -0.6 TER from post-editing the entire training corpus, esp the new 2015 additions (~25000 lines)
+% +0.1 BLEU -0.2 TER from ``snoop-editing'' Moses output lexical items only. cheating takes you FAR...
+\begin{table}[htb]
+\begin{center}
+%\begin{footnotesize}
+\begin{tabular}{|l|l|l|l|l|l|}
+\hline
+\bf Metric & \bf System (all with post-edited {\small \tt tst2010}) & \bf Avg & \bf $\overline{s}_{\text{sel}}$ & \bf $s_{\text{Test}}$ & \bf $p$-value \\
+\hline
+\multirow{4}{*}{BLEU $\uparrow$}
+& Our baseline & 11.4 & 0.3 & 0.2 & - \\
+& Tuning: {\small \tt dev2010} (post-edited) & \textbf{11.9} & 0.3 & 0.1 & 0.00 \\
+& Training: 2015 corpus & 11.4 & 0.3 & 0.2 & 0.44 \\
+& Training: 2015 corpus (post-edited) & 11.6 & 0.3 & 0.2 & 0.01 \\
+%& Post-edited: {\small \tt train2015} and {\small \tt dev2010} & \textbf{12.2} & 0.3 & 0.1 & 0.00 \\
+\hline
+\multirow{4}{*}{TER $\downarrow$}
+& Our baseline & 70.6 & 0.4 & 0.5 & - \\
+& Tuning: {\small \tt dev2010} (post-edited) & 75.1 & 0.5 & 0.3 & 0.00 \\
+& Training: 2015 corpus & 70.1 & 0.4 & 0.3 & 0.00 \\
+& Training: 2015 corpus (post-edited) & \textbf{69.7} & 0.4 & 0.2 & 0.00 \\
+& + generated & & & &  \\
+%& Post-edited: {\small \tt train2015} and {\small \tt dev2010} & 74.7 & 0.5 & 0.2 & 0.00 \\
+\hline
+\end{tabular}
+%\end{footnotesize}
+\end{center}
+
+\caption{\label{tab:multeval_tst2010_fixed} % \caption[LoF entry] would allow multi-paragraph caption... \hspace{0.6 cm} to indent
+%\-\caption[LoF entry]{ \label{tab:multeval_tst2010}
+MultEval results obtained as described in Table~\ref{tab:multeval_tst2010}, but evaluated on the post-edited {\small \tt tst2010} data set.
+} % caption
+\end{table}
+
+%\pagebreak
+Now we discuss Table~\ref{tab:multeval_tst2010_fixed}, for which we now as a test set the previously post-edited tst2010.
+Unlike the tuning set, post-editing the test set is expected to have a more predictable effect.
+Namely, correcting lines in the test set as described in Section~\ref{subsec:case_study} is expected to improve scores compared with Table~\ref{tab:multeval_tst2010}, as it should generally be easier for the MT system to reproduce lines that have not been grossly mistranslated.
+This does indeed seem to be the case, with a more pronounced effect for TER scores.% with most runs except for the uncorrected 2015 training corpus
+
+% no, this was if I HAND corrected dev set translations to be more literal
+%\footnote{This is different from 
+%which in our experiments seemed to lead to worse results, due to premature termination of optimization (not shown).
+%}
+
+%Baseline's TER score changes by about 10\% relative to Table~\ref{tab:multeval_tst2010}, but BLEU doesn't change as much.
+This time there is a statistically significant improvement for train2015, but only in the TER score and not for BLEU.
+Apparently, BLEU scores are more robust than TER with respect to changes in training, tuning, and test corpora.
+Post-editing train2015 again leads to further improvement in both metrics.
+
+Unlike Table~\ref{tab:multeval_tst2010}, post-edited dev2010 no longer makes everything uniformly worse. 
+Instead, and as in Table~\ref{tab:tst2010}, post-editing of the tuning set obtains slightly better BLEU at the expense of much worse TER. 
+This again emphasizes the fact that tuning set noise can affect bottom-line performance in relatively unpredictable ways.
+
+Table~\ref{tab:multeval_tst2010_fixed} also has some initial results from the ``example-based'' mode of our generator.
+[Should I add an example of the meta-templates, to show how easy they are to write?]
+These examples focus on quality instead of quantity and so there aren't many of them.
+Started by looking for non-compositional sentences in poorly-scoring training pairs.
+Alternative translations to some liberal (but not incorrect) translations were also included.
+This was only possible after post-editing to clear out the noise.
+They were simply concatenated to the end of the training corpus.
+The results almost certainly suck.
+To better direct future efforts, we perform some data snooping.
+We are looking for at least a partial answer to the question, what makes a corpus in-domain?
+
+Maybe we just to generate more data at scale. 
+But should we scale up lexical data or syntactic data? 
+To address this question, we resort to data snooping.
+
+\subsection{Data Snooping}
 % Data Snooping - dedicate this to its own section, to make clear that these are not ``REAL'' results
-% we perform snooping on a dedicated snooping set (as if that makes it right...)
+\textbf{Warning: This section is dedicated to ``data snooping runs'' that train on the test set in one way or another.}
+The scores given here must not considered indicative of actual out-of-sample system performance .
+The reason for the snooping is to try to isolate the most effective characteristics of in-domain data, whose power was seen in Section~\ref{sec:experiments}.
+It is hoped that these characteristics can guide future development of our generator from Section~\ref{sec:generator}.
+
+The snooping is performed on {\small \tt tst2010}, to keep {\small \tt tst2014} relatively pristine for future work.
+Since they were so much smaller (\textasciitilde 1500 lines vs.
+\textasciitilde 180000 lines), snooping data sets were simply concatenated to the end of the IWSLT 2014 training corpus instead of using {\small \tt tmcombine}, with the rationale being that unlike our large, ``rule-based'' generated corpora from Section~\ref{sec:experiments}, these snooping additions should only perturb the translation tables slightly.
+
+\begin{table}[htb]
+\begin{center}
+%\begin{footnotesize}
+\begin{tabular}{|l|l|l|l|l|l|}
+\hline
+\bf Metric & \bf Snooping Type & \bf Avg & \bf $\overline{s}_{\text{sel}}$ & \bf $s_{\text{Test}}$ & \bf $p$-value \\
+\hline
+\multirow{4}{*}{BLEU $\uparrow$}
+& Our baseline & 11.4 & 0.3 & 0.2 & - \\
+& Full snooping & \textbf{36.8} & 0.6 & 0.0 & 0.00 \\
+& Syntactic snooping & 17.6 & 0.4 & 1.0 & 0.00 \\
+& Lexical snooping & 11.2 & 0.3 & 0.1 & 0.02 \\
+\hline
+\multirow{4}{*}{TER $\downarrow$}
+& Our baseline & 76.2 & 0.8 & 0.3 & - \\
+& Full snooping & \textbf{48.3} & 0.7 & 0.0 & 0.00 \\
+& Syntactic snooping & 69.8 & 0.7 & 1.3 & 0.00 \\
+& Lexical snooping & 76.3 & 0.7 & 0.1 & 0.48 \\
+\hline
+\end{tabular}
+%\end{footnotesize}
+\end{center}
+
+\caption{ \label{tab:snooping} % \caption[LoF entry] would allow multi-paragraph caption... \hspace{0.6 cm} to indent
+%\-\caption[LoF entry]{ \label{tab:multeval_tst2010}
+Data snooping results for systems trained on information from the {\small \tt tst2010} data set concatenated to the end of the IWSLT 2014 training corpus and also evaluated with MultEval on {\small \tt tst2010}.
+The different types of snooping are defined in the text.
+} % caption
+\end{table}
+
+For full snooping, the corpus was concatenated directly with {\small \tt tst2010}. 
+This establishes an ``upper bound'' of sorts of how much performance boost you could get by simply adding more data.
+Can think of this as the limit of infinite data (with in-domain distribution)?
+It's also a measure of the inherent limitation (bias?) of the model being used.
+Even snooped data is only on-par with out-of-sample Spanish-English performance, illustrating the difficulty of the task.
+
+For syntactic snooping, before concatenation, rare tokens in {\small \tt tst2010} were replaced by their part of speech (POS) from the Stanford POS tagger \mycitep{postagger:2003}. 
+ Non-rare tokens were arbitrarily defined for each language as the most common 80\% in the IWSLT 2014 corpus: 670 tokens for English, 1406 tokens for Chinese, tokenized using the latest version of the Stanford Segmenter \mycitep{segmenter:2005,segmenter:2008}.
+You get a significant increase in performance, even though this essentially performs zero snooping on out-of-vocabulary (OOV) words that the baseline system failed to translate.
+
+For lexical snooping, we selected \textasciitilde 100 mistranslated or OOV words from the baseline translation and used the rule-based mode of our generator to generate \textasciitilde 30000 sentences with OOV words for concatenation.
+This snooping-generated data doesn't make a dent, and in fact makes things slightly worse (although TER is within error bars).
+So nonsensical sentences of OOV words does NOT help - or even if it does, the nonsense-noise drowns it out.
+
+Taken together, these snooping results reinforce the sentiment from the end of Section~\ref{sec:experiments} that future scaled-up work on our generator should focus on the ``example-based'' approach.
+Thus  you should focus on the example-based half.
+
+%%% TODO: see OneNote
+
+%
 
@@
 
+% what the heck, let's do this - as it makes it transparent that I'm not over reference length either
+% but ugh, huge blank space looks silly
+%\pagebreak
+
 % TODO: I suspect MT requirements are very similar to that of, say, AP Latin - where the best translations are the most LITERAL, not necessarily the most fluent. 
@@
 % TODO: SMT as dynamic-length or variable-length EBMT. just a qualitative observation of mine...but I suppoose, thinking about SMT mechanistically, that's what it is, really
+
+% TODO: I think another thing that SHOULD be done, but isn't commonly done, is that for a given DOMAIN, characterize inter-annotator differences in terms of , what kind of BLEU/TER scores do typical human translations get?
 

### 2016-07-11 template_maker.py
forgot to handle NR for zh

@@ -51,6 +51,8 @@ def get_type_from_pos(pos, lang):
     elif lang == 'zh':
         if pos in ['JJ', 'VA']:
             return 'adjective'
+        elif pos == 'NR':
+            return 'name'
         elif pos == 'NN':
             return 'noun'
         elif pos == 'VV':

### 2016-07-11 doc/histogram.m
quick script to make histograms (currently unused anyway...) - remember octave?

@@ -1,0 +1,7 @@
+data = load('fixed.csv');
+hist(data);
+
+
+
+print -depslatex 'fixed.eps'
+print -dpng 'fixed.png'

### 2016-07-11 doc/paper.tex
AMTA was extended AGAIN

@@
 & Training: 2015 corpus & 11.5 & 0.3 & 0.3 & 0.30 \\
-& Training: 2015 corpus (post-edited) & \textbf{11.7} & 0.3 & 0.1 & 0.01 \\
+& Training: 2015 corpus (post-edited) & \textbf{11.9} & 0.3 & 0.0 & 0.00 \\
 %& Post-edited: {\small \tt train2015} and {\small \tt dev2010} & 11.0 & 0.3 & 0.2 & 0.00 \\
@@
 & Training: 2015 corpus & 76.4 & 0.8 & 0.5 & 0.26 \\
-& Training: 2015 corpus (post-edited) & \textbf{75.6} & 0.8 & 0.0 & 0.00 \\
+& Training: 2015 corpus (post-edited) & \textbf{75.2} & 0.8 & 0.0 & 0.00 \\
 %& system 4 & 82.7 & 1.0 & 0.8 & 0.00 \\
@@
 Although {\small \tt train2015} includes an additional \textasciitilde 30,000 lines of in-domain data, increasing the corpus size by \textasciitilde 17\%, it is also plagued by the new problem of missing fragments and resulting misaligned sentences mentioned in Section~\ref{subsec:post_edited}.
-Accordingly, post-editing \textasciitilde 7500 lines (3.6\%) of {\small \tt train2015} produced a small but significant improvement in scores,\footnote{
+Accordingly, post-editing \textasciitilde 7500 lines (3.6\%) of {\small \tt train2015} produced a small but significant improvement in scores, \footnote{
 This might look like a proportional improvement in BLEU, but gains were found to come in discrete steps during the editing process. 
 This is thought to be due to the non-parametric nature of SMT, with changes in scores only occurring when there are appreciable changes to the parts of the probability tables that are relevant to the test set.
+It remains to be seen what effect post-editing of training, tuning, and test corpora would have on parametric systems like recent deep-learning based MT methods \mycitep{sutskever:2014,bengio:2014}, in which all training examples contribute to the same shared connection weight matrices.
 } demonstrating the importance of corpus quality, not just quantity.
@@
 Whereas input-domain noise can play a role similar to regularization in methods such as denoising autoencoders \mycitep{bengio:2008}, the effect of ``large-amplitude'' noise from mistranslated or misaligned parallel corpora is unclear, and it may warrant further investigation in other models.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\noindent\makebox[\linewidth]{\rule{\paperwidth}{0.4pt}}
-TODO: I am here. the rest of the stuff from here to ``Proposed Future Work'' is not finished yet. Just draft entries...
 
@@
 \hline
-\multirow{4}{*}{BLEU $\uparrow$}
+\multirow{5}{*}{BLEU $\uparrow$}
 & Our baseline & 11.4 & 0.3 & 0.2 & - \\
@@
 & Training: 2015 corpus & 11.4 & 0.3 & 0.2 & 0.44 \\
-& Training: 2015 corpus (post-edited) & 11.6 & 0.3 & 0.2 & 0.01 \\
+& Training: 2015 corpus (post-edited) & \textbf{11.9} & 0.3 & 0.2 & 0.00 \\
 %& Post-edited: {\small \tt train2015} and {\small \tt dev2010} & \textbf{12.2} & 0.3 & 0.1 & 0.00 \\
+& + generated & \textbf{11.9} & 0.3 & 0.1 & 0.00 \\ 
 \hline
-\multirow{4}{*}{TER $\downarrow$}
+\multirow{5}{*}{TER $\downarrow$}
 & Our baseline & 70.6 & 0.4 & 0.5 & - \\
@@
 & Training: 2015 corpus & 70.1 & 0.4 & 0.3 & 0.00 \\
-& Training: 2015 corpus (post-edited) & \textbf{69.7} & 0.4 & 0.2 & 0.00 \\
-& + generated & & & &  \\
+& Training: 2015 corpus (post-edited) & 69.4 & 0.4 & 0.2 & 0.00 \\
 %& Post-edited: {\small \tt train2015} and {\small \tt dev2010} & 74.7 & 0.5 & 0.2 & 0.00 \\
+& + generated & \textbf{69.2} & 0.4 & 0.1 & 0.00 \\ 
 \hline
@@
 MultEval results obtained as described in Table~\ref{tab:multeval_tst2010}, but evaluated on the post-edited {\small \tt tst2010} data set.
+For the ``+ generated'' system, \textasciitilde 5000 lines from the ``example-based'' mode of our generator were concatenated to the post-edited 2015 training corpus.
 } % caption
@@
 %\pagebreak
-Now we discuss Table~\ref{tab:multeval_tst2010_fixed}, for which we now as a test set the previously post-edited tst2010.
-Unlike the tuning set, post-editing the test set is expected to have a more predictable effect.
-Namely, correcting lines in the test set as described in Section~\ref{subsec:case_study} is expected to improve scores compared with Table~\ref{tab:multeval_tst2010}, as it should generally be easier for the MT system to reproduce lines that have not been grossly mistranslated.
+For Table~\ref{tab:multeval_tst2010_fixed}, the test set is changed to the previously post-edited tst2010.
+Unlike its use as a tuning set in Section~\ref{subsec:case_study}, post-editing the test set is expected to have a more predictable effect.
+Namely, translation corrections are expected to improve scores compared with Table~\ref{tab:multeval_tst2010}, as it should generally be easier for the MT system to reproduce lines that have not been grossly mistranslated.
 This does indeed seem to be the case, with a more pronounced effect for TER scores.% with most runs except for the uncorrected 2015 training corpus
@@
 %Baseline's TER score changes by about 10\% relative to Table~\ref{tab:multeval_tst2010}, but BLEU doesn't change as much.
-This time there is a statistically significant improvement for train2015, but only in the TER score and not for BLEU.
-Apparently, BLEU scores are more robust than TER with respect to changes in training, tuning, and test corpora.
-Post-editing train2015 again leads to further improvement in both metrics.
+With the post-edited test set, there is now a small but statistically significant improvement in the TER score when using the unedited 2015 training corpus.
+BLEU scores, on the other hand, appear to be more robust than TER with respect to changes in the test corpus.
+Post-editing the training corpus again leads to further improvement in both metrics.
 
-Unlike Table~\ref{tab:multeval_tst2010}, post-edited dev2010 no longer makes everything uniformly worse. 
-Instead, and as in Table~\ref{tab:tst2010}, post-editing of the tuning set obtains slightly better BLEU at the expense of much worse TER. 
+With the post-edited test set, post-editing of the tuning set {\small \tt dev2010} no longer makes everything uniformly worse, unlike Table~\ref{tab:multeval_tst2010}.
+Instead, and as was the case when evaluating on {\small \tt tst2014} in Table~\ref{tab:tst2010}, post-editing of the tuning set in this case results in slightly improved BLEU at the expense of much worsened TER.
 This again emphasizes the fact that tuning set noise can affect bottom-line performance in relatively unpredictable ways.
 
-Table~\ref{tab:multeval_tst2010_fixed} also has some initial results from the ``example-based'' mode of our generator.
-[Should I add an example of the meta-templates, to show how easy they are to write?]
-These examples focus on quality instead of quantity and so there aren't many of them.
-Started by looking for non-compositional sentences in poorly-scoring training pairs.
+The ``+ generated'' entry in Table~\ref{tab:multeval_tst2010_fixed} shows initial results using the quality-oriented ``example-based'' mode of our generator.
+The starting point for template construction was non-compositional translations from the 2015 training corpus with high TER scores, which was only practical after post-editing to remove the clerical-like errors described in Section~\ref{subsec:case_study}.
+The basic rationale is that poorly scoring training sentences are ones that the model has not fully ``learned'' yet.
 Alternative translations to some liberal (but not incorrect) translations were also included.
-This was only possible after post-editing to clear out the noise.
-They were simply concatenated to the end of the training corpus.
-The results almost certainly suck.
+As there were relatively few of them compared with Table~\ref{tab:results}, the generated sentences here were simply concatenated to the end of the training corpus.
+
 To better direct future efforts, we perform some data snooping.
@@
 But should we scale up lexical data or syntactic data? 
-To address this question, we resort to data snooping.
+To address this question, we consider the taboo practice of data snooping.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\noindent\makebox[\linewidth]{\rule{\paperwidth}{0.4pt}}
+TODO: I am here. the rest of the stuff from here to ``Proposed Future Work'' is not finished yet. Just draft entries...
 
@@
 So nonsensical sentences of OOV words does NOT help - or even if it does, the nonsense-noise drowns it out.
+More generally, it seems that although nonsensical sentences like Chomsky's ``colorless green ideas sleep furiously'' are fully licensed by English grammar, they can be detrimental practical NLP applications like MT that operate on naturally occurring sentences at test time.
+Even when using an in-domain lexicon, their syntax can be sufficiently out-of-domain to result in a net loss of performance.
 
@@
 Thus you should focus on the example-based half.
+Moreover, the relative unimportance of OOV words suggests that the lexical databse need not be THAT exhaustive.
 

### 2016-07-14 doc/paper.tex
pretty much done with MY edits...

--- 2016-07-11 doc/paper.tex
@@
 
+    \usepackage{graphicx}
+
 \fi
@@
 \begin{abstract}
-The large-scale generation of synthetic parallel corpora for data-driven machine translation is proposed.
+The large-scale generation of synthetic parallel corpora for data-driven machine translation is proposed\amtaonly{ in this demo paper}.
 A rudimentary system for doing this is presented, with mechanisms for syntactic complexity in the spirit of classical rule-based machine translation and high-quality sentence pairs in the spirit of example-based machine translation.
@@
 As early as Jelinek's (in)famous quote during the 1980's that ``{\em Every time I fire a linguist, my performance goes up},'' both linguists and linguistics have been increasingly marginalized in natural language processing (NLP).
-More recently, for example, \mynewcite{collobert:2011} mention the possibility ``{\em [ideally]...to learn from letter sequences rather than words},'' illustrating an AI-motivated desire to move away from even the most fundamental of linguistic concepts.
+More recently, for example, \mynewcite{collobert:2011} mention the possibility ``{\em [ideally]...to learn from letter sequences rather than words},'' illustrating an AI-motivated desire to move away from even fundamental linguistic concepts.
 A specific instance of this overall trend is the gradual shift of machine translation (MT) away from classical linguistic rule-based approaches to contemporary machine learning approaches \mycitep{hutchins:2005}, which underlie statistical phrase-based systems like Moses \mycitep{moses} and more recent neural network systems \mycitep{sutskever:2014,bengio:2014}.
-But in the end, even the most cutting-edge machine learning MT system relies crucially on multilingual parallel corpora, typically taken from external sources like multilingual parliamentary proceedings.
+But in the end, even the most cutting-edge machine learning MT system relies crucially on bilingual parallel corpora, typically taken from external sources like multilingual parliamentary proceedings.
 This results in what we perceive to be a disproportionate amount of effort being dedicated to the machine learning component of machine translation, with input corpora being treated almost as an afterthought.\footnote{
@@
 %An example is shown in Figure~\ref{fig:yaml_template}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\noindent\makebox[\linewidth]{\rule{\paperwidth}{0.4pt}} % full-width horizontal line
+TODO: detailed system description
+
+\noindent\makebox[\linewidth]{\rule{\paperwidth}{0.4pt}} % full-width horizontal line
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 } % amtaonly
+
+% if i'm gonna do source code snippets with zh, maybe it'd be best to use IMAGES? it's not for camera-ready anyway...
+% well, i can do the meta-template easily... the problem is with REAL templates
+
+%here is the figure
+%\begin{figure}[ht]
+%    \centering
+%    \includegraphics{Chick1}
+%    \caption{\label{fig:test} chick}
+%\end{figure}
+%Hola chica, Figure~\ref{fig:test}.
 
@@
 \section{Experimental Results}
+\label{supersec:experimental_results}
 
@@
 \caption{ \label{tab:results} % oh, for reference numbering, label goes INSIDE caption
-    (Left) Chinese-to-English BLEU scores for our baseline Moses configuration with phrase and reordering tables trained on a random 25\% subset of the IWSLT15 training corpus and interpolated (+) with tables for the remaining 75\% (\textasciitilde 150K) of the in-domain corpus, the MultiUN corpus (a \textasciitilde 1M line subset), and a synthetic corpus from our generator (\textasciitilde 5M lines).
+    (Left) Chinese-to-English BLEU scores for our baseline Moses configuration with phrase and reordering tables trained on a random 25\% subset of the IWSLT15 training corpus and interpolated (+) with tables for the remaining 75\% ($\sim$150K) of the in-domain corpus, the MultiUN corpus (a $\sim$1M line subset), and a synthetic corpus from our generator ($\sim$5M lines).
     (Right) Same as the left, but starting from tables trained on the entire IWSLT15 corpus.
@@
 
-For the rest of this section, we show the results of applying the same post-editing procedure to the test and training sets as well, with the additional step of restoring missing segments for both languages.
+For the rest of this section, we show the results of applying post-editing to corpora used for multiple purposes.
 To preserve the integrity of {\small \tt tst2014}, evaluation is now performed using MultEval \mycitep{multeval} on the {\small \tt tst2010} data set itself, with the MERT procedure tuned on the {\small \tt dev2010} data set instead.
 Our baseline model is also trained on the IWSLT corpus Chinese-English training corpus from 2014 instead of 2015.
-These data sets and this evaluation software were used by the official benchmark cited in footnote \ref{foot:tst2010}.
+These data sets and this evaluation software were used by the official WIT$^3$ benchmark linked in footnote \ref{foot:tst2010}.
 MultEval has the added benefit of providing bootstrapped error bar estimates.
@@
 \hline
-\multirow{5}{*}{BLEU $\uparrow$}
+\multirow{6}{*}{BLEU $\uparrow$}
 & Official IWSLT14 baseline  & 11.2 & -   & -   & - \\
@@
 & Training: 2015 corpus & 11.5 & 0.3 & 0.3 & 0.30 \\
-& Training: 2015 corpus (post-edited) & \textbf{11.7} & 0.3 & 0.1 & 0.01 \\
+& Training: 2015 corpus (post-edited) & \textbf{12.1} & 0.3 & 0.1 & 0.00 \\
+& + generated & 11.9 & 0.3 & 0.0 & 0.00 \\ 
 %& Post-edited: {\small \tt train2015} and {\small \tt dev2010} & 11.0 & 0.3 & 0.2 & 0.00 \\
 \hline
-\multirow{5}{*}{TER $\downarrow$}
+\multirow{6}{*}{TER $\downarrow$}
 & Official IWSLT14 baseline & 77.0 & -   & -   & - \\
@@
 & Training: 2015 corpus & 76.4 & 0.8 & 0.5 & 0.26 \\
-& Training: 2015 corpus (post-edited) & \textbf{75.6} & 0.8 & 0.0 & 0.00 \\
+& Training: 2015 corpus (post-edited) & \textbf{74.8} & 0.7 & 0.1 & 0.00 \\
+& + generated & 74.9 & 0.7 & 0.0 & 0.00 \\ 
 %& system 4 & 82.7 & 1.0 & 0.8 & 0.00 \\
@@
 %\-\caption[LoF entry]{ \label{tab:multeval_tst2010}
-MultEval results for systems evaluated on the unmodified {\small \tt tst2010} data set, trained on the 2014 training corpus, and tuned on {\small \tt dev2010} unless indicated otherwise.
+MultEval results for systems evaluated on the {\small \tt tst2010} data set, tuned on the {\small \tt dev2010} data set,  and trained on the 2014 training corpus unless indicated otherwise.
 Official baseline results (with IRSTLM instead of KenLM as discussed previously) are also given for comparison.
+Post-editing of the 2015 training corpus involved correcting $\sim$7500 lines (3.6\% of the corpus) as described in Section~\ref{subsec:case_study}, as well as adding $\sim$750 sentence pairs obtained by splitting lines that were longer than 100 tokens and hence removed from the training procedure by the {\small \tt clean-corpus-n.perl} script in the Moses toolchain.
 %UHHHH do I have anything else to say here in the immediate caption? 
@@
 
-Table~\ref{tab:multeval_tst2010} gives results evaluated on the original {\small \tt tst2010} data set.
-The IWSLT 2015 training corpus {\small \tt train2015} did not produce a significant improvement over the IWSLT 2014 baseline.
-Although {\small \tt train2015} includes an additional \textasciitilde 30,000 lines of in-domain data, increasing the corpus size by \textasciitilde 17\%, it is also plagued by the new problem of missing fragments and resulting misaligned sentences mentioned in Section~\ref{subsec:post_edited}.
-Accordingly, post-editing \textasciitilde 7500 lines (3.6\%) of {\small \tt train2015} produced a small but significant improvement in scores,\footnote{
-This might look like a proportional improvement in BLEU, but gains were found to come in discrete steps during the editing process.
+Table~\ref{tab:multeval_tst2010} gives results evaluated on the {\small \tt tst2010} data set.
+The IWSLT 2015 training corpus did not produce a significant improvement over the IWSLT 2014 baseline, even though the 2015 corpus includes an additional $\sim$30,000 lines of in-domain data, increasing the corpus size by $\sim$17\%.
+Post-editing the 2015 corpus to correct for the missing fragments and resulting misaligned sentences mentioned in Section~\ref{subsec:case_study} resulted in a modest but statistically significant improvement in scores,\footnote{
+By comparison, the best scoring systems from the IWSLT 2014 evaluation campaign \mycitep{iwslt2014} scored $\sim$4 BLEU points and $\sim$3 TER points better than the baseline.
+Also, this might look like a proportional improvement in BLEU, but gains were found to come in discrete steps during the editing process. 
 This is thought to be due to the non-parametric nature of SMT, with changes in scores only occurring when there are appreciable changes to the parts of the probability tables that are relevant to the test set.
+It remains to be seen what effect post-editing of training, tuning, and test corpora would have on parametric systems like recent deep-learning based MT methods \mycitep{sutskever:2014,bengio:2014}, in which all training examples contribute to the same shared connection weight matrices.
 } demonstrating the importance of corpus quality, not just quantity.
@@
 The effect of corrective post-editing tuning corpora appears to be relatively unpredictable, guiding the MERT optimizer to different local optima with varying out-of-sample performance.
-Whereas input-domain noise can play a role similar to regularization in methods such as denoising autoencoders \mycitep{bengio:2008}, the effect of ``large-amplitude'' noise from mistranslated or misaligned parallel corpora is unclear, and it may warrant further investigation in other models.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\noindent\makebox[\linewidth]{\rule{\paperwidth}{0.4pt}}
-TODO: I am here. the rest of the stuff from here to ``Proposed Future Work'' is not finished yet. Just draft entries...
+%Whereas input-domain noise can play a role similar to regularization in methods such as denoising autoencoders \mycitep{bengio:2008}, the effect of ``large-amplitude'' noise from mistranslated or misaligned parallel corpora is unclear, and it may warrant further investigation in other models.
+% relatively pointless filler, and commenting it out should buy me one reference!
 
@@
 
-% NOTE
-% +0.3 BLEU -0.6 TER from post-editing the entire training corpus, esp the new 2015 additions (~25000 lines)
-% +0.1 BLEU -0.2 TER from ``snoop-editing'' Moses output lexical items only. cheating takes you FAR...
-\begin{table}[htb]
-\begin{center}
-%\begin{footnotesize}
-\begin{tabular}{|l|l|l|l|l|l|}
-\hline
-\bf Metric & \bf System (all with post-edited {\small \tt tst2010}) & \bf Avg & \bf $\overline{s}_{\text{sel}}$ & \bf $s_{\text{Test}}$ & \bf $p$-value \\
-\hline
-\multirow{4}{*}{BLEU $\uparrow$}
-& Our baseline & 11.4 & 0.3 & 0.2 & - \\
-& Tuning: {\small \tt dev2010} (post-edited) & \textbf{11.9} & 0.3 & 0.1 & 0.00 \\
-& Training: 2015 corpus & 11.4 & 0.3 & 0.2 & 0.44 \\
-& Training: 2015 corpus (post-edited) & 11.6 & 0.3 & 0.2 & 0.01 \\
-%& Post-edited: {\small \tt train2015} and {\small \tt dev2010} & \textbf{12.2} & 0.3 & 0.1 & 0.00 \\
-\hline
-\multirow{4}{*}{TER $\downarrow$}
-& Our baseline & 70.6 & 0.4 & 0.5 & - \\
-& Tuning: {\small \tt dev2010} (post-edited) & 75.1 & 0.5 & 0.3 & 0.00 \\
-& Training: 2015 corpus & 70.1 & 0.4 & 0.3 & 0.00 \\
-& Training: 2015 corpus (post-edited) & \textbf{69.7} & 0.4 & 0.2 & 0.00 \\
-& + generated & & & &  \\
-%& Post-edited: {\small \tt train2015} and {\small \tt dev2010} & 74.7 & 0.5 & 0.2 & 0.00 \\
-\hline
-\end{tabular}
-%\end{footnotesize}
-\end{center}
-
-\caption{\label{tab:multeval_tst2010_fixed} % \caption[LoF entry] would allow multi-paragraph caption... \hspace{0.6 cm} to indent
-%\-\caption[LoF entry]{ \label{tab:multeval_tst2010}
-MultEval results obtained as described in Table~\ref{tab:multeval_tst2010}, but evaluated on the post-edited {\small \tt tst2010} data set.
-} % caption
-\end{table}
+%\begin{table}[htb]
+%\begin{center}
+%%\begin{footnotesize}
+%\begin{tabular}{|l|l|l|l|l|l|}
+%\hline
+%\bf Metric & \bf System (all with post-edited {\small \tt tst2010}) & \bf Avg & \bf $\overline{s}_{\text{sel}}$ & \bf $s_{\text{Test}}$ & \bf $p$-value \\
+%\hline
+%\multirow{5}{*}{BLEU $\uparrow$}
+%& Our baseline & 11.4 & 0.3 & 0.2 & - \\
+%& Tuning: {\small \tt dev2010} (post-edited) & \textbf{11.9} & 0.3 & 0.1 & 0.00 \\
+%& Training: 2015 corpus & 11.4 & 0.3 & 0.2 & 0.44 \\
+%& Training: 2015 corpus (post-edited) & \textbf{12.0} & 0.3 & 0.2 & 0.00 \\
+%%& Post-edited: {\small \tt train2015} and {\small \tt dev2010} & \textbf{12.2} & 0.3 & 0.1 & 0.00 \\
+%& + generated & 11.8 & 0.3 & 0.1 & 0.00 \\ 
+%\hline
+%\multirow{5}{*}{TER $\downarrow$}
+%& Our baseline & 70.6 & 0.4 & 0.5 & - \\
+%& Tuning: {\small \tt dev2010} (post-edited) & 75.1 & 0.5 & 0.3 & 0.00 \\
+%& Training: 2015 corpus & 70.1 & 0.4 & 0.3 & 0.00 \\
+%& Training: 2015 corpus (post-edited) & \textbf{69.3} & 0.4 & 0.2 & 0.00 \\
+%%& Post-edited: {\small \tt train2015} and {\small \tt dev2010} & 74.7 & 0.5 & 0.2 & 0.00 \\
+%& + generated & \textbf{69.3} & 0.4 & 0.1 & 0.00 \\ 
+%\hline
+%\end{tabular}
+%%\end{footnotesize}
+%\end{center}
+%
+%\caption{\label{tab:multeval_tst2010_fixed} % \caption[LoF entry] would allow multi-paragraph caption... \hspace{0.6 cm} to indent
+%%\-\caption[LoF entry]{ \label{tab:multeval_tst2010}
+%MultEval results obtained as described in Table~\ref{tab:multeval_tst2010}, but evaluated on the post-edited {\small \tt tst2010} data set.
+%For the ``+ generated'' system, $\sim$5000 lines from the ``example-based'' mode of our generator were concatenated to the post-edited 2015 training corpus.
+%} % caption
+%\end{table}
 
 %\pagebreak
-Now we discuss Table~\ref{tab:multeval_tst2010_fixed}, for which we now as a test set the previously post-edited tst2010.
-Unlike the tuning set, post-editing the test set is expected to have a more predictable effect.
-Namely, correcting lines in the test set as described in Section~\ref{subsec:case_study} is expected to improve scores compared with Table~\ref{tab:multeval_tst2010}, as it should generally be easier for the MT system to reproduce lines that have not been grossly mistranslated.
-This does indeed seem to be the case, with a more pronounced effect for TER scores.% with most runs except for the uncorrected 2015 training corpus
+%For Table~\ref{tab:multeval_tst2010_fixed}, the test set is changed to the previously post-edited tst2010.
+
+%This does indeed seem to be the case, with a more pronounced effect for TER scores.% with most runs except for the uncorrected 2015 training corpus
 
@@
 
-%Baseline's TER score changes by about 10\% relative to Table~\ref{tab:multeval_tst2010}, but BLEU doesn't change as much.
-This time there is a statistically significant improvement for train2015, but only in the TER score and not for BLEU.
-Apparently, BLEU scores are more robust than TER with respect to changes in training, tuning, and test corpora.
-Post-editing train2015 again leads to further improvement in both metrics.
+%%Baseline's TER score changes by about 10\% relative to Table~\ref{tab:multeval_tst2010}, but BLEU doesn't change as much.
+%With the post-edited test set, there is now a small but statistically significant improvement in the TER score when using the unedited 2015 training corpus.
+%BLEU scores, on the other hand, appear to be more robust than TER with respect to changes in the test corpus.
+%Post-editing the training corpus again leads to further improvement in both metrics.
+%
+%With the post-edited test set, post-editing of the tuning set {\small \tt dev2010} no longer makes everything uniformly worse, unlike Table~\ref{tab:multeval_tst2010}.
+%Instead, and as was the case when evaluating on {\small \tt tst2014} in Table~\ref{tab:tst2010}, post-editing of the tuning set in this case results in slightly improved BLEU at the expense of much worsened TER.
+%This again emphasizes the fact that tuning set noise can affect bottom-line performance in relatively unpredictable ways.
 
-Unlike Table~\ref{tab:multeval_tst2010}, post-edited dev2010 no longer makes everything uniformly worse. 
-Instead, and as in Table~\ref{tab:tst2010}, post-editing of the tuning set obtains slightly better BLEU at the expense of much worse TER. 
-This again emphasizes the fact that tuning set noise can affect bottom-line performance in relatively unpredictable ways.
+Finally, the ``+ generated'' entry %in Table~\ref{tab:multeval_tst2010_fixed} 
+shows results obtained by concatenating sentence pairs from the quality-oriented ``example-based'' mode of our generator to the post-edited 2015 corpus.
+The starting point for writing template pairs was non-compositional translations from the 2015 training corpus with high TER scores, which were only practical to hand-inspect after post-editing to remove the clerical-like errors described in Section~\ref{subsec:case_study}.
+The basic rationale was that poorly scoring training pairs are ones that the model has not fully ``learned'' yet.
+These pairs were amplified by replacing one or more words in the templates with items from a lexical database, as proposed by \mynewcite{brown:99}.
+Unamplified alternative translations to some liberal (but not incorrect) translations were also included.
+As there were relatively few of them compared with Table~\ref{tab:results}, the generated sentences here were simply concatenated to the end of the training corpus.
+Not unlike the ``rule-based'' generated corpora from Table~\ref{tab:results}, these ``example-based'' generated corpora do not make a significant impact on system performance. 
+This may be because these atypical translations wound up being overamplified and thus skewing the model's probability distributions unfavorably.
+Alternatively, this can be interpreted that a crucial component of in-domain data is syntactic diversity.
+% TODO: 10 nouns and 10 adjectives gives you 100 sentences.
+but that skews the tables too much?
+BUT TO VERIFY THIS, I SHOULD RUN MORE EXPERIMENTS - like, how much does it skew if I don't amplify and only do like 1 or maybe 10 sentences for that pair?
 
-Table~\ref{tab:multeval_tst2010_fixed} also has some initial results from the ``example-based'' mode of our generator.
-[Should I add an example of the meta-templates, to show how easy they are to write?]
-These examples focus on quality instead of quantity and so there aren't many of them.
-Started by looking for non-compositional sentences in poorly-scoring training pairs.
-Alternative translations to some liberal (but not incorrect) translations were also included.
-This was only possible after post-editing to clear out the noise.
-They were simply concatenated to the end of the training corpus.
-The results almost certainly suck.
-To better direct future efforts, we perform some data snooping.
-We are looking for at least a partial answer to the question, what makes a corpus in-domain?
-
-Maybe we just to generate more data at scale. 
-But should we scale up lexical data or syntactic data? 
-To address this question, we resort to data snooping.
+Unlike its use as a tuning set in Section~\ref{subsec:case_study}, post-editing {\small \tt tst2010} as the test set had a more predictable effect.
+Namely, translation corrections improved TER scores by $\sim$5 points compared with Table~\ref{tab:multeval_tst2010}; it was easier for the MT system to reproduce lines that have not been grossly mistranslated.
+Unlike TER scores, the BLEU scores did not change appreciably. % uh, why not? any intuition behind this? ugh
+Interestingly, in Table~{tab:tst2010}, BLEU scores also changed much less than TER scores as a result of post-editing {\small \tt tst2010}, even though it was being used as a tuning set there.
+Scores evaluated on post-edited {\small \tt tst2010} are not shown, for the sake of clarity, and since the basic qualitative trends between the baseline, post-edited {\small \tt dev2010}, etc.
+were observed to be the same as in Table~\ref{tab:multeval_tst2010}.
 
 \subsection{Data Snooping}
-% Data Snooping - dedicate this to its own section, to make clear that these are not ``REAL'' results
-\textbf{Warning: This section is dedicated to ``data snooping runs'' that train on the test set in one way or another.}
-The scores given here must not considered indicative of actual out-of-sample system performance .
-The reason for the snooping is to try to isolate the most effective characteristics of in-domain data, whose power was seen in Section~\ref{sec:experiments}.
-It is hoped that these characteristics can guide future development of our generator from Section~\ref{sec:generator}.
+\textbf{
+Warning: This section is dedicated to ``data snooping'' experiments that included information from the test set during training in one way or another.
+The scores given here must not considered indicative of actual out-of-sample system performance.
+}
 
+Our reason for using the taboo practice of snooping is to try to isolate the most effective characteristics of in-domain data, whose power was seen in Section~\ref{sec:experiments}.
+It is hoped that these characteristics can guide future development of our generator from Section~\ref{sec:generator}, in particular, whether future efforts are better spent on scaling up the lexical database or the template database.
 The snooping is performed on {\small \tt tst2010}, to keep {\small \tt tst2014} relatively pristine for future work.
-Since they were so much smaller (\textasciitilde 1500 lines vs.
-\textasciitilde 180000 lines), snooping data sets were simply concatenated to the end of the IWSLT 2014 training corpus instead of using {\small \tt tmcombine}, with the rationale being that unlike our large, ``rule-based'' generated corpora from Section~\ref{sec:experiments}, these snooping additions should only perturb the translation tables slightly.
+Since they were so much smaller ($\sim$1500 lines vs.
+$\sim$180000 lines), snooping data sets were simply concatenated to the end of the IWSLT 2014 training corpus instead of using {\small \tt tmcombine}, with the rationale being that unlike our large, ``rule-based'' generated corpora from Section~\ref{sec:experiments}, these snooping additions should only perturb the resulting translation tables slightly.
 
@@
 
-For full snooping, the corpus was concatenated directly with {\small \tt tst2010}. 
-This establishes an ``upper bound'' of sorts of how much performance boost you could get by simply adding more data.
-Can think of this as the limit of infinite data (with in-domain distribution)?
-It's also a measure of the inherent limitation (bias?) of the model being used.
-Even snooped data is only on-par with out-of-sample Spanish-English performance, illustrating the difficulty of the task.
+For full snooping, {\small \tt tst2010} was directly concatenated to the training, establishing an ``upper bound'' of sorts on how much performance boost is obtainable by simply adding more data.
+For syntactic snooping, before concatenation, rare tokens in {\small \tt tst2010} were replaced by their part of speech (POS) from the Stanford POS tagger \mycitep{postagger:2003}. 
+Non-rare tokens were arbitrarily defined for each language as the most common 80\% in the IWSLT 2014 corpus: 670 tokens for English and 1406 tokens for Chinese, with the latter tokenized using the latest version of the Stanford Segmenter \mycitep{segmenter:2005,segmenter:2008}.
+This achieves a much smaller but still substantial increase in performance,\footnote{
+    The performance gain of 5 BLEU points or 6 TER points over the baseline is comparable with the gains obtained by the best scoring (and of course non-snooping) systems from IWSLT 2014 \mycitep{iwslt2014}.
+} even though this essentially performs zero snooping on out-of-vocabulary (OOV) words that the baseline system failed to translate.
+Finally, for lexical snooping, we hand-selected $\sim$100 mistranslated or OOV words from the baseline translation and used the rule-based mode of our generator to generate $\sim$30000 sentences with OOV words for concatenation with the training corpus.
+This method of lexical snooping in fact makes performance slightly worse (although the TER score is within error bars).
+This is thought to be because any gains from the translated OOV words are offset by the noise introduced into the phrase table by the nonsensical artificial sentences, which are otherwise composed of very common words.
+% UGH, I should probably do a qualitative analysis here...
+More generally, we speculate that although nonsensical sentences like Chomsky's ``colorless green ideas sleep furiously'' are fully licensed by English grammar, they can be detrimental in practical NLP applications like MT that operate on naturally occurring sentences at test time.
+Even when using an in-domain lexicon, their syntax may be sufficiently out-of-domain to result in a net loss of performance.
 
-For syntactic snooping, before concatenation, rare tokens in {\small \tt tst2010} were replaced by their part of speech (POS) from the Stanford POS tagger \mycitep{postagger:2003}. 
- Non-rare tokens were arbitrarily defined for each language as the most common 80\% in the IWSLT 2014 corpus: 670 tokens for English, 1406 tokens for Chinese, tokenized using the latest version of the Stanford Segmenter \mycitep{segmenter:2005,segmenter:2008}.
-You get a significant increase in performance, even though this essentially performs zero snooping on out-of-vocabulary (OOV) words that the baseline system failed to translate.
+Taken together, these snooping results reinforce the sentiment from the end of Section~\ref{sec:experiments} that future scaled-up work on our generator should eschew complex ``rule-based'' recursive modifications in favor of focus on the ``example-based'' approach.
+Alternatively, much more work on semantic constraints in the ``rule-based'' component is warranted, before trying to model more complex syntactic linguistic phenomena.
+Also, the observed relative unimportance of OOV words suggests it is less of a priority for the lexical database to be exhaustive.
 
-For lexical snooping, we selected \textasciitilde 100 mistranslated or OOV words from the baseline translation and used the rule-based mode of our generator to generate \textasciitilde 30000 sentences with OOV words for concatenation.
-This snooping-generated data doesn't make a dent, and in fact makes things slightly worse (although TER is within error bars).
-So nonsensical sentences of OOV words does NOT help - or even if it does, the nonsense-noise drowns it out.
-
-Taken together, these snooping results reinforce the sentiment from the end of Section~\ref{sec:experiments} that future scaled-up work on our generator should focus on the ``example-based'' approach.
-Thus  you should focus on the example-based half.
-
-%%% TODO: see OneNote
-
-%
+% TODO: full snooping comments. cut it shorter? model description is more important at this point?
+%It's also a measure of the inherent limitation (bias?) of the model being used. 
+%Can think of this as the limit of infinite data (with in-domain distribution)?
+%Even snooped data is only on-par with out-of-sample Spanish-English performance, illustrating the difficulty of the task.
 
@@
 Large quantities of supervised data underlie state-of-the-art performance in many NLP tasks.
-While current paid approaches like Amazon's Mechanical Turk are already a viable way of obtaining crowdsourced data, we believe it would be well worth the initial investment of time and resources to build a educational app compelling enough to become widely adopted.
+While current paid approaches like Amazon's Mechanical Turk are already a viable way of obtaining crowdsourced data, we believe it would be well worth the initial investment of time and resources to build an educational app compelling enough to become widely adopted.
 This could then serve as a platform from which to harvest supervised natural language data from its natural source. 
@@
 
+\section*{Acknowledgments}
+We are very grateful to the anonymous EMNLP 2016 reviewers for many helpful comments on an earlier version of this paper.
+
+%\pagebreak a man a plan a canal a panama. yes, it looks like ANY new reference will cause AMTA to go over limit.
+
 % these are so short that it's probably better to have it in literally instead of using \input

### 2016-07-14 doc/paper.tex
addressed all short EMNLP reviewer comments except for syntax pre-(re)ordering...

@@
     Ideally the generator would ``know'' {\em a priori} whether it is writing a sentence about, say, (financial) banks or (river) banks.
-} segmentation of languages like Chinese \mycitep{segmenter:2005,segmenter:2008} becomes a non-issue, and there is the potential for generating annotated corpora with zero annotation error.
+} segmentation of languages like Chinese \mycitep{segmenter:2008} becomes a non-issue, and there is the potential for generating annotated corpora with zero annotation error.
 Parallel corpora also provide a natural avenue to interface contemporary MT systems with the countless linguist-years of expertise that have been invested into rule-based MT systems.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\noindent\makebox[\linewidth]{\rule{\paperwidth}{0.4pt}} % full-width horizontal line
-TODO: detailed system description
+%\noindent\makebox[\linewidth]{\rule{\paperwidth}{0.4pt}} % full-width horizontal line
+%TODO: detailed system description
+% code section - do this first, since it MAY shed light on what linguistic phenomena are easy to implement?
+    % oh yes, definitely, because then you can explain things briefly in terms of their implementation
+%
+% Simplistic Linguistics
+% maybe use a bulleted list for the different linguistic phenomena supported?
+%Clause: transitive, meta, modal
+%basic support for NP, ADJP, ADVP (targeting ADJP or Clause), pronouns with antecedents 
+%prepositions, which are really relations, allowing semantic constraints on both object and target
+%indirect object as a special case of ``null preposition''
+%- not necessarily linguistically well-founded, but good enough
+%past tenses via transformation of present tense sentence
+% other transformations: topicalization, and some hokier stuff like \zh{把} 
+%Custom (``template-based mode''): lots of iterals, maybe mark an NP somewhere
 
-\noindent\makebox[\linewidth]{\rule{\paperwidth}{0.4pt}} % full-width horizontal line
+% big question: do I want to show a source code example??
+%\noindent\makebox[\linewidth]{\rule{\paperwidth}{0.4pt}} % full-width horizontal line
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 For the remaining unspecified model settings, we used the {\small \tt grow-diag} alignment heuristic recommended by \mynewcite{segmenter:2008} and the {\small \tt mslr-fe} configuration for lexicalized reordering.
-Chinese text was preprocessed using the latest version of the Stanford Segmenter \mycitep{segmenter:2005,segmenter:2008}.
+Chinese text was preprocessed using the latest version of the Stanford Segmenter \mycitep{segmenter:2008}.
 Phrase and lexicalized reordering tables were interpolated using the {\small \tt tmcombine} script included with Moses \mycitep{tmcombine}, with the interpolation optimized on the {\small \tt dev2010} data set.
@@
 For syntactic snooping, before concatenation, rare tokens in {\small \tt tst2010} were replaced by their part of speech (POS) from the Stanford POS tagger \mycitep{postagger:2003}. 
-Non-rare tokens were arbitrarily defined for each language as the most common 80\% in the IWSLT 2014 corpus: 670 tokens for English and 1406 tokens for Chinese, with the latter tokenized using the latest version of the Stanford Segmenter \mycitep{segmenter:2005,segmenter:2008}.
+Non-rare tokens were arbitrarily defined for each language as the most common 80\% in the IWSLT 2014 corpus: 670 tokens for English and 1406 tokens for Chinese, with the latter tokenized using the latest version of the Stanford Segmenter \mycitep{segmenter:2008}.
 This achieves a much smaller but still substantial increase in performance,\footnote{
@@
 The data format (English itself) is widely agreed upon, and there is even an existing market, a vast untapped resource of student problem sets that is yet to be harnessed by machine learning.
-\amtaonly{As a back of the envelope estimate, at a translation rate of about 1 noun per minute (based on our own data input experience), even for a modest user base of 10,000 users, a ten-minute problem set could in principle yield 100,000 data samples.
-It seems like the bottleneck would be judicious problem set design.}
+\amtaonly{ % Uh, should I bother with this stupidity, 
+%As a back of the envelope estimate, at a translation rate of about 1 noun per minute (based on our own data input experience), even for a modest user base of 10,000 users, a ten-minute problem set could in principle yield 100,000 data samples.
+%It seems like the bottleneck would be judicious problem set design.
+It is also worth noting that in a sense, this could regarded as a ``massively-online'' version of elicitation-based techniques  from fieldwork linguistics \mycitep{elicitation}, with an emphasis on acquiring unprocessed parallel sentences, and doing so at scale.
+}
 
 Large quantities of supervised data underlie state-of-the-art performance in many NLP tasks.
-While current paid approaches like Amazon's Mechanical Turk are already a viable way of obtaining crowdsourced data, we believe it would be well worth the initial investment of time and resources to build an educational app compelling enough to become widely adopted.
-This could then serve as a platform from which to harvest supervised natural language data from its natural source. 
+While current paid approaches like Amazon's Mechanical Turk are already a viable way of obtaining crowdsourced data, we believe 
+compelling, widely adopted educational apps like Duolingo\footnote{https://schools.duolingo.com/} to be much more cost-effective yet presently untapped platforms from which to harvest supervised natural language data from its natural source.
 Compared with current paid approaches, the cost per datum could be significantly reduced or perhaps even become negative (a profitable app), although the value of the harvested data itself might be worth subsidizing.
-We consider the construction of such a potentially game-changing source of supervised NLP data to be an open challenge to the NLP community in general.
+We consider the effective exploitation of such potentially game-changing sources of supervised NLP data to be an open challenge to the NLP community in general.
+
+% TODO: mention somehow that Duolingo is from the creator of reCAPTCHA? this doesn't seem to fit anywhere... and I don't have an academic reference for Duolingo, do I?
+% TODO: mention that Duolingo only does short phrases?
+but they USED to manage to piece them together to do professional translation work...
+BUT they stopped that?
+plus, there's no zh-en support?
+is that the main reason i stopped?
 
@@
 
-%\pagebreak a man a plan a canal a panama. yes, it looks like ANY new reference will cause AMTA to go over limit.
+%\pagebreak %a man a plan a canal a panama. yes, it looks like ANY new reference will cause AMTA to go over limit.
 

### 2016-07-14 doc/paper.tex
oops, forgot the transduction grammar and prereordering comments. running out of time...

@@
 } segmentation of languages like Chinese \mycitep{segmenter:2008} becomes a non-issue, and there is the potential for generating annotated corpora with zero annotation error.
+In principle, high-quality alignments could be obtained at the word or phrase level for use in the statistical MT process, and these alignments 
+could also be potentially useful in training methods for MT subtasks such as syntactic pre-reordering \mycitep{prereordering}.
 Parallel corpora also provide a natural avenue to interface contemporary MT systems with the countless linguist-years of expertise that have been invested into rule-based MT systems.
@@
     % oh yes, definitely, because then you can explain things briefly in terms of their implementation
+
+%Language-specific rules (e.g., word order) are confined to ad hoc bits of code the generator module.\footnote{
+%    Formally, this might be considered an extremely rudimentary synchronous (bilingual) context-free grammar of sorts \mycitep{transduction}, one that also admits long-range dependencies and semantic constraints.
+%    But it would be a disservice to those other well-developed methods to refer to our system as anything other than an ad hoc way to generate multilingual sentences.
+%}
+
 %
@@
 TED talks, which are prepared lectures, fall somewhere between the conversational language of movie subtitles and the formal written language of UN resolutions.
-As such, the OpenSubtitles\footnote{http://www.opensubtitles.org/} 2016 corpus \mycitep{opensubtitles:2016} from the OPUS project was also out-of-domain; results for this corpus were similar to MultiUN and are not shown.
+As such, the OpenSubtitles\footnote{http://www.opensubtitles.org/} 2016 corpus \emnlponly{\mycitep{opensubtitles:2016}} from the OPUS project was also out-of-domain; results for this corpus were similar to MultiUN and are not shown.
+% AMTA: removed citation to keep reference list under length
 For the table on the right, note the diminishing returns of out-of-domain data as the amount of in-domain data increases.
@@
 In tables with MultEval scores below, metric scores for all systems were generated using MultEval 0.5.1 \mycitep{multeval}: jBLEU V0.1.1 (an exact reimplementation of NIST's mteval-v13.pl without tokenization); Translation Error Rate (TER) V0.8.0.
-The $p$-values are relative to the baseline and indicate whether a difference of this magnitude (between the baseline and the system on that line) is likely to be generated again by some random process (a randomized optimizer).
+The $p$-values are relative to our baseline and indicate whether a difference of this magnitude (between the baseline and the system on that line) is likely to be generated again by some random process (a randomized optimizer).
 Metric scores are averages over three runs per system. 
@@
 
-%\pagebreak %a man a plan a canal a panama. yes, it looks like ANY new reference will cause AMTA to go over limit.
+\pagebreak
 

### 2016-07-14 doc/paper.tex
finished most of section 3 except tree figure and list of linguistic phenomena

@@
     \usepackage{graphicx}
+    %\usepackage{caption}
+    %\usepackage{subcaption}
 
@@
 
-\section{A Nascent Multilingual Generator}
+\section{A Rudimentary Multilingual Generator}
 \label{sec:generator}
@@
 Semantic constraints can be imposed from the verb {\small \tt V} as well as externally.
-The generator can then vary the template lexically\footnote{
-    Rewrite probabilities were set by hand; a more principled approach might use, say, a generative PCFG parser.} and syntactically, allowing a large number of sentence pairs to be generated from a single template.
+The generator can then vary the template lexically and syntactically,\footnote{\label{footnote:probabilities}
+    Rewrite probabilities were set by hand; a more principled approach might use, say, a generative synchronous CFG.} allowing a large number of sentence pairs to be generated from a single template.
 We can also trade quantity for quality by adding more literals to the template, in the spirit of example-based MT.
 
+%This Section is organized as follows. 
+%Section~\ref{subsec:software}, we describe our software architecture and internal data representation. 
+%Section~\ref{subsec:linguistics} then covers the linguistic phenomena we implemented in this framework.
+
 % TODO: maybe these nonsense sentences shed some light as to what is NOT a sensible, ``natural'' sentence...
-%\subsection{Simplistic Linguistics}
+\amtaonly{
+\subsection{Simplistic Linguistics}
+\label{subsec:linguistics}
+}
 We adopt an extremely simplistic view of language as nouns, verbs, and their modifications and elaborations.
@@
 }The fact that verb phrases can also serve as nouns (gerundives) and modifiers (participles) appears to be a major source of syntactic complexity, and we made it a priority to implement participial modifiers.
+
+\amtaonly{
+
+\subsection{Software Architecture}
+\label{subsec:software}
+
+The bulk of our software, which we plan on eventually open-sourcing, along with its accompanying data, comprises four Python modules:\footnote{
+    In particular, we used Python 3 for its improved Unicode support over Python 2.
+}
+
+\begin{itemize}
+\item {\small \tt data}: data abstraction layer
+\item {\small \tt nodes}: internal tree representation
+\item {\small \tt generator}: converts internal trees to surface forms (sentence pairs)
+\item {\small \tt main}: outermost logic
+\end{itemize}
+
+The {\small \tt data} module implements an abstraction layer that mediates between data resources on disk and the rest of the program.
+These data resources include bilingual (extensible to multilingual) lexicons for different parts of speech, monolingual morphology, templates, and a small taxonomy used to tag nouns.
+All of our data resources were written in YAML, a ``human-readable'' language particularly well-suited to manual editing.
+In particular, YAML has native Unicode support and does not even require quotation marks to enclose strings.
+An example of YAML code is shown in Figure~\ref{fig:yaml-rb}.
+
+% TODO: probably going to have to resize/crop this image to keep this paper from going over length
+% TODO: YAML code examples. couldn't get verbatim + zh to work... just use an image? see if there's room.
+% TODO: manually indent YAML. use placeholder graphic for now instead.
+% TODO: get proper left/right subfigures working? just slapping in a single image for now.
+\begin{figure}[ht]
+    \centering
+    \includegraphics{yaml-rb}
+    \caption{An example of YAML code, showing syntactic (left) and semantic (right) components of a clause template used for the ``rule-based'' mode of generation mentioned in the discussion of the {\small \tt main} module.}
+    \label{fig:yaml-rb}
+\end{figure}
+
+% the internal data representation. the bulk of the discussion?
+The {\small \tt nodes} module implements the data structures used internally by the program.
 To capture recursive structures like participles, we adopt a rudimentary tree representation that is something intermediate between constituency and dependency trees, with constituent-like modifiers that also maintain dependency-like links to their targets.
-We plan on eventually open-sourcing both our generator code and its accompanying data.
+Following our highly simplified picture of linguistics in Section~\ref{subsec:linguistics}, the basic backbone of the tree is constituency-like, as shown in Figure~\ref{fig:tree}.
 
-%\subsection{Implementation Details}
-\amtaonly{
-% TODO: mention Python?
-One implementation detail worth mentioning is the choice of YAML for writing sentence templates and lexical data. 
-In particular, YAML has native Unicode support and does not even require quotation marks to enclose strings.
+% TODO: nodes class diagram? seems like an implementation detail...
+% TODO: the fastest way i have to get dotted lines is through Word or PowerPoint, but I wanted to do the figures in OneNote, at least the camera-unready ones...
+% TODO: it would be NICE if the tree matched a YAML example...
+\begin{figure}[ht]
+    \centering
+    placeholder graphic, used to estimate space remaining in the paper
+    \includegraphics{yaml-rb-rotated}
 
-% TODO: YAML code examples. couldn't get verbatim + zh to work...
-%To maximize human translator productivity, 
+    \caption{
+        An example tree made of nodes from the {\small \tt nodes} module.
+        Thick lines indicate the ``constituent'' backbone that dictates the order of traversal in the {\small \tt generator} module, while thin lines indicate longer-range dependencies that are stored as links between nodes.
+    }
+    \label{fig:tree}
+
+\end{figure}
+
+% TODO: class hierarchy for nodes module only??
+% - data is too mundane, and i'll just describe it in words.
+% - Generator is only 2 classes, and i'll also just describe it in words.
+%- ad hoc semantic taxonomy that allows a template to request
+
+The {\small \tt generator} module traverses the backbone of the ``interlingual'' tree from the {\small \tt nodes} module and recursively builds up parallel surface forms (bilingual sentence pairs).
+Language-specific rules (e.g., word order) are implemented as ad hoc bits of code in this module,\footnote{
+    Formally, the ``rule-based'' component of our system can be considered an extremely rudimentary synchronous (bilingual) context-free grammar of sorts \mycitep{transduction} that also admits long-range dependencies and semantic constraints.
+    But it would be a disservice to those other well-developed methods to refer to our system as anything other than an ad hoc way to generate multilingual sentences.
+%    MEH: Also, example-based mode admits the possibility of just writing down translations and then generating parse trees on PORTIONS of those templates.
+%    Although then I guess you have a transductive grammar on those portions?
+} reading long-range dependency information from the tree as needed.
+All language-dependent code is confined to this module, although this is coupled with language-dependent information from the YAML data (e.g., the tags and punctuation in Figure~\ref{fig:yaml-rb}). 
+% TODO: More ideally, A desirable longer-term goal would be to move language rules to data, but that's beyond the scope of this paper.
+
+Finally, the outermost {\small \tt main} module is responsible for tree instantiation and modification.
+To generate a sentence pair in ``rule-based mode,'' a syntactic clause template is randomly selected, followed by a semantic verb category that uses that syntactic template, as in Figure~\ref{fig:yaml-rb}.
+Modifier nodes are then randomly attached, recursively so for constructs like participles and prepositions, in principle admitting arbitrarily long sentences.
+The probabilities for these random processes are chosen heuristically as mentioned in Footnote~\ref{footnote:probabilities}.
+The system also allows for semantic constraints by adding semantic tags to nodes and querying the taxonomy through the {\small \tt data} module for eligible lexical choices. 
+%For example, the semantic template component in Figure~\ref{fig:yaml-rb} requires a person for the subject node, and it should be able to query the noun database for UGH
+Building a comprehensive taxonomy and labeling the lexical database exhaustively is a labor-intensive task; results in Section~\ref{supersec:experimental_results} below were obtained with relatively minimal semantic constraints.
+
+% example-based amplification: loop over different lexical choices exhaustively
+One of the things we wanted to do differently from KPML \mycitep{kpml} was to implement an ``example-based mode'' in which sentence pairs that would be difficult to generate in ``rule-based mode'' could simply be written down and then amplified as proposed by \mynewcite{brown:99}.
+This would trade the quantity available through arbitrarily complex sentences from ``rule-based'' mode for the quality of sentence pairs with a larger degree of hand-translation.
+Our software achieves this by allowing the {\small \tt template} entries as on the left side of Figure~\ref{fig:yaml-rb} to contain literal strings in addition to symbols.
+These ``example-based'' templates are currently selected by hand for the experiments described in Section~\ref{supersec:experimental_results}, with all amplification resulting from varying the lexical choices (i.e., without adding ).
+Whether it helps to add modifier nodes in this mode is a potential area for future work.
+% TODO: meta-template picture (if space allows)?
+or is this description adequate?
+this is kind of like the tradeoff I faced in the EMNLP submission, where I was like, uh oh, out of space...
+meh, this system sucks anyways, so I'll just omit its details
+A helper script also supports the use of ``meta-templates'' to facilitate hand-construction of templates, allowing a minimal annotation of a sentence pair to produce a full YAML template similar to Figure~\ref{fig:yaml-rb}.
+
+%% TODO: ugh, out of room? this stuff is pretty pointless anyway
+%It is hoped that these ``example-based'' templates prove useful for tasks like training set amplification, which we consider in Sections~\ref{sec:experiments}~and~\ref{subsec:post_edited}. 
+%But it would be very easy to specify modifiable nodes in the meta-templates, and you could add modifiers as in ``rule-based'' mode.
+%The point of rule-based templates (and indeed linguistics) is to be precise and write a few rules that can generate a lot of sentences.
+%But for example-based templates, we want to do lots of sentences, fast, and then maybe amplify little pieces of it.
+%Meta-templates let you iterate quickly over sentences.
+
+%An example of such a meta-template is given in Figure~XXX, which corresponds to the example-based template given above in Figure~YYY
+%\begin{small}
+%\noindent {\tt S V O .} \\
+%{\tt S \zh{把} O V \zh{。}}
+%\end{small}
+
+% figure: example-based template, and its associated 
 %meta-templates, which can then be easily parsed into templates to be fed into the generator.
+% one big image, top/bottom?
 %An example is shown in Figure~\ref{fig:yaml_template}.
 
+%} \hcomponly{ % disable last bit of unfinished stuff for temp upload
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\noindent\makebox[\linewidth]{\rule{\paperwidth}{0.4pt}} % full-width horizontal line
-%TODO: detailed system description
-% code section - do this first, since it MAY shed light on what linguistic phenomena are easy to implement?
-    % oh yes, definitely, because then you can explain things briefly in terms of their implementation
+\noindent\makebox[\linewidth]{\rule{\paperwidth}{0.4pt}} % full-width horizontal line
+TODO
 
-%Language-specific rules (e.g., word order) are confined to ad hoc bits of code the generator module.\footnote{
-%    Formally, this might be considered an extremely rudimentary synchronous (bilingual) context-free grammar of sorts \mycitep{transduction}, one that also admits long-range dependencies and semantic constraints.
-%    But it would be a disservice to those other well-developed methods to refer to our system as anything other than an ad hoc way to generate multilingual sentences.
-%}
+\subsection{Implemented Linguistic Phenomena}
 
-%
-% Simplistic Linguistics
-% maybe use a bulleted list for the different linguistic phenomena supported?
-%Clause: transitive, meta, modal
-%basic support for NP, ADJP, ADVP (targeting ADJP or Clause), pronouns with antecedents 
-%prepositions, which are really relations, allowing semantic constraints on both object and target
-%indirect object as a special case of ``null preposition''
-%- not necessarily linguistically well-founded, but good enough
-%past tenses via transformation of present tense sentence
-% other transformations: topicalization, and some hokier stuff like \zh{把} 
-%Custom (``template-based mode''): lots of iterals, maybe mark an NP somewhere
+We briefly list
+and how they are implemented in terms of the tree representation from Section~\ref{subsec:software}
 
-% big question: do I want to show a source code example??
-%\noindent\makebox[\linewidth]{\rule{\paperwidth}{0.4pt}} % full-width horizontal line
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+a couple of reasons for keeping linguistic phenomena to the most basic ones (aside from manageability)
+- we believe that for the purposes of generating parallel corpora, is most effective to treat very difficult, non-compositional phenomena (example?!) by simply writing them down, in tradition of example-based MT
+- keeping the data format simpler makes it potentially easier to scale up to admit contributions from non-linguists
 
-} % amtaonly
+- emphasize that we prioritize generating in-domain sentences above linguistic soundness
+
+% transformations were chosen with an eye toward scalability, so that things like semantic constraints (<animal> eats <food>) could be used (eating <food>, and should modify <animal>)
 
@@
 
-%here is the figure
-%\begin{figure}[ht]
-%    \centering
-%    \includegraphics{Chick1}
-%    \caption{\label{fig:test} chick}
-%\end{figure}
-%Hola chica, Figure~\ref{fig:test}.
+% maybe use a bulleted list for the different linguistic phenomena supported? yeah, let's just do that.
+
+%Oh, maybe I should make this a table. that should cut down on the vertical spacing
+%
+%\begin{table}[ht]
+%\centering
+%\begin{tabular}{|l|l|}
+%\hline
+%Gold en & 60,000. \\ \hline
+%Gold en & 60,000. \\ \hline
+%Gold en & 60,000. \\ \hline
+%Gold en & 60,000. \\ \hline
+%Gold en & 60,000. \\ \hline
+%Gold en & 60,000. \\ \hline
+%Gold en & 60,000. \\ \hline
+%Gold en & 60,000. \\ \hline
+%Gold en & 60,000. \\ \hline
+%Gold en & 60,000. \\ \hline
+%\end{tabular}
+%\caption{ 
+%    Linguistic constructs implemented in our multilingual generator.
+%}
+%\label{tab:implemented_linguistics}
+%\end{table}
+
+\begin{itemize}
+\item clauses: transitive, meta, modal. a word on modalish. 
+\item NP
+\item ADJP
+\item participles as transformations of clauses that convert subject into a linked node, allowing reuse of semantic constraints
+\item prepositions as relations between two nodes (noun/noun or noun/verb), and accompanying semantic requirements on both object and target
+\item custom nodes - because sometimes, the most effective way to represent a translation is to just write it down. (``template-based mode''): lots of iterals, maybe mark an NP somewhere
+
+% other (if there's room)
+\item ADVP (targeting ADJP or Clause), pronouns with antecedents 
+\item indirect object as a special case of ``null preposition'' - not necessarily linguistically well-founded, but good enough to make sentences
+\item past tense as transformation of present tense  % MEH
+\item other transformations: topicalization, and some hokier stuff like \zh{把} 
+\end{itemize}
+
+% big question: do I want to show a source code example?? yeah, i probably have room
+\noindent\makebox[\linewidth]{\rule{\paperwidth}{0.4pt}} % full-width horizontal line
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+} % amtaonly
 
@@
 TED talks, which are prepared lectures, fall somewhere between the conversational language of movie subtitles and the formal written language of UN resolutions.
-As such, the OpenSubtitles\footnote{http://www.opensubtitles.org/} 2016 corpus \emnlponly{\mycitep{opensubtitles:2016}} from the OPUS project was also out-of-domain; results for this corpus were similar to MultiUN and are not shown.
+As such, the OpenSubtitles\footnote{http://www.opensubtitles.org/} 2016 corpus \emnlponly{\mycitep{opensubtitles:2016} }from the OPUS project was also out-of-domain; results for this corpus were similar to MultiUN and are not shown.
 % AMTA: removed citation to keep reference list under length
@@
 Alternatively, one can interpret these errors as an additional source of noise in the human encoding of English into Chinese.
-It is not entirely clear how the MT process is affected by this kind of noise.
+It is not immediately obvious how the MT process is affected by this kind of noise.
 
@@
 
-% \pagebreak % soooooo ugly
+% TODO: didn't run MultiUN + tst2010 - but that's okay I guess, I have the out-of-domain in-domain comparison in Table 1
 
@@
 Also, this might look like a proportional improvement in BLEU, but gains were found to come in discrete steps during the editing process. 
-This is thought to be due to the non-parametric nature of SMT, with changes in scores only occurring when there are appreciable changes to the parts of the probability tables that are relevant to the test set.
+This is thought to be due to the non-parametric nature of statistical MT, with changes in scores only occurring when there are appreciable changes to the parts of the probability tables that are relevant to the test set.
 It remains to be seen what effect post-editing of training, tuning, and test corpora would have on parametric systems like recent deep-learning based MT methods \mycitep{sutskever:2014,bengio:2014}, in which all training examples contribute to the same shared connection weight matrices.
@@
 Not unlike the ``rule-based'' generated corpora from Table~\ref{tab:results}, these ``example-based'' generated corpora do not make a significant impact on system performance. 
-This may be because these atypical translations wound up being overamplified and thus skewing the model's probability distributions unfavorably.
-Alternatively, this can be interpreted that a crucial component of in-domain data is syntactic diversity.
+These atypical translations may have wound up being overamplified and thus also ``teaching'' spurious phrases to the statistical MT system (which can be thought of as variable-length example-based MT of sorts).
+Alternatively, this can be interpreted as indicating that a crucial component of in-domain data is syntactic diversity.
 % TODO: 10 nouns and 10 adjectives gives you 100 sentences.
@@
 
-\pagebreak
+%\pagebreak  % use this to gauge reference list length
 

### 2016-07-15 doc/paper.tex
I guess that's it for text - my GOODNESS the table in section 3.3 is sloppy and impenetrable...

@@
 
+    \usepackage{tabularx}
+
 \fi
@@
 A rudimentary system for doing this is presented, with mechanisms for syntactic complexity in the spirit of classical rule-based machine translation and high-quality sentence pairs in the spirit of example-based machine translation.
-Preliminary experiments integrating these synthetic corpora into a shared task are reported.
+Preliminary experiments integrating these synthetic corpora into the IWSLT shared task are reported.
+In the process, it was incidentally found that correcting $\sim$4\% of the shared task's training corpus for obvious, clerical-type errors resulted in a $\sim$6\% relative increase in BLEU scores.
 Promising future directions are discussed, in particular, large-scale contributions from ``amateur linguists'' via crowdsourcing.
@@
     Formally, the ``rule-based'' component of our system can be considered an extremely rudimentary synchronous (bilingual) context-free grammar of sorts \mycitep{transduction} that also admits long-range dependencies and semantic constraints.
-    But it would be a disservice to those other well-developed methods to refer to our system as anything other than an ad hoc way to generate multilingual sentences.
+    But it would be a disservice to those other well-developed methods to refer to our system as anything other than a makeshift way to generate multilingual sentences.
 %    MEH: Also, example-based mode admits the possibility of just writing down translations and then generating parse trees on PORTIONS of those templates.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\noindent\makebox[\linewidth]{\rule{\paperwidth}{0.4pt}} % full-width horizontal line
-TODO
+
+%TODO
 
@@
 
-We briefly list
-and how they are implemented in terms of the tree representation from Section~\ref{subsec:software}
+Finally, to conclude this section, we briefly list linguistic constructs and their implementations in terms of the tree representation from Section~\ref{subsec:software}.
+Note that since we are interested primarily in the generated surface forms, the linguistic soundness of how they are implemented in our tree representation is of relatively secondary importance.
+Aside from manageability, there are a few other reasons for focusing on just the most basic linguistic phenomena for the ``rule-based'' mode.
+We strongly believe that for the purposes of generating large and high-quality parallel corpora, it is most effective to treat difficult, non-compositional phenomena by simply writing them down, in tradition of example-based MT.
+Also, minimizing the learning curve makes it potentially easier to scale up to admit contributions from non-linguists.
 
-a couple of reasons for keeping linguistic phenomena to the most basic ones (aside from manageability)
-- we believe that for the purposes of generating parallel corpora, is most effective to treat very difficult, non-compositional phenomena (example?!) by simply writing them down, in tradition of example-based MT
-- keeping the data format simpler makes it potentially easier to scale up to admit contributions from non-linguists
-
-- emphasize that we prioritize generating in-domain sentences above linguistic soundness
-
-% transformations were chosen with an eye toward scalability, so that things like semantic constraints (<animal> eats <food>) could be used (eating <food>, and should modify <animal>)
-
-% if i'm gonna do source code snippets with zh, maybe it'd be best to use IMAGES? it's not for camera-ready anyway...
-% well, i can do the meta-template easily... the problem is with REAL templates
+% transformations - with scalability in mind, so that things like semantic constraints (<animal> eats <food>) could be used (eating <food>, and should modify <animal>)
 
@@
 %
-%\begin{table}[ht]
-%\centering
-%\begin{tabular}{|l|l|}
-%\hline
-%Gold en & 60,000. \\ \hline
-%Gold en & 60,000. \\ \hline
-%Gold en & 60,000. \\ \hline
-%Gold en & 60,000. \\ \hline
-%Gold en & 60,000. \\ \hline
-%Gold en & 60,000. \\ \hline
-%Gold en & 60,000. \\ \hline
-%Gold en & 60,000. \\ \hline
-%Gold en & 60,000. \\ \hline
-%Gold en & 60,000. \\ \hline
-%\end{tabular}
-%\caption{ 
-%    Linguistic constructs implemented in our multilingual generator.
-%}
-%\label{tab:implemented_linguistics}
-%\end{table}
+\begin{table}[ht]
+\centering
+%\begin{tabular}{|l|l|} % tabularx allows automatic wrapping
+\begin{tabularx}{\linewidth}{ |l|X| }
+\hline
+Clauses & indicative present active; intransitive, transitive, ``meta'' (clause subnode), ``modalish'' (e.g., ``Alice wants to kick Bob'' - with ``Alice'' linked to ``kick'' as in Figure~\ref{fig:tree} for subject-verb semantics) \\ \hline % see with a ``dependency link'' for subject-verb syntactic agreement and
+NP & noun phrases headed by nouns, names, or pronouns; optional modifiers as child nodes; pronoun antecedents via links \\ \hline
+ADJP & adjectives as NP modifiers, with links to head nouns (e.g., this vs. these) \\ \hline
+ADVP & adverbs as Clause or ADJP modifiers \\ \hline
+Participles & transformations of clauses that convert subject child node into a linked node, allowing reuse of semantic constraints \\ \hline
+PP & prepositional phrases (relations) between two nodes (noun/noun or noun/verb), with relation-dependent semantic requirements on both nodes; indirect objects as ``null preposition'' nodes \\ \hline
+%Custom nodes & ``example-based'' nodes as discussed in Section~\ref{subsec:software} \\ \hline
+%Transformations & alterations of existing templates - past tense, topicalization, \zh{把}-construction \\ \hline
+\end{tabularx}
+\caption{ 
+    Linguistic constructs implemented for the ``rule-based'' mode of our generator.
+}
+\label{tab:implemented_linguistics}
+\end{table}
 
-\begin{itemize}
-\item clauses: transitive, meta, modal. a word on modalish. 
-\item NP
-\item ADJP
-\item participles as transformations of clauses that convert subject into a linked node, allowing reuse of semantic constraints
-\item prepositions as relations between two nodes (noun/noun or noun/verb), and accompanying semantic requirements on both object and target
-\item custom nodes - because sometimes, the most effective way to represent a translation is to just write it down. (``template-based mode''): lots of iterals, maybe mark an NP somewhere
+%\begin{itemize}
+%\item 
+%\item participles as
 
-% other (if there's room)
-\item ADVP (targeting ADJP or Clause), pronouns with antecedents 
-\item indirect object as a special case of ``null preposition'' - not necessarily linguistically well-founded, but good enough to make sentences
-\item past tense as transformation of present tense  % MEH
-\item other transformations: topicalization, and some hokier stuff like \zh{把} 
-\end{itemize}
+%
+%% other (if there's room)
+%\item indirect object as a special case of ``null preposition'' - not necessarily linguistically well-founded, but good enough to make sentences
 
-% big question: do I want to show a source code example?? yeah, i probably have room
-\noindent\makebox[\linewidth]{\rule{\paperwidth}{0.4pt}} % full-width horizontal line
+%\end{itemize}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 Warning: This section is dedicated to ``data snooping'' experiments that included information from the test set during training in one way or another.
-The scores given here must not considered indicative of actual out-of-sample system performance.
+The scores given here must not be considered indicative of actual out-of-sample system performance.
 }
@@
 %It seems like the bottleneck would be judicious problem set design.
-It is also worth noting that in a sense, this could regarded as a ``massively-online'' version of elicitation-based techniques  from fieldwork linguistics \mycitep{elicitation}, with an emphasis on acquiring unprocessed parallel sentences, and doing so at scale.
+It is also worth noting that in a sense, this could regarded as a ``massively-online'' version of elicitation-based techniques from fieldwork linguistics \mycitep{elicitation} for low-resource language pairs, with an emphasis on acquiring unprocessed parallel sentences, and doing so at scale.
 }

### 2016-07-15 doc/paper.tex
amta submission

@@
 % TODO: the fastest way i have to get dotted lines is through Word or PowerPoint, but I wanted to do the figures in OneNote, at least the camera-unready ones...
+% TODO: preposition (or participle!) might be illustrative, but PP is just WAY too messy and involved to draw in a small space...
 % TODO: it would be NICE if the tree matched a YAML example...
+% TODO: oh, I could have made a tree diagram using some LaTeX nlp tree package? Word + pdf wound up being pretty fast though
 \begin{figure}[ht]
     \centering
-    placeholder graphic, used to estimate space remaining in the paper
-    \includegraphics{yaml-rb-rotated}
-
+    \includegraphics{tree}
     \caption{
         An example tree made of nodes from the {\small \tt nodes} module.
-        Thick lines indicate the ``constituent'' backbone that dictates the order of traversal in the {\small \tt generator} module, while thin lines indicate longer-range dependencies that are stored as links between nodes.
+        Solid lines indicate the ``constituent'' backbone that dictates the order of traversal in the {\small \tt generator} module, while dashed lines indicate longer-range syntactic or semantic dependencies that are stored as links between nodes.
+        % In this example, V$_1$ can be a ``modalish'' verb like ``want'' that acts like a non-defective modal verb of sorts % meh, let's just bury it in the table below
     }
@@
 All language-dependent code is confined to this module, although this is coupled with language-dependent information from the YAML data (e.g., the tags and punctuation in Figure~\ref{fig:yaml-rb}). 
+For example, it is left to the generator to order prepositional phrases after their targets in English but before their targets in Chinese.
 % TODO: More ideally, A desirable longer-term goal would be to move language rules to data, but that's beyond the scope of this paper.
@@
 
-% maybe use a bulleted list for the different linguistic phenomena supported? yeah, let's just do that.
-
 %Oh, maybe I should make this a table. that should cut down on the vertical spacing
@@
 ADVP & adverbs as Clause or ADJP modifiers \\ \hline
-Participles & transformations of clauses that convert subject child node into a linked node, allowing reuse of semantic constraints \\ \hline
+VP & transformations of clauses into participles or modal(ish) complements; convert subject child node into a linked node, allowing reuse of semantic constraints \\ \hline
 PP & prepositional phrases (relations) between two nodes (noun/noun or noun/verb), with relation-dependent semantic requirements on both nodes; indirect objects as ``null preposition'' nodes \\ \hline
@@
 \end{table}
-
-%\begin{itemize}
-%\item 
-%\item participles as
-
-%
-%% other (if there's room)
-%\item indirect object as a special case of ``null preposition'' - not necessarily linguistically well-founded, but good enough to make sentences
-
-%\end{itemize}
 
@@
 % but this creates a natural break for the discussion 
-Table~\ref{tab:tst2010} shows the scores obtained by ``lightly'' post-editing the MERT tuning set {\small \tt tst2010} as follows. 
+Table~\ref{tab:tst2010} below shows the scores obtained by ``lightly'' post-editing the MERT tuning set {\small \tt tst2010} as follows. 
 On the Chinese side, duplicated segments, applause lines, and translator notes were removed. 
@@
 \subsection{Data Snooping}
-\textbf{
-Warning: This section is dedicated to ``data snooping'' experiments that included information from the test set during training in one way or another.
+\textbf{Warning: This subsection is dedicated to ``data snooping'' experiments that included information from the test set during training in one way or another.
 The scores given here must not be considered indicative of actual out-of-sample system performance.
